### 2018-02-26 paper/dkt.tex
added initial python files and .gitignore

\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amssymb,calc,ifthen}
\usepackage{float}
%\usepackage{cancel}
\usepackage[table,usenames,dvipsnames]{xcolor} % for coloured cells in tables
\usepackage{tikz}
% Allows us to click on links and references!
\usepackage{hyperref}
\usepackage{url}
\hypersetup{
colorlinks,
citecolor=black,
filecolor=black,
linkcolor=black,
urlcolor=black
}
% Nice package for plotting graphs
% See excellent guide:
% http://www.tug.org/TUGboat/tb31-1/tb97wright-pgfplots.pdf
\usetikzlibrary{plotmarks,shapes}
\usepackage{amsmath,graphicx}
\usepackage{mathtools}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
% highlight - useful for TODOs and similar
\usepackage{color}
\newcommand{\hilight}[1]{\colorbox{yellow}{#1}}
\newcommand\ci{\perp\!\!\!\perp} % perpendicular sign
\newcommand*\rfrac[2]{{}^{#1}\!/_{#2}} % diagonal fraction
\newcommand\SLASH{\char`\\}
\usepackage{listings}
% margin size
\usepackage[margin=0.5in]{geometry}
\usepackage{pdfpages}
\usepackage{enumitem} % for nested enumerate numbers 1 1.1 1.1.1

\usepackage{titlesec} % reduce spacing after subsections

% \titlespacing\subsection{0pt}{4pt plus 4pt minus 2pt}{4pt plus 2pt minus 2pt}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\begin{document}
\belowdisplayskip=12pt plus 3pt minus 9pt
\belowdisplayshortskip=7pt plus 3pt minus 4pt

% \rowcolors{2}{gray!25}{white}

% \begin{figure}[H]
% \centering
% \includegraphics[scale=0.5]{../code/synthetic/14/figures/liksTrueFittedParams}
% \caption{14 clear biomk}
% \end{figure}
% 
% \begin{figure}[H]
% \centering
% \includegraphics[scale=0.5]{../code/synthetic/14/figures/liksTrueFittedParams}
% \caption{11 clear + 3 unclear biomk}
% \end{figure}

\definecolor{blue3}{HTML}{86B7FC} % med blue
\definecolor{blue1}{HTML}{B5F1FF} % light blue
\definecolor{blue2}{HTML}{E0F9FF} % very light blue

% \section{Motivation}
% 
% It is currently believed that neurodegenerative diseases target specific brain networks (Seeley, 2009). Moreover, different symptoms assiciated with different syndromes are thus believed to be a result of different underlying brain areas being affected.

% TITLE Disease Knowledge Transfer for Alzheimer's Variants

\section{Abstract}

We present Disease Knowledge Transfer (DKT), a technique for transferring biomarker correlations between Alzheimer's disease (AD) variants. DKT allows, for the first time, to infer biomarker values and their progressions in rare variants of AD for which data on such biomarkers is not available, by learning them from a different disease. For doing this, we formulate a new dementias paradigm called "Affected Overlap", which proposes that dementias are diseases that affect overlapping brain regions, and thus present certain biomarker characteristics that are shared and can thus be transferred across diseases. We then formulate the DKT model implementing this paradigm as a joint-disease generative model of biomarker progression which disentangles the biomarker correlations that are disease-specific from those correlations that are disease-agnostic. We demonstrate DKT on three datasets: (1) the TADPOLE Challenge dataset containing typical AD subjects (tAD) from the Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset who had MRI, FDG, AV45, AV1451 or DTI scans, as well as two datasets from our local centre containing subjects with (2) typical AD and (3) Posterior Cortical Atrophy (PCA) that have only undergone MRI scans. We train the DKT model on all three datasets at once, and show that it is able to predict, in PCA subjects, plausible population-level biomarker for FDG, DTI, AV45 and AV1451 biomarkers, for which no data was available. We validate the predictions of DKT using a small set of DTI scans for the PCA cohort. Finally, DKT can be a useful tool to analyse and understand rare forms of dementias for which multimodal data is not available or is limited (e.g. cross-sectional). Finally, by leveraging data from multiple diseases, DKT also has the potential to provide more accurate disease staging compared to traditional disease progression models, which is useful for patient stratification in clinical trials.

% 1. Statement of Novelty/Impact *

We present a novel dementias paradigm (Affected Overlap) and technique (DKT) that can be used to predict, for the first time, the population-level progression of biomarkers in a disease for which no such biomarker data is available, by transferring them from a different disease where they are available.

\section{Hierarchical model of disease progression}

The framework/model assumes that within a dataset of patients with a particular form of dementia, there are certain correlations between biomarkers that are \emph{disease agnostic}. For example, atrophy in the temporal lobe usually correlates with amyloid/tau-levels within the same region or with memory impairment. This is because we believe that amyloid/tau deposition cause neurodegeneration, which then causes cognitive decline in functions related to the underlying affected areas. One can then view different neurodenegenerative diseases as affecting a different set of brain regions and their underlying connectivity networks, a view supported by previous studies (Seeley, 2009, Zhou 2012). This view suggests that one can uniquely define a disease in terms of the different subset of brain regions that it affects. However, sometimes there are \emph{digo dktseasesease-agnostic} correlations that are not confined to a single region, e.g. disconnected brains regions are known to be responsible for certain cognitive functions, or biomarker measurements in different brain regions could correlate due to underlying structural or functional connectivity. One can therefore extend the idea of disease-agostic correlations between biomarkers to region-nonspecific sets of biomarkers which we will call \emph{functional units}, because the biomarkers being related through similar function.

We propose a hierarchical model of disease progression that disentangles \emph{disease-specific} biomarker correlations from \emph{disease-agnostic} correlations. For modelling disease-agnostic correlations, we assume biomarkers are grouped into \emph{functional units}, where corelations within each unit are disease-agnostic (e.g. biomarkers corresponding to the same ROI).  Each biomarker is assigned a-priori to a specific functional unit, which corresponds to a brain region, say temporal lobe (Fig 1., bottom).  For example, biomarkers assigned to the temporal unit could be: amyloid in temporal lobe, tau in temporal lobe, MRI atrophy in temporal lobe and certain memory tests. Using a disease progression model of our choice, for each region-specific functional unit we can model the correlation of biomarkers within that functional unit (Figure 1, bottom), which allows us to reconstruct a region- or unit-specific \emph{dysfunction} progression axis which can be used for staging subjects. Finally, we express the disease specific correlations (Figure 1, top) not directly with each biomarkers, but via the dysfunction scores estimated within each functional unit.

The proposed model is parsimonious, easily interpretable by abstracting away biomarker information intothe functional units, and can be used for \emph{disease knowledge transfer}, i.e. learning correlations in one disease from another. Our approach is inspired by transfer learning methods from machine learning, but the model is generative, easing interpretability.

\begin{figure}[h]
 \centering
 \includegraphics[height=11cm]{figures/disease_knowledge_transfer.png}
 \caption{Outline of the proposed framework for joint modelling of multiple diseases. }
\end{figure}

\subsection{Mathematical formulation}

Variable names:
\begin{itemize}
 \item $\Lambda$ = set of \emph{functional units}, i.e. \{temporal, parietal, occipital, ..\}. Biomarkers belonging to the same functional unit are asusmed to have disease-agnostic correlations.
 \item $y_{ijk}$ = measurement in subject $i$, visit $j$, biomarker $k$
%  \item $d_{ijl}$ = dysfunctionality score in subject $i$, visit $j$ in \emph{functional unit} $l$, where $l \in \Lambda$
%  \item $s_{ij}$ = disease progression score in subject $i$, visit $j$
 \item $\theta_k$ = parameters of trajectory for biomarker $k$
 \item $\lambda^l$ = parameters of trajectory for functional unit $l$, where $l \in \Lambda$
 \item $\psi$ : \{1, ..., K\} $ \rightarrow Lambda$ is a function that maps beach biomarker to its corresponding functional unit
 \item $\beta_i$: time shift parameter for subject $i$ (disease specific)
 \item $\beta_i^{l}$: time shift parameter for subject $i$ used in functional unit $l$, where $l \in \Lambda$
 \item $\Omega$: set ${(i,j,k)}$ of measurements available from every subject $i$, visit $j$ in biomarker $k$
\end{itemize}

Likelihood for one single measurement in one subject visit: 
\begin{equation}
 p(y_{ijk}|\theta_k, \lambda^{\psi(k)}, \beta_i) = \sum_{\beta_i^{\psi(k)}} p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) p(\beta_i^{\psi(k)}| \lambda^{\psi(k)}, \beta_i)
\end{equation}

where $\beta_i^{\psi(k)}$ is a latent variable denoting the stage of subject $i$ in functional unit $\psi(k)$, where biomarker $k$ was assigned.

Extending the above to multiple, subjects, visits and biomarkers, we get the final model likelihood:
\begin{equation}
 p(y_{.,.,.}|\theta_1, ..., \theta_K, \{\lambda^{\psi(l)} | l \in \Lambda \}, \beta_1, ..., \beta_N) = \prod_{(i,j,k) \in \Omega} \sum_{\beta_i^{\psi(k)}} p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) p(\beta_i^{\psi(k)}| \lambda^{\psi(k)}, \beta_i)
\end{equation}

Modelling of $p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k)$ and $p(\beta_i^{\psi(k)}| \lambda^{\psi(k)}, \beta_i)$ will be done with a non-parametric model (Gaussian Process, Lorenzi et al., 2017)

\subsection*{Application 1: Disease Knowledge Transfer}

Let's assume we have MRI and PET data in tAD (from ADNI), but only MRI data for another disease, e.g. PCA. We can model the disease-agnostic region-specific relationship between MRI and PET, and also build the disease specific progression models for tAD (using MRI and PET) and PCA (using MRI only). Finally, our model allows us to propagate the knowledge of PET dynamics in tAD to PCA subjects, without using any PET scans from PCA patients. This can be done by maximising the following likelihood (assuming we already have a model for tAD):

\begin{equation}
 p(y_{.,.,.}^{PCA}|\theta_1^{tAD}, ..., \theta_K^{tAD}, \{\lambda^{\psi(l)^{PCA}} | l \in \Lambda \}, \beta_1^{PCA}, ..., \beta_N^{PCA}) = \prod_{(i,j,k) \in \Omega} \sum_{\beta_i^{\psi(k)}} p(y_{ijk}^{PCA}| \beta_i^{\psi(k)}, \theta_k^{tAD}) p(\beta_i^{\psi(k)}| \lambda^{\psi(k)^{PCA}}, \beta_i^{PCA})
\end{equation}

\subsection*{Application 2: Improved biomarker predictions and staging}

We also hope to gain improved predictions and subject staging both in all the datasets used, both the larger ones as well as the smaller ones. We can fit the models for every functional unit using all the data available in order to learn the disease-agnostic correlations between biomarkers, and then fit the disease specific progression models. Our belief is that, by using more data to inform the disease-agnostic correlations, we will get better predictions.

\end{document}

### 2018-02-26 DisProgBuilder.py

@@ -1,0 +1,1 @@
+../diffEqModel/DisProgBuilder.py

### 2018-02-26 GPHierModel.py

@@ -1,0 +1,221 @@
+import os
+import sys
+import numpy as np
+import math
+from aux import *
+import scipy
+#import scipy.cluster.hierarchy
+import pickle
+import gc
+from matplotlib import pyplot as pl
+import Plotter
+import copy
+import colorsys
+
+import MarcoModel
+
+import DisProgBuilder
+
+class GPHierBuilder(DisProgBuilder.DPMBuilder):
+  # builds a voxel-wise disease progression model
+
+  def __init__(self, plotTrajParams):
+    self.plotterObj = Plotter.PlotterJDM(plotTrajParams)
+
+  def setPlotter(self, plotterObj):
+    self.plotterObj = plotterObj
+
+  def generate(self, dataIndices, expName, params):
+    return JointModel(dataIndices, expName, params, self.plotterObj)
+
+class GPHierModel(DisProgBuilder.DPMInterface):
+
+  def __init__(self, dataIndices, expName, params, plotterObj):
+    self.dataIndices = dataIndices
+    self.expName = expName
+    self.params = params
+    self.outFolder = params['outFolder']
+    os.system('mkdir -p %s' % self.outFolder)
+    self.params['plotTrajParams']['outFolder'] = self.outFolder
+    self.plotterObj = plotterObj
+    self.nrBiomk = len(params['X'])
+    self.nrFuncUnits = params['nrFuncUnits']
+    self.mapBiomkToFuncUnits = params['mapBiomkToFuncUnits']
+
+    self.gpModels = None
+    self.grandModel = None
+
+
+  def runStd(self, runPart):
+    self.run(runPart)
+
+  def run(self, runPart):
+    filePath = '%s/unitModels.npz' % self.outFolder
+    if runPart[0] == 'R':
+      N_global_iterations = 50
+      iterParams = 50
+      iterShifts = 30
+      N = int(10)  # Number of random features for kernel approximation
+      Xfilt, Yfilt = filterDataListFormat(self.params, self.dataIndices)
+
+      # print(adsa)
+
+      self.gpModels = [_ for _ in range(self.nrFuncUnits)]
+
+      # functional units
+      for u in range(self.nrFuncUnits):
[…]

### 2018-02-26 GPModel.py

@@ -1,0 +1,74 @@
+import numpy as np
+import scipy as sp
+
+class GPModel:
+
+  def __init__(self, params):
+    self.etaB = params['etaB']
+    self.lB = params['lB']
+    self.epsB = params['epsB']
+    self.sigmaSB = params['sigmaSB']
+
+    self.sigmaGfunc = params['sigmaGfunc']
+    self.nrBiomk = len(params['etaB'])
+
+
+  def predPopBiomk(self, dps, b):
+    '''
+    Predict population-level effect f_b (X) evaluated at dps (X) for biomarker b
+
+    :param dps: 1D disease progression scores for all subjects and visits, or X
+    :param b: biomarker nr
+    :return: Y = f_b (X)
+    '''
+
+    sigmaGdps = self.sigmaGfunc(dps, nuB[b], lB[b])
+    print(sigmaGdps.shape)
+    meanGP = np.zeros(sigmaGdps.shape[0], float) # set mean of GP to zero
+    fDps = np.random.multivariate_normal(meanGP, sigmaGdps)
+
+    return fDps
+
+  def predPop(self, dps):
+    '''
+    Predict population-level effect f(X) = [f_1(X) .. f_n(X)] evaluated at dps (X) for all biomarkers
+
+    :param dps: 1D disease progression scores for all subjects and visits, or X
+    :return: Y = f(X)
+    '''
+
+    fDpsCrossSB = np.zeros((dps.shape[0], self.nrBiomk), float)
+
+    for b in range(self.nrBiomk):
+      fDpsCrossSB = self.predPopBiomk(dps, b)
+
+    return fDpsCrossSB
+
+  def predIndiv(self, dps):
+    '''
+    Predict individual-level effect nu(X)  evaluated at dps (X) for all biomarkers
+
+    :param dps: 1D disease progression scores for all subjects and visits, or X
+    :return: Y = f(X)
+    '''
+
+    return 0
+
+  def pred(self, dps):
+    '''
+    Predict full model f(X)+nu(X) evaluated at dps (X), for all biomarkers
+
+    :param dps: 1D disease progression scores for all subjects and visits, or X
+    :return: Y = f(X) + nu(X)
+    '''
+
+    return self.predPop(dps) + self.predIndiv(dps)
+
+
+def genSigmaG(ts, nuB, lb):
+
[…]

### 2018-02-26 IncompleteModel.py

@@ -1,0 +1,190 @@
+import os
+import sys
+import numpy as np
+import math
+from aux import *
+import scipy
+#import scipy.cluster.hierarchy
+import pickle
+import gc
+from matplotlib import pyplot as pl
+import Plotter
+import copy
+import colorsys
+
+import MarcoModel
+import JointModel
+import DisProgBuilder
+
+class IncompleteBuilder(DisProgBuilder.DPMBuilder):
+  # builds a voxel-wise disease progression model
+
+  def __init__(self, plotTrajParams, gpModels):
+    self.plotterObj = Plotter.PlotterJDM(plotTrajParams)
+    self.gpModels = gpModels
+
+  def setPlotter(self, plotterObj):
+    self.plotterObj = plotterObj
+
+  # set complete model
+  def setGPModels(self, gpModels):
+    self.gpModels = gpModels
+
+  def generate(self, dataIndices, expName, params):
+    return IncompleteModel(dataIndices, expName, params, self.plotterObj, self.gpModels)
+
+class IncompleteModel(JointModel.JointModel):
+
+  def __init__(self, dataIndices, expName, params, plotterObj, gpModels):
+    self.dataIndices = dataIndices
+    self.expName = expName
+    self.params = params
+    self.outFolder = params['outFolder']
+    os.system('mkdir -p %s' % self.outFolder)
+    self.params['plotTrajParams']['outFolder'] = self.outFolder
+    self.plotterObj = plotterObj
+    self.nrBiomk = len(params['X'])
+    self.nrFuncUnits = params['nrFuncUnits']
+    self.mapBiomkToFuncUnits = params['mapBiomkToFuncUnits']
+
+    self.gpModels = gpModels
+    self.grandModel = None
+
+
+  def runStd(self, runPart):
+    self.run(runPart)
+
+  def run(self, runPart):
+
+    unitsStagingFile = '%s/unitsStaging.npz' % self.outFolder
+    if runPart[0] == 'R':
+      meanStagesUS = [0 for x in range(self.nrFuncUnits)]#
+      xsU = [0 for x in range(self.nrFuncUnits)]
+      for u in range(self.nrFuncUnits):
+        deltaX = (self.gpModels[u].maxX - self.gpModels[u].minX)
[…]

### 2018-02-26 JMIter.py

@@ -1,0 +1,308 @@
+import os
+import sys
+import numpy as np
+import math
+from aux import *
+import scipy
+#import scipy.cluster.hierarchy
+import pickle
+import gc
+from matplotlib import pyplot as pl
+import Plotter
+import copy
+import colorsys
+
+import MarcoModel
+
+import DisProgBuilder
+import JointModel
+
+class JMIterBuilder(DisProgBuilder.DPMBuilder):
+  # builds a Joint Disease model
+
+  def __init__(self, plotTrajParams):
+    self.plotterObj = Plotter.PlotterJDM(plotTrajParams)
+
+  def setPlotter(self, plotterObj):
+    self.plotterObj = plotterObj
+
+  def generate(self, dataIndices, expName, params):
+    return JMIter(dataIndices, expName, params, self.plotterObj)
+
+class JMIter(DisProgBuilder.DPMInterface):
+
+  def __init__(self, dataIndices, expName, params, plotterObj):
+    self.dataIndices = dataIndices
+    self.expName = expName
+    self.params = params
+    self.outFolder = params['outFolder']
+    os.system('mkdir -p %s' % self.outFolder)
+    self.params['plotTrajParams']['outFolder'] = self.outFolder
+    self.params['plotTrajParams']['expName'] = expName
+    self.plotterObj = plotterObj
+    self.nrBiomk = len(params['X'])
+    self.nrFuncUnits = params['nrFuncUnits']
+    self.mapBiomkToFuncUnits = params['mapBiomkToFuncUnits']
+
+    self.unitModels = None # functional unit models
+    self.disModels = None # disease specific models
+
+
+  def runStd(self, runPart):
+    self.run(runPart)
+
+  def run(self, runPart):
+    filePath = '%s/unitModels.npz' % self.outFolder
+    disModelsFile = '%s/disModels.npz' % self.outFolder
+
+    if runPart[0] == 'R':
+      nrGlobIterUnits = 10
+      iterParamsUnits = 80
+      iterShifts = -1 # doesn't matter anymore
+      nrGlobIterDIs = 20
+      iterParamsDis = 80
+
+      N = int(3)  # Number of random features for kernel approximation
[…]

### 2018-02-26 JointModel.py

@@ -1,0 +1,338 @@
+import os
+import sys
+import numpy as np
+import math
+from aux import *
+import scipy
+#import scipy.cluster.hierarchy
+import pickle
+import gc
+from matplotlib import pyplot as pl
+import Plotter
+import copy
+import colorsys
+
+import MarcoModel
+
+import DisProgBuilder
+
+class JMDBuilder(DisProgBuilder.DPMBuilder):
+  # builds a Joint Disease model
+
+  def __init__(self, plotTrajParams):
+    self.plotterObj = Plotter.PlotterJDM(plotTrajParams)
+
+  def setPlotter(self, plotterObj):
+    self.plotterObj = plotterObj
+
+  def generate(self, dataIndices, expName, params):
+    return JointModel(dataIndices, expName, params, self.plotterObj)
+
+class JointModel(DisProgBuilder.DPMInterface):
+
+  def __init__(self, dataIndices, expName, params, plotterObj):
+    self.dataIndices = dataIndices
+    self.expName = expName
+    self.params = params
+    self.outFolder = params['outFolder']
+    os.system('mkdir -p %s' % self.outFolder)
+    self.params['plotTrajParams']['outFolder'] = self.outFolder
+    self.params['plotTrajParams']['expName'] = expName
+    self.plotterObj = plotterObj
+    self.nrBiomk = len(params['X'])
+    self.nrFuncUnits = params['nrFuncUnits']
+    self.mapBiomkToFuncUnits = params['mapBiomkToFuncUnits']
+
+    self.unitModels = None # functional unit models
+    self.disModels = None # disease specific models
+
+
+  def runStd(self, runPart):
+    self.run(runPart)
+
+  def run(self, runPart):
+    filePath = '%s/unitModels.npz' % self.outFolder
+    if runPart[0] == 'R':
+      nrGlobIter = 10
+      iterParams = 80
+      iterShifts = 100
+      N = int(3)  # Number of random features for kernel approximation
+      Xfilt, Yfilt = filterDataListFormat(self.params, self.dataIndices)
+
+      # print(adsa)
+
+      self.unitModels = [_ for _ in range(self.nrFuncUnits)]
+
+      # functional units
+      for u in range(self.nrFuncUnits):
[…]

### 2018-02-26 MarcoDataGenerator.py

@@ -1,0 +1,103 @@
+__author__ = 'mlorenzi'
+import numpy as np
+import matplotlib.pyplot as plt
+
+
+class DataGenerator(object):
+    def __init__(self, Nbiom, interval, param, Nsubs):
+        self.Nbiom = Nbiom
+        self.YData = []
+        self.YDataNoNoise = []
+        self.XData = []
+        self.model = []
+        self.Nsubs = Nsubs
+        self.interval = interval
+        self.param = param
+        self.shiftData = []
+
+        np.random.seed(1)
+
+        for i in range(Nbiom):
+            shift = np.random.randint(interval[0], interval[1])
+            self.shiftData.append([shift + l for l in range(np.int(np.float(interval[0])/2), np.int(np.float(interval[1])/2))])
+            self.model.append(self.f(self.shiftData[i], param[i][0], param[i][1]))
+            #self.model.append(self.f(range(interval[0], interval[1]), param[i][0], param[i][1]))
+
+        self.XData.append([])
+        for k in range(self.Nsubs):
+            start = np.random.randint(interval[1] - 5)
+            Nel = np.random.randint(1, 5)
+            # sequence = np.sort(random.sample(range(5), Nel))
+            sequence = np.sort(np.random.choice(range(5), Nel, replace=False))
+            Xd = [l + start for l in sequence]
+            self.XData[0].append(Xd)
+
+        for i in range(Nbiom-1):
+            self.XData.append([])
+            self.XData[i+1] = self.XData[0]
+
+        for i in range(Nbiom):
+            self.YData.append([])
+            self.YDataNoNoise.append([])
+            for l in range(self.Nsubs):
+                obs = np.array(self.model[i])[self.XData[i][l]] + param[i][2] * np.random.randn(len(self.XData[i][l]))
+                self.YData[i].append(obs)
+                self.YDataNoNoise[i].append(np.array(self.model[i])[self.XData[i][l]] )
+
+        self.ZeroXData = []
+        for i in range(self.Nbiom):
+            self.ZeroXData.append([])
+            for k in range(self.Nsubs):
+                self.ZeroXData[i].append(np.array(self.XData[i][k]) - np.float (self.XData[i][k][len(self.XData[i][k])-1]+self.XData[i][k][0])/2)
+
+
+    def f(self, X, L, k):
+        return [L / (1 + np.exp(-k * i)) for i in X]
+
+    def predPop(self, X, b):
[…]

### 2018-02-26 MarcoModel.py

@@ -1,0 +1,1124 @@
+### Gaussian process-based disease progression modeling and time-shift estimation.
+#
+#  - The software iteratively estimates monotonic progressions for each biomarker and realigns the individual observations in time
+#   Basic usage:
+#
+#       model = GP_progression_model.GP_progression_model(input_X,input_N,N_random_features)
+#
+#   X and Y should be A list of biomarkers arrays. Each entry "i" of the list is a list of individuals' observations for the biomarker i
+#   The monotonicity is enforced by the parameter self.penalty (higher -> more monotonic)
+#
+# - The class comes with an external method for transforming a given .csv file in the required input X and Y:
+#
+#       X,Y,list_biomarker = GP_progression_model.convert_csv(file_path)
+#
+# - The method Save(folder_path) saves the model parameters to an external folder, that can be subsequently read with the
+# method Load(folder_path)
+#
+# - Optimization can be done with the method Optimize:
+#
+#       model.Optimize()
+#
+# This software is based on the publication:
+#
+# Disease Progression Modeling and Prediction through Random Effect Gaussian Processes and Time Transformation
+# Marco Lorenzi, Maurizio Filippone, Daniel C. Alexander, Sebastien Ourselin
+# arXiv:1701.01668
+#
+# Gaussian process regression based on random features approximations is based on the paper:
+#
+# Random Feature Expansions for Deep Gaussian Processes (ICML 2017, Sydney)
+# K. Cutajar, E. V. Bonilla, P. Michiardi, M. Filippone
+# arXiv:1610.04386
+
+
+import pandas as pd
+import matplotlib.pyplot as plt
+import numpy as np
+import sys
+#from autograd import grad
+#import autograd.numpy as np
+#import autograd.numpy.random as npr
+#from autograd import value_and_grad
+import scipy.optimize
+
+class GP_progression_model(object):
+    plt.interactive(False)
+    def __init__(self, X,Y, N_rnd_features, outFolder, plotter, names_biomarkers = [], group = []):
+
+        #Initializing variables
+        self.plotter = plotter
+        self.outFolder = outFolder
+        self.expName = plotter.plotTrajParams['expName']
+        self.names_biomarkers = names_biomarkers
+        self.group = group
[…]

### 2018-02-26 MarcoModelWrapper.py

@@ -1,0 +1,91 @@
+import os
+import sys
+import numpy as np
+import math
+from aux import *
+import scipy
+#import scipy.cluster.hierarchy
+import pickle
+import gc
+from matplotlib import pyplot as pl
+import Plotter
+
+import MarcoModel
+
+import DisProgBuilder
+
+class MarcoModelBuilder(DisProgBuilder.DPMBuilder):
+  # builds a voxel-wise disease progression model
+
+  def __init__(self, plotTrajParams):
+    self.plotterObj = Plotter.PlotterGP(plotTrajParams)
+
+  def setPlotter(self, plotterObj):
+    self.plotterObj = plotterObj
+
+  def generate(self, dataIndices, expName, params):
+    return MarcoModelWrapper(dataIndices, expName, params, self.plotterObj)
+
+class MarcoModelWrapper(DisProgBuilder.DPMInterface):
+
+  def __init__(self, dataIndices, expName, params, plotterObj):
+    self.dataIndices = dataIndices
+    self.expName = expName
+    self.params = params
+    self.outFolder = params['outFolder']
+    os.system('mkdir -p %s' % self.outFolder)
+    self.params['plotTrajParams']['outFolder'] = self.outFolder
+    self.plotterObj = plotterObj
+
+
+    Xfilt, Yfilt = filterDataListFormat(params, dataIndices)
+
+    #changed to 2 for testing!!
+    N = int(10)  # Number of random features for kernel approximation
+    self.gpModel = MarcoModel.GP_progression_model(Xfilt,Yfilt, N, self.outFolder, plotterObj, self.params['labels'])
+
+    # set penalty for decreasing trajectories
+    self.gpModel.Set_penalty(self.params['penalty'])
+
+  def runStd(self, runPart):
+    self.run(runPart)
+
+  def run(self, runPart):
+    filePath = '%s/fittedGPModel.npz' % self.outFolder
+    if runPart[0] == 'R':
+      N_global_iterations = 10
+      iterParams = 80
+      iterShifts = 50
+      self.gpModel.Optimize(N_global_iterations, [iterParams,iterShifts], Plot=True)
+      pickle.dump(self.gpModel, open(filePath, 'wb'), protocol = pickle.HIGHEST_PROTOCOL)
+    else:
+      self.gpModel = pickle.load(open(filePath, 'rb'))
+
+
+
+    res = None
+    return res
+
[…]

### 2018-02-26 MarcoTestADNI.py

@@ -1,0 +1,66 @@
+#!/usr/bin/env python
+
+import numpy as np
+import sys
+sys.path.append('../')
+import MarcoModel
+import matplotlib.pyplot as plt
+# import MarcoDataGenerator
+import Plotter
+import os
+import pickle
+import colorsys
+
+# remember to change test_synthetic/test_ADNI/test_TADPOLE in GP_progression_model
+
+### load the biomarkers
+X,Y,RID,list_biomarker = MarcoModel.convert_csv("./table_APOEpos.csv")
+
+N=int(10)   # Number of random features for kernel approximation
+nrBiomk = len(list_biomarker)
+
+### optimise the model
+plotTrajParams = dict(nrRows=2,nrCols=3, SubfigTrajWinSize=(1600,900), isSynth=False, labels=list_biomarker)
+plotTrajParams['colorsTraj'] = [colorsys.hsv_to_rgb(hue, 1, 1) for hue in np.linspace(0, 1, num=nrBiomk, endpoint=False)]
+plotTrajParams['allTrajOverlap'] = False
+gpPlotter = Plotter.PlotterGP(plotTrajParams)
+outFolder = 'resfiles/marcoADNI'
+os.system('mkdir -p %s' % outFolder)
+gp  = MarcoModel.GP_progression_model(X,Y,N, outFolder, gpPlotter, list_biomarker)
+
+N_global_iterations = 100
+iterParams = 50
+iterShifts = 30
+runPart = ['R', 'R']
+gpFile = '%s/fittedGPModel.npz' % outFolder
+if runPart[0] == 'R':
+  gp.Optimize(N_global_iterations, [iterParams,iterShifts], Plot = True)
+
+  pickle.dump(gp, open(gpFile, 'wb'), protocol = pickle.HIGHEST_PROTOCOL)
+else:
+  gp = pickle.load(open(gpFile, 'rb'))
+  gp.plotter.plotTraj(gp, replaceFig=False)
+
+
+### prediction
+gpPredFile = '%s/preds.npz' % outFolder
+Xrange = np.linspace(gp.minX, gp.maxX, 30).reshape([30, 1])
+if runPart[1] == 'R':
+  pred_prob, pred_exp = gp.StageSubjects(X,Y,Xrange)
+  datastruct = dict(pred_prob=pred_prob, pred_exp=pred_exp)
[…]

### 2018-02-26 MarcoTestSynthetic.py

@@ -1,0 +1,224 @@
+#!/usr/bin/env python
+
+import numpy as np
+import sys
+sys.path.append('../')
+import MarcoModel
+import MarcoDataGenerator
+import matplotlib.pyplot as plt
+import Plotter
+import os
+import pickle
+import colorsys
+
+import numpy as np
+
+np.random.seed(1)
+
+# Creating synthetic multivariate progressions data with monotonic behaviour
+L = 1
+k = 0.3
+interval = [-15,15]
+
+# Number of biomarkers
+nrBiomk = 9
+# Number of individuals
+nrSubj = 300
+# Gaussian observational noise
+noise = 0.05
+
+flag = 0
+while (flag!=1):
+    CurveParam = []
+    for i in range(nrBiomk):
+        CurveParam.append([L,0.8*np.random.rand(),noise])
+        if CurveParam[i][1] > 0.0:
+            flag = 1
+
+dg = MarcoDataGenerator.DataGenerator(nrBiomk, interval, CurveParam, nrSubj)
+
+# Number of random features for kernel approximation
+N=int(10)
+
+X = []
+Y = []
+
+for biom in range(len(dg.ZeroXData)):
+    X.append([])
+    Y.append([])
+    for sub in range(len(dg.ZeroXData[biom])):
+        # X[biom].append([0])
+        # Y[biom].append([dg.YData[biom][sub][0]])
+        X[biom].append(dg.XData[biom][sub] - dg.XData[biom][sub][0])
+        Y[biom].append(dg.YData[biom][sub])
+
+
+
+#gp  = GP_progression_modelSYN.GP_progression_model(dg.ZeroXData,dg.YData,N)
+labels = ['b%d' % b for b in range(nrBiomk)]
+plotTrajParams = dict(nrRows=3,nrCols=4, SubfigTrajWinSize=(1600,900), isSynth=True, labels=labels)
+plotTrajParams['colorsTraj'] = [colorsys.hsv_to_rgb(hue, 1, 1) for hue in np.linspace(0, 1, num=nrBiomk, endpoint=False)]
+plotTrajParams['allTrajOverlap'] = False
+gpPlotter = Plotter.PlotterGP(plotTrajParams)
+outFolder = 'resfiles/marcoSynth'
+os.system('mkdir -p %s' % outFolder)
+
[…]

### 2018-02-26 ParHierModel.py

@@ -1,0 +1,141 @@
+import numpy as np
+import scipy as sp
+
+class ParHierModel:
+
+  def __init__(self, dysfuncParams, thetas, mapBiomkToFuncUnits, parFunc, sigmaB):
+    self.dysfuncParams = dysfuncParams
+    self.thetas = thetas
+    self.mapBiomkToFuncUnits = mapBiomkToFuncUnits
+    self.parFunc = parFunc # parametric function
+    self.sigmaB = sigmaB
+    self.nrBiomk = thetas.shape[0]
+    self.nrFuncUnits = np.unique(mapBiomkToFuncUnits).shape[0]
+
+    # self.subShiftsLongTrue = subShiftsLongTrue
+    # self.subShiftsCrossTrue = subShiftsLongTrue[long2crossInd]
+
+
+  def timeShiftFuncCross(self, ageAtVisitCross):
+
+    return ageAtVisitCross + self.subShiftsCrossTrue
+
+
+  def predPopBiomk(self, dpsCross, b):
+    '''
+    Predict population-level effect f_b (X) evaluated at dps (X) for biomarker b
+
+    :param dpsCross: 1D disease progression scores for all subjects and visits, or X
+    :param b: biomarker nr
+    :return: Y = f_b (X)
+    '''
+
+
+    # find dysfunction scores for each subject
+    dysScoresS = self.parFunc(dpsCross, self.dysfuncParams[self.mapBiomkToFuncUnits[b],:])
+    modelPredS = self.parFunc(dysScoresS, self.thetas[b,:])
+
+    # print('self.dysfuncParams', self.dysfuncParams)
+    print('dpsCross', dpsCross)
+    print('dysScoresS', dysScoresS)
+    print('params', self.dysfuncParams[self.mapBiomkToFuncUnits[b],:])
+    print('modelPredS', modelPredS)
+    # print(ads)
+
+    return modelPredS
+
+  def predPopDys(self, dpsCross):
+    '''
+    Predict population-level effect f_b (X) evaluated at dps (X) for biomarker b
+
+    :param dpsCross: 1D disease progression scores for all subjects and visits, or X
+    :param b: biomarker nr
+    :return: Y = f_b (X)
+    '''
+
+    dysScoresSU = np.zeros((dpsCross.shape[0], self.nrFuncUnits), float)
+
+    # find dysfunction scores for each subject
+    for u in range(self.nrFuncUnits):
[…]

### 2018-02-26 Plotter.py

@@ -1,0 +1,634 @@
+from matplotlib import pyplot as pl
+import numpy as np
+import sys
+import colorsys
+from env import *
+
+class PlotterJDM:
+
+  def __init__(self, plotTrajParams):
+    self.plotTrajParams = plotTrajParams
+
+  #
+  # def plotTrajData(self, longData, longDiag, longDPS, model,
+  #   replaceFigMode=True, yLimUseData=False,showConfInt=False, adjustBottomHeight=0.1):
+  #   figSizeInch = (self.plotTrajParams['SubfigTrajWinSize'][0] / 100, self.plotTrajParams['SubfigTrajWinSize'][1] / 100)
+  #   fig = pl.figure(1, figsize=figSizeInch)
+  #   pl.clf()
+  #   nrRows = self.plotTrajParams['nrRows']
+  #   nrCols = self.plotTrajParams['nrCols']
+  #
+  #   nrBiomk = longData[0].shape[1]
+  #
+  #   minX = np.min([np.min(dpsCurr) for dpsCurr in longDPS])
+  #   maxX = np.max([np.max(dpsCurr) for dpsCurr in longDPS])
+  #
+  #   xs = np.linspace(minX, maxX, 100)
+  #   diagNrs = self.plotTrajParams['diagNrs']
+  #
+  #   nrSubjLong = len(longData)
+  #
+  #   dysScoresSF = model.predPopDys(xs)
+  #   modelPredSB = model.predPop(xs)
+  #
+  #   print('xs', xs)
+  #   print('modelPredSB[:,0]', modelPredSB[:,0])
+  #
+  #   lw = 3.0
+  #
+  #   # first plot the dysfunctionality biomarkers
+  #   ax = pl.subplot(nrRows, nrCols, 1)
+  #   ax.set_title('dysfunc all')
+  #   moveTicksInside(ax)
+  #   for f in range(model.nrFuncUnits):
+  #     ax.plot(xs, dysScoresSF[:,f], 'k-', linewidth=lw)
+  #
+  #   # for each unit, plot all biomarkers against the dysfunctional scores
+  #   for f in range(model.nrFuncUnits):
+  #     ax2 = pl.subplot(nrRows, nrCols, f+2)
+  #     ax2.set_title('dysfunc %d' % f)
+  #     moveTicksInside(ax2)
+  #     biomkInCurrUnit = np.where(self.plotTrajParams['mapBiomkToFuncUnits'] == f)[0]
+  #     for b in range(len(biomkInCurrUnit)):
[…]

### 2018-02-26 aux.py

@@ -1,0 +1,172 @@
+
+from env import *
+import numpy as np
+import ParHierModel
+import sklearn.metrics
+
+
+
+def convert_csv(file):
+  table = pd.read_csv(file)
+
+  return convert_table_marco(table)
+
+def convert_table_marco(table, biomkStartCol=3, list_biomarkers=None):
+    X = []
+    Y = []
+
+    # list of individuals
+    list_RID = np.unique(table[['RID']])
+    print(len(list_RID))
+    # list of biomarkers
+    if list_biomarkers is None:
+      list_biomarkers = table.columns[range(biomkStartCol, len(table.columns))]
+
+    RID = []
+
+    for id_biom, biomarker in enumerate(list_biomarkers):
+        X.append([])
+        Y.append([])
+
+    # Parsing every biomarker and assigning to the list
+    print('list_RID', list_RID)
+    diagLong = []
+    # print(adsas)
+    for id_sub, sub in enumerate(list_RID):
+        flag_missing = 0
+        indices = np.where(np.in1d(table.loc[:, 'RID'], sub))[0]
+        for id_biom, biomarker in enumerate(list_biomarkers):
+            X[id_biom].append(np.array(table[['Month_bl']])[indices].flatten())
+            Y[id_biom].append(np.array(table[[biomarker]])[indices].flatten())
+
+            idx_to_keep = ~np.isnan(Y[id_biom][id_sub])
+
+            Y[id_biom][id_sub] = Y[id_biom][id_sub][idx_to_keep]
+            X[id_biom][id_sub] = X[id_biom][id_sub][idx_to_keep]
+
+            if len(Y[id_biom][id_sub]) < 1:
+                flag_missing = flag_missing + 1
+
+
+        # print(sub, monthsSinceBlCurrSub)
+        diagCurrSub = np.array(table['diag'][indices])
+        diagNNind = np.logical_not(np.isnan(diagCurrSub))
+        # print(diagCurrSub)
+        monthsSinceBlCurrSub = np.array(table['Month_bl'])[indices][diagNNind]
+
+        # if flag_missing == 0:
+        #     print('missing data for sub: ', sub)
+
+        diagExistsForAtLeastOneVisit = monthsSinceBlCurrSub.shape[0] > 0
+        if diagExistsForAtLeastOneVisit:
[…]

### 2018-02-26 env.py

@@ -1,0 +1,25 @@
+
+
+# ADNI dataset
+CTL = 1
+MCI = 2
+AD = 3
+
+# DRC dataset
+CTL2 = 4 # control diagnosis from a separate dataset as CTL
+PCA = 5
+AD2 = 6 # AD diagnosis from a separate dataset as AD
+
+
+
+
+# MORE ADNI
+EMCI = 4
+LMCI = 5
+SMC = 6
+
+DECR = 1
+INCR = 2
+
+
+

### 2018-02-26 evaluationFramework.py

@@ -1,0 +1,1 @@
+../diffEqModel/evaluationFramework.py

### 2018-02-26 genSynthData.py

@@ -1,0 +1,237 @@
+import numpy as np
+import os
+import pickle
+from GPModel import *
+from env import *
+import scipy
+import scipy.stats
+from aux import *
+import ParHierModel
+import pandas as pd
+
+import MarcoModel
+import aux
+
+
+def generateDataJMD(nrSubjLong, nrBiomk, nrTimepts, shiftsLowerLim, shiftsUpperLim, model,
+  outFolder, fileName, forceRegenerate, localParams, ctlDiagNr, patDiagNr):
+  ''' generates data from a hierarchical model of disease '''
+
+
+  if os.path.isfile(fileName) and not forceRegenerate:
+    dataStruct = pickle.load(open(fileName, 'rb'))
+    localParams = dataStruct['localParams']
+
+  else:
+    np.random.seed(1)
+    # generate subject data
+    subShiftsLongTrue = np.random.uniform(shiftsLowerLim, shiftsUpperLim, (nrSubjLong,))
+    nrSubjCross = nrTimepts * nrSubjLong
+
+    # ageAtBlScanLong = np.random.uniform(lowerAgeLim,upperAgeLim, (nrSubjLong,))
+    # ageAtScanCross = np.zeros(nrSubjCross, float)
+    # ageAtBlScanCross = np.zeros(nrSubjCross, float)
+    yearsSinceBlScanCross = np.zeros(nrSubjCross, float)
+
+    dataCrossSB = np.zeros((nrSubjCross, nrBiomk), float)
+    subShiftsCrossTrue = np.zeros(nrSubjCross, float)
+
+    partCodeCross = np.zeros(nrSubjCross, float)
+    partCodeLong = np.array(range(nrSubjLong)) # unique id for every participant
+    scanTimeptsCross = np.zeros(nrSubjCross, float)
+
+    counter = 0
+    long2crossInd = np.zeros(nrSubjCross, int)
+    cross2longInd = [0 for s in range(nrSubjLong)]
+
+    for s in range(nrSubjLong):
+
+      cross2longInd[s] = np.array(range(counter,counter+nrTimepts))
+
+      for tp in range(nrTimepts):
+        # get currTimept, age at curr Timepints, and partCodeCross
+        partCodeCross[counter] = partCodeLong[s]
+        scanTimeptsCross[counter] = tp
+        yearsSinceBlScanCross[counter] = tp
[…]

### 2018-02-26 hierSynth.py

@@ -1,0 +1,305 @@
+import sys
+import numpy
+import numpy as np
+import colorsys
+from socket import gethostname
+import time
+import argparse
+import os
+import colorsys
+import copy
+
+# sys.path.append(os.path.abspath("../diffEqModel/"))
+
+
+parser = argparse.ArgumentParser(description='Launches two diseases hierarchical model ...'
+'i.e. runs two separate models on datasets distinctively')
+
+parser.add_argument('--agg', dest='agg', type=int, default=0,
+  help='agg=1 => plot figures without using Xwindows, for use on cluster where the plots cannot be displayed '
+       ' agg=0 => plot with Xwindows (for use on personal machine)')
+
+parser.add_argument('--runIndex', dest='runIndex', type=int,
+  default=1, help='index of run instance/process .. for cross-validation')
+
+parser.add_argument('--nrProc', dest='nrProc', type=int,
+  default=1, help='# of processes')
+
+parser.add_argument('--modelToRun', dest='modelToRun', type=int,
+  help='index of model to run')
+
+parser.add_argument('--cluster', action="store_true",
+  help='need to include this flag if runnin on cluster')
+
+parser.add_argument('--nrRows', dest='nrRows', type=int,
+  help='nr of subfigure rows to plot at every iteration')
+
+parser.add_argument('--nrCols', dest='nrCols', type=int,
+  help='nr of subfigure columns to plot at every iteration')
+
+parser.add_argument('--penalty', dest='penalty', type=float,
+  help='penalty value for non-monotonic trajectories. between 0 (no effect) and 10 (strong effect). ')
+
+args = parser.parse_args()
+
+if args.agg:
+  # print(matplotlib.__version__)
+  import matplotlib
+  matplotlib.use('Agg')
+  # print(asds)
+
+import genSynthData
+import GPModel
+import ParHierModel
+import Plotter
+from aux import *
+import evaluationFramework
+
[…]

### 2018-02-26 jointSynth.py

@@ -1,0 +1,360 @@
+import sys
+import numpy
+import numpy as np
+import colorsys
+from socket import gethostname
+import time
+import argparse
+import os
+import colorsys
+import copy
+
+# sys.path.append(os.path.abspath("../diffEqModel/"))
+
+
+parser = argparse.ArgumentParser(description='Launches voxel-wise/point-wise DPM on ADNI'
+                                             'using cortical thickness maps derived from MRI')
+
+parser.add_argument('--agg', dest='agg', type=int, default=0,
+  help='agg=1 => plot figures without using Xwindows, for use on cluster where the plots cannot be displayed '
+       ' agg=0 => plot with Xwindows (for use on personal machine)')
+
+parser.add_argument('--runIndex', dest='runIndex', type=int,
+  default=1, help='index of run instance/process .. for cross-validation')
+
+parser.add_argument('--nrProc', dest='nrProc', type=int,
+  default=1, help='# of processes')
+
+parser.add_argument('--modelToRun', dest='modelToRun', type=int,
+  help='index of model to run')
+
+parser.add_argument('--cluster', action="store_true",
+  help='need to include this flag if runnin on cluster')
+
+parser.add_argument('--nrRows', dest='nrRows', type=int,
+  help='nr of subfigure rows to plot at every iteration')
+
+parser.add_argument('--nrCols', dest='nrCols', type=int,
+  help='nr of subfigure columns to plot at every iteration')
+
+parser.add_argument('--penalty', dest='penalty', type=float,
+  help='penalty value for non-monotonic trajectories. between 0 (no effect) and 10 (strong effect). ')
+
+args = parser.parse_args()
+
+if args.agg:
+  # print(matplotlib.__version__)
+  import matplotlib
+  matplotlib.use('Agg')
+  # print(asds)
+
+import genSynthData
+import GPModel
+import ParHierModel
+import Plotter
+from aux import *
[…]

### 2018-02-26 launchCommon.py

@@ -1,0 +1,94 @@
+import os
+import glob
+import sys
+import argparse
+from socket import gethostname
+from subprocess import check_output as qx
+import subprocess
+import nibabel as nib
+import numpy as np
+import csv
+from datetime import datetime
+import pickle
+
+
+def addParserArgs(parser):
+  parser.add_argument('--firstInstance', dest='firstInstance', type=int,
+                      default=1,help='index of first run instance/process')
+
+  parser.add_argument('--lastInstance', dest='lastInstance', type=int,
+                     default=10,help='index of last run instance/process')
+
+  parser.add_argument('--nrProc', dest='nrProc', type=int,
+                     default=10,help='# of processes')
+
+  parser.add_argument('--firstModel', dest='firstModel', type=int,
+                     help='index of first experiment to run')
+
+  parser.add_argument('--lastModel', dest='lastModel', type=int,
+                     help='index of last experiment to run')
+
+  parser.add_argument('--models', dest='models',
+                     help='index of first experiment to run')
+
+  parser.add_argument('--launcherScript', dest='launcherScript',
+                     help='launcher script, such as adniThick.py, etc ..')
+
+  parser.add_argument('--cluster', action="store_true", help='set this if the program should run on the CMIC cluster')
+  parser.add_argument('--timeLimit', dest='timeLimit',
+                     help='timeLimit for the job in hours')
+  parser.add_argument('--printOnly', action="store_true", help='only print experiment to be run, not actualy run it')
+
+  # specific to the VDPM
+
+
+  parser.add_argument('--nrParamsIt', dest='nrParamsIt', type=int, default=50,
+                     help='# of outer iterations to run, for estimating clustering probabilities')
+
+  parser.add_argument('--nrShiftsIt', dest='nrShiftsIt', type=int, default=30,
+                     help='# of inner iterations to run, for fitting the model parameters and subj. shifts')
+
+  parser.add_argument('--mem', dest='mem', type=int, default=15,
+                     help='memory limit of process in GB')
+
+  parser.add_argument('--agg', dest='agg', type=int, default=0,
[…]

### 2018-02-26 launchSynth.py

@@ -1,0 +1,108 @@
+from launchCommon import *
+
+
+parser = argparse.ArgumentParser(
+  description=r'''Launches in synthetic experiments with different parameters.
+  Scripts can run in three modes: parallel, serial or on cluster
+
+  ''')
+
+parser.add_argument('--penalty', dest='penalty', type=int,
+  help='penalty value for non-monotonic trajectories. between 0 (no effect) and 10 (strong effect). ')
+
+
+addParserArgs(parser)
+
+args = parser.parse_args()
+
+launchParams = initCommonLaunchParams(args)
+
+def getRunCmd(model, runIndex, nrClust, agg):
+  """
+  :param model:
+  :param runIndex:
+  :param nrClust:
+  :param agg: if 1 then plot without Xwindows (for cluster)
+              if 0 then plot with Xwindows (for personal machine)
+  :return:
+  runCmdList command representation as a list of pairs
+  runCmdStr command representatition as a string
+  """
+  runCmdList = [('python3', args.launcherScript),
+    ('--runIndex', runIndex),
+    ('--nrProc',  args.nrProc),
+    ('--modelToRun', model),
+    ('--nrOuterIt', args.nrOuterIt),
+    ('--nrInnerIt', args.nrInnerIt), ('--penalty', penalty),
+
+  ]
+
+  runCmdStr = ' '.join(['%s %s' % (s,str(a)) for (s,a) in runCmdList])
+
+  return runCmdList, runCmdStr
+
+def getQsubCmd(model, runIndex, nrClust):
+  ''' Creates full command for cluster'''
+  # if there's an error about tty, add & in the last parameter
+  runCmd = ('cd %s; /share/apps/python-3.5.1/bin/' %
+            launchParams['REPO_DIR']) + \
+           getRunCmd(model,runIndex, nrClust, agg=1)[1]
+  runCmd += ' --cluster '
+  jobName = "%s_i%d_m%d_c%d_%s_r%.1f_p%d" % (args.launcherScript.split('.')[0],
+    runIndex,model,nrClust,args.initClustering,args.rangeFactor,
+    args.informPrior)
+  qsubCmd = getQsubCmdPart2(launchParams, jobName)
+  cmdClust = 'echo "%s" | %s' % (runCmd, qsubCmd) # echo the matlab cmd then pipe it to qsub
+  return cmdClust, runCmd
+
+pList = []
+#instanceIndices = [2,3,4,5,6,7,8];
+instanceIndices = range(args.firstInstance, args.lastInstance+1)
[…]

### 2018-02-26 simpleSynth.py

@@ -1,0 +1,230 @@
+import sys
+import numpy
+import numpy as np
+import colorsys
+from socket import gethostname
+import time
+import argparse
+import os
+import colorsys
+
+# sys.path.append(os.path.abspath("../diffEqModel/"))
+
+
+parser = argparse.ArgumentParser(description='Launches voxel-wise/point-wise DPM on ADNI'
+                                             'using cortical thickness maps derived from MRI')
+
+parser.add_argument('--agg', dest='agg', type=int, default=0,
+  help='agg=1 => plot figures without using Xwindows, for use on cluster where the plots cannot be displayed '
+       ' agg=0 => plot with Xwindows (for use on personal machine)')
+
+parser.add_argument('--runIndex', dest='runIndex', type=int,
+  default=1, help='index of run instance/process .. for cross-validation')
+
+parser.add_argument('--nrProc', dest='nrProc', type=int,
+  default=1, help='# of processes')
+
+parser.add_argument('--modelToRun', dest='modelToRun', type=int,
+  help='index of model to run')
+
+parser.add_argument('--cluster', action="store_true",
+  help='need to include this flag if runnin on cluster')
+
+parser.add_argument('--nrRows', dest='nrRows', type=int,
+  help='nr of subfigure rows to plot at every iteration')
+
+parser.add_argument('--nrCols', dest='nrCols', type=int,
+  help='nr of subfigure columns to plot at every iteration')
+
+parser.add_argument('--penalty', dest='penalty', type=float,
+  help='penalty value for non-monotonic trajectories. between 0 (no effect) and 10 (strong effect). ')
+
+args = parser.parse_args()
+
+if args.agg:
+  # print(matplotlib.__version__)
+  import matplotlib
+  matplotlib.use('Agg')
+  # print(asds)
+
+import genSynthData
+import GPModel
+import ParHierModel
+import Plotter
+from aux import *
+import evaluationFramework
+
+hostName = gethostname()
[…]

### 2018-02-26 tadpoleDrc.py

@@ -1,0 +1,1104 @@
+import sys
+import numpy
+import numpy as np
+import colorsys
+from socket import gethostname
+import time
+import argparse
+import os
+import colorsys
+import copy
+import pandas as pd
+import pickle
+import random
+import aux
+import scipy
+import scipy.io as sio
+import scipy.stats
+
+# sys.path.append(os.path.abspath("../diffEqModel/"))
+
+
+parser = argparse.ArgumentParser(description='Launches voxel-wise/point-wise DPM on ADNI'
+                                             'using cortical thickness maps derived from MRI')
+
+parser.add_argument('--agg', dest='agg', type=int, default=0,
+  help='agg=1 => plot figures without using Xwindows, for use on cluster where the plots cannot be displayed '
+       ' agg=0 => plot with Xwindows (for use on personal machine)')
+
+parser.add_argument('--runIndex', dest='runIndex', type=int,
+  default=1, help='index of run instance/process .. for cross-validation')
+
+parser.add_argument('--nrProc', dest='nrProc', type=int,
+  default=1, help='# of processes')
+
+parser.add_argument('--modelToRun', dest='modelToRun', type=int,
+  help='index of model to run')
+
+parser.add_argument('--cluster', action="store_true",
+  help='need to include this flag if runnin on cluster')
+
+parser.add_argument('--nrRows', dest='nrRows', type=int,
+  help='nr of subfigure rows to plot at every iteration')
+
+parser.add_argument('--nrCols', dest='nrCols', type=int,
+  help='nr of subfigure columns to plot at every iteration')
+
+parser.add_argument('--penalty', dest='penalty', type=float,
[…]

### 2018-02-26 tadpoleOnly.py

@@ -1,0 +1,1032 @@
+import sys
+import numpy
+import numpy as np
+import colorsys
+from socket import gethostname
+import time
+import argparse
+import os
+import colorsys
+import copy
+import pandas as pd
+import pickle
+import random
+import aux
+import scipy
+import scipy.io as sio
+import scipy.stats
+
+# sys.path.append(os.path.abspath("../diffEqModel/"))
+
+
+parser = argparse.ArgumentParser(description='Launches voxel-wise/point-wise DPM on ADNI'
+                                             'using cortical thickness maps derived from MRI')
+
+parser.add_argument('--agg', dest='agg', type=int, default=0,
+  help='agg=1 => plot figures without using Xwindows, for use on cluster where the plots cannot be displayed '
+       ' agg=0 => plot with Xwindows (for use on personal machine)')
+
+parser.add_argument('--runIndex', dest='runIndex', type=int,
+  default=1, help='index of run instance/process .. for cross-validation')
+
+parser.add_argument('--nrProc', dest='nrProc', type=int,
+  default=1, help='# of processes')
+
+parser.add_argument('--modelToRun', dest='modelToRun', type=int,
+  help='index of model to run')
+
+parser.add_argument('--cluster', action="store_true",
+  help='need to include this flag if runnin on cluster')
+
+parser.add_argument('--nrRows', dest='nrRows', type=int,
+  help='nr of subfigure rows to plot at every iteration')
+
+parser.add_argument('--nrCols', dest='nrCols', type=int,
+  help='nr of subfigure columns to plot at every iteration')
+
+parser.add_argument('--penalty', dest='penalty', type=float,
[…]

### 2018-02-26 DisProgBuilder.py
added actual files instead of soft links for two python scripts

@@ -1,1 +1,35 @@
-../diffEqModel/DisProgBuilder.py
+from abc import ABC, abstractmethod
+
+class DPMBuilder(ABC):
+
+  def __init__(self):
+    pass
+
+  @abstractmethod
+  def generate(self, dataIndices, expName, params):
+    pass
+
+
+class DPMInterface(ABC):
+
+  @abstractmethod
+  def run(self, runPart):
+    raise NotImplementedError("Should have implemented this")
+
+  @abstractmethod
+  def stageSubjects(self, indices):
+    raise NotImplementedError("Should have implemented this")
+
+  @abstractmethod
+  def stageSubjectsData(self, data):
+    raise NotImplementedError("Should have implemented this")
+
+  @abstractmethod
+  def plotTrajectories(self, res):
+    raise NotImplementedError("Should have implemented this")
+
+  @abstractmethod
+  def plotTrajSummary(self, res):
+    raise NotImplementedError("Should have implemented this")
+
+

### 2018-02-26 tadpoleDrc.py
fixed everything for regenerating data

@@ -149,8 +149,8 @@ def loadTadpole(tadpoleFile):
 
   cols = list(df.loc[:, ['RID', 'AGE', 'Month_bl', 'DXCHANGE']]) + cols
 
-  # df[cols] = df[cols].apply(pd.to_numeric, errors='coerce', axis=1)
-  # pickle.dump(df, open('tadpoleCleanDf.npz', 'wb'), protocol=pickle.HIGHEST_PROTOCOL)
+  df[cols] = df[cols].apply(pd.to_numeric, errors='coerce', axis=1)
+  pickle.dump(df, open('tadpoleCleanDf.npz', 'wb'), protocol=pickle.HIGHEST_PROTOCOL)
   df = pickle.load(open('tadpoleCleanDf.npz', 'rb'))
 
   dataDf = df[['RID', 'Month_bl']]
@@ -421,7 +421,7 @@ def addMetadataDrc(dfAll):
   ############ load medatada - age, gender, ... ###########
 
   # load metadata in order to get the timepoint information for each subj id
-  metaData = sio.loadmat('../data/DRC/pcaData.mat')
+  metaData = sio.loadmat('pcaData.mat')
 
   print(metaData.keys())
   partCode = metaData['participantCode']
@@ -987,7 +987,7 @@ def main():
   else:
     expName = '%sPen%.1f' % (expName, args.penalty)
 
-  params['runPartStd'] = ['L', 'R']
+  params['runPartStd'] = ['R', 'R']
   params['runPartMain'] = ['R', 'I', 'I'] # [mainPart, plot, stage]
   params['masterProcess'] = args.runIndex == 0
 

### 2018-02-26 tadpoleDrc.py
bug in paramsOtherBiomk

@@ -868,7 +868,7 @@ def main():
   random.seed(1)
   pd.set_option('display.max_columns', 50)
   tinyData = True
-  regenerateData = True
+  regenerateData = False
   if tinyData:
     finalDataFile = 'tadpoleDrcTiny.npz'
   else:
@@ -974,7 +974,7 @@ def main():
   # first disease has CTL+AD, second disease has CTL2+PCA
   params['diagsSetInDis'] = [np.array([CTL, MCI, AD]), np.array([CTL2, PCA])]
   params['disLabels'] = ['tAD', 'PCA']
-  params['otherBiomkPerDisease'] = {[nrBiomk-3,nrBiomk-2, nrBiomk-1], []}
+  params['otherBiomkPerDisease'] = [[nrBiomk-3,nrBiomk-2, nrBiomk-1], []]
 
   print('diag', params['diag'].shape[0])
   print('X[0]',len(params['X'][0]))

### 2018-02-26 GPHierModel.py
changed to auxFunc

@@ -2,7 +2,7 @@ import os
 import sys
 import numpy as np
 import math
-from aux import *
+from auxFunc import *
 import scipy
 #import scipy.cluster.hierarchy
 import pickle

### 2018-02-26 aux.py

@@ -1,172 +1,0 @@
-
-from env import *
-import numpy as np
-import ParHierModel
-import sklearn.metrics
-
-
-
-def convert_csv(file):
-  table = pd.read_csv(file)
-
-  return convert_table_marco(table)
-
-def convert_table_marco(table, biomkStartCol=3, list_biomarkers=None):
-    X = []
-    Y = []
-
-    # list of individuals
-    list_RID = np.unique(table[['RID']])
-    print(len(list_RID))
-    # list of biomarkers
-    if list_biomarkers is None:
-      list_biomarkers = table.columns[range(biomkStartCol, len(table.columns))]
-
-    RID = []
-
-    for id_biom, biomarker in enumerate(list_biomarkers):
-        X.append([])
-        Y.append([])
-
-    # Parsing every biomarker and assigning to the list
-    print('list_RID', list_RID)
-    diagLong = []
-    # print(adsas)
-    for id_sub, sub in enumerate(list_RID):
-        flag_missing = 0
-        indices = np.where(np.in1d(table.loc[:, 'RID'], sub))[0]
-        for id_biom, biomarker in enumerate(list_biomarkers):
-            X[id_biom].append(np.array(table[['Month_bl']])[indices].flatten())
-            Y[id_biom].append(np.array(table[[biomarker]])[indices].flatten())
-
-            idx_to_keep = ~np.isnan(Y[id_biom][id_sub])
-
-            Y[id_biom][id_sub] = Y[id_biom][id_sub][idx_to_keep]
-            X[id_biom][id_sub] = X[id_biom][id_sub][idx_to_keep]
-
-            if len(Y[id_biom][id_sub]) < 1:
-                flag_missing = flag_missing + 1
-
-
-        # print(sub, monthsSinceBlCurrSub)
-        diagCurrSub = np.array(table['diag'][indices])
-        diagNNind = np.logical_not(np.isnan(diagCurrSub))
-        # print(diagCurrSub)
-        monthsSinceBlCurrSub = np.array(table['Month_bl'])[indices][diagNNind]
-
-        # if flag_missing == 0:
-        #     print('missing data for sub: ', sub)
-
-        diagExistsForAtLeastOneVisit = monthsSinceBlCurrSub.shape[0] > 0
-        if diagExistsForAtLeastOneVisit:
[…]

### 2018-02-26 evaluationFramework.py

@@ -4,7 +4,7 @@ import math
 import pickle
 
 import os.path
-from aux import *
+from auxFunc import *
 import sys
 from matplotlib import pyplot as pl
 import gc

### 2018-02-26 genSynthData.py

@@ -5,12 +5,12 @@ from GPModel import *
 from env import *
 import scipy
 import scipy.stats
-from aux import *
+from auxFunc import *
 import ParHierModel
 import pandas as pd
 
 import MarcoModel
-import aux
+import auxFunc
 
 
 def generateDataJMD(nrSubjLong, nrBiomk, nrTimepts, shiftsLowerLim, shiftsUpperLim, model,
@@ -181,7 +181,7 @@ def convertToMarcoFormat(data, labels, yearsSinceBlScan, partCode, diag):
   df.insert(0, 'RID', partCode)
   df.insert(0, 'SUB', np.array(range(data.shape[0])))
 
-  X,Y,RID,list_biomarkers, diag = aux.convert_table_marco(df, list_biomarkers=labels)
+  X,Y,RID,list_biomarkers, diag = auxFunc.convert_table_marco(df, list_biomarkers=labels)
 
   return X,Y,np.array(RID)
 

### 2018-02-26 hierSynth.py

@@ -52,7 +52,7 @@ import genSynthData
 import GPModel
 import ParHierModel
 import Plotter
-from aux import *
+from auxFunc import *
 import evaluationFramework
 
 hostName = gethostname()

### 2018-02-26 jointSynth.py

@@ -52,7 +52,7 @@ import genSynthData
 import GPModel
 import ParHierModel
 import Plotter
-from aux import *
+from auxFunc import *
 import evaluationFramework
 from matplotlib import pyplot as pl
 

### 2018-02-26 simpleSynth.py

@@ -51,7 +51,7 @@ import genSynthData
 import GPModel
 import ParHierModel
 import Plotter
-from aux import *
+from auxFunc import *
 import evaluationFramework
 
 hostName = gethostname()

### 2018-02-26 tadpoleDrc.py

@@ -11,7 +11,7 @@ import copy
 import pandas as pd
 import pickle
 import random
-import aux
+import auxFunc
 import scipy
 import scipy.io as sio
 import scipy.stats
@@ -59,7 +59,7 @@ import genSynthData
 import GPModel
 import ParHierModel
 import Plotter
-from aux import *
+from auxFunc import *
 import evaluationFramework
 from matplotlib import pyplot as pl
 
@@ -830,7 +830,7 @@ def prepareData(finalDataFile, tinyData):
   # print(adsa)
 
   X, Y, RID, list_biomarkers, diag = \
-    aux.convert_table_marco(dataDfAll, list_biomarkers=selectedBiomk)
+    auxFunc.convert_table_marco(dataDfAll, list_biomarkers=selectedBiomk)
 
   ds = dict(X=X, Y=Y, RID=RID, list_biomarkers=list_biomarkers,
     dataDfAll=dataDfAll, regParamsICV=regParamsICV,

### 2018-02-26 tadpoleOnly.py

@@ -11,7 +11,7 @@ import copy
 import pandas as pd
 import pickle
 import random
-import aux
+import auxFunc
 import scipy
 import scipy.io as sio
 import scipy.stats
@@ -59,7 +59,7 @@ import genSynthData
 import GPModel
 import ParHierModel
 import Plotter
-from aux import *
+from auxFunc import *
 import evaluationFramework
 from matplotlib import pyplot as pl
 
@@ -834,7 +834,7 @@ def prepareData(finalDataFile, tinyData):
 
 
   X, Y, RID, list_biomarkers, diag = \
-    aux.convert_table_marco(dataDfAll, list_biomarkers=selectedBiomk)
+    auxFunc.convert_table_marco(dataDfAll, list_biomarkers=selectedBiomk)
 
   ds = dict(X=X, Y=Y, RID=RID, list_biomarkers=list_biomarkers,
     dataDfAll=dataDfAll, regParamsICV=regParamsICV,

### 2018-02-26 MarcoModel.py
forgot to add auxFunc

@@ -940,6 +940,43 @@ class GP_progression_model(object):
         scaleX = self.max_X[0] * self.mean_std_X[0][1]
         return np.array(individual_time) *  scaleX + self.mean_std_X[0][0]
 
+    def ReturnTimeShiftLong(self):
+      for sub in range(self.unitModels[u].N_samples):
+        for b in range(self.unitModels[u].N_biom):
+          xDysfunSubjUCurrSubj = self.unitModels[u].X[b][sub]  # Xs in the unit model
+          xDysfunSubjU[u][sub] += list(xDysfunSubjUCurrSubj)
+
+          # dysfuncScoresCurrSubExtr = [XarraysScaledB[b][k][0] for k in range(int(np.sum(self.gpModels[
+          # u].N_obs_per_sub[b][:sub])), np.sum(self.gpModels[u].N_obs_per_sub[b][:sub + 1]))]
+          dysfuncScoresCurrSubExtr = [self.unitModels[u].X_array[b][k][0] for k in range(int(np.sum(
+            self.unitModels[u].N_obs_per_sub[b][:sub])), np.sum(self.unitModels[u].N_obs_per_sub[b][:sub + 1]))]
+
+          dysfuncScoresU[u][sub] += dysfuncScoresCurrSubExtr  # (Xs + timeShift) in the unit model
+
+          # xsNewGpTestCurrSub = [newGPTest.X_array[b][k][0] for k in range(int(np.sum(
+          #   newGPTest.N_obs_per_sub[b][:sub])), np.sum(newGPTest.N_obs_per_sub[b][:sub + 1]))]
+
+
+
+          #
+          # print('dysfuncScoresUCurrSubCalc', dysfuncScoresCurrSubCalc)
+          # print('xsNewGpTestCurrSub', xsNewGpTestCurrSub, np.array(xsNewGpTestCurrSub) + self.unitModels[u].params_time_shift[0][sub])
+          # print('dysfuncScoresCurrSubExtr', dysfuncScoresCurrSubExtr)
+          # print('params_time_shift[0][sub]',
+          #       self.unitModels[u].params_time_shift[0][sub])
+          # print('xDysfunSubjU[u][sub]', xDysfunSubjU[u][sub])
+          # print(adsa) they are indeed equal if you standardize them.
+
+        # apply the forward scaling transform
+
+
+        print('xDysfunSubjU[u][sub]', xDysfunSubjU[u][sub])
[…]

### 2018-02-26 JointModel.py
removed dependencies that were not in repo

@@ -171,7 +171,7 @@ class JointModel(DisProgBuilder.DPMInterface):
 
         # make the functional scores be between [0,1]
         dysfuncScoresU[u] = [ (x - minDys) / (maxDys - minDys) for x in  dysfuncScoresU[u]]
-        # print('dysfuncScoresU[u]', dysfuncScoresU[u])
+        print('dysfuncScoresU[u]', dysfuncScoresU[u])
         # print(adsa)
 
 
@@ -187,13 +187,13 @@ class JointModel(DisProgBuilder.DPMInterface):
       disLabels = self.params['disLabels']
       nrDis = len(disLabels)
       self.disModels = [_ for _ in range(nrDis)]
-      for disNr in range(nrDis):
+      for disNr in [1]: #range(nrDis):
 
         xDysfunSubjUCopy = copy.deepcopy(xDysfunSubjU)
         dysfuncScoresUCopy = copy.deepcopy(dysfuncScoresU)
 
-        xDysfunSubjUCopy += [self.params['X'][i] for i in params['otherBiomkPerDisease'][disNr]]
-        dysfuncScoresUCopy += [self.params['Y'][i] for i in params['otherBiomkPerDisease'][disNr]]
+        xDysfunSubjUCopy += [self.params['X'][i] for i in self.params['otherBiomkPerDisease'][disNr]]
+        dysfuncScoresUCopy += [self.params['Y'][i] for i in self.params['otherBiomkPerDisease'][disNr]]
 
         # first filter the data .. keep only subj in current disease
         xDysfunSubjCurrDisU = [_ for _ in range(nrBiomkDisModel)]
@@ -207,7 +207,7 @@ class JointModel(DisProgBuilder.DPMInterface):
           print(self.params['diag'].shape)
           xDysfunSubjCurrDisU[b] = [xDysfunSubjUCopy[b][s] for s in range(nrSubj)
             if self.params['diag'][s] in self.params['diagsSetInDis'][disNr]]
-          dysfuncScoresCurrDisU[b] = [xDysfunSubjUCopy[b][s] for s in range(nrSubj)
+          dysfuncScoresCurrDisU[b] = [dysfuncScoresUCopy[b][s] for s in range(nrSubj)
             if self.params['diag'][s] in self.params['diagsSetInDis'][disNr]]
 
 
[…]

### 2018-02-26 MarcoModel.py

@@ -198,6 +198,8 @@ class GP_progression_model(object):
       meanAll = np.mean(XarrayAllBiomk)
       stdAll = np.std(XarrayAllBiomk)
 
+      print('self.Y_array', self.Y_array)
+      print('self.Y_array', self.Y_array)
 
       # print('meanAll', meanAll)
       # print('stdAll', stdAll)
@@ -229,7 +231,7 @@ class GP_progression_model(object):
 
         self.X_array[l] = self.X_array[l]/self.max_X[l]
 
-
+      # print('self.Y_array', self.Y_array)
       # print('self.X_array', self.X_array)
       # print(adsa)
 

### 2018-02-26 evaluationFramework.py

@@ -9,28 +9,11 @@ import sys
 from matplotlib import pyplot as pl
 import gc
 
-sys.path.append(os.path.abspath("../voxelwiseDPM/"))
-sys.path.append(os.path.abspath("../jointModellingDisease/"))
-sys.path.append(os.path.abspath("../diffEqModel/"))
 
 
 # Differential equation models
-from DEM import *
 from DisProgBuilder import *
-from aligners import *
-from diffEqModel import *
-from plotFunc import *
-
-# voxelwise models
-import voxelDPM
-import VDPMLinear
-import VDPMMean
-import VDPM_MRF
-import VDPMSimplified
-import VDPMNan
-import PlotterVDPM
-import VDPMNanMasks
-import VDPMNanNonMean
+#from aligners import *
 
 
 # joint disease models
@@ -47,132 +30,6 @@ plotFuncDiffEq = SourceFileLoader("*",
 def runModels(params, expName, modelToRun, runAllExpFunc):
   modelNames = []
   res = []
-
-  if np.any(modelToRun == 0) or np.any(modelToRun == 1):
-    fittingFunc = fitGaussianProc
-    aligner = AlignerBaseVisit()  # aligns to baseline visit
-    dpmBuilder = DEMBuilder(fittingFunc, aligner)  # disease progression model builder
-    modelName = 'DEMStd'
-    expNameCurrModel = '%s_%s' % (expName, modelName)
-    params['currModel'] = 1
-    res += [runAllExpFunc(params, expNameCurrModel, dpmBuilder)]
-    modelNames += [modelName]
-
-  # if np.any(modelToRun == 0) or np.any(modelToRun == 2):
-  #   fittingFunc = fitGaussianProc
-  #   aligner = AlignerEM()  # aligns using Expectation-Maximisation
-  #   dpmBuilder = DEMBuilder(fittingFunc, aligner)  # disease progression model builder
-  #   modelName = 'DEM-EM'
-  #   expNameCurrModel = '%s_%s' % (expName, modelName)
-  #   params['currModel'] = 2
-  #   res += [runAllExpFunc(params, expNameCurrModel, dpmBuilder)]
-  #   modelNames += [modelName]
-
-  if np.any(modelToRun == 0) or np.any(modelToRun == 3):
-    fittingFunc = fitGaussianProc
-    aligner = AlignerLogLOpt()  # aligns using direct optimisation of incomplete logL (marginal over stages)
[…]

### 2018-02-26 tadpoleDrc.py

@@ -987,7 +987,7 @@ def main():
   else:
     expName = '%sPen%.1f' % (expName, args.penalty)
 
-  params['runPartStd'] = ['R', 'R']
+  params['runPartStd'] = ['L', 'R']
   params['runPartMain'] = ['R', 'I', 'I'] # [mainPart, plot, stage]
   params['masterProcess'] = args.runIndex == 0
 

### 2018-02-26 JointModel.py
minor

@@ -187,7 +187,7 @@ class JointModel(DisProgBuilder.DPMInterface):
       disLabels = self.params['disLabels']
       nrDis = len(disLabels)
       self.disModels = [_ for _ in range(nrDis)]
-      for disNr in [1]: #range(nrDis):
+      for disNr in range(nrDis):
 
         xDysfunSubjUCopy = copy.deepcopy(xDysfunSubjU)
         dysfuncScoresUCopy = copy.deepcopy(dysfuncScoresU)

### 2018-02-26 evaluationFramework.py
removed plotFunc dep

@@ -24,8 +24,6 @@ import IncompleteModel
 
 from importlib.machinery import SourceFileLoader
 from sklearn import linear_model
-plotFuncDiffEq = SourceFileLoader("*",
-  os.path.abspath("../diffEqModel/plotFunc.py")).load_module()
 
 def runModels(params, expName, modelToRun, runAllExpFunc):
   modelNames = []
@@ -75,17 +73,7 @@ def runStdDPM(params, expNameCurrModel, dpmBuilder, runPart):
   if runPart[1] == 'R':
     dpmObj.plotTrajectories(res)
 
-  if runPart[2] == 'R':
-    # dataIndicesNN = np.logical_and(dataIndices, np.sum(np.isnan(params['data']),1) == 0)
-    (maxLikStages, maxStagesIndex, stagingProb, stagingLik, tsStages, _) = dpmObj.stageSubjects(dataIndices)
-    print(params['diag'].shape, dataIndices.shape)
-    print('maxLikStages min max', np.min(maxLikStages), np.max(maxLikStages))
-    fig, lgd = plotFuncDiffEq.plotStagingHist(maxLikStages, diag=params['diag'][dataIndices],
-                    plotTrajParams=params['plotTrajParams'], expNameCurrModel=expNameCurrModel)
-    stagingHistFigName = '%s/stagingHist.png' % dpmObj.outFolder
-    fig.savefig(stagingHistFigName, bbox_extra_artists=(lgd,), bbox_inches='tight')
-
-
+  
   return dpmObj, res
 
 

### 2018-02-26 tadpoleDrc.py
for pere to test

@@ -868,7 +868,7 @@ def main():
   random.seed(1)
   pd.set_option('display.max_columns', 50)
   tinyData = True
-  regenerateData = False
+  regenerateData = True
   if tinyData:
     finalDataFile = 'tadpoleDrcTiny.npz'
   else:

### 2018-02-26 JointModel.py
m

@@ -246,6 +246,9 @@ class JointModel(DisProgBuilder.DPMInterface):
     elif runPart[1] == 'L':
       self.disModels = pickle.load(open(disModelsFile, 'rb'))
 
+    print('self.disModels', self.disModels)
+    print(adsa)
+
     res = None
     return res
 

### 2018-02-26 tadpoleDrc.py
Pere commit

@@ -76,6 +76,10 @@ elif args.cluster:
   freesurfPath = '/share/apps/freesurfer-5.3.0'
   homeDir = '/home/rmarines'
   blenderPath = '/share/apps/blender-2.75/blender'
+elif hostName == 'planell-VirtualBox':
+  homeDir = '/home/planell'
+  freesurfPath = ""
+  blenderPath = ""
 else:
   raise ValueError('Wrong hostname. If running on new machine, add '
                    'application paths in python code above')
@@ -93,9 +97,9 @@ plotTrajParams['legendCols'] = 4
 plotTrajParams['diagLabels'] = {CTL:'CTL ADNI', MCI:'MCI ADNI', AD:'AD ADNI',
   CTL2:'CTL DRC', PCA:'PCA DRC', AD2:'AD DRC'}
 
-plotTrajParams['freesurfPath'] = freesurfPath
+#plotTrajParams['freesurfPath'] = freesurfPath
 # plotTrajParams['ylimitsRandPoints'] = (-3,2)
-plotTrajParams['blenderPath'] = blenderPath
+#plotTrajParams['blenderPath'] = blenderPath
 plotTrajParams['isSynth'] = True
 
 
@@ -119,6 +123,10 @@ elif hostName == 'razvan-Precision-T1700':
 elif args.cluster:
   homeDir = '/home/rmarines'
   freesurfPath = '/home/rmarines/src/freesurfer-6.0.0'
+elif hostName == 'planell-VirtualBox':
+  homeDir = '/home/planell'
+  freesurfPath = ""
+  blenderPath = ""
 else:
   raise ValueError('wrong hostname or cluster flag')
 
@@ -868,7 +876,7 @@ def main():
   random.seed(1)
   pd.set_option('display.max_columns', 50)
   tinyData = True
-  regenerateData = True
+  regenerateData = False#True
   if tinyData:
     finalDataFile = 'tadpoleDrcTiny.npz'
   else:
@@ -987,7 +995,7 @@ def main():
   else:
     expName = '%sPen%.1f' % (expName, args.penalty)
 
-  params['runPartStd'] = ['L', 'R']
+  params['runPartStd'] = ['R', 'R']
   params['runPartMain'] = ['R', 'I', 'I'] # [mainPart, plot, stage]
   params['masterProcess'] = args.runIndex == 0
 

### 2018-02-26 JointModel.py
added run and regData flags, starting predictions

@@ -252,6 +252,23 @@ class JointModel(DisProgBuilder.DPMInterface):
     res = None
     return res
 
+  def predictBiomkSubjGivenXs(self, newXs, disNr):
+    # newXs is an array as with np.linspace(minX, maxX)
+    # not a longitudinal list
+
+    # first predict the dysfunctionality scores in the disease specific model
+    # self.disModels[disNr].predictBiomk
+
+
+    # then predict the inidividual biomarkers in the disease agnostic model
+
+    for b in range(self.nrBiomk):
+      u = self.mapBiomkToFuncUnits[b]
+
+      self.unitModels[u].predictBiomk(newXs)
+
+
+
   def createPlotTrajParamsFuncUnit(self, nrCurrFuncUnit):
 
     plotTrajParamsFuncUnit = copy.deepcopy(self.params['plotTrajParams'])

### 2018-02-26 jointSynth.py

@@ -309,7 +309,7 @@ def main():
   else:
     expName = '%sPen%.1f' % (expName, args.penalty)
 
-  params['runPartStd'] = ['L', 'R']
+  params['runPartStd'] = ['L', 'L']
   params['runPartMain'] = ['R', 'I', 'I'] # [mainPart, plot, stage]
   params['masterProcess'] = args.runIndex == 0
 

### 2018-02-26 tadpoleDrc.py

@@ -47,6 +47,15 @@ parser.add_argument('--nrCols', dest='nrCols', type=int,
 parser.add_argument('--penalty', dest='penalty', type=float,
   help='penalty value for non-monotonic trajectories. between 0 (no effect) and 10 (strong effect). ')
 
+parser.add_argument('--regData', action="store_true", default=False,
+  help=' add this flag to regenerate the data')
+
+parser.add_argument('--runPartStd', dest='runPartStd', default='RR',
+  help=' choose whether to (R) run or (L) load from the checkpoints: '
+  'either LL, RR, LR or RL. ')
+
+
+
 args = parser.parse_args()
 
 if args.agg:
@@ -868,7 +877,7 @@ def main():
   random.seed(1)
   pd.set_option('display.max_columns', 50)
   tinyData = True
-  regenerateData = True
+  regenerateData = args.regData
   if tinyData:
     finalDataFile = 'tadpoleDrcTiny.npz'
   else:
@@ -987,7 +996,8 @@ def main():
   else:
     expName = '%sPen%.1f' % (expName, args.penalty)
 
-  params['runPartStd'] = ['L', 'R']
+  # params['runPartStd'] = ['L', 'L']
+  params['runPartStd'] = args.runPartStd
   params['runPartMain'] = ['R', 'I', 'I'] # [mainPart, plot, stage]
   params['masterProcess'] = args.runIndex == 0
 

### 2018-02-26 MarcoModel.py
more on predicting biomarkers ... will merge scale+pred+scale into 1 func

@@ -156,6 +156,14 @@ class GP_progression_model(object):
       scaleY = self.max_Y[biomk] * self.mean_std_Y[biomk][1]
       return scaleY * y_data + self.mean_std_Y[biomk][0]
 
+    def applyScalingYAllBiomk(self, biomksXB):
+      biomksNewXB = np.zeros(biomksXB.shape)
+      for b in range(gpModel.N_biom):
+        biomksNewXB[:, b] = self.applyScalingY(biomksXB[:, b], b)
+
+      return biomksNewXB
+
+
     def applyScalingXForward(self, x_data, biomk):
       scaleX = self.max_X[biomk] * self.mean_std_X[biomk][1]
       return (x_data - self.mean_std_X[biomk][0])/scaleX

### 2018-02-26 Plotter.py

@@ -467,10 +467,7 @@ class PlotterGP:
     newXTrajScaledZeroOne = (newXTraj - np.min(newXTraj)) / (np.max(newXTraj) - np.min(newXTraj))
 
     predictedBiomksXB = gpModel.predictBiomk(newXTraj)
-    predTrajScaledXB = np.zeros(predictedBiomksXB.shape)
-
-    for b in range(gpModel.N_biom):
-      predTrajScaledXB[:, b] = gpModel.applyScalingY(predictedBiomksXB[:, b], b)
+    predTrajScaledXB = gpModel.applyScalingYAllBiomk(predictedBiomksXB)
 
       if self.plotTrajParams['zScoreTraj']:
         meanCtlB = np.zeros(gpModel.N_biom)

### 2018-03-01 paper/paper.tex
re-fitting everything. fixed the scaling problem ... need to check validation results after fitting

--- 2018-02-26 paper/dkt.tex
@@
-\documentclass[11pt,a4paper]{article}
+\documentclass{llncs}
 \usepackage{amsmath,amssymb,calc,ifthen}
@@
 \usepackage{amsmath,graphicx}
-\usepackage{mathtools}
 \usepackage{epstopdf}
@@
 % margin size
-\usepackage[margin=0.5in]{geometry}
 \usepackage{pdfpages}
 \usepackage{enumitem} % for nested enumerate numbers 1 1.1 1.1.1
+% \usepackage{breqn}
+% \usepackage[linesnumbered]{algorithm2e}
+% \usepackage{algorithmicx,algpseudocode}
+% \usepackage{wrapfig} % for allowing text wrapped around the algorithm
+% \newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{blue}{#1}}
+% \SetCommentSty{mycommfont}
 
-\usepackage{titlesec} % reduce spacing after subsections
-
-% \titlespacing\subsection{0pt}{4pt plus 4pt minus 2pt}{4pt plus 2pt minus 2pt}
+% \usepackage{titlesec}
+% \titlespacing*{\section}
+% {0pt}{5.5ex plus 1ex minus .2ex}{4.3ex plus .2ex}
 
@@
 \begin{document}
-\belowdisplayskip=12pt plus 3pt minus 9pt
-\belowdisplayshortskip=7pt plus 3pt minus 4pt
-
-% \rowcolors{2}{gray!25}{white}
-
-% \begin{figure}[H]
-% \centering
-% \includegraphics[scale=0.5]{../code/synthetic/14/figures/liksTrueFittedParams}
-% \caption{14 clear biomk}
-% \end{figure}
-% 
-% \begin{figure}[H]
-% \centering
-% \includegraphics[scale=0.5]{../code/synthetic/14/figures/liksTrueFittedParams}
-% \caption{11 clear + 3 unclear biomk}
-% \end{figure}
 
@@
 
-% \section{Motivation}
+\title{Disease Knowledge Transfer for Alzheimer's Variants}
 %
-% It is currently believed that neurodegenerative diseases target specific brain networks (Seeley, 2009).
-Moreover, different symptoms assiciated with different syndromes are thus believed to be a result of different underlying brain areas being affected.
+\titlerunning{Disease Knowledge Transfer for Alzheimer's Variants}  % abbreviated title (for running head)
+%                                     also used for the TOC unless
+%                                     \toctitle is used
+%
+\author{R\u{a}zvan V. Marinescu\inst{1} \and Pere P. Morell\inst{1} \and
+Marco Lorenzi\inst{1,4} \and Neil P.
+Oxtoby\inst{1} \and Alexandra L.
+Young\inst{1} \and Arman Eshaghi\inst{1,2} \and Timothy J.
+Shakespeare\inst{3} \and Kier X.
+X.
+Yong\inst{3} \and Sebastian J.
+Crutch\inst{3} \and Daniel C.
+Alexander\inst{1}, for
+the  Alzheimer’s  Disease  Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu).
+As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report.
+A complete listing of ADNI investigators can be found at: http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
 
-% TITLE Disease Knowledge Transfer for Alzheimer's Variants
+%
+\authorrunning{R\u{a}zvan V. Marinescu et al.} % abbreviated author list (for running head)
+%
+%%%% list of authors for the TOC (use if author list has to be modified)
+% \tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
+% Craig Chambers, Kim B. Bruce, and Elisa Bertino}
+%
+\institute{Centre for Medical Image Computing, Computer Science Department, University College London, UK
+\and 
+Queen Square MS Centre, UCL Institute of Neurology, London
+\and 
+Dementia Research Centre, UCL Institute of Neurology, University College London, UK
+\and
+University of C\^{o}te d'Azur, Inria Sophia Antipolis, Asclepios Research Project
+% Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
+% F-91405 Orsay Cedex, France}
+}
 
-\section{Abstract}
+\maketitle              % typeset the title of the contribution
+
+\begin{abstract}
+% The abstract should summarize the contents of the paper
+% using at least 70 and at most 150 words. It will be set in 9-point
+% font size and be inset 1.0 cm from the right and left margins.
+% There will be two blank lines before and after the Abstract. \dots
+
+%5. achievement: we built a voxelwise disease progression model that uses sigmoidal traj
+%6. model groups together vertices into clusters and builds a common trajectory for all vertices in the cluster
+%7. Simulations show bla bla ...
+%8. Tested the model on ADNI and stages correlate well with cognitive tests.
+%9. Compared the patterns of typical and atypical AD with DRC data
+
+% The model highlights, for the first time, groups of brain vertices that exhibit a similar temporal trajectory over the population.
+This provides a new way to parcellate the brain that is specific to the temporal trajectory of a particular disease
+
+%%% MAX NR OF WORDS USED, DO NOT ADD MORE %%%%%%
 
@@
 
-% 1. Statement of Novelty/Impact *
+\keywords{Disease Progression Model, Transfer Learning, Manifold Learning, Alzheimer's Disease, Posterior Cortical Atrophy}
+\end{abstract}
 
-We present a novel dementias paradigm (Affected Overlap) and technique (DKT) that can be used to predict, for the first time, the population-level progression of biomarkers in a disease for which no such biomarker data is available, by transferring them from a different disease where they are available.
+%%%%%%%%% NO LONGER THAN 12 pages!! otherwise paper will be instantly rejected %%%%%%%%
+
+\section{Introduction}
 
 \section{Hierarchical model of disease progression}
+
+% alzheimer's -> not homogeneous, there are many varieties -> PCA -> 
+There exist several image-based biomarkers for Alzheimer's disease that can be used to track its progression: AV45 Positron Emission Tomography (PET) measuring amyloid-plaque aggregation, AV1451 PET measuring tau tangle aggregation, Flourodeoxyglucose (FDG) PET measuring brain hypometabolism, Magnetic Resonance Imaging (MRI) measuring structural integrity and Diffusion Tensor Imaging (DTI) measuring connectivity integrity.
+Measuring the exact evolution of these biomarkers over the disease progression can lead to better disease staging of patients, which helps stratification in clinical trials.
+
+% add another disease example at the end
+A hypothetical model of disease progression has been previously published by \ref{jack2010hypothetical}, which proposes that the first biomarkers to become abnormal are measures of amyloid beta aggregation, followed by tau abnormalities, hypometabolism, structural MRI-based measures and finally cognitive decline.
+While this hypothetical model proposes a modality-specific ordering of biomarkers in typical AD, it has also been observed that within the same modality, biomarker measurements also have a spatial ordering of abnormality that correlated with Braak stages: hippocampal volumes and entorhinal measures become abnormal first, followed by other structures within the temporal lobe, followed by parietal and frontal abnormalities.
+Moreover, this type of spatial ordering of abnormality has been observed to differ in other Alzheimer's variants.
+For example, in Posterior Cortical Atrophy, posterior regions such as the occipital lobe and superior parietal regions have been shown to become affected.
+The network vulnerability theory proposed by \cite{seeley2009neurodegenerative} suggests that damage to specific underlying connectivity networks leads to spatially distinct patterns of neurodegeneration.
+However, the current understanding is that within the same brain regions,
 
@@
  \centering
- \includegraphics[height=11cm]{figures/disease_knowledge_transfer.png}
+ \includegraphics[height=9cm]{figures/disease_knowledge_transfer.png}
  \caption{Outline of the proposed framework for joint modelling of multiple diseases. }
@@
 
+\section{Acknowledgements}
+
+This work was supported by the EPSRC Centre For Doctoral Training in Medical Imaging with grant EP/L016478/1.
+AE received a McDonald Fellowship from the Multiple Sclerosis International Federation (MSIF, www.msif.org), and the ECTRIMS - MAGNIMS Fellowship.
+ALY was supported through EPSRC grant EP/J020990/01.
+NPO and SG received funding from the EU Horizon 2020 research and innovation programme under grant agreement No 666992.
+SJC was supported by an Alzheimer’s Research UK Senior Research Fellowship and ESRC/NIHR (ES/L001810/1) and EPSRC (EP/M006093/1) grants.
+DCA's work on this topic has funding from the EU Horizon 2020 research and innovation programme under grant agreement No 666992, as well as EPSRC grants J020990, M006093 and M020533.
+Data collection and sharing for this project was funded by the Alzheimer's Disease Neuroimaging Initiative (ADNI) (National Institutes of Health Grant U01 AG024904) and DOD ADNI (Department of Defense award number W81XWH-12-2-0012).
+The Dementia Research Centre is an ARUK coordination center.
+
+%
+% ---- Bibliography ---- 
+% USE HARVARD STANDARD
+
+\bibliographystyle{unsrtnat}
+\begin{thebibliography}{5}
+
+\bibitem{jack2010hypothetical}
+Jack, C.R., Knopman, D.S., Jagust, W.J., Shaw, L.M., Aisen, P.S., Weiner, M.W., Petersen, R.C.
+and Trojanowski, J.Q., 2010.
+Hypothetical model of dynamic biomarkers of the Alzheimer's pathological cascade.
+The Lancet Neurology, 9(1), pp.119-128.
+
+\bibitem{bateman2012clinical}
+Bateman, R.J., Xiong, C., Benzinger, T.L., Fagan, A.M., Goate, A., Fox, N.C., Marcus, D.S., Cairns, N.J., Xie, X., Blazey, T.M.
+and Holtzman, D.M., 2012.
+Clinical and biomarker changes in dominantly inherited Alzheimer's disease.
+New England Journal of Medicine, 367(9), pp.795-804.
+
+\bibitem{richberg2015multi}
+Schmidt-Richberg, A., Guerrero, R., Ledig, C., Molina-Abril, H., Frangi, A.F., Rueckert, D.
+and Alzheimers Disease Neuroimaging Initiative, 2015, June.
+Multi-stage Biomarker Models for Progression Estimation in Alzheimer’s Disease.
+In International Conference on Information Processing in Medical Imaging (pp.
+387-398).
+Springer International Publishing.
+
+\bibitem{fonteijn2012event}
+Fonteijn, H.M., Modat, M., Clarkson, M.J., Barnes, J., Lehmann, M., Hobbs, N.Z., Scahill, R.I., Tabrizi, S.J., Ourselin, S., Fox, N.C.
+and Alexander, D.C., 2012.
+An event-based model for disease progression and its application in familial Alzheimer's disease and Huntington's disease.
+NeuroImage, 60(3), pp.1880-1889.
+
+\bibitem {jedynak2012}
+Jedynak, B.M., Lang, A., Liu, B., Katz, E., Zhang, Y., Wyman, B.T., Raunig, D., Jedynak, C.P., Caffo, B., Prince, J.L.
+and Alzheimer's Disease Neuroimaging Initiative, 2012.
+A computational neurodegenerative disease progression score: method and results with the Alzheimer's Disease Neuroimaging Initiative cohort.
+Neuroimage, 63(3), pp.1478-1486.
+
+\bibitem{donohue2014estimating}
+Donohue, M.C., Jacqmin-Gadda, H., Le Goff, M., Thomas, R.G., Raman, R., Gamst, A.C., Beckett, L.A., Jack, C.R., Weiner, M.W., Dartigues, J.F.
+and Aisen, P.S., 2014.
+Estimating long-term multivariate progression from short-term data.
+Alzheimer's \& Dementia, 10(5), pp.S400-S410.
+
+\bibitem{schiratti2015mixed}
+Schiratti, J.B., Allassonniere, S., Routier, A., Colliot, O., Durrleman, S.
+and Alzheimers Disease Neuroimaging Initiative, 2015, June.
+A mixed-effects model with time reparametrization for longitudinal univariate manifold-valued data.
+In International Conference on Information Processing in Medical Imaging (pp.
+564-575).
+Springer International Publishing.
+
+\bibitem{bilgel2016multivariate}
+Bilgel, M., Prince, J.L., Wong, D.F., Resnick, S.M.
+and Jedynak, B.M., 2016.
+A multivariate nonlinear mixed effects model for longitudinal image analysis: Application to amyloid imaging.
+NeuroImage, 134, pp.658-670.
+
+\bibitem{seeley2009neurodegenerative}
+Seeley, W.W., Crawford, R.K., Zhou, J., Miller, B.L. and Greicius, M.D., 2009. Neurodegenerative diseases target large-scale human brain networks. Neuron, 62(1), pp.42-52.
+
+% \bibitem{bishop2006}
+% Bishop, C.M., 2006. Pattern recognition. Machine Learning, 128.
+
+\bibitem{reuter2012longitudinal}
+Reuter, M., Schmansky, N.J., Rosas, H.D. and Fischl, B., 2012. Within-subject template estimation for unbiased longitudinal image analysis. Neuroimage, 61(4), pp.1402-1418.
+
+% \bibitem{reuter2010inverse}
+% Reuter, M., Rosas, H.D. and Fischl, B., 2010. Highly accurate inverse consistent registration: a robust approach. Neuroimage, 53(4), pp.1181-1196.
+
+\bibitem{dickerson2009cortical}
+Dickerson, B.C., Bakkour, A., Salat, D.H., Feczko, E., Pacheco, J., Greve, D.N., Grodstein, F., Wright, C.I., Blacker, D., Rosas, H.D.
+and Sperling, R.A., 2009.
+The cortical signature of Alzheimer's disease: regionally specific cortical thinning relates to symptom severity in very mild to mild AD dementia and is detectable in asymptomatic amyloid-positive individuals.
+Cerebral cortex, 19(3), pp.497-510.
+
+% \bibitem{thompson2001cortical}
+% Thompson, P.M., Mega, M.S., Woods, R.P., Zoumalan, C.I., Lindshield, C.J., Blanton, R.E., Moussai, J., Holmes, C.J., Cummings, J.L.
+and Toga, A.W., 2001.
+Cortical change in Alzheimer's disease detected with a disease-specific population-based brain atlas.
+Cerebral Cortex, 11(1), pp.1-16.
+
+\bibitem{crutch2012posterior}
+Crutch, S.J., Lehmann, M., Schott, J.M., Rabinovici, G.D., Rossor, M.N. and Fox, N.C., 2012. Posterior cortical atrophy. The Lancet Neurology, 11(2), pp.170-178.
+
+\bibitem{young2015multiple}
+Young, A.L., Oxtoby, N.P., Huang, J., Marinescu, R.V., Daga, P., Cash, D.M., Fox, N.C., Ourselin, S., Schott, J.M., Alexander, D.C.
+and Alzheimers Disease Neuroimaging Initiative, 2015, June.
+Multiple Orderings of Events in Disease Progression.
+In International Conference on Information Processing in Medical Imaging (pp.
+711-722).
+Springer International Publishing.
+
+\end{thebibliography}
+
+\clearpage
+
 \end{document}

### 2018-03-01 JointModel.py

@@ -41,7 +41,7 @@ class JointModel(DisProgBuilder.DPMInterface):
     self.plotterObj = plotterObj
     self.nrBiomk = len(params['X'])
     self.nrFuncUnits = params['nrFuncUnits']
-    self.mapBiomkToFuncUnits = params['mapBiomkToFuncUnits']
+    self.biomkInFuncUnit = params['biomkInFuncUnit']
 
     self.unitModels = None # functional unit models
     self.disModels = None # disease specific models
@@ -76,8 +76,8 @@ class JointModel(DisProgBuilder.DPMInterface):
         plotTrajParamsFuncUnit = self.createPlotTrajParamsFuncUnit(nrCurrFuncUnit=u)
         plotterObjCurrFuncUnit = Plotter.PlotterGP(plotTrajParamsFuncUnit)  # set separate plotter for the
 
-        XfiltCurrUnit = [Xfilt[b] for b in range(self.nrBiomk) if self.mapBiomkToFuncUnits[b] == u]
-        YfiltCurrUnit = [Yfilt[b] for b in range(self.nrBiomk) if self.mapBiomkToFuncUnits[b] == u]
+        XfiltCurrUnit = [Xfilt[b] for b in self.biomkInFuncUnit[u]]
+        YfiltCurrUnit = [Yfilt[b] for b in self.biomkInFuncUnit[u]]
         outFolderCurrUnit = '%s/unit%d' % (self.outFolder, u)
         os.system('mkdir -p %s' % outFolderCurrUnit)
         self.unitModels[u] = MarcoModel.GP_progression_model(XfiltCurrUnit, YfiltCurrUnit, nrRandFeatures, outFolderCurrUnit,
@@ -93,12 +93,17 @@ class JointModel(DisProgBuilder.DPMInterface):
       # make sure the data you used has not been changed since fitting this model
       for b in range(len(self.unitModels[0].X)):
         # print('--------------- b', b)
-        idxInAllBiomk = np.where(self.mapBiomkToFuncUnits == 0)[0][b]
+        idxInAllBiomk = self.biomkInFuncUnit[0][b]
         for s in range(len(self.unitModels[0].X[b])):
           # print(self.unitModels[0].X[b][s])
           # print(self.params['X'][idxInAllBiomk][s])
           assert np.all(self.unitModels[0].X[b][s] == self.params['X'][idxInAllBiomk][s])
[…]

### 2018-03-01 MarcoModel.py

@@ -82,16 +82,21 @@ class GP_progression_model(object):
         # print('X_array', self.X_array[0])
         # print(np.where(np.isnan(self.X_array[0]))[0])
         self.rescale()
+
         self.minX = np.float128(np.min([el for sublist in self.X_array for item in sublist for el in item]))
         self.maxX = np.float128(np.max([el for sublist in self.X_array for item in sublist for el in item]))
 
-        self.addMinXMaxXExtraRange()
+        self.minScX = self.applyScalingX(self.minX)
+        self.maxScX = self.applyScalingX(self.maxX)
+
+        nrBiomk = len(self.X)
+        scaledYarrayB = [self.applyScalingY(self.Y_array[b], b) for b in range(nrBiomk)]
+        self.min_yB = np.array([np.min(scaledYarrayB[b].reshape(-1)) for b in range(nrBiomk)])
+        self.max_yB = np.array([np.max(scaledYarrayB[b].reshape(-1)) for b in range(nrBiomk)])
 
         # Number of derivative points uniformely distributed on the X axis
         self.N_Dpoints = 10
-        minXforDXs, maxXforDXs = self.addMinXMaxXExtraRange(deltaRangeFactor=0.0)
-        self.DX = np.linspace(minXforDXs, maxXforDXs, self.N_Dpoints).reshape([self.N_Dpoints, 1])
-        # self.DX = np.linspace(self.minX,self.maxX,self.N_Dpoints).reshape([self.N_Dpoints,1])
+        self.DX = np.linspace(self.minX,self.maxX,self.N_Dpoints).reshape([self.N_Dpoints,1])
 
         # Initializing random features for kernel approximation
         self.perturbation_Omega = np.random.randn(self.N_rnd_features)
@@ -141,14 +146,14 @@ class GP_progression_model(object):
 
 
 
-    def addMinXMaxXExtraRange(self, deltaRangeFactor = 0.0):
-        deltaXRange = deltaRangeFactor * (self.maxX - self.minX)
-        minX =self.minX - deltaXRange
-        maxX =self.maxX + deltaXRange
+    # def addMinXMaxXExtraRange(self, deltaRangeFactor = 0.0):
+    #     deltaXRange = deltaRangeFactor * (self.maxX - self.minX)
+    #     minX =self.minX - deltaXRange
+    #     maxX =self.maxX + deltaXRange
+    #
+    #     return minX,maxX
 
-        return minX,maxX
-
[…]

### 2018-03-01 Plotter.py

@@ -11,127 +11,6 @@ class PlotterJDM:
   def __init__(self, plotTrajParams):
     self.plotTrajParams = plotTrajParams
 
-  #
-  # def plotTrajData(self, longData, longDiag, longDPS, model,
-  #   replaceFigMode=True, yLimUseData=False,showConfInt=False, adjustBottomHeight=0.1):
-  #   figSizeInch = (self.plotTrajParams['SubfigTrajWinSize'][0] / 100, self.plotTrajParams['SubfigTrajWinSize'][1] / 100)
-  #   fig = pl.figure(1, figsize=figSizeInch)
-  #   pl.clf()
-  #   nrRows = self.plotTrajParams['nrRows']
-  #   nrCols = self.plotTrajParams['nrCols']
-  #
-  #   nrBiomk = longData[0].shape[1]
-  #
-  #   minX = np.min([np.min(dpsCurr) for dpsCurr in longDPS])
-  #   maxX = np.max([np.max(dpsCurr) for dpsCurr in longDPS])
-  #
-  #   xs = np.linspace(minX, maxX, 100)
-  #   diagNrs = self.plotTrajParams['diagNrs']
-  #
-  #   nrSubjLong = len(longData)
-  #
-  #   dysScoresSF = model.predPopDys(xs)
-  #   modelPredSB = model.predPop(xs)
-  #
-  #   print('xs', xs)
-  #   print('modelPredSB[:,0]', modelPredSB[:,0])
-  #
-  #   lw = 3.0
-  #
-  #   # first plot the dysfunctionality biomarkers
-  #   ax = pl.subplot(nrRows, nrCols, 1)
-  #   ax.set_title('dysfunc all')
-  #   moveTicksInside(ax)
-  #   for f in range(model.nrFuncUnits):
-  #     ax.plot(xs, dysScoresSF[:,f], 'k-', linewidth=lw)
-  #
-  #   # for each unit, plot all biomarkers against the dysfunctional scores
-  #   for f in range(model.nrFuncUnits):
-  #     ax2 = pl.subplot(nrRows, nrCols, f+2)
-  #     ax2.set_title('dysfunc %d' % f)
-  #     moveTicksInside(ax2)
-  #     biomkInCurrUnit = np.where(self.plotTrajParams['mapBiomkToFuncUnits'] == f)[0]
-  #     for b in range(len(biomkInCurrUnit)):
[…]

### 2018-03-01 jointSynth.py

@@ -133,14 +133,21 @@ def main():
   # should give smth like [0,1,2,3,0,1,2,3,0,1,2,3]
   print('mapBiomkToFuncUnits', mapBiomkToFuncUnits)
 
-  plotTrajParams['mapBiomkToFuncUnits'] = mapBiomkToFuncUnits
+  biomkInFuncUnit = [0 for u in range(nrFuncUnits+1)]
+  for u in range(nrFuncUnits):
+    biomkInFuncUnit[u] = list(np.where(mapBiomkToFuncUnits == u)[0])
+
+  biomkInFuncUnit[nrFuncUnits] = [] # need to leave this as empty list
+
+  plotTrajParams['biomkInFuncUnit'] = biomkInFuncUnit
   plotTrajParams['labels'] = ['b%d' % n for n in range(nrBiomk)]
-  plotTrajParams['nrRowsFuncUnit'] = 2
-  plotTrajParams['nrColsFuncUnit'] = 3
+  plotTrajParams['nrRowsFuncUnit'] = 3
+  plotTrajParams['nrColsFuncUnit'] = 4
   plotTrajParams['colorsTraj'] = [colorsys.hsv_to_rgb(hue, 1, 1) for hue in np.linspace(0, 1, num=nrBiomk, endpoint=False)]
 
-  # plotTrajParams['yNormMode'] = 'zScoreCtl'
+  # plotTrajParams['yNormMode'] = 'zScoreTraj'
   plotTrajParams['yNormMode'] = 'zScoreEarlyStageTraj'
+  # plotTrajParams['yNormMode'] = 'unscaled'
 
   # if False, plot estimated traj. in separate plot from true traj.
   plotTrajParams['allTrajOverlap'] = False

### 2018-03-01 tadpoleDrc.py

@@ -111,7 +111,7 @@ plotTrajParams['diagLabels'] = {CTL:'CTL ADNI', MCI:'MCI ADNI', AD:'AD ADNI',
 plotTrajParams['freesurfPath'] = freesurfPath
 # plotTrajParams['ylimitsRandPoints'] = (-3,2)
 plotTrajParams['blenderPath'] = blenderPath
-plotTrajParams['isSynth'] = True
+# plotTrajParams['isSynth'] = False
 
 
 if args.agg:
@@ -168,26 +168,26 @@ def loadTadpole(tadpoleFile):
 
   cols = list(df.loc[:, ['RID', 'AGE', 'Month_bl', 'DXCHANGE']]) + cols
 
-  df[cols] = df[cols].apply(pd.to_numeric, errors='coerce', axis=1)
-  pickle.dump(df, open('tadpoleCleanDf.npz', 'wb'), protocol=pickle.HIGHEST_PROTOCOL)
+  # df[cols] = df[cols].apply(pd.to_numeric, errors='coerce', axis=1)
+  # pickle.dump(df, open('tadpoleCleanDf.npz', 'wb'), protocol=pickle.HIGHEST_PROTOCOL)
   df = pickle.load(open('tadpoleCleanDf.npz', 'rb'))
 
   dataDf = df[['RID', 'Month_bl']]
-  dataDf['scanID'] = np.nan
+  dataDf.loc[:,'scanID'] = np.nan
   print('dataDf', dataDf.loc[:20,:])
 
   mapGender = {'Female' : 0, 'Male' : 1}
-  df['PTGENDER'] = df['PTGENDER'].map(mapGender)
+  df.loc[:, 'PTGENDER'] = df['PTGENDER'].map(mapGender)
 
   # print(ads)
 
-  dataDf['gender-0f1m'] = df['PTGENDER']
-  dataDf['age'] = df['AGE'] + (df['Month_bl'] / 12)
-  dataDf['dataset'] = 1
+  dataDf.loc[:, 'gender-0f1m'] = df['PTGENDER']
+  dataDf.loc[:, 'age'] = df['AGE'] + (df['Month_bl'] / 12)
+  dataDf.loc[:, 'dataset'] = 1
   ssTagMRI = 'UCSFFSX_11_02_15_UCSFFSX51_08_01_16'
   ICV = df['ST10CV_%s' % ssTagMRI]
-  dataDf['ICV'] = ICV
[…]

### 2018-03-01 JointModel.py
fixed scaling issue

@@ -116,73 +116,36 @@ class JointModel(DisProgBuilder.DPMInterface):
       dysfuncScoresU = [0 for x in range(self.nrFuncUnits)]
       xDysfunSubjU = [0 for x in range(self.nrFuncUnits)]
       for u in range(self.nrFuncUnits):
-        xs = np.linspace(self.unitModels[u].minX, self.unitModels[u].maxX, num=100).reshape([100, 1])
-
         dysfuncScoresU[u] = [[] for _ in range(nrSubj)]
         xDysfunSubjU[u] = [[] for _ in range(nrSubj)]
-        xsNewGpTest = [[] for _ in range(nrSubj)]
+
+        XshiftedUnitModel, XunitModel, YunitModel = self.unitModels[u].getData()
 
         for sub in range(self.unitModels[u].N_samples):
           for b in range(self.unitModels[u].N_biom):
-            xDysfunSubjUCurrSubj = self.unitModels[u].X[b][sub]  # Xs in the unit model
+            xDysfunSubjUCurrSubj = XunitModel[b][sub]  # Xs in the unit model
             xDysfunSubjU[u][sub] += list(xDysfunSubjUCurrSubj)
 
-            dysfuncScoresCurrSubExtr = [self.unitModels[u].X_array[b][k][0] for k in range(int(np.sum(
-              self.unitModels[u].N_obs_per_sub[b][:sub])), np.sum(self.unitModels[u].N_obs_per_sub[b][:sub + 1]))]
+            # dysfuncScoresCurrSubExtr = [self.unitModels[u].X_array[b][k][0] for k in range(int(np.sum(
+            #   self.unitModels[u].N_obs_per_sub[b][:sub])), np.sum(self.unitModels[u].N_obs_per_sub[b][:sub + 1]))]
 
-            dysfuncScoresU[u][sub] += dysfuncScoresCurrSubExtr # (Xs + timeShift) in the unit model
-
-            # xsNewGpTestCurrSub = [newGPTest.X_array[b][k][0] for k in range(int(np.sum(
-            #   newGPTest.N_obs_per_sub[b][:sub])), np.sum(newGPTest.N_obs_per_sub[b][:sub + 1]))]
-
-
-
-            #
-            # print('dysfuncScoresUCurrSubCalc', dysfuncScoresCurrSubCalc)
-            # print('xsNewGpTestCurrSub', xsNewGpTestCurrSub, np.array(xsNewGpTestCurrSub) + self.unitModels[u].params_time_shift[0][sub])
[…]

### 2018-03-01 tadpoleDrc.py

@@ -1115,11 +1115,11 @@ def main():
 
   biomkInFuncUnit = [0 for u in range(nrFuncUnits+1)]
   for u in range(nrFuncUnits):
-    biomkInFuncUnit[u] = list(np.where(mapBiomkToFuncUnits == u)[0])
+    biomkInFuncUnit[u] = np.where(mapBiomkToFuncUnits == u)[0]
     # biomkInFuncUnit[u] += [nrBiomk-3, nrBiomk-2] # also add CDRSOB and ADAS in order hlp disentangle the trajectories and get better staging
 
   # add extra entry with other biomks to be added in the disease models
-  biomkInFuncUnit[nrFuncUnits] = [nrBiomk-3, nrBiomk-2, nrBiomk-1]
+  biomkInFuncUnit[nrFuncUnits] = np.array([nrBiomk-3, nrBiomk-2, nrBiomk-1])
 
 
   plotTrajParams['biomkInFuncUnit'] = biomkInFuncUnit

### 2018-03-01 JointModel.py
m

@@ -129,7 +129,7 @@ class JointModel(DisProgBuilder.DPMInterface):
             # dysfuncScoresCurrSubExtr = [self.unitModels[u].X_array[b][k][0] for k in range(int(np.sum(
             #   self.unitModels[u].N_obs_per_sub[b][:sub])), np.sum(self.unitModels[u].N_obs_per_sub[b][:sub + 1]))]
 
-            dysfuncScoresU[u][sub] += XshiftedUnitModel[b][sub] # (Xs + timeShift) in the unit model
+            dysfuncScoresU[u][sub] += list(XshiftedUnitModel[b][sub]) # (Xs + timeShift) in the unit model
 
           xDysfunSubjU[u][sub] = np.sort(np.unique(xDysfunSubjU[u][sub]))
           dysfuncScoresU[u][sub] = np.sort(np.unique(dysfuncScoresU[u][sub]))

### 2018-03-01 MarcoModel.py
plotting tAD disease space

@@ -1090,7 +1090,12 @@ class GP_progression_model(object):
 
     def predictBiomk(self, newX):
 
-      assert self.minScX <= np.min(newX) <= self.maxScX
+      deltaX = 0.2 * (self.maxScX - self.minScX)
+      if not (self.minScX - deltaX <= np.min(newX) <= self.maxScX + deltaX):
+        print('newX', newX)
+        print('self.minScX', self.minScX)
+        print('self.maxScX', self.maxScX)
+        raise ValueError('newX not in the correct renge. check the scaling')
 
       xsScaled = self.applyScalingXForward(newX.reshape(-1, 1), biomk=0) # arbitrary space ->[0,1]
 

### 2018-03-01 Plotter.py

@@ -455,7 +455,6 @@ class PlotterGP:
     nrRows, nrCols = auxFunc.findOptimalRowsCols(nrBiomk)
 
     nrSamples = trajSamplesBXS.shape[2]
-    nrSubValid = len(XsubjValidBSX[0])
 
     min_yB = np.zeros(nrBiomk)
     max_yB = np.zeros(nrBiomk)
@@ -463,10 +462,14 @@ class PlotterGP:
       # print([np.min(yS) for yS in YsubjBSX[b] if len(yS) > 0])
       # print([np.min(predTrajXB[:,b])])
       # print([np.min(yS) for yS in YsubjBSX[b] if len(yS) > 0] + [np.min(predTrajXB[:,b])])
-      min_yB[b] = np.min([np.min(yS) for yS in YsubjBSX[b] if len(yS) > 0] +
-        [np.min(predTrajXB[:,b])] + [np.min(yS) for yS in YsubjValidBSX[b] if len(yS) > 0])
-      max_yB[b] = np.max([np.max(yS) for yS in YsubjBSX[b] if len(yS) > 0] +
-        [np.max(predTrajXB[:,b])] + [np.max(yS) for yS in YsubjValidBSX[b] if len(yS) > 0])
+      listsMin = [np.min(yS) for yS in YsubjBSX[b] if len(yS) > 0] + [np.min(predTrajXB[:,b])]
+      listsMax = [np.max(yS) for yS in YsubjBSX[b] if len(yS) > 0] + [np.max(predTrajXB[:,b])]
+      if YsubjValidBSX is not None:
+        listsMin += [np.min(yS) for yS in YsubjValidBSX[b] if len(yS) > 0]
+        listsMax += [np.max(yS) for yS in YsubjValidBSX[b] if len(yS) > 0]
+
+      min_yB[b] = np.min(listsMin)
+      max_yB[b] = np.max(listsMax)
 
     deltaB = (max_yB - min_yB)/5
 
@@ -481,7 +484,8 @@ class PlotterGP:
 
       self.plotSubjData(ax, XsubjBSX[b], YsubjBSX[b], diagS, labelExtra = '')
       # print('-------------- validation data')
-      self.plotSubjData(ax, XsubjValidBSX[b], YsubjValidBSX[b], diagValidS, labelExtra = '')
+      if XsubjValidBSX is not None:
[…]

### 2018-03-01 tadpoleDrc.py

@@ -765,8 +765,8 @@ def applyRegFromParams(data, regressorVector, diag, params, diagsCTL = (CTL, CTL
 def prepareData(finalDataFile, tinyData):
 
   tadpoleFile = 'TADPOLE_D1_D2.csv'
-  dataDfTadpole = loadTadpole(tadpoleFile)
-  dataDfTadpole.to_csv('tadpoleCleanDf.csv', sep=',', quotechar='"')
+  # dataDfTadpole = loadTadpole(tadpoleFile)
+  # dataDfTadpole.to_csv('tadpoleCleanDf.csv', sep=',', quotechar='"')
   dataDfTadpole = pd.read_csv('tadpoleCleanDf.csv')
 
   # print(dsa)
@@ -795,8 +795,6 @@ def prepareData(finalDataFile, tinyData):
   # print('validDf', validDf)
   #validDf = addDRCValidDataMock(validDf) # generate random numbers for now
   validDf = addDRCValidData(validDf) # change to this real dataset one when ready
-
-  # visValidDf(validDf)
 
   validDf.to_csv('validDf.csv')
 
@@ -876,14 +874,26 @@ def prepareData(finalDataFile, tinyData):
   # Note that the previous dataset normalisation doesn't work, because in the training
   # set there were no DTI biomarkers in dataset 2.
   for c in range(len(dtiCols)):
-    meanADNI = np.nanmean(dataDfAll.loc[:, dtiCols[c]])
+
     stdADNI =np.nanstd(dataDfAll.loc[:, dtiCols[c]])
+    stdDRC = np.nanstd(validDf.loc[:, dtiCols[c]])
+    stdRatio = stdDRC / stdADNI
+    validDf.loc[:, dtiCols[c]] = validDf.loc[:, dtiCols[c]] / stdRatio
+
+    meanADNI = np.nanmean(dataDfAll.loc[:, dtiCols[c]])
+    meanDRC = np.nanmean(validDf.loc[:, dtiCols[c]])
+    meanDiff = (meanDRC - meanADNI)
+    validDf.loc[:, dtiCols[c]] = validDf.loc[:, dtiCols[c]] - meanDiff
+
     meanDRC = np.nanmean(validDf.loc[:, dtiCols[c]])
     stdDRC = np.nanstd(validDf.loc[:, dtiCols[c]])
-    meanDiff = (meanDRC - meanADNI)
-    stdRatio = stdDRC / stdADNI
[…]

### 2018-03-02 paper/paper.tex
wrote intro paper, finalised figs. writing results in overleaf

--- 2018-02-26 paper/dkt.tex
@@
 
-% \section{Motivation}
+\title{Disease Knowledge Transfer for Alzheimer's Variants}
 %
-% It is currently believed that neurodegenerative diseases target specific brain networks (Seeley, 2009).
-Moreover, different symptoms assiciated with different syndromes are thus believed to be a result of different underlying brain areas being affected.
+\titlerunning{Disease Knowledge Transfer across Alzheimer's Variants}  % abbreviated title (for running head)
+%                                     also used for the TOC unless
+%                                     \toctitle is used
+%
 
-% TITLE Disease Knowledge Transfer for Alzheimer's Variants
+% * <mrazvan22@gmail.com> 2018-03-02T16:41:47.321Z:
+% 
+% Author list: Me, Pere (helped with validation), Marco (used his code), Alex (writing+tadpole), Neil (tadpole),  Arman, Keir (DTI data), Seb, Danny
+% 
+% ^ <mrazvan22@gmail.com> 2018-03-02T16:55:06.003Z.
 
-\section{Abstract}
+\maketitle              % typeset the title of the contribution
+
+\begin{abstract}
 
@@
 We then formulate the DKT model implementing this paradigm as a joint-disease generative model of biomarker progression which disentangles the biomarker correlations that are disease-specific from those correlations that are disease-agnostic.
-We demonstrate DKT on three datasets: (1) the TADPOLE Challenge dataset containing typical AD subjects (tAD) from the Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset who had MRI, FDG, AV45, AV1451 or DTI scans, as well as two datasets from our local centre containing subjects with (2) typical AD and (3) Posterior Cortical Atrophy (PCA) that have only undergone MRI scans.
+We demonstrate DKT on three datasets: (1) the TADPOLE Challenge dataset containing typical AD subjects (tAD) from the Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset who had MRI, FDG-PET, AV45, AV1451 or DTI scans, as well as two datasets from our local centre containing subjects with (2) typical AD and (3) Posterior Cortical Atrophy (PCA) that have only undergone MRI scans.
 We train the DKT model on all three datasets at once, and show that it is able to predict, in PCA subjects, plausible population-level biomarker for FDG, DTI, AV45 and AV1451 biomarkers, for which no data was available.
 We validate the predictions of DKT using a small set of DTI scans for the PCA cohort.
-Finally, DKT can be a useful tool to analyse and understand rare forms of dementias for which multimodal data is not available or is limited (e.g.
+We demonstrate that DKT may be a useful tool to analyse and understand rare forms of dementias for which multimodal data is not available or is limited (e.g.
 cross-sectional).
 Finally, by leveraging data from multiple diseases, DKT also has the potential to provide more accurate disease staging compared to traditional disease progression models, which is useful for patient stratification in clinical trials.
+% * <alexandra.young@ucl.ac.uk> 2018-03-02T17:28:57.289Z:
+% 
+% > Finally,
+% You've got finally twice
+% 
+% ^ <alexandra.young@ucl.ac.uk> 2018-03-02T17:29:38.345Z:
+% 
+% Could put we demonstrate (or something a bit weaker)
+%
+% ^ <alexandra.young@ucl.ac.uk> 2018-03-02T19:02:52.149Z.
 
-% 1. Statement of Novelty/Impact *
+\keywords{Disease Progression Model, Transfer Learning, Manifold Learning, Alzheimer's Disease, Posterior Cortical Atrophy}
+\end{abstract}
 
-We present a novel dementias paradigm (Affected Overlap) and technique (DKT) that can be used to predict, for the first time, the population-level progression of biomarkers in a disease for which no such biomarker data is available, by transferring them from a different disease where they are available.
+\section{Introduction}
 
-\section{Hierarchical model of disease progression}
+% biomarkers in alzheimer's -> measuring the evolution helps staging in clinical trials
+There currently exist several image-based biomarkers for Alzheimer's disease that can be used to track its progression: AV45 Positron Emission Tomography (PET) measuring amyloid-plaque aggregation, AV1451 PET measuring tau tangle aggregation, Flourodeoxyglucose (FDG) PET measuring brain hypometabolism, Magnetic Resonance Imaging (MRI) measuring structural integrity and Diffusion Tensor Imaging (DTI) measuring connectivity integrity.
+Measuring the exact evolution of these biomarkers over the disease progression can lead to better disease staging of patients, which helps stratification in clinical trials.
 
-The framework/model assumes that within a dataset of patients with a particular form of dementia, there are certain correlations between biomarkers that are \emph{disease agnostic}.
-For example, atrophy in the temporal lobe usually correlates with amyloid/tau-levels within the same region or with memory impairment.
-This is because we believe that amyloid/tau deposition cause neurodegeneration, which then causes cognitive decline in functions related to the underlying affected areas.
-One can then view different neurodenegenerative diseases as affecting a different set of brain regions and their underlying connectivity networks, a view supported by previous studies (Seeley, 2009, Zhou 2012).
-This view suggests that one can uniquely define a disease in terms of the different subset of brain regions that it affects.
-However, sometimes there are \emph{digo dktseasesease-agnostic} correlations that are not confined to a single region, e.g.
-disconnected brains regions are known to be responsible for certain cognitive functions, or biomarker measurements in different brain regions could correlate due to underlying structural or functional connectivity.
-One can therefore extend the idea of disease-agostic correlations between biomarkers to region-nonspecific sets of biomarkers which we will call \emph{functional units}, because the biomarkers being related through similar function.
+% hypothetical model has specific ordering of modality biomarkers -> ordering or spatial biomarkers (different dimension) -> spatial differences in different dementias -> network vulnerability hypothesis.
+A hypothetical model of disease progression has been previously published by \cite{jack2010hypothetical}, which proposes that the first biomarkers to become abnormal are measures of amyloid beta aggregation, followed by tau abnormalities, hypometabolism, structural MRI-based measures and finally cognitive decline.
+While this hypothetical model proposes a modality-specific ordering of biomarkers in typical AD, it has also been observed that within the same modality, biomarker measurements have a spatial sequence of abnormality that correlates with Braak stages: hippocampal volumes and entorhinal measures become abnormal first, followed by other structures within the temporal lobe, followed by parietal and frontal abnormalities.
+However, the spatial ordering of abnormality differs in other Alzheimer's variants.
+For example, in Posterior Cortical Atrophy, posterior regions such as the occipital lobe and superior parietal regions have been shown to become affected.
+The network vulnerability theory by Seeley et al.
+\cite{seeley2009neurodegenerative} suggests that damage to specific underlying connectivity networks leads to spatially distinct patterns of neurodegeneration.
+There is nevertheless a great degree of heterogeneity within these diseases, which led some to suggest that subjects with different Alzheimer's variants lie on a continuum of phenotypical variation \cite{crutch2012posterior}.
+% * <alexandra.young@ucl.ac.uk> 2018-03-02T19:01:16.243Z:
+% 
+% > The network vulnerability theory by Seeley et al.
+\cite{seeley2009neurodegenerative} suggests that damage to specific underlying connectivity networks leads to spatially distinct patterns of neurodegeneration.
+There is nevertheless a great degree of heterogeneity within these diseases, which led some to suggest that subjects with different Alzheimer's variants lie on a continuum of phenotypical variation \cite{crutch2012posterior}.
+% The relationship between these two statements is unclear from the way its written
+% 
+% ^.
 
-We propose a hierarchical model of disease progression that disentangles \emph{disease-specific} biomarker correlations from \emph{disease-agnostic} correlations.
-For modelling disease-agnostic correlations, we assume biomarkers are grouped into \emph{functional units}, where corelations within each unit are disease-agnostic (e.g.
+The amount of heterogeneity within each variant and the network vulnerability hypothesis by \cite{seeley2009neurodegenerative} suggest that it should be theoretically possible to perform transfer learning across Alzheimer's variants.
+While recent studies \cite{hon2017towards} tried to perform transfer learning by transferring knowledge from generic image datasets to Alzheimer's disease, we are not aware of any studies that have tried to transfer knowledge across different types of Alzheimer's variants or other dementias.
+Moreover, current transfer learning approaches focused only on improving diagnostic classification using supervised approaches.
+We are not aware of any study to date that has tried to estimate, through transfer learning, a continuous disease progression manifold that doesn't rely on diagnoses.
+There are two key benefits to performing transfer learning across dementia variants: 1.
+biomarker evolutions can be estimated in rare dementias for which there is not enough data (uni-modal datasets, few subjects, cross-sectional data only) and 2.
+adding extra information from other datasets can help with staging and biomarker trajectory estimation.
+
+We propose Disease Knowledge Transfer, a generative model that estimates progression of multiple dementias simultaneously and which inherently performs transfer learning from one type of dementia to another.
+The model achieves this by disentangling \emph{disease-specific} biomarker correlations from \emph{disease-agnostic} correlations.
+We fit the model simultaneously to three datasets: 1.
+the TADPOLE Challenge dataset containing subjects from the ADNI study containing MRI, FDG-PET, DTI, AV45 and AV1451 scans 2.
+MRI scans from typical AD subjects in our local centre and 3.
+MRI scans from patients with Posterior Cortical Atrophy (PCA) from our local centre.
+We then used the fitted model to predict non-MRI trajectories for PCA patients, which have been learned from the ADNI dataset.
+We finally validated the DTI trajectories in PCA using a small test set of 20 DTI scans from the PCA patients and controls from our local centre.
+% * <alexandra.young@ucl.ac.uk> 2018-03-02T17:54:07.562Z:
+% 
+% > unsupervised
+% Is it unsupervised or does it know which disease/dataset is which, in which case it's semi-supervised?
+% 
+% ^ <mrazvan22@gmail.com> 2018-03-02T18:05:17.625Z:
+% 
+% Model is indeed supervised with respect to dataset/disease, but unsupervised with respect to diagnosis within same disease (e.g.
+Control, MCI and AD in ADNI).
+I removed the supervised-ness though in order not to create confusion.
+%
+% ^.
+
+\begin{figure}[h]
+ \centering
+ \includegraphics[width=\textwidth]{figures/disease_knowledge_transfer.png}
+ \caption{Outline of the proposed framework for joint modelling of multiple diseases.
+We assume that each disease can be modelled as the evolution of some abstract dysfunctionality scores, each related to a brain region.
+Each dysfunctionality score is then assumed to be modelled as the progression of several biomarkers within that same region, but having different modalities.
+One key assumption of the model is that the biomarkers within each unit (bottom row) are disease agnostic (i.e.
+have the same dynamics for every disease), while the dysfunction scores (top row) are disease specific.}
+\end{figure}
+
+\subsection{Methods}
+
+For modelling disease-agnostic correlations, we assume biomarkers are grouped into \emph{functional units}, where correlations within each unit are disease-agnostic (e.g.
 biomarkers corresponding to the same ROI).
@@
 
-The proposed model is parsimonious, easily interpretable by abstracting away biomarker information intothe functional units, and can be used for \emph{disease knowledge transfer}, i.e.
+The proposed model is parsimonious, easily interpretable by abstracting away biomarker information into the functional units, and can be used for \emph{disease knowledge transfer}, i.e.
 learning correlations in one disease from another.
 Our approach is inspired by transfer learning methods from machine learning, but the model is generative, easing interpretability.
-
-\begin{figure}[h]
- \centering
- \includegraphics[height=11cm]{figures/disease_knowledge_transfer.png}
- \caption{Outline of the proposed framework for joint modelling of multiple diseases. }
-\end{figure}
-
-\subsection{Mathematical formulation}
 
@@
 \begin{equation}
- p(y_{.,.,.}|\theta_1, ..., \theta_K, \{\lambda^{\psi(l)} | l \in \Lambda \}, \beta_1, ..., \beta_N) = \prod_{(i,j,k) \in \Omega} \sum_{\beta_i^{\psi(k)}} p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) p(\beta_i^{\psi(k)}| \lambda^{\psi(k)}, \beta_i)
+ p(y_{.,.,.}|\theta_1, ..., \theta_K, \{\lambda^{\psi(l)} | l \in \Lambda \}, \beta_1, ..., \beta_N) = \\ \prod_{(i,j,k) \in \Omega} \sum_{\beta_i^{\psi(k)}} p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) p(\beta_i^{\psi(k)}| \lambda^{\psi(k)}, \beta_i)
 \end{equation}
@@
 Modelling of $p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k)$ and $p(\beta_i^{\psi(k)}| \lambda^{\psi(k)}, \beta_i)$ will be done with a non-parametric model (Gaussian Process, Lorenzi et al., 2017)
-
-\subsection*{Application 1: Disease Knowledge Transfer}
 
@@
 \begin{equation}
- p(y_{.,.,.}^{PCA}|\theta_1^{tAD}, ..., \theta_K^{tAD}, \{\lambda^{\psi(l)^{PCA}} | l \in \Lambda \}, \beta_1^{PCA}, ..., \beta_N^{PCA}) = \prod_{(i,j,k) \in \Omega} \sum_{\beta_i^{\psi(k)}} p(y_{ijk}^{PCA}| \beta_i^{\psi(k)}, \theta_k^{tAD}) p(\beta_i^{\psi(k)}| \lambda^{\psi(k)^{PCA}}, \beta_i^{PCA})
+ p(y_{.,.,.}^{PCA}|\theta_1^{tAD}, ..., \theta_K^{tAD}, \{\lambda^{\psi(l)^{PCA}} | l \in \Lambda \}, \beta_1^{PCA}, ..., \beta_N^{PCA}) = 
 \end{equation}
 
-\subsection*{Application 2: Improved biomarker predictions and staging}
+ \begin{equation}
+ \prod_{(i,j,k) \in \Omega} \sum_{\beta_i^{\psi(k)}} p(y_{ijk}^{PCA}| \beta_i^{\psi(k)}, \theta_k^{tAD}) p(\beta_i^{\psi(k)}| \lambda^{\psi(k)^{PCA}}, \beta_i^{PCA})
+\end{equation}
 
-We also hope to gain improved predictions and subject staging both in all the datasets used, both the larger ones as well as the smaller ones.
-We can fit the models for every functional unit using all the data available in order to learn the disease-agnostic correlations between biomarkers, and then fit the disease specific progression models.
-Our belief is that, by using more data to inform the disease-agnostic correlations, we will get better predictions.
+\section{Results}
+
+Fig.
+\ref{fig:occipUnit} shows the estimated biomarker trajectories within the \emph{occipital unit}, along with samples from the model posterior and aligned subject data.
+The model shows a good data fit, and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunctionality scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}.
+Fig.
+\ref{fig:occipUnit} shows the progression of dysfunctionality scores for (a) typical AD and (b) PCA.
+In typical AD, we see that temporal dysfunction becomes abnormal
+
+\newcommand{\expFld}{figures}
+
+% data fitting within one unit
+\begin{figure}[H]
+\includegraphics[width=\textwidth, trim=90 0 110 0, clip]{figures/unit1_allTraj_tad-drcTinyPen1_JMD.png} 
+\caption{Estimated biomarker trajectories in the Occipital Unit.
+Subject data from ADNI and our local cohort are also shown.
+The X-axis represents, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject.
+Red lines represent samples from the trajectory posterior.
+The Y-axis represents biomarker values which have been normalised to the [0,1] range.}
+\label{fig:occipUnit}
+\end{figure}
+
+% PCA vs tAD disease space
+\begin{figure}[H]
+\begin{subfigure}{0.47\textwidth}
+\centering
+typical AD\\
+\includegraphics[width=\textwidth, trim=0 0 0 20, clip]{figures/tAD_trajSameSpace_tad-drcTinyPen1_JMD.png} 
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.47\textwidth}
+\centering
+PCA\\
+\includegraphics[width=\textwidth, trim=0 0 0 20, clip]{figures/PCA_trajSameSpace_tad-drcTinyPen1_JMD.png} 
+\caption{}
+\end{subfigure}
+\caption{Progression of dysfunctionality scores for (a) typical AD and (b) PCA. In typical AD, we notice that the temporal scores are most affected after the disease }
+\label{fig:pcaTadDisSpace}
+\end{figure}
+
+% estimated (hypothetical) trajectories in PCA: DTI, FDG, AV45, AV1451.Volumetric trajectories were based on PCA MRI data.
+\begin{figure}[H]
+ \includegraphics[width=\textwidth, trim=0 0 0 0, clip]{figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen1_JMD.png}
+ \caption{Estimated trajectories for the PCA cohort.
+The only data that was available was the MRI volumetric data.
+The dynamics of the other biomarkers has been inferred by the model using data from typical AD, and taking into account the different spatial distribution of pathology in PCA as compared to typical AD.}
+ \label{fig:PCAtrajByModality}
+\end{figure}
+% * <mrazvan22@gmail.com> 2018-03-02T15:58:16.565Z:
+% 
+% some trajectories in AV45 screwed up due to bad initial starting point. Will try to re-fit the model if I have time.
+% 
+% ^.
+
+\section*{Validation}
+
+% LOCAL DTI validation
+
+\begin{figure}[H]
+ \includegraphics[width=\textwidth, trim=0 0 0 0, clip]{figures/validDtiPCA.png}
+ \caption{Validation of the Disease Knowledge Transfer Model using DTI data of PCA subjects from our local cohort.
+No DTI data, or other non-MRI data from PCA subjects was used to fit the model.
+The estimated trajectories seem to approximate the data reasonably well in temporal, hippocampus and cingulate regions, and less well in the other regions.}
+\label{fig:DTIvalid}
+\end{figure}
+
+\section{Acknowledgements}
+
+This work was supported by the EPSRC Centre For Doctoral Training in Medical Imaging with grant EP/L016478/1.
+AE received a McDonald Fellowship from the Multiple Sclerosis International Federation (MSIF, www.msif.org), and the ECTRIMS - MAGNIMS Fellowship.
+ALY was supported through an EPSRC Doctoral Prize Fellowship.
+NPO and SG received funding from the EU Horizon 2020 research and innovation programme under grant agreement No 666992.
+SJC was supported by an Alzheimer’s Research UK Senior Research Fellowship and ESRC/NIHR (ES/L001810/1) and EPSRC (EP/M006093/1) grants.
+DCA's work on this topic has funding from the EU Horizon 2020 research and innovation programme under grant agreement No 666992, as well as EPSRC grants J020990, M006093 and M020533.
+Data collection and sharing for this project was funded by the Alzheimer's Disease Neuroimaging Initiative (ADNI) (National Institutes of Health Grant U01 AG024904) and DOD ADNI (Department of Defense award number W81XWH-12-2-0012).
+The Dementia Research Centre is an ARUK coordination center.
+
+%
+% ---- Bibliography ---- 
+% USE HARVARD STANDARD
+
+\bibliographystyle{unsrtnat}
+\begin{thebibliography}{5}
+
+\bibitem{jack2010hypothetical}
+Jack, C.R., Knopman, D.S., Jagust, W.J., Shaw, L.M., Aisen, P.S., Weiner, M.W., Petersen, R.C.
+and Trojanowski, J.Q., 2010.
+Hypothetical model of dynamic biomarkers of the Alzheimer's pathological cascade.
+The Lancet Neurology, 9(1), pp.119-128.
+
+\bibitem{fonteijn2012event}
+Fonteijn, H.M., Modat, M., Clarkson, M.J., Barnes, J., Lehmann, M., Hobbs, N.Z., Scahill, R.I., Tabrizi, S.J., Ourselin, S., Fox, N.C.
+and Alexander, D.C., 2012.
+An event-based model for disease progression and its application in familial Alzheimer's disease and Huntington's disease.
+NeuroImage, 60(3), pp.1880-1889.
+
+\bibitem {jedynak2012}
+Jedynak, B.M., Lang, A., Liu, B., Katz, E., Zhang, Y., Wyman, B.T., Raunig, D., Jedynak, C.P., Caffo, B., Prince, J.L.
+and Alzheimer's Disease Neuroimaging Initiative, 2012.
+A computational neurodegenerative disease progression score: method and results with the Alzheimer's Disease Neuroimaging Initiative cohort.
+Neuroimage, 63(3), pp.1478-1486.
+
+\bibitem{donohue2014estimating}
+Donohue, M.C., Jacqmin-Gadda, H., Le Goff, M., Thomas, R.G., Raman, R., Gamst, A.C., Beckett, L.A., Jack, C.R., Weiner, M.W., Dartigues, J.F.
+and Aisen, P.S., 2014.
+Estimating long-term multivariate progression from short-term data.
+Alzheimer's \& Dementia, 10(5), pp.S400-S410.
+
+\bibitem{schiratti2015mixed}
+Schiratti, J.B., Allassonniere, S., Routier, A., Colliot, O., Durrleman, S.
+and Alzheimers Disease Neuroimaging Initiative, 2015, June.
+A mixed-effects model with time reparametrization for longitudinal univariate manifold-valued data.
+In International Conference on Information Processing in Medical Imaging (pp.
+564-575).
+Springer International Publishing.
+
+\bibitem{seeley2009neurodegenerative}
+Seeley, W.W., Crawford, R.K., Zhou, J., Miller, B.L. and Greicius, M.D., 2009. Neurodegenerative diseases target large-scale human brain networks. Neuron, 62(1), pp.42-52.
+
+\bibitem{crutch2012posterior}
+Crutch, S.J., Lehmann, M., Schott, J.M., Rabinovici, G.D., Rossor, M.N. and Fox, N.C., 2012. Posterior cortical atrophy. The Lancet Neurology, 11(2), pp.170-178.
+
+\bibitem{hon2017towards}
+Hon, M. and Khan, N., 2017. Towards Alzheimer's Disease Classification through Transfer Learning. arXiv preprint arXiv:1711.11117.
+
+\end{thebibliography}
+
+\clearpage
 

### 2018-03-02 JointModel.py

@@ -104,6 +104,13 @@ class JointModel(DisProgBuilder.DPMInterface):
         self.unitModels[u].minScX = self.unitModels[u].applyScalingX(self.unitModels[u].minX)
         self.unitModels[u].maxScX = self.unitModels[u].applyScalingX(self.unitModels[u].maxX)
 
+        plotTrajParamsFuncUnit = self.createPlotTrajParamsFuncUnit(nrCurrFuncUnit=u)
+        plotterObjCurrFuncUnit = Plotter.PlotterGP(plotTrajParamsFuncUnit)  # set separate plotter for the
+
+        # fig = plotterObjCurrFuncUnit.plotTraj(self.unitModels[u], replaceFig=True, legendExtraPlot=True)
+        # fig.savefig('%s/unit%s_allTraj_%s.png' % (self.outFolder, u, self.expName))
+
+
 
       # print(asda)
 
@@ -217,6 +224,12 @@ class JointModel(DisProgBuilder.DPMInterface):
         self.disModels[disNr].minScX = self.disModels[disNr].applyScalingX(self.disModels[disNr].minX)
         self.disModels[disNr].maxScX = self.disModels[disNr].applyScalingX(self.disModels[disNr].maxX)
 
+        plotTrajParamsDis = self.createPlotTrajParamsDis(disNr)
+        plotterCurrDis = Plotter.PlotterGP(plotTrajParamsDis)  # set separate plotter for the
+
+        # fig = plotterCurrDis.plotTrajSameSpace(self.disModels[disNr])
+        # fig.savefig('%s/%s_trajSameSpace_%s.png' % (self.outFolder, self.params['disLabels'][disNr],
+        #   self.expName))
 
     res = None
     return res
@@ -348,6 +361,8 @@ class JointModel(DisProgBuilder.DPMInterface):
       # print(plotTrajParamsDis['trueParams']['trueTrajXB'].shape)
       # print(adsa)
 
+    print(self.params['nrBiomkDisModel'])
+    # print(adssa)
 
     plotTrajParamsDis['labels'] = self.params['plotTrajParams']['unitNames']
     plotTrajParamsDis['colorsTraj'] = [colorsys.hsv_to_rgb(hue, 1, 1) for hue in

### 2018-03-02 Plotter.py

@@ -155,9 +155,15 @@ class PlotterGP:
   def __init__(self, plotTrajParams):
     self.plotTrajParams = plotTrajParams
 
-  def plotTraj(self, gpModel, replaceFig=True):
+  def plotTraj(self, gpModel, replaceFig=True, legendExtraPlot=False):
     nrBiomk = gpModel.N_biom
     # Plot method
+
+    font = {'family': 'normal',
+      'size': 13}
+
+    import matplotlib
+    matplotlib.rc('font', **font)
 
     figSizeInch = (self.plotTrajParams['SubfigTrajWinSize'][0] / 100, self.plotTrajParams['SubfigTrajWinSize'][1] / 100)
     fig = pl.figure(1, figsize = figSizeInch)
@@ -174,8 +180,11 @@ class PlotterGP:
     min_yB, max_yB = gpModel.getMinMaxY_B(extraDelta=0)
     deltaB = [(max_yB[b] - min_yB[b]) * 0.2 for b in range(nrBiomk)]
 
-    nrRows = self.plotTrajParams['nrRows']
-    nrCols = self.plotTrajParams['nrCols']
+    # nrRows = self.plotTrajParams['nrRows']
+    # nrCols = self.plotTrajParams['nrCols']
+
+    nrRows = 2
+    nrCols = 3
 
     newX = gpModel.getXsMinMaxRange()
     predBiomksYscaledXB = gpModel.predictBiomk(newX)
@@ -212,12 +221,49 @@ class PlotterGP:
 
         diagCounters[diagCurrSubj] += 1
 
-      ax.plot(newX,predBiomksYscaledXB[:,b], lw=2, color='black', label='estim traj')
+      ax.plot(newX,predBiomksYscaledXB[:,b], lw=2, color='black', label='estimated trajectory')
 
-      ax.plot(np.array([np.min(newX), np.max(newX)]), [min_yB[b], max_yB[b]],
-        color=(0.5,0.5,0.5), lw=2)
+      # ax.plot(np.array([np.min(newX), np.max(newX)]), [min_yB[b], max_yB[b]],
+      #   color=(0.5,0.5,0.5), lw=2)
 
       ax.set_ylim([min_yB[b]-deltaB[b], max_yB[b]+deltaB[b]])
+
+      # ax.legend(ncol=1)
+
+    if legendExtraPlot:
+      ax = pl.subplot(nrRows, nrCols, nrBiomk+1)
+      pl.axis('off')
+
+      # plot subject data
[…]

### 2018-03-02 auxFunc.py

@@ -184,6 +184,6 @@ def findOptimalRowsCols(nrBiomk):
   potentialLayout = [(2, 2), (2, 3), (3, 3), (3, 4), (4, 4), (4, 5), (4, 6), (5, 6), (5,7), (6,7)]
 
   optimalRows, optimalCols = ([potentialLayout[i] for i in range(len(potentialLayout)) if
-                               potentialLayout[i][0] * potentialLayout[i][1] > nrBiomk])[0]
+                               potentialLayout[i][0] * potentialLayout[i][1] >= nrBiomk])[0]
 
   return optimalRows, optimalCols

### 2018-03-02 tadpoleDrc.py

@@ -105,8 +105,8 @@ plotTrajParams['diagColors'] = {CTL:'g', MCI:'y', AD:'r',
 plotTrajParams['diagScatterMarkers'] = {CTL:'o', MCI:'o', AD:'o',
   CTL2:'x', PCA:'x', AD2:'x'}
 plotTrajParams['legendCols'] = 4
-plotTrajParams['diagLabels'] = {CTL:'CTL ADNI', MCI:'MCI ADNI', AD:'AD ADNI',
-  CTL2:'CTL DRC', PCA:'PCA DRC', AD2:'AD DRC'}
+plotTrajParams['diagLabels'] = {CTL:'CTL ADNI', MCI:'MCI ADNI', AD:'tAD ADNI',
+  CTL2:'CTL LOCAL', PCA:'PCA LOCAL', AD2:'tAD LOCAL'}
 
 plotTrajParams['freesurfPath'] = freesurfPath
 # plotTrajParams['ylimitsRandPoints'] = (-3,2)
@@ -1376,12 +1376,13 @@ def runAllExpTadpoleDrc(params, expName, dpmBuilder, compareTrueParamsFunc = Non
   params['excludeStaging'] = [-1]
 
   params['outFolder'] = 'resfiles/%s' % expName
+  params['expName'] = expName
 
   dpmObjStd, res['std'] = evaluationFramework.runStdDPM(params,
     expName, dpmBuilder, params['runPartMain'])
 
 
-  plotAllBiomkDisSpace(dpmObjStd, params, disNr=0)
+  # plotAllBiomkDisSpace(dpmObjStd, params, disNr=0)
 
   # perform the validation against DRC data
   validateDRCBiomk(dpmObjStd, params)
@@ -1580,12 +1581,34 @@ def validateDRCBiomk(dpmObj, params):
   # print(ads)
 
 
-  fig = dpmObj.plotterObj.plotTrajInDisSpace(xsTrajX, predTrajXB, trajSamplesBXS,
-    XshiftedDisModelBS, Yfilt, diagSubjCurrDis,
-    XvalidShifFilt, YvalidFilt, diagValidFilt, replaceFig=True)
-  fig.savefig('%s/validPCA.png' % params['outFolder'])
+  # fig = dpmObj.plotterObj.plotTrajInDisSpace(xsTrajX, predTrajXB, trajSamplesBXS,
+  #   XshiftedDisModelBS, Yfilt, diagSubjCurrDis,
[…]

### 2018-03-02 paper/paper.tex
m

@@
 
+% write background in abstract. motivation. take out
+
 We present Disease Knowledge Transfer (DKT), a technique for transferring biomarker correlations between Alzheimer's disease (AD) variants.
@@
 A hypothetical model of disease progression has been previously published by \cite{jack2010hypothetical}, which proposes that the first biomarkers to become abnormal are measures of amyloid beta aggregation, followed by tau abnormalities, hypometabolism, structural MRI-based measures and finally cognitive decline.
-While this hypothetical model proposes a modality-specific ordering of biomarkers in typical AD, it has also been observed that within the same modality, biomarker measurements have a spatial sequence of abnormality that correlates with Braak stages: hippocampal volumes and entorhinal measures become abnormal first, followed by other structures within the temporal lobe, followed by parietal and frontal abnormalities.
+While this hypothetical model proposes a inter-modality ordering of biomarkers in typical AD, it has also been observed that within the same modality, biomarker measurements have a spatial sequence of abnormality that correlates with Braak stages: hippocampal volumes and entorhinal measures become abnormal first, followed by other structures within the temporal lobe, followed by parietal and frontal abnormalities.
 However, the spatial ordering of abnormality differs in other Alzheimer's variants.
 For example, in Posterior Cortical Atrophy, posterior regions such as the occipital lobe and superior parietal regions have been shown to become affected.
-The network vulnerability theory by Seeley et al.
+More generally, the network vulnerability theory by Seeley et al.
 \cite{seeley2009neurodegenerative} suggests that damage to specific underlying connectivity networks leads to spatially distinct patterns of neurodegeneration.
-There is nevertheless a great degree of heterogeneity within these diseases, which led some to suggest that subjects with different Alzheimer's variants lie on a continuum of phenotypical variation \cite{crutch2012posterior}.
+Overall, there is a great degree of heterogeneity within these diseases, which led some to suggest that subjects with different Alzheimer's variants lie on a continuum of phenotypical variation \cite{crutch2012posterior}.
 % * <alexandra.young@ucl.ac.uk> 2018-03-02T19:01:16.243Z:
@@
 \subsection{Methods}
-
+% TODO: this was copy-pasted from a draft 2 weeks ago. needs re-writing!
 For modelling disease-agnostic correlations, we assume biomarkers are grouped into \emph{functional units}, where correlations within each unit are disease-agnostic (e.g.
@@
 For example, biomarkers assigned to the temporal unit could be: amyloid in temporal lobe, tau in temporal lobe, MRI atrophy in temporal lobe and certain memory tests.
-Using a disease progression model of our choice, for each region-specific functional unit we can model the correlation of biomarkers within that functional unit (Figure 1, bottom), which allows us to reconstruct a region- or unit-specific \emph{dysfunction} progression axis which can be used for staging subjects.
+Using an arbitrary disease progression model (e.g.
+CITE), for each region-specific functional unit we can model the correlation of biomarkers within that functional unit (Figure 1, bottom), which allows us to reconstruct a region- or unit-specific \emph{dysfunction} progression axis which can be used for staging subjects.
 Finally, we express the disease specific correlations (Figure 1, top) not directly with each biomarkers, but via the dysfunction scores estimated within each functional unit.
@@
  \item $\lambda^l$ = parameters of trajectory for functional unit $l$, where $l \in \Lambda$
- \item $\psi$ : \{1, ..., K\} $ \rightarrow Lambda$ is a function that maps beach biomarker to its corresponding functional unit
+ \item $\psi$ : \{1, ..., K\} $ \rightarrow \Lambda$ is a function that maps beach biomarker to its corresponding functional unit
  \item $\beta_i$: time shift parameter for subject $i$ (disease specific)
@@
 
+% Fig1: biomarker traj.
+over dysfunction scores in one functional unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
 Fig.
-\ref{fig:occipUnit} shows the estimated biomarker trajectories within the \emph{occipital unit}, along with samples from the model posterior and aligned subject data.
+\ref{fig:occipUnit} shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data.
+The X-axis shows the dysfunctionality scores within the occipital unit, which represent estimated time-shifts, in months, from an arbitrary reference X=0, while the Y-axis shows biomarker values normalised to [0,1] range.
 The model shows a good data fit, and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunctionality scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}.
 Fig.
-\ref{fig:occipUnit} shows the progression of dysfunctionality scores for (a) typical AD and (b) PCA.
-In typical AD, we see that temporal dysfunction becomes abnormal
+\ref{fig:occipUnit} shows the progression of dysfunctionality scores over the disease stage for (a) typical AD and (b) PCA (disease ).
+In typical AD, we see that temporal dysfunction becomes abnormal earliest, while PCA shows early parietal dysfunction, again in line with previous findings in the literature \cite{crutch2012posterior} (J.C.
+Baron, Neuroimage, 2001).
+In Fig.
+\ref{fig:PCAtrajByModality} we plot the inferred biomarker trajectories for PCA across the five different modalities: MRI Volumes, DTI, FDG, AV45 and AV1451.
+The trajectories
 
@@
 % data fitting within one unit
-\begin{figure}[H]
+\begin{figure}
 \includegraphics[width=\textwidth, trim=90 0 110 0, clip]{figures/unit1_allTraj_tad-drcTinyPen1_JMD.png} 
@@
 Red lines represent samples from the trajectory posterior.
-The Y-axis represents biomarker values which have been normalised to the [0,1] range.}
+The Y-axis measures biomarker values (normalised).}
 \label{fig:occipUnit}
@@
 % PCA vs tAD disease space
-\begin{figure}[H]
+\begin{figure}
 \begin{subfigure}{0.47\textwidth}

### 2018-03-03 paper/paper.tex
commited current version

@@
 
-\title{Disease Knowledge Transfer for Alzheimer's Variants}
+\title{Disease Knowledge Transfer across Alzheimer's Variants}
 %
@@
 % 
-% Author list: Me, Pere (helped with validation), Marco (used his code), Alex (writing+tadpole), Neil (tadpole),  Arman, Keir (DTI data), Seb, Danny
+% Author list: Me, Marco (writing+model code), Pere (helped with validation), Alex (writing+tadpole), Neil (tadpole),  Arman, Keir (DTI data), Seb, Danny
 % 
@@
 
-% write background in abstract. motivation. take out
-
-We present Disease Knowledge Transfer (DKT), a technique for transferring biomarker correlations between Alzheimer's disease (AD) variants.
-DKT allows, for the first time, to infer biomarker values and their progressions in rare variants of AD for which data on such biomarkers is not available, by learning them from a different disease.
-For doing this, we formulate a new dementias paradigm called "Affected Overlap", which proposes that dementias are diseases that affect overlapping brain regions, and thus present certain biomarker characteristics that are shared and can thus be transferred across diseases.
-We then formulate the DKT model implementing this paradigm as a joint-disease generative model of biomarker progression which disentangles the biomarker correlations that are disease-specific from those correlations that are disease-agnostic.
-We demonstrate DKT on three datasets: (1) the TADPOLE Challenge dataset containing typical AD subjects (tAD) from the Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset who had MRI, FDG-PET, AV45, AV1451 or DTI scans, as well as two datasets from our local centre containing subjects with (2) typical AD and (3) Posterior Cortical Atrophy (PCA) that have only undergone MRI scans.
-We train the DKT model on all three datasets at once, and show that it is able to predict, in PCA subjects, plausible population-level biomarker for FDG, DTI, AV45 and AV1451 biomarkers, for which no data was available.
-We validate the predictions of DKT using a small set of DTI scans for the PCA cohort.
-We demonstrate that DKT may be a useful tool to analyse and understand rare forms of dementias for which multimodal data is not available or is limited (e.g.
-cross-sectional).
-Finally, by leveraging data from multiple diseases, DKT also has the potential to provide more accurate disease staging compared to traditional disease progression models, which is useful for patient stratification in clinical trials.
-% * <alexandra.young@ucl.ac.uk> 2018-03-02T17:28:57.289Z:
-% 
-% > Finally,
-% You've got finally twice
-% 
-% ^ <alexandra.young@ucl.ac.uk> 2018-03-02T17:29:38.345Z:
-% 
-% Could put we demonstrate (or something a bit weaker)
-%
-% ^ <alexandra.young@ucl.ac.uk> 2018-03-02T19:02:52.149Z.
+The availability of multimodal biomarker measurements in large scale clinical data is fundamental for the application of current disease progression models of neurodegenerative disorders.
+For this reason, disease progression modeling of rare dementias is currently very challenging, due to the presence of missing data and to the low sample size.
+To overcome this problem, in this study we introduce Disease Knowledge Transfer (DKT), a technique for transferring biomarker's progression models between Alzheimer's disease (AD) variants.
+We assume that different type of dementias affect overlapping brain regions ("Affected Overlap"), and thus present shared biomarker characteristics that can be transferred across diseases.
+We then implement this paradigm as a joint-disease generative model of biomarker progressions, disentangling disease-specific from  disease-agnostic biomarker relationships.
+We demonstrate DKT on independent datasets of 1) typical AD data (tAD), with large sample size and several biomarkers available over time, and 2) Posterior Cortical Atrophy (PCA) data, for which only MRI scans are available.
+DKT is able to predict, in testing PCA subjects, plausible population-level biomarkers for structural and molecular imaging biomarkers, for which no data was available.
+DKT may be a useful tool to analyse and understand rare forms of dementias for which multimodal data is not available or is limited.
+Moreover, by leveraging data from multiple diseases, DKT also has the potential to provide more accurate disease staging compared to traditional disease progression models.
 
@@
 % biomarkers in alzheimer's -> measuring the evolution helps staging in clinical trials
-There currently exist several image-based biomarkers for Alzheimer's disease that can be used to track its progression: AV45 Positron Emission Tomography (PET) measuring amyloid-plaque aggregation, AV1451 PET measuring tau tangle aggregation, Flourodeoxyglucose (FDG) PET measuring brain hypometabolism, Magnetic Resonance Imaging (MRI) measuring structural integrity and Diffusion Tensor Imaging (DTI) measuring connectivity integrity.
-Measuring the exact evolution of these biomarkers over the disease progression can lead to better disease staging of patients, which helps stratification in clinical trials.
+Several image-based biomarkers for Alzheimer's disease (AD) are currently used to track the progression of the pathology: AV45 Positron Emission Tomography (PET) measuring amyloid-plaque aggregation, AV1451 PET measuring Tau tangle aggregation, Flourodeoxyglucose (FDG) PET measuring brain hypometabolism, Magnetic Resonance Imaging (MRI) measuring structural integrity and Diffusion Tensor Imaging (DTI) measuring connectivity integrity.
+Measuring the exact evolution of these biomarkers over the disease progression is fundamental for better understanding the pathology, as well as for improving the stratification in clinical trials through patient staging.
 
-% hypothetical model has specific ordering of modality biomarkers -> ordering or spatial biomarkers (different dimension) -> spatial differences in different dementias -> network vulnerability hypothesis.
+% hypothetical model -> dimention 1: ordering across modalities -> dimention 2: spatial ordering
 A hypothetical model of disease progression has been previously published by \cite{jack2010hypothetical}, which proposes that the first biomarkers to become abnormal are measures of amyloid beta aggregation, followed by tau abnormalities, hypometabolism, structural MRI-based measures and finally cognitive decline.
 While this hypothetical model proposes a inter-modality ordering of biomarkers in typical AD, it has also been observed that within the same modality, biomarker measurements have a spatial sequence of abnormality that correlates with Braak stages: hippocampal volumes and entorhinal measures become abnormal first, followed by other structures within the temporal lobe, followed by parietal and frontal abnormalities.
-However, the spatial ordering of abnormality differs in other Alzheimer's variants.
-For example, in Posterior Cortical Atrophy, posterior regions such as the occipital lobe and superior parietal regions have been shown to become affected.
-More generally, the network vulnerability theory by Seeley et al.
-\cite{seeley2009neurodegenerative} suggests that damage to specific underlying connectivity networks leads to spatially distinct patterns of neurodegeneration.
-Overall, there is a great degree of heterogeneity within these diseases, which led some to suggest that subjects with different Alzheimer's variants lie on a continuum of phenotypical variation \cite{crutch2012posterior}.
-% * <alexandra.young@ucl.ac.uk> 2018-03-02T19:01:16.243Z:
-% 
-% > The network vulnerability theory by Seeley et al.
-\cite{seeley2009neurodegenerative} suggests that damage to specific underlying connectivity networks leads to spatially distinct patterns of neurodegeneration.
-There is nevertheless a great degree of heterogeneity within these diseases, which led some to suggest that subjects with different Alzheimer's variants lie on a continuum of phenotypical variation \cite{crutch2012posterior}.
-% The relationship between these two statements is unclear from the way its written
-% 
-% ^.
 
-The amount of heterogeneity within each variant and the network vulnerability hypothesis by \cite{seeley2009neurodegenerative} suggest that it should be theoretically possible to perform transfer learning across Alzheimer's variants.
-While recent studies \cite{hon2017towards} tried to perform transfer learning by transferring knowledge from generic image datasets to Alzheimer's disease, we are not aware of any studies that have tried to transfer knowledge across different types of Alzheimer's variants or other dementias.
-Moreover, current transfer learning approaches focused only on improving diagnostic classification using supervised approaches.
-We are not aware of any study to date that has tried to estimate, through transfer learning, a continuous disease progression manifold that doesn't rely on diagnoses.
+Several data-driven methods have been proposed in order to reconstruct group-wise long term biomarker progressions from collections of short term individual biomarker measurements (\cite{lorenzi2017disease}, Young, Donohue).
+These approaches mostly rely on the estimation of a latent time reparameterization associated with each individual, often encoded by a translation (time shift) of the subject's measurements over the temporal axis.
+This time reparameterization can be used as a proxy of disease staging associated with an individual biomarker profile relatively to the global disease progression model.
+
+% Flow: multimodal data needed for disease progression -> disease progression on rare dementias is challenging -> cannot apply model learned from other disease because of different spatial patterns -> but studies show at least some partial spatial overlap -> therefore transfer learning possible due to presence of "some overlap" -> benefits of transfer learning -> existing transfer learning literature
+The availability of multimodal collections of biomarker's measurements across clinical groups is fundamental to the application of such disease progression models.
+For this reason, the application of disease progression modeling to rare dementia type is currently very challenging, due to the associated problem of missing biomarkers and low sample size.
+Moreover, an average model of disease progression estimated from sporadic AD cases may not generalize to specific disease variants.
+For example, in Posterior Cortical Atrophy (PCA), posterior regions such as the occipital lobe and superior parietal regions have been shown to become affected early.
+While the spatial patterns are distinct between various forms of dementia, recent studies (Ossenkoppele, HBM, 2015) have showed at least partial spatial-overlap between different clinical phenotypes, while others have even suggested that PCA and typical AD subjects lie on a continuum of phenotypical variation \cite{crutch2012posterior}.
+The presence of overlap in pathology patterns across different dementias suggests that it should be theoretically possible to perform transfer learning across the diseases.
 There are two key benefits to performing transfer learning across dementia variants: 1.
-biomarker evolutions can be estimated in rare dementias for which there is not enough data (uni-modal datasets, few subjects, cross-sectional data only) and 2.
+biomarker evolutions can be estimated for rare dementias for which there is not enough data (unimodal datasets, few subjects, cross-sectional data only) and 2.
 adding extra information from other datasets can help with staging and biomarker trajectory estimation.
+While recent studies \cite{hon2017towards} tried to transfer knowledge from generic image datasets to Alzheimer's disease, we are not aware of any studies that have tried to transfer knowledge across different types of Alzheimer's variants or other dementias.
+Furthermore, we are not aware of any studies to have done transfer learning in medical imaging for estimating a continuous disease progression without reliance on diagnostic classes, which are often biased.
 
-We propose Disease Knowledge Transfer, a generative model that estimates progression of multiple dementias simultaneously and which inherently performs transfer learning from one type of dementia to another.
-The model achieves this by disentangling \emph{disease-specific} biomarker correlations from \emph{disease-agnostic} correlations.
-We fit the model simultaneously to three datasets: 1.
-the TADPOLE Challenge dataset containing subjects from the ADNI study containing MRI, FDG-PET, DTI, AV45 and AV1451 scans 2.
-MRI scans from typical AD subjects in our local centre and 3.
-MRI scans from patients with Posterior Cortical Atrophy (PCA) from our local centre.
-We then used the fitted model to predict non-MRI trajectories for PCA patients, which have been learned from the ADNI dataset.
+% To overcome this problem, in this study we aim at defining a probabilistic framework for the estimation of disease progression trajectories in small datasets of specific disease subtypes, in presence of missing biomarkers and low numerosity.
+To this end we introduce the idea of \emph{transfer learning} of disease progression models across datasets.
+In particular, our framework is formulated in a probabilistic fashion, in order to quantify the uncertainty associated with low sample size and high variability characterizing the problem.
+
+We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal progressions for multiple dementias simultaneously and which inherently performs transfer learning between the modelled dementia phenotypes.
+This is achieved by disentangling \emph{disease-specific} from \emph{disease-agnostic} biomarker relationship.
+We fit the DKT model to three datasets simultaneously: (1) the TADPOLE Challenge dataset containing subjects from the ADNI study with MRI, FDG-PET, DTI, AV45 and AV1451 scans, (2) MRI scans from typical AD subjects in our local centre and (3) MRI scans from patients with Posterior Cortical Atrophy (PCA) from our local centre.
+We then used the fitted model to predict non-MRI trajectories for PCA patients, which have been learned from the multimodal TADPOLE dataset.
 We finally validated the DTI trajectories in PCA using a small test set of 20 DTI scans from the PCA patients and controls from our local centre.
+
 % * <alexandra.young@ucl.ac.uk> 2018-03-02T17:54:07.562Z:
@@
  \caption{Outline of the proposed framework for joint modelling of multiple diseases.
-We assume that each disease can be modelled as the evolution of some abstract dysfunctionality scores, each related to a brain region.
+We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores, each related to different brain regions.
 Each dysfunctionality score is then assumed to be modelled as the progression of several biomarkers within that same region, but having different modalities.
@@
 have the same dynamics for every disease), while the dysfunction scores (top row) are disease specific.}
+ \label{fig:diagram}
 \end{figure}
@@
 \subsection{Methods}
-% TODO: this was copy-pasted from a draft 2 weeks ago. needs re-writing!
-For modelling disease-agnostic correlations, we assume biomarkers are grouped into \emph{functional units}, where correlations within each unit are disease-agnostic (e.g.
-biomarkers corresponding to the same ROI).
-Each biomarker is assigned a-priori to a specific functional unit, which corresponds to a brain region, say temporal lobe (Fig 1., bottom).
-For example, biomarkers assigned to the temporal unit could be: amyloid in temporal lobe, tau in temporal lobe, MRI atrophy in temporal lobe and certain memory tests.
-Using an arbitrary disease progression model (e.g.
-CITE), for each region-specific functional unit we can model the correlation of biomarkers within that functional unit (Figure 1, bottom), which allows us to reconstruct a region- or unit-specific \emph{dysfunction} progression axis which can be used for staging subjects.
+
+Fig.
+\ref{fig:diagram} shows the overall diagram of our proposed framework for joint modelling of disease.
+We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores, each related to different brain regions (top row).
+Each dysfunctionality score is then assumed to be modelled as the progression of several biomarkers within that same region, but having different modalities (bottom row).
+Each group of biomarkers in the bottom row will be called a \emph{functional unit}, because the correlations between biomarkers are related though common "function" in a disease agnostic way, because they are related to the same underlying brain region.
+Biomarker groupings into functional units are defined a-priori.
+We model the correlations within each unit using a Gaussian-Process based disease progression model \cite{lorenzi2017disease}, for each region-specific functional unit we can model the correlation of biomarkers within that functional unit (Figure 1, bottom), which allows us to reconstruct a region- or unit-specific \emph{dysfunction} progression axis which can be used for staging subjects.
 Finally, we express the disease specific correlations (Figure 1, top) not directly with each biomarkers, but via the dysfunction scores estimated within each functional unit.
@@
 Fig.
-\ref{fig:occipUnit} shows the progression of dysfunctionality scores over the disease stage for (a) typical AD and (b) PCA (disease ).
+\ref{fig:occipUnit} shows the progression of dysfunctionality scores over the disease stage for (a) typical AD and (b) PCA.
 In typical AD, we see that temporal dysfunction becomes abnormal earliest, while PCA shows early parietal dysfunction, again in line with previous findings in the literature \cite{crutch2012posterior} (J.C.
@@
 In Fig.
-\ref{fig:PCAtrajByModality} we plot the inferred biomarker trajectories for PCA across the five different modalities: MRI Volumes, DTI, FDG, AV45 and AV1451.
-The trajectories
+\ref{fig:PCAtrajByModality}, we plot the inferred biomarker trajectories for PCA directly across the disease progression.
+We do this for five different modalities: MRI Volumes, DTI, FDG, AV45 and AV1451.
+The results again recapitulate known patterns in PCA, where trajectories from posterior regions become abnormal first in MRI, DTI and AV1451, although in FDG we instead see temporal and hippocampal regions more affected.
+This might be due to the normalisation of the FDG trajectories on the Y-axis to the [0,1] range, which results in less scaling to trajectories that contain more disease signal.
 
@@
 % 
-% some trajectories in AV45 screwed up due to bad initial starting point. Will try to re-fit the model if I have time.
+% two trajectories in AV45 screwed up due to bad initial starting point.
+Will take around 2-3h to fix, by implementing a more informed starting point, and re-fit the model.
+I'll try to see if I have time, otherwise might drop that modality?!
+Or take out the plot entirely.
 % 
@@
 \section*{Validation}
+
+We performed validation using a separate test set of 20 DTI scans from controls and PCA patients from our local cohort.
+Fig.
+\ref{fig:DTIvalid} shows the estimated DTI biomarker trajectories plotted directly against the disease progression time-shifts, along with the DTI data from the test set.
+The time shift of each subject was estimated only based on their MRI scans.
+The inferred trajectories in PCA show good agreement with the data in the cingulate, hippocampus and temporal lobes, but less agreement in the other regions.
+This might be because the dataset that was used to train the model had better disease signal in DTI regions more traditionally associated with typical AD.
 
@@
  \caption{Validation of the Disease Knowledge Transfer Model using DTI data of PCA subjects from our local cohort.
-No DTI data, or other non-MRI data from PCA subjects was used to fit the model.
-The estimated trajectories seem to approximate the data reasonably well in temporal, hippocampus and cingulate regions, and less well in the other regions.}
+Each subject has been shifted along the X-axis only according to their MRI data.
+No DTI data from PCA subjects was used to fit the model.}
 \label{fig:DTIvalid}
@@
 
-\section{Acknowledgements}
+\section{Discussion}
 
-This work was supported by the EPSRC Centre For Doctoral Training in Medical Imaging with grant EP/L016478/1.
-AE received a McDonald Fellowship from the Multiple Sclerosis International Federation (MSIF, www.msif.org), and the ECTRIMS - MAGNIMS Fellowship.
-ALY was supported through an EPSRC Doctoral Prize Fellowship.
-NPO and SG received funding from the EU Horizon 2020 research and innovation programme under grant agreement No 666992.
-SJC was supported by an Alzheimer’s Research UK Senior Research Fellowship and ESRC/NIHR (ES/L001810/1) and EPSRC (EP/M006093/1) grants.
-DCA's work on this topic has funding from the EU Horizon 2020 research and innovation programme under grant agreement No 666992, as well as EPSRC grants J020990, M006093 and M020533.
-Data collection and sharing for this project was funded by the Alzheimer's Disease Neuroimaging Initiative (ADNI) (National Institutes of Health Grant U01 AG024904) and DOD ADNI (Department of Defense award number W81XWH-12-2-0012).
-The Dementia Research Centre is an ARUK coordination center.
+% summary: what we developed -> what the model can do
+We present here, for the first time, a framework for joint modelling of biomarker progression in multiple dementias simultaneously.
+The framework allows transferring biomarker trajectories to rare dementias for which there is not enough data to allow estimation of such trajectories, accounting for a different spatial distribution of pathology between dementia varieties.
+
+% limitations: each region needs to have disease signal -> validation did not work well in some regions -> not accounted for subject specific effects. 
+The model and the results presented here have several limitations.
+First of all, the model assumes that within each region there is enough disease signal within each region to propely estimate the dysfunctionality scores.
+Another limitation is that, in the validation, the inferred trajectories in posterior regions traditionally associted with PCA do not match well with the validation data.
+One explanation for this might be the lack of disease signal in posterior regions in DTI data during training, as per the previous limitation.
+Finally, in terms of modelling, our framework did not account for subject specific effects, as we were only interested in estimating population-level trajectories.
+
+% how to address aforementioned limitations + future work
+There are several potential avenues for further research.
+During training, we can account for missing disease signal in some brain regions by adding other types of dementias that affect other regions, such as various types of Fronto-temporal Dementia or Primary Progressive Aphasia.
+The estimated PCA trajectories also need to be thoroughly validated using data from modalities other than DTI.
+The model formulation can also be naturally extended to include subject-specific effects, like in the work of Lorenzi et al., 2017.
 
@@
 
+\bibitem{lorenzi2017disease}
+Lorenzi, M., Filippone, M., Alexander, D.C.
+and Ourselin, S., 2017.
+Disease Progression Modeling and Prediction through Random Effect Gaussian Processes and Time Transformation.
+arXiv preprint arXiv:1701.01668.
+
 \end{thebibliography}

### 2018-03-03 paper/paper.tex
8 pagesgst

@@
 
+\author{*******************************************}
+
+\institute{***************************************}
+
 \maketitle              % typeset the title of the contribution
@@
 \begin{abstract}
-
 The availability of multimodal biomarker measurements in large scale clinical data is fundamental for the application of current disease progression models of neurodegenerative disorders.
-For this reason, disease progression modeling of rare dementias is currently very challenging, due to the presence of missing data and to the low sample size.
+Due to the lack of some data and to the low sample size, disease progression modeling of rare dementias is currently very challenging.
 To overcome this problem, in this study we introduce Disease Knowledge Transfer (DKT), a technique for transferring biomarker's progression models between Alzheimer's disease (AD) variants.
-We assume that different type of dementias affect overlapping brain regions ("Affected Overlap"), and thus present shared biomarker characteristics that can be transferred across diseases.
+We assume that different types of dementia affect overlapping brain regions, and thus present shared biomarker characteristics that can be transferred across diseases.
 We then implement this paradigm as a joint-disease generative model of biomarker progressions, disentangling disease-specific from  disease-agnostic biomarker relationships.
-We demonstrate DKT on independent datasets of 1) typical AD data (tAD), with large sample size and several biomarkers available over time, and 2) Posterior Cortical Atrophy (PCA) data, for which only MRI scans are available.
-DKT is able to predict, in testing PCA subjects, plausible population-level biomarkers for structural and molecular imaging biomarkers, for which no data was available.
-DKT may be a useful tool to analyse and understand rare forms of dementias for which multimodal data is not available or is limited.
+We demonstrate DKT on independent datasets of 1) multimodal typical AD data (tAD), with large sample size and several biomarkers available over time, and 2) Posterior Cortical Atrophy (PCA) data, for which only MRI scans are available.
+DKT is able to predict, in tested PCA subjects, plausible population-level biomarkers for structural and molecular imaging biomarkers, for which no data were available.
+DKT may be a useful tool to analyse and understand rare forms of dementia for which multimodal data is not available or is limited.
 Moreover, by leveraging data from multiple diseases, DKT also has the potential to provide more accurate disease staging compared to traditional disease progression models.
@@
 Several image-based biomarkers for Alzheimer's disease (AD) are currently used to track the progression of the pathology: AV45 Positron Emission Tomography (PET) measuring amyloid-plaque aggregation, AV1451 PET measuring Tau tangle aggregation, Flourodeoxyglucose (FDG) PET measuring brain hypometabolism, Magnetic Resonance Imaging (MRI) measuring structural integrity and Diffusion Tensor Imaging (DTI) measuring connectivity integrity.
-Measuring the exact evolution of these biomarkers over the disease progression is fundamental for better understanding the pathology, as well as for improving the stratification in clinical trials through patient staging.
+Measuring the exact evolution of these biomarkers over the disease progression is fundamental for better understanding of the pathology, as well as for improving the stratification in clinical trials through patient staging.
 
@@
 A hypothetical model of disease progression has been previously published by \cite{jack2010hypothetical}, which proposes that the first biomarkers to become abnormal are measures of amyloid beta aggregation, followed by tau abnormalities, hypometabolism, structural MRI-based measures and finally cognitive decline.
-While this hypothetical model proposes a inter-modality ordering of biomarkers in typical AD, it has also been observed that within the same modality, biomarker measurements have a spatial sequence of abnormality that correlates with Braak stages: hippocampal volumes and entorhinal measures become abnormal first, followed by other structures within the temporal lobe, followed by parietal and frontal abnormalities.
+While this hypothetical model proposes a inter-modality ordering of biomarkers in typical AD, it has also been observed that within the same modality, biomarker measurements have a spatial sequence of abnormality that correlates with Braak stages: hippocampal volumes and entorhinal measures become abnormal first, followed by other structures within the temporal, parietal and frontal lobes.
 
-Several data-driven methods have been proposed in order to reconstruct group-wise long term biomarker progressions from collections of short term individual biomarker measurements (\cite{lorenzi2017disease}, Young, Donohue).
+Several data-driven methods have been proposed in order to reconstruct group-wise long term biomarker progressions from collections of short term individual biomarker measurements (\cite{lorenzi2017disease,young2014data,donohue2014estimating}).
 These approaches mostly rely on the estimation of a latent time reparameterization associated with each individual, often encoded by a translation (time shift) of the subject's measurements over the temporal axis.
@@
 Moreover, an average model of disease progression estimated from sporadic AD cases may not generalize to specific disease variants.
-For example, in Posterior Cortical Atrophy (PCA), posterior regions such as the occipital lobe and superior parietal regions have been shown to become affected early.
-While the spatial patterns are distinct between various forms of dementia, recent studies (Ossenkoppele, HBM, 2015) have showed at least partial spatial-overlap between different clinical phenotypes, while others have even suggested that PCA and typical AD subjects lie on a continuum of phenotypical variation \cite{crutch2012posterior}.
+For example, in Posterior Cortical Atrophy (PCA), posterior regions such as the occipital lobe and superior parietal regions have been shown to be affected early.
+While the spatial patterns are distinct between various forms of dementia, recent studies \cite{Ossenkoppele2014atrophy} have showed at least partial spatial-overlap between different clinical phenotypes, while others have further suggested that PCA and typical AD subjects lie on a continuum of phenotypical variation \cite{crutch2012posterior}.
 The presence of overlap in pathology patterns across different dementias suggests that it should be theoretically possible to perform transfer learning across the diseases.
@@
 adding extra information from other datasets can help with staging and biomarker trajectory estimation.
-While recent studies \cite{hon2017towards} tried to transfer knowledge from generic image datasets to Alzheimer's disease, we are not aware of any studies that have tried to transfer knowledge across different types of Alzheimer's variants or other dementias.
-Furthermore, we are not aware of any studies to have done transfer learning in medical imaging for estimating a continuous disease progression without reliance on diagnostic classes, which are often biased.
+While recent studies \cite{hon2017towards} tried to transfer knowledge from generic image datasets to Alzheimer's disease, we are not aware of any studies that have tried to transfer knowledge across different types of Alzheimer's variants, either for diagnosis classification or for estimating continuous biomarker trajectories.
 
@@
 We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal progressions for multiple dementias simultaneously and which inherently performs transfer learning between the modelled dementia phenotypes.
-This is achieved by disentangling \emph{disease-specific} from \emph{disease-agnostic} biomarker relationship.
-We fit the DKT model to three datasets simultaneously: (1) the TADPOLE Challenge dataset containing subjects from the ADNI study with MRI, FDG-PET, DTI, AV45 and AV1451 scans, (2) MRI scans from typical AD subjects in our local centre and (3) MRI scans from patients with Posterior Cortical Atrophy (PCA) from our local centre.
+This is achieved by disentangling \emph{disease-specific} from \emph{disease-agnostic} biomarker relationships.
+We fit the DKT model to two datasets simultaneously: (1) the TADPOLE Challenge dataset containing subjects from the ADNI study with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy (PCA) from our local centre.
 We then used the fitted model to predict non-MRI trajectories for PCA patients, which have been learned from the multimodal TADPOLE dataset.
@@
  \centering
- \includegraphics[width=\textwidth]{figures/disease_knowledge_transfer.png}
+ \includegraphics[width=0.8\textwidth,trim=0 0 0 80,clip]{figures/disease_knowledge_transfer.png}
  \caption{Outline of the proposed framework for joint modelling of multiple diseases.
-We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores, each related to different brain regions.
-Each dysfunctionality score is then assumed to be modelled as the progression of several biomarkers within that same region, but having different modalities.
-One key assumption of the model is that the biomarkers within each unit (bottom row) are disease agnostic (i.e.
-have the same dynamics for every disease), while the dysfunction scores (top row) are disease specific.}
+We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores (Y-axis, top row), each one related to different brain regions.
+Each region-specific dysfunctionality score then further models (X-axis, bottom row) the progression of several modality-specific biomarkers within that same region.
+The biomarker correlations within the bottom units are assumed to be disease agnostic and shared across all diseases modelled.
+Disease knowledge transfer can then be achieved via the disease-agnostic units.}
  \label{fig:diagram}
@@
 
-\subsection{Methods}
+\section{Method}
 
 Fig.
-\ref{fig:diagram} shows the overall diagram of our proposed framework for joint modelling of disease.
-We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores, each related to different brain regions (top row).
-Each dysfunctionality score is then assumed to be modelled as the progression of several biomarkers within that same region, but having different modalities (bottom row).
-Each group of biomarkers in the bottom row will be called a \emph{functional unit}, because the correlations between biomarkers are related though common "function" in a disease agnostic way, because they are related to the same underlying brain region.
+\ref{fig:diagram} shows the overall diagram of our proposed framework for joint modelling of diseases.
+We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores, each one related to different brain regions (top row).
+Each dysfunctionality score is then modelled as the progression of several biomarkers within that same region, but having different modalities (bottom row).
+Each group of biomarkers in the bottom row will be called a \emph{functional unit}, because the correlations between biomarkers are related though common "function" in a disease agnostic way, since they are related to the same underlying brain region.
 Biomarker groupings into functional units are defined a-priori.
-We model the correlations within each unit using a Gaussian-Process based disease progression model \cite{lorenzi2017disease}, for each region-specific functional unit we can model the correlation of biomarkers within that functional unit (Figure 1, bottom), which allows us to reconstruct a region- or unit-specific \emph{dysfunction} progression axis which can be used for staging subjects.
-Finally, we express the disease specific correlations (Figure 1, top) not directly with each biomarkers, but via the dysfunction scores estimated within each functional unit.
+We model the correlations within each unit using a Gaussian-Process (GP) based disease progression model (DPM) \cite{lorenzi2017disease}, which allows us to reconstruct unit-specific \emph{dysfunction} progression manifold (X axis), which can be used for staging subjects.
+Finally, we use the same GP model to express the progression within each disease (Figure 1, top) in terms of the dysfunction scores estimated within each functional unit.
+More precisely, the X-axis dysfunction scores from the functional units become Y-axis measurements in the disease specific models.
 
-The proposed model is parsimonious, easily interpretable by abstracting away biomarker information into the functional units, and can be used for \emph{disease knowledge transfer}, i.e.
-learning correlations in one disease from another.
-Our approach is inspired by transfer learning methods from machine learning, but the model is generative, easing interpretability.
+The model has a simple mathematical formulation.
+We assume a set of given biomarkers measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers in at various visits.
+We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the subject-spefic time shift.
+We futher denote by $\theta_k$ the parameters used to represent the trajectory for biomarker $k \in K$ within its functional unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ is a function that maps each biomarker $k$ to a unique functional unit $l \in \Lambda$ and $\Lambda$ is the set of functional units.
+Moreover, we denote by $\lambda_d^l$ the parameters for the trajectory of the dysfunction score corresponding to functional unit $l \in \Lambda$ in the space of disease $d$.
+These definitions allow us to formulate the likelihood for a single measurement $y_{ijk}$ as follows:
 
-Variable names:
-\begin{itemize}
- \item $\Lambda$ = set of \emph{functional units}, i.e. \{temporal, parietal, occipital, ..\}. Biomarkers belonging to the same functional unit are asusmed to have disease-agnostic correlations.
- \item $y_{ijk}$ = measurement in subject $i$, visit $j$, biomarker $k$
-%  \item $d_{ijl}$ = dysfunctionality score in subject $i$, visit $j$ in \emph{functional unit} $l$, where $l \in \Lambda$
-%  \item $s_{ij}$ = disease progression score in subject $i$, visit $j$
- \item $\theta_k$ = parameters of trajectory for biomarker $k$
- \item $\lambda^l$ = parameters of trajectory for functional unit $l$, where $l \in \Lambda$
- \item $\psi$ : \{1, ..., K\} $ \rightarrow \Lambda$ is a function that maps beach biomarker to its corresponding functional unit
- \item $\beta_i$: time shift parameter for subject $i$ (disease specific)
- \item $\beta_i^{l}$: time shift parameter for subject $i$ used in functional unit $l$, where $l \in \Lambda$
- \item $\Omega$: set ${(i,j,k)}$ of measurements available from every subject $i$, visit $j$ in biomarker $k$
-\end{itemize}
-
-Likelihood for one single measurement in one subject visit: 
 \begin{equation}
- p(y_{ijk}|\theta_k, \lambda^{\psi(k)}, \beta_i) = \sum_{\beta_i^{\psi(k)}} p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) p(\beta_i^{\psi(k)}| \lambda^{\psi(k)}, \beta_i)
+ p(y_{ijk}|\theta_k, \lambda_{d_i}^{\psi(k)}, \beta_i) = \sum_{\beta_i^{\psi(k)}} p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) p(\beta_i^{\psi(k)}| \lambda_{d_i}^{\psi(k)}, \beta_i)
 \end{equation}
 
-where $\beta_i^{\psi(k)}$ is a latent variable denoting the stage of subject $i$ in functional unit $\psi(k)$, where biomarker $k$ was assigned.
-
-Extending the above to multiple, subjects, visits and biomarkers, we get the final model likelihood:
+where $\beta_i^{\psi(k)}$ is a latent variable denoting the dysfunctionality score of subject $i$ in functional unit $\psi(k)$, where biomarker $k$ was assigned.
+Moreover, $d_i \in \mathbb{D}$ represents the index of the disease space where subject $i$ belongs, where $\mathbb{D}$ is the set of all diseases modelled.
+For example, MCI and tAD subjects from ADNI as well as tAD subjects from our local cohort can all be assigned $d_i=1$, while PCA subjects from our local dataset can be assigned $d_i=2$.
+Healthy controls can be assigned to either disease space.
+Extending the above model to multiple subjects, visits and biomarkers, we get the final model likelihood:
 \begin{equation}
- p(y_{.,.,.}|\theta_1, ..., \theta_K, \{\lambda^{\psi(l)} | l \in \Lambda \}, \beta_1, ..., \beta_N) = \\ \prod_{(i,j,k) \in \Omega} \sum_{\beta_i^{\psi(k)}} p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) p(\beta_i^{\psi(k)}| \lambda^{\psi(k)}, \beta_i)
+ p(y_{.,.,.}|\theta, \lambda, \beta) = \\ \prod_{(i,j,k) \in \Omega} \sum_{\beta_i^{\psi(k)}} p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) p(\beta_i^{\psi(k)}| \lambda_{d_i}^{\psi(k)}, \beta_i)
 \end{equation}
-
-Modelling of $p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k)$ and $p(\beta_i^{\psi(k)}| \lambda^{\psi(k)}, \beta_i)$ will be done with a non-parametric model (Gaussian Process, Lorenzi et al., 2017)
-
-Let's assume we have MRI and PET data in tAD (from ADNI), but only MRI data for another disease, e.g.
-PCA.
-We can model the disease-agnostic region-specific relationship between MRI and PET, and also build the disease specific progression models for tAD (using MRI and PET) and PCA (using MRI only).
-Finally, our model allows us to propagate the knowledge of PET dynamics in tAD to PCA subjects, without using any PET scans from PCA patients.
-This can be done by maximising the following likelihood (assuming we already have a model for tAD):
+where $\theta = [\theta_1, ..., \theta_K]$, $\lambda = [\lambda_d^{l}|l \in \Lambda, d \in \mathbb{D}]$ and $\beta =[\beta_1, ..., \beta_N]$ represents the stacked parameters for the trajectories of biomarker, dysfunctionality scores in disease $d$ and subject-specific time shifts.
+Modelling of the likelihood of measurements $y_{ijk}$ within each functional unit is done using the non-parametric GP model by \cite{lorenzi2017disease}:
 
 \begin{equation}
- p(y_{.,.,.}^{PCA}|\theta_1^{tAD}, ..., \theta_K^{tAD}, \{\lambda^{\psi(l)^{PCA}} | l \in \Lambda \}, \beta_1^{PCA}, ..., \beta_N^{PCA}) = 
+p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) = f(m_{ij} + \beta_i^{\psi(k)}) + \epsilon
 \end{equation}
+where $f \approx GP(0, \Sigma_G)$ models the population-level trajectory.
+The trajectory is also constrained to be monotonically increasing by enforcing a prohibit-likelihood on the derivative process $\frac{df(t)}{dt}$ \cite{lorenzi2017disease}.
+The same model is also applied to express the likelihood of $p(\beta_i^{\psi(k)}| \lambda_d^{\psi(k)}, \beta_i)$, which can give an estimate of the trajectories of dysfunctionality scores in the model specific to disease $d$.
+As already explained in \cite{lorenzi2017disease}, model fitting is done by Expectation-Propagation through conjugate gradient descent.
+In our hierarchical design, we first fit the models within each functional units independently, and then fit the disease specific models.
 
- \begin{equation}
- \prod_{(i,j,k) \in \Omega} \sum_{\beta_i^{\psi(k)}} p(y_{ijk}^{PCA}| \beta_i^{\psi(k)}, \theta_k^{tAD}) p(\beta_i^{\psi(k)}| \lambda^{\psi(k)^{PCA}}, \beta_i^{PCA})
-\end{equation}
+\subsection{Data acquisition and Preprocessing}
+
+We used 87 controls and 76 PCA subjects from our local dataset, which only had MRI scans.
+From TADPOLE, out of the subjects that contained at least one FDG PET, AV45, AV1451 or DTI scan, we randomly selected a subset of 230 subjects, in order not to bias the numbers towards the typical AD disease.
+Volumetric measures for each subject have been obtained using the Freeusurfer software in both ADNI and our local dataset.
+For FDG, AV45 and AV1451 PET, we used already extracted SUVR measures from ADNI which were included in the TADPOLE dataset.
+For DTI, we used fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe.
+For every lobe, we averaged the biomarker values for regions of interest within each lobe, regressed out age, gender, TIV and dataset (ADNI vs our local dataset) and then normalised the final biomarker values to lie within the [0,1] range.
 
@@
 Fig.
-\ref{fig:occipUnit} shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data.
+\ref{fig:pcaTadDisSpace}A shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data.
 The X-axis shows the dysfunctionality scores within the occipital unit, which represent estimated time-shifts, in months, from an arbitrary reference X=0, while the Y-axis shows biomarker values normalised to [0,1] range.
@@
 Fig.
-\ref{fig:occipUnit} shows the progression of dysfunctionality scores over the disease stage for (a) typical AD and (b) PCA.
+\ref{fig:pcaTadDisSpace} also shows the progression of dysfunctionality scores over the disease stage for (B) typical AD and (C) PCA.
 In typical AD, we see that temporal dysfunction becomes abnormal earliest, while PCA shows early parietal dysfunction, again in line with previous findings in the literature \cite{crutch2012posterior} (J.C.
@@
 % data fitting within one unit
-\begin{figure}
-\includegraphics[width=\textwidth, trim=90 0 110 0, clip]{figures/unit1_allTraj_tad-drcTinyPen1_JMD.png} 
-\caption{Estimated biomarker trajectories in the Occipital Unit.
+% \begin{figure}
+% \centering
+% \includegraphics[width=0.8\textwidth, trim=90 0 110 0, clip]{figures/unit1_allTraj_tad-drcTinyPen1_JMD.png} 
+% \caption{Estimated biomarker trajectories in the Occipital Unit.
 Subject data from ADNI and our local cohort are also shown.
-The X-axis represents, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject.
+The X-axis, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject.
 Red lines represent samples from the trajectory posterior.
 The Y-axis measures biomarker values (normalised).}
-\label{fig:occipUnit}
-\end{figure}
+% \label{fig:occipUnit}
+% \end{figure}
 
 % PCA vs tAD disease space
-\begin{figure}
+% \vspace{-2em}
+\begin{picture}(5,5)
+\put(0,-70){\textbf{\Large{A}}}
+\end{picture}
+\begin{figure}[H]
+\centering
+\begin{subfigure}{\textwidth}
+\centering
+\includegraphics[width=0.8\textwidth, trim=90 0 110 0, clip]{figures/unit1_allTraj_tad-drcTinyPen1_JMD.png} 
+% \caption{}
+% \label{fig:occipUnit}
+\end{subfigure}
+
+\begin{picture}(5,5)
+\put(0,20){\textbf{\Large{B}}}
+\end{picture}
 \begin{subfigure}{0.47\textwidth}
 \centering
-typical AD\\
-\includegraphics[width=\textwidth, trim=0 0 0 20, clip]{figures/tAD_trajSameSpace_tad-drcTinyPen1_JMD.png} 
-\caption{}
+% typical AD\\
+\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{figures/tAD_trajSameSpace_tad-drcTinyPen1_JMD.png} 
+% \caption{C}
 \end{subfigure}
+\begin{picture}(5,5)
+\put(0,20){\textbf{\Large{C}}}
+\end{picture}
 \begin{subfigure}{0.47\textwidth}
 \centering
-PCA\\
-\includegraphics[width=\textwidth, trim=0 0 0 20, clip]{figures/PCA_trajSameSpace_tad-drcTinyPen1_JMD.png} 
-\caption{}
+% PCA\\
+\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{figures/PCA_trajSameSpace_tad-drcTinyPen1_JMD.png} 
+% \caption{PCA}
 \end{subfigure}
-\caption{Progression of dysfunctionality scores for (a) typical AD and (b) PCA. In typical AD, we notice that the temporal scores are most affected after the disease }
+\caption{(A) Estimated biomarker trajectories in the Occipital Unit.
+Subject data from ADNI and our local cohort are also shown.
+The X-axis, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject.
+Red lines represent samples from the trajectory posterior.
+The Y-axis measures biomarker values (normalised).
+(B-C) Progression of dysfunctionality scores for (B) typical AD and (C) PCA.}
 \label{fig:pcaTadDisSpace}
@@
 % estimated (hypothetical) trajectories in PCA: DTI, FDG, AV45, AV1451.Volumetric trajectories were based on PCA MRI data.
-\begin{figure}[H]
- \includegraphics[width=\textwidth, trim=0 0 0 0, clip]{figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen1_JMD.png}
+\begin{figure}
+ \includegraphics[width=\textwidth, trim=0 20 0 0, clip]{figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen1_JMD.png}
  \caption{Estimated trajectories for the PCA cohort.
-The only data that was available was the MRI volumetric data.
+The only data that were available were the MRI volumetric data.
 The dynamics of the other biomarkers has been inferred by the model using data from typical AD, and taking into account the different spatial distribution of pathology in PCA as compared to typical AD.}
@@
 
-\section*{Validation}
+\section{Validation}
 
@@
 
-% LOCAL DTI validation
+% DTI validation with local dataset
 
-\begin{figure}[H]
+\begin{figure}
  \includegraphics[width=\textwidth, trim=0 0 0 0, clip]{figures/validDtiPCA.png}
@@
 % limitations: each region needs to have disease signal -> validation did not work well in some regions -> not accounted for subject specific effects. 
-The model and the results presented here have several limitations.
-First of all, the model assumes that within each region there is enough disease signal within each region to propely estimate the dysfunctionality scores.
-Another limitation is that, in the validation, the inferred trajectories in posterior regions traditionally associted with PCA do not match well with the validation data.
-One explanation for this might be the lack of disease signal in posterior regions in DTI data during training, as per the previous limitation.
-Finally, in terms of modelling, our framework did not account for subject specific effects, as we were only interested in estimating population-level trajectories.
+The model and the results presented here have several limitations: (1) the model assumes that within each region there is enough disease signal to properly estimate the dysfunctionality scores (2) in the validation, the inferred trajectories in posterior regions traditionally associated with PCA do not match well with the data and (3) in terms of modelling, our framework did not account for subject-specific effects.
 
@@
 There are several potential avenues for further research.
-During training, we can account for missing disease signal in some brain regions by adding other types of dementias that affect other regions, such as various types of Fronto-temporal Dementia or Primary Progressive Aphasia.
-The estimated PCA trajectories also need to be thoroughly validated using data from modalities other than DTI.
-The model formulation can also be naturally extended to include subject-specific effects, like in the work of Lorenzi et al., 2017.
+During training, we can account for missing disease signal in some brain regions by adding other types of dementias that affect other regions.
+The estimated PCA trajectories also need to be thoroughly validated using data from other modalities.
+The model formulation can also be naturally extended to include subject-specific effects.
 
@@
 
-\bibitem{fonteijn2012event}
-Fonteijn, H.M., Modat, M., Clarkson, M.J., Barnes, J., Lehmann, M., Hobbs, N.Z., Scahill, R.I., Tabrizi, S.J., Ourselin, S., Fox, N.C.
-and Alexander, D.C., 2012.
-An event-based model for disease progression and its application in familial Alzheimer's disease and Huntington's disease.
-NeuroImage, 60(3), pp.1880-1889.
+\bibitem{lorenzi2017disease}
+Lorenzi, M., Filippone, M., Alexander, D.C.
+and Ourselin, S., 2017.
+Disease Progression Modeling and Prediction through Random Effect Gaussian Processes and Time Transformation.
+arXiv preprint arXiv:1701.01668.
 
-\bibitem {jedynak2012}
-Jedynak, B.M., Lang, A., Liu, B., Katz, E., Zhang, Y., Wyman, B.T., Raunig, D., Jedynak, C.P., Caffo, B., Prince, J.L.
-and Alzheimer's Disease Neuroimaging Initiative, 2012.
-A computational neurodegenerative disease progression score: method and results with the Alzheimer's Disease Neuroimaging Initiative cohort.
-Neuroimage, 63(3), pp.1478-1486.
+\bibitem{young2014data}
+Young, A.L., Oxtoby, N.P., Daga, P., Cash, D.M., Fox, N.C., Ourselin, S., Schott, J.M.
+and Alexander, D.C., 2014.
+A data-driven model of biomarker changes in sporadic Alzheimer's disease.
+Brain, 137(9), pp.2564-2577.
 
@@
 
-\bibitem{schiratti2015mixed}
-Schiratti, J.B., Allassonniere, S., Routier, A., Colliot, O., Durrleman, S.
-and Alzheimers Disease Neuroimaging Initiative, 2015, June.
-A mixed-effects model with time reparametrization for longitudinal univariate manifold-valued data.
-In International Conference on Information Processing in Medical Imaging (pp.
-564-575).
-Springer International Publishing.
-
-\bibitem{seeley2009neurodegenerative}
-Seeley, W.W., Crawford, R.K., Zhou, J., Miller, B.L. and Greicius, M.D., 2009. Neurodegenerative diseases target large-scale human brain networks. Neuron, 62(1), pp.42-52.
+\bibitem{Ossenkoppele2014atrophy}
+Ossenkoppele, R., Cohn‐Sheehy, B.I., La Joie, R., Vogel, J.W., Möller, C., Lehmann, M., van Berckel, B.N., Seeley, W.W., Pijnenburg, Y.A., Gorno‐Tempini, M.L.
+and Kramer, J.H., 2015.
+Atrophy patterns in early clinical stages across distinct phenotypes of Alzheimer's disease.
+Human brain mapping, 36(11), pp.4421-4437.
 
@@
 
-\bibitem{lorenzi2017disease}
-Lorenzi, M., Filippone, M., Alexander, D.C.
-and Ourselin, S., 2017.
-Disease Progression Modeling and Prediction through Random Effect Gaussian Processes and Time Transformation.
-arXiv preprint arXiv:1701.01668.
-
 \end{thebibliography}

### 2018-03-03 tadpoleDrc.py

@@ -1095,7 +1095,15 @@ def main():
   diag = ds['diag']
 
   # visDataHist(dataDfAll)
+  nrUnqDiags = np.unique(dataDfAll.diag)
+  print(dataDfAll)
+  for d in nrUnqDiags:
+    idxCurrDiag = ds['diag'] == d
+    print('nr subj %s %d' % (plotTrajParams['diagLabels'][d], np.sum(idxCurrDiag)))
+    # avgScans = []
+    # print('avg scans %s %d' % plotTrajParams['diagLabels'][d])
 
+  print(ads)
 
   meanVols = np.array([np.mean(Y[0][s]) for s in range(RID.shape[0])])
   meanVols[diag != CTL2] = np.inf

### 2018-03-03 paper/paper.tex
m
minor

@@
 The availability of multimodal biomarker measurements in large scale clinical data is fundamental for the application of current disease progression models of neurodegenerative disorders.
-Due to the lack of some data and to the low sample size, disease progression modeling of rare dementias is currently very challenging.
-To overcome this problem, in this study we introduce Disease Knowledge Transfer (DKT), a technique for transferring biomarker's progression models between Alzheimer's disease (AD) variants.
+Due to the lack of data and to the low sample size, disease progression modeling of rare dementias is currently very challenging.
+To overcome this problem, we introduce Disease Knowledge Transfer (DKT), a technique for transferring biomarker's progression models between Alzheimer's disease (AD) variants.
 We assume that different types of dementia affect overlapping brain regions, and thus present shared biomarker characteristics that can be transferred across diseases.
@@
 % biomarkers in alzheimer's -> measuring the evolution helps staging in clinical trials
-Several image-based biomarkers for Alzheimer's disease (AD) are currently used to track the progression of the pathology: AV45 Positron Emission Tomography (PET) measuring amyloid-plaque aggregation, AV1451 PET measuring Tau tangle aggregation, Flourodeoxyglucose (FDG) PET measuring brain hypometabolism, Magnetic Resonance Imaging (MRI) measuring structural integrity and Diffusion Tensor Imaging (DTI) measuring connectivity integrity.
-Measuring the exact evolution of these biomarkers over the disease progression is fundamental for better understanding of the pathology, as well as for improving the stratification in clinical trials through patient staging.
+Several image-based biomarkers for Alzheimer's disease (AD) are currently used to track the progression of pathology: AV45 Positron Emission Tomography (PET) measuring amyloid-plaque aggregation, AV1451 PET measuring tau tangle aggregation, Flourodeoxyglucose (FDG) PET measuring brain hypometabolism, Magnetic Resonance Imaging (MRI) measuring structural integrity and Diffusion Tensor Imaging (DTI) measuring connectivity integrity.
+Measuring the exact evolution of these biomarkers over the disease progression is fundamental for better understanding of the pathology, as well as for improving stratification in clinical trials through patient staging.
 
@@
 adding extra information from other datasets can help with staging and biomarker trajectory estimation.
-While recent studies \cite{hon2017towards} tried to transfer knowledge from generic image datasets to Alzheimer's disease, we are not aware of any studies that have tried to transfer knowledge across different types of Alzheimer's variants, either for diagnosis classification or for estimating continuous biomarker trajectories.
-
-% To overcome this problem, in this study we aim at defining a probabilistic framework for the estimation of disease progression trajectories in small datasets of specific disease subtypes, in presence of missing biomarkers and low numerosity.
-To this end we introduce the idea of \emph{transfer learning} of disease progression models across datasets.
-In particular, our framework is formulated in a probabilistic fashion, in order to quantify the uncertainty associated with low sample size and high variability characterizing the problem.
+While recent studies \cite{hon2017towards} performed transfer learning from generic image datasets to Alzheimer's disease, we are not aware of any studies that have tried to transfer knowledge across different types of Alzheimer's variants, either for diagnosis classification or for estimating continuous biomarker trajectories.
 
@@
 We fit the DKT model to two datasets simultaneously: (1) the TADPOLE Challenge dataset containing subjects from the ADNI study with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy (PCA) from our local centre.
-We then used the fitted model to predict non-MRI trajectories for PCA patients, which have been learned from the multimodal TADPOLE dataset.
+We then used the fitted model to predict plausible non-MRI trajectories for PCA patients.
 We finally validated the DTI trajectories in PCA using a small test set of 20 DTI scans from the PCA patients and controls from our local centre.
@@
 We used 87 controls and 76 PCA subjects from our local dataset, which only had MRI scans.
-From TADPOLE, out of the subjects that contained at least one FDG PET, AV45, AV1451 or DTI scan, we randomly selected a subset of 230 subjects, in order not to bias the numbers towards the typical AD disease.
-Volumetric measures for each subject have been obtained using the Freeusurfer software in both ADNI and our local dataset.
-For FDG, AV45 and AV1451 PET, we used already extracted SUVR measures from ADNI which were included in the TADPOLE dataset.
+From TADPOLE/ADNI, out of the subjects that contained at least one FDG PET, AV45, AV1451 or DTI scan, we randomly selected a subset of 230 subjects, in order not to bias the numbers towards the typical AD disease.
+Volumetric measures for each subject have been obtained using the Freesurfer software in both ADNI and our local dataset.
+For FDG, AV45 and AV1451 PET, we used already extracted SUVR measures from ADNI.
 For DTI, we used fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe.
@@
 We do this for five different modalities: MRI Volumes, DTI, FDG, AV45 and AV1451.
-The results again recapitulate known patterns in PCA, where trajectories from posterior regions become abnormal first in MRI, DTI and AV1451, although in FDG we instead see temporal and hippocampal regions more affected.
-This might be due to the normalisation of the FDG trajectories on the Y-axis to the [0,1] range, which results in less scaling to trajectories that contain more disease signal.
+The results again recapitulate known patterns in PCA, where trajectories from posterior regions become abnormal first in MRI, DTI and AV1451, with the exception of FDG.
+This might be due to the normalisation of the FDG trajectories on the Y-axis to the [0,1] range, which results in artificially more scaling to trajectories that contain less disease signal.
+We also notice two non-monotonic trajectories in the AV45 plot, which are due to bad starting points within parameter optimisation combined with low disease signal within these biomarkers.
 
@@
 \end{figure}
-% * <mrazvan22@gmail.com> 2018-03-02T15:58:16.565Z:
-% 
-% two trajectories in AV45 screwed up due to bad initial starting point.
-Will take around 2-3h to fix, by implementing a more informed starting point, and re-fit the model.
-I'll try to see if I have time, otherwise might drop that modality?!
-Or take out the plot entirely.
-% 
-% ^.
 
@@
 % limitations: each region needs to have disease signal -> validation did not work well in some regions -> not accounted for subject specific effects. 
-The model and the results presented here have several limitations: (1) the model assumes that within each region there is enough disease signal to properly estimate the dysfunctionality scores (2) in the validation, the inferred trajectories in posterior regions traditionally associated with PCA do not match well with the data and (3) in terms of modelling, our framework did not account for subject-specific effects.
+Our work has several limitations: (1) the model assumes that within each region there is enough disease signal to properly estimate the dysfunctionality scores (2) in the validation, the inferred trajectories in posterior regions traditionally associated with PCA do not match well with the data and (3) in terms of modelling, our framework did not account for subject-specific effects.
 

### 2018-03-03 MarcoModel.py

@@ -144,6 +144,7 @@ class GP_progression_model(object):
                 obs = np.array([self.X_array[biom][k][0] for k in range(int(np.sum(self.N_obs_per_sub[biom][:sub])),
                                                                np.sum(self.N_obs_per_sub[biom][:sub + 1]))])
 
+        # self.addInitTimeShifts()
 
 
     # def addMinXMaxXExtraRange(self, deltaRangeFactor = 0.0):
@@ -735,6 +736,46 @@ class GP_progression_model(object):
 
       # objective_grad = lambda test_params_time_shift: self.grad_time_shift(test_params_time_shift)
       # fun_value, fun_grad = objective_grad(params_time_shift)
+
+
+    def addInitTimeShifts(self):
+
+        convTimeOnlyToTimePlusAcc = lambda params_time_shift_only_shift: \
+          np.concatenate((params_time_shift_only_shift.reshape(1,-1),
+          np.ones((1, params_time_shift_only_shift.shape[0]))),axis=0)
+
+        initTimeShifts = []
+        for s in range(len(self.Y[0])):
+          initTimeShifts += [[self.Y[b][s][0] for b in range(len(self.Y)) if self.Y[b][s].shape[0] > 0][0] * 50] # 50 months
+
+
+        print('initTimeShifts', initTimeShifts)
+        initTimeShifts = np.array(initTimeShifts)
+        initTimeShifts = initTimeShifts - np.mean(initTimeShifts)
+        initTimeShifts = convTimeOnlyToTimePlusAcc(np.array(initTimeShifts))
+        print('initTimeShifts', initTimeShifts)
+        # print(ads)
+
+        # import pdb
+        # pdb.set_trace()
+
+        for l in range(1):
+            self.params_time_shift[l] = self.params_time_shift[l] + initTimeShifts[l]
+
+        for i in range(self.N_biom):
+            Xdata = np.array([[100]])
+            for sub in range(self.N_samples):
+                temp = self.X_array[i][int(np.sum(self.N_obs_per_sub[i][:sub])):np.sum(self.N_obs_per_sub[i][:sub+1])]
+                shifted_temp = (temp + initTimeShifts[0][sub])
+                Xdata = np.hstack([Xdata,shifted_temp.T])
+
+            self.X_array[i] = Xdata[0][1:].reshape([len(Xdata[0][1:]),1])
+
+
[…]

### 2018-03-03 tadpoleDrc.py

@@ -1103,7 +1103,7 @@ def main():
     # avgScans = []
     # print('avg scans %s %d' % plotTrajParams['diagLabels'][d])
 
-  print(ads)
+  # print(ads)
 
   meanVols = np.array([np.mean(Y[0][s]) for s in range(RID.shape[0])])
   meanVols[diag != CTL2] = np.inf

### 2018-03-03 paper/paper.tex
final version for MICCAI ... my brain is dead

@@
 Due to the lack of data and to the low sample size, disease progression modeling of rare dementias is currently very challenging.
-To overcome this problem, we introduce Disease Knowledge Transfer (DKT), a technique for transferring biomarker's progression models between Alzheimer's disease (AD) variants.
+To overcome this problem, we introduce Disease Knowledge Transfer (DKT), a technique for transferring biomarkers' progression models between Alzheimer's disease (AD) variants.
 We assume that different types of dementia affect overlapping brain regions, and thus present shared biomarker characteristics that can be transferred across diseases.
@@
 
-Several data-driven methods have been proposed in order to reconstruct group-wise long term biomarker progressions from collections of short term individual biomarker measurements (\cite{lorenzi2017disease,young2014data,donohue2014estimating}).
-These approaches mostly rely on the estimation of a latent time reparameterization associated with each individual, often encoded by a translation (time shift) of the subject's measurements over the temporal axis.
-This time reparameterization can be used as a proxy of disease staging associated with an individual biomarker profile relatively to the global disease progression model.
+Several data-driven methods have been proposed in order to reconstruct group-wise long term biomarker progressions from collections of short term individual biomarker measurements \cite{lorenzi2017disease,oxtoby2018}.
+These approaches mostly rely on the estimation of a latent time re-parameterization associated with each individual, often encoded by a translation (time shift) of the subject's measurements over the temporal axis.
+This time re-parameterization can be used as a proxy of disease staging associated with an individual biomarker profile relatively to the global disease progression model.
 
 % Flow: multimodal data needed for disease progression -> disease progression on rare dementias is challenging -> cannot apply model learned from other disease because of different spatial patterns -> but studies show at least some partial spatial overlap -> therefore transfer learning possible due to presence of "some overlap" -> benefits of transfer learning -> existing transfer learning literature
-The availability of multimodal collections of biomarker's measurements across clinical groups is fundamental to the application of such disease progression models.
+The availability of multimodal collections of biomarkers' measurements across clinical groups is fundamental to the application of such disease progression models.
 For this reason, the application of disease progression modeling to rare dementia type is currently very challenging, due to the associated problem of missing biomarkers and low sample size.
@@
 Biomarker groupings into functional units are defined a-priori.
-We model the correlations within each unit using a Gaussian-Process (GP) based disease progression model (DPM) \cite{lorenzi2017disease}, which allows us to reconstruct unit-specific \emph{dysfunction} progression manifold (X axis), which can be used for staging subjects.
+We model the correlations within each unit using a Gaussian-Process (GP) based disease progression model (DPM) \cite{lorenzi2017disease}, which allows us to reconstruct unit-specific \emph{dysfunction} progression manifolds (bottom row, X axis), which can be used for staging subjects.
 Finally, we use the same GP model to express the progression within each disease (Figure 1, top) in terms of the dysfunction scores estimated within each functional unit.
@@
 The model has a simple mathematical formulation.
-We assume a set of given biomarkers measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers in at various visits.
-We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the subject-spefic time shift.
-We futher denote by $\theta_k$ the parameters used to represent the trajectory for biomarker $k \in K$ within its functional unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ is a function that maps each biomarker $k$ to a unique functional unit $l \in \Lambda$ and $\Lambda$ is the set of functional units.
+We assume a set of given biomarkers measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers at various visits.
+We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the time shift of subject $i$.
+We further denote by $\theta_k$ the parameters used to represent the trajectory for biomarker $k \in K$ within its functional unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ is a function that maps each biomarker $k$ to a unique functional unit $l \in \Lambda$, where $\Lambda$ is the set of functional units.
 Moreover, we denote by $\lambda_d^l$ the parameters for the trajectory of the dysfunction score corresponding to functional unit $l \in \Lambda$ in the space of disease $d$.
@@
 For DTI, we used fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe.
-For every lobe, we averaged the biomarker values for regions of interest within each lobe, regressed out age, gender, TIV and dataset (ADNI vs our local dataset) and then normalised the final biomarker values to lie within the [0,1] range.
+For every lobe, we averaged the biomarker values for regions of interest within each lobe, regressed out age, gender, total intracranial volume (TIV) and dataset (ADNI vs our local dataset) and then normalised the final biomarker values to lie within the [0,1] range.
 
@@
 \ref{fig:pcaTadDisSpace} also shows the progression of dysfunctionality scores over the disease stage for (B) typical AD and (C) PCA.
-In typical AD, we see that temporal dysfunction becomes abnormal earliest, while PCA shows early parietal dysfunction, again in line with previous findings in the literature \cite{crutch2012posterior} (J.C.
-Baron, Neuroimage, 2001).
+In typical AD, we see that temporal dysfunction becomes abnormal earliest, while PCA shows early parietal dysfunction, again in line with previous findings in the literature \cite{crutch2012posterior,Baron2001}.
 In Fig.
@@
 The results again recapitulate known patterns in PCA, where trajectories from posterior regions become abnormal first in MRI, DTI and AV1451, with the exception of FDG.
-This might be due to the normalisation of the FDG trajectories on the Y-axis to the [0,1] range, which results in artificially more scaling to trajectories that contain less disease signal.
+This might be due to the normalisation of the FDG trajectories on the Y-axis to the [0,1] range, which results in artificially more scaling to trajectories with less disease signal.
 We also notice two non-monotonic trajectories in the AV45 plot, which are due to bad starting points within parameter optimisation combined with low disease signal within these biomarkers.
@@
 \newcommand{\expFld}{figures}
-
-% data fitting within one unit
-% \begin{figure}
-% \centering
-% \includegraphics[width=0.8\textwidth, trim=90 0 110 0, clip]{figures/unit1_allTraj_tad-drcTinyPen1_JMD.png} 
-% \caption{Estimated biomarker trajectories in the Occipital Unit.
-Subject data from ADNI and our local cohort are also shown.
-The X-axis, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject.
-Red lines represent samples from the trajectory posterior.
-The Y-axis measures biomarker values (normalised).}
-% \label{fig:occipUnit}
-% \end{figure}
 
@@
 \end{subfigure}
-\caption{(A) Estimated biomarker trajectories in the Occipital Unit.
+\caption{(A) Estimated biomarker trajectories in the occipital functional unit.
 Subject data from ADNI and our local cohort are also shown.
@@
 The inferred trajectories in PCA show good agreement with the data in the cingulate, hippocampus and temporal lobes, but less agreement in the other regions.
-This might be because the dataset that was used to train the model had better disease signal in DTI regions more traditionally associated with typical AD.
+This might be because the tAD DTI dataset that was used to train the model had better disease signal in regions traditionally associated with typical AD.
 
@@
 The estimated PCA trajectories also need to be thoroughly validated using data from other modalities.
-The model formulation can also be naturally extended to include subject-specific effects.
+The model formulation also extends naturally to include subject-specific effects.
 
@@
 
-\bibitem{young2014data}
-Young, A.L., Oxtoby, N.P., Daga, P., Cash, D.M., Fox, N.C., Ourselin, S., Schott, J.M.
-and Alexander, D.C., 2014.
-A data-driven model of biomarker changes in sporadic Alzheimer's disease.
-Brain, 137(9), pp.2564-2577.
-
-\bibitem{donohue2014estimating}
-Donohue, M.C., Jacqmin-Gadda, H., Le Goff, M., Thomas, R.G., Raman, R., Gamst, A.C., Beckett, L.A., Jack, C.R., Weiner, M.W., Dartigues, J.F.
-and Aisen, P.S., 2014.
-Estimating long-term multivariate progression from short-term data.
-Alzheimer's \& Dementia, 10(5), pp.S400-S410.
+\bibitem{oxtoby2018}
+Oxtoby, N.P., Young, A.L., Cash, D.M., Benzinger, T.L., Fagan, A.M., Morris, J.C., Bateman, R.J., Fox, N.C., Schott, J.M.
+and Alexander, D.C., 2018.
+Data-driven models of dominantly-inherited Alzheimer's disease progression.
+bioRxiv, p.250654.
 
@@
 
+\bibitem{Baron2001}
+Baron, J.C., Ch\'{e}telat, G., Desgranges, B., Perchey, G., Landeau, B., {de la Sayette}, V. and Eustache, F., 2001. NeuroImage 14(2), pp.298-309.
+
 \end{thebibliography}

### 2018-10-09 paper/paper.tex
updated figures, added quantification in paper

@@
 
+\newcommand{\expFld}{../resfiles/tad-drcTiny_JMD}
+\newcommand{\jmdFld}{..}
+
 \begin{abstract}
@@
  \centering
- \includegraphics[width=0.8\textwidth,trim=0 0 0 80,clip]{figures/disease_knowledge_transfer.png}
+ \includegraphics[width=0.8\textwidth,trim=0 0 0 0,clip]{figures/disease_knowledge_transfer.pdf}
  \caption{Outline of the proposed framework for joint modelling of multiple diseases.
@@
 
-\newcommand{\expFld}{figures}
-
 % PCA vs tAD disease space
@@
 \centering
-\includegraphics[width=0.8\textwidth, trim=90 0 110 0, clip]{figures/unit1_allTraj_tad-drcTinyPen1_JMD.png} 
+\includegraphics[width=0.8\textwidth, trim=90 0 110 0, clip]{\expFld/unit1_allTraj.png} 
 % \caption{}
@@
 % typical AD\\
-\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{figures/tAD_trajSameSpace_tad-drcTinyPen1_JMD.png} 
+\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis0_tAD_allTrajZeroOne.png} 
 % \caption{C}
@@
 % PCA\\
-\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{figures/PCA_trajSameSpace_tad-drcTinyPen1_JMD.png} 
+\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis1_PCA_allTrajZeroOne.png} 
 % \caption{PCA}
@@
 \begin{figure}
- \includegraphics[width=\textwidth, trim=0 20 0 0, clip]{figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen1_JMD.png}
+ \includegraphics[width=\textwidth, trim=0 20 0 0, clip]{\expFld/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
  \caption{Estimated trajectories for the PCA cohort.
@@
 
-\begin{figure}
- \includegraphics[width=\textwidth, trim=0 0 0 0, clip]{figures/validDtiPCA.png}
- \caption{Validation of the Disease Knowledge Transfer Model using DTI data of PCA subjects from our local cohort.
-Each subject has been shifted along the X-axis only according to their MRI data.
-No DTI data from PCA subjects was used to fit the model.}
-\label{fig:DTIvalid}
-\end{figure}
+\begin{table}
+\begin{tabular}{c | c c c c c c}
+\textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippocampus} & \textbf{Occipital} & \textbf{Parietal} & \textbf{Temporal}\\
+& \multicolumn{6}{c}{\textbf{Prediction Error (MSE)}}\\
+DKT & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.18$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
+Latent stage model & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.17$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
+Linear Model & 0.05$\pm$0.02* & 0.15$\pm$0.04* & 0.09$\pm$0.03* & 0.07$\pm$0.03* & 0.07$\pm$0.02* & 0.07$\pm$0.02*\\
+& \multicolumn{6}{c}{\textbf{Rank Correlation (Spearman rho)}}\\
+DKT  & 0.76  & 0.48  & 0.76  & 0.55  & 0.55  & 0.33 \\
+Latent stage model  & 0.76  & 0.49  & 0.80*  & 0.56  & 0.51*  & 0.33 \\
+Linear Model  & 0.48*  & 0.31*  & 0.64*  & 0.61*  & 0.57*  & 0.27* \\
+\end{tabular}
+\caption[Performance evaluation of DKT and two simpler models]{Performance evaluation of DKT and two simpler models: a latent stage disease progression model and a linear univariate model of DTI predicted from MRI.
+The top table shows the prediction error (MSE) while the bottom table shows the rank correlation measured by Spearman's rho.
+(*) Statistically significant difference between the current model performance and DKT, based on a two-tailed t-test, Bonferroni corrected.}
+\label{sec:dktPerfMetrics}
+\end{table}
 

### 2018-10-09 JointModel.py

@@ -152,7 +152,13 @@ class JointModel(DisProgBuilder.DPMInterface):
       # self.loadCheckpoint(i, 2)
       # self.makePlots(plotFigs, i, 2)
 
-      self.loadCheckpoint(0, 2) # for the real data
+      i = 0
+
+      self.loadCheckpoint(i, 2) # for the real data
+      # self.makePlots(plotFigs, i, 2)
+      # print(adsa)
+
+
 
     res = None
     return res
@@ -183,6 +189,7 @@ class JointModel(DisProgBuilder.DPMInterface):
       if self.params['plotTrajParams']['isSynth']:
         fig = self.plotter.plotCompWithTrueParams(self.unitModels, self.disModels, replaceFig=True)
         fig.savefig('%s/compTrueParams%d%d_%s.pdf' % (self.outFolder, iterNr, picNr, self.expName))
+        fig.savefig('%s/compTrueParams%d%d_%s.png' % (self.outFolder, iterNr, picNr, self.expName))
         pl.clf()
         pl.cla()
         pl.close()
@@ -197,10 +204,10 @@ class JointModel(DisProgBuilder.DPMInterface):
         fig = self.plotter.plotAllBiomkDisSpace(self, self.params, d)
         fig.savefig('%s/trajDisSpace%s_%d%d_%s.pdf' % (self.outFolder, self.params['disLabels'][d],
           iterNr, picNr, self.expName))
+        fig.savefig('%s/trajDisSpace%s_%d%d_%s.png' % (self.outFolder, self.params['disLabels'][d],
+                                                       iterNr, picNr, self.expName))
 
-        pl.clf()
-        pl.cla()
-        pl.close()
+
 
   def initParams(self, runPartOnePass):
     paramsCopy = copy.deepcopy(self.params)

### 2018-10-09 Plotter.py

@@ -7,6 +7,8 @@ import copy
 import auxFunc
 from abc import ABC, ABCMeta, abstractmethod
 import DPMModelGeneric
+import sklearn.metrics
+import scipy.interpolate
 
 class PlotterJDM:
 
@@ -565,7 +567,6 @@ class PlotterJDM:
         replaceFig)
 
 
-
 class PlotterGP(ABC):
 
   def __init__(self, plotTrajParams):
@@ -643,8 +644,11 @@ class PlotterGP(ABC):
 
       ax.plot(newX,predBiomksYscaledXB[:,b], lw=2, color='black', label='estimated trajectory')
 
-      # ax.plot(np.array([np.min(newX), np.max(newX)]), [min_yB[b], max_yB[b]],
-      #   color=(0.5,0.5,0.5), lw=2)
+      Y_arrayPredXB = gpModel.predictBiomk(gpModel.X_array[b].reshape(-1))
+      # rSq = sklearn.metrics.r2_score(Y_arrayPredXB[:,b], gpModel.Y_array[b])
+      maeTraj = np.mean(np.abs(Y_arrayPredXB[:,b] - gpModel.Y_array[b]))
+      # ax.text(0.05, 0.9, '$R^2$ = %.2f' % rSq)
+      ax.text(0.05, 1, 'MAE = %.2f' % maeTraj)
 
       ax.set_ylim([min_yB[b]-deltaB[b], max_yB[b]+deltaB[b]])
 
@@ -688,7 +692,6 @@ class PlotterGP(ABC):
     else:
       pl.show()
     pl.pause(0.05)
-
 
     return fig
 
@@ -878,7 +881,10 @@ class PlotterGP(ABC):
     # pl.tight_layout(pad=0.4, w_pad=0.5, h_pad=1.0)
 
 
-    fig.text(0.5, 0.04, 'Disease Progression (months)', ha='center')
+    # xLabel = 'Disease Progression (β, normalised)'
+    xLabel = 'Disease Progression (months)'
+
+    fig.text(0.5, 0.04, xLabel, ha='center')
     fig.text(0.08, 0.5, 'Biomarker Value (normalised)', va='center', rotation='vertical')
 
     h, labels = ax.get_legend_handles_labels()
@@ -972,13 +978,27 @@ class PlotterGP(ABC):
 
[…]

### 2018-10-09 SigmoidModel.py

@@ -85,6 +85,9 @@ class SigmoidModel(DPMModelGeneric.DPMModelGeneric):
   def applyScalingXForward(self, x_data, biomk=0):
     return x_data
 
+  def applyScalingXInv(self, x_data, biomk=0):
+    return x_data
+
 
   def sigFunc(self, xs, theta):
     # print('theta', theta)

### 2018-10-09 drcValidFuncs.py

@@ -603,13 +603,13 @@ def validateDRCBiomk(dpmObj, params):
   plotFigs = True
   if plotFigs:
 
-    # for u in range(dpmObj.nrFuncUnits):
-    #   trajStructUnitModel = dpmObj.unitModels[u].plotter.getTrajStructWithTrueParams(dpmObj.unitModels[u])
-    #   fig = dpmObj.unitModels[u].plotter.plotTraj(dpmObj.unitModels[u], trajStructUnitModel,
-    #     legendExtraPlot=True, rowsAuto=True)
-    #   fig.savefig('%s/unit%d_allTraj.png' % (params['outFolder'], u))
-    #
-    #
+    for u in range(dpmObj.nrFuncUnits):
+      trajStructUnitModel = dpmObj.unitModels[u].plotter.getTrajStructWithTrueParams(dpmObj.unitModels[u])
+      fig = dpmObj.unitModels[u].plotter.plotTraj(dpmObj.unitModels[u], trajStructUnitModel,
+        legendExtraPlot=True, rowsAuto=True)
+      fig.savefig('%s/unit%d_allTraj.png' % (params['outFolder'], u))
+
+
     # for d in range(dpmObj.nrDis):
     #   # yNormMode = dpmObj.params['plotTrajParams']['yNormMode']
     #   yNormMode = 'unscaled'

### 2018-10-09 jointSynth.py

@@ -430,6 +430,7 @@ def transferProgression(dpmObjStdDisOne, paramsDisTwo,
 
 
 
+
 if __name__ == '__main__':
   main()
 

### 2018-11-02 paper/paper.tex
fixed things around DKT, adressing danny and marco feedbakc

@@
 \usepackage{epstopdf}
-\usepackage{caption}
+%\usepackage{caption}
 \usepackage{subcaption}
@@
 
+\usepackage[linesnumbered]{algorithm2e}
+\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{blue}{#1}}
+\SetCommentSty{mycommfont}
+
 \DeclareMathOperator*{\argmin}{arg\,min}
@@
 
-\title{Disease Knowledge Transfer across Alzheimer's Variants}
+\title{Disease Knowledge Transfer across Neurodegenerative Diseases}
 %
@@
 \begin{abstract}
-The availability of multimodal biomarker measurements in large scale clinical data is fundamental for the application of current disease progression models of neurodegenerative disorders.
-Due to the lack of data and to the low sample size, disease progression modeling of rare dementias is currently very challenging.
-To overcome this problem, we introduce Disease Knowledge Transfer (DKT), a technique for transferring biomarkers' progression models between Alzheimer's disease (AD) variants.
-We assume that different types of dementia affect overlapping brain regions, and thus present shared biomarker characteristics that can be transferred across diseases.
-We then implement this paradigm as a joint-disease generative model of biomarker progressions, disentangling disease-specific from  disease-agnostic biomarker relationships.
-We demonstrate DKT on independent datasets of 1) multimodal typical AD data (tAD), with large sample size and several biomarkers available over time, and 2) Posterior Cortical Atrophy (PCA) data, for which only MRI scans are available.
-DKT is able to predict, in tested PCA subjects, plausible population-level biomarkers for structural and molecular imaging biomarkers, for which no data were available.
-DKT may be a useful tool to analyse and understand rare forms of dementia for which multimodal data is not available or is limited.
-Moreover, by leveraging data from multiple diseases, DKT also has the potential to provide more accurate disease staging compared to traditional disease progression models.
+We introduce Disease Knowledge Transfer (DKT), a novel technique for transferring biomarker information between related neurodegenerative diseases.
+DKT infers robust multimodal biomarker trajectories in rare neurodegenerative diseases even when only limited, unimodal data is available, by transferring information from larger multimodal datasets from common neurodegenerative diseases.
+DKT is a joint-disease generative model of biomarker progressions, which exploits biomarker relationships that are shared across diseases.
+As opposed to current deep learning approaches, DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects instead of solving simpler control vs diseased classification tasks.
+Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects.
+For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD dataset (tAD) from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own centre, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
+We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature.
+We further validate DKT in two situations: (1) on synthetic data, showing that it can accurately estimate the ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable performance compared to standard approaches.
+While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
+Our code is available online: \url{https://www.dropbox.com/s/yotur0ohvgwc8om/dkt.zip?dl=0} (anonymised link).
 
@@
 
-% biomarkers in alzheimer's -> measuring the evolution helps staging in clinical trials
-Several image-based biomarkers for Alzheimer's disease (AD) are currently used to track the progression of pathology: AV45 Positron Emission Tomography (PET) measuring amyloid-plaque aggregation, AV1451 PET measuring tau tangle aggregation, Flourodeoxyglucose (FDG) PET measuring brain hypometabolism, Magnetic Resonance Imaging (MRI) measuring structural integrity and Diffusion Tensor Imaging (DTI) measuring connectivity integrity.
-Measuring the exact evolution of these biomarkers over the disease progression is fundamental for better understanding of the pathology, as well as for improving stratification in clinical trials through patient staging.
+The estimation of accurate biomarker signatures in Alzheimer's disease (AD) and related neurodegenerative diseases is crucial for understanding underlying disease mechanisms, predicting subjects' progressions, and enrichment in clinical trials.
+Recently, several data-driven disease progression models were proposed to reconstruct long term biomarker signatures from collections of short term individual measurements \cite{lorenzi2017disease,oxtoby2018}.
+When applied to large datasets of typical AD, disease progression models have shown important benefits in understanding the earliest events in the AD cascade \cite{iturria2016early}, quantifying biomarkers' heterogeneity \cite{young2018uncovering}, helped discover novel genes involved in AD \cite{scelsi2018genetic} and they showed improved predictions over standard approaches \cite{oxtoby2018}.
+However, by necessity these models require large datasets -- in addition they must be both multimodal and longitudinal.
+Such data is not available in rare neurodegenerative diseases.
+In particular, most datasets for rare neurodegenerative diseases come from local clinical centres, are unimodal (e.g.
+MRI only) and limited both cross-sectionally and longitudinally -- this makes the application of disease progression models extremely difficult.
+Moreover, such a model estimated from common diseases such as typical AD may not generalise to specific variants.
+For example, in Posterior Cortical Atrophy (PCA) -- a neurodegenerative syndrome causing visual disruption -- posterior regions such as the occipital lobe and superior parietal regions are affected early, instead of the hippocampus and temporal regions that are affected early in typical AD.
 
-% hypothetical model -> dimention 1: ordering across modalities -> dimention 2: spatial ordering
-A hypothetical model of disease progression has been previously published by \cite{jack2010hypothetical}, which proposes that the first biomarkers to become abnormal are measures of amyloid beta aggregation, followed by tau abnormalities, hypometabolism, structural MRI-based measures and finally cognitive decline.
-While this hypothetical model proposes a inter-modality ordering of biomarkers in typical AD, it has also been observed that within the same modality, biomarker measurements have a spatial sequence of abnormality that correlates with Braak stages: hippocampal volumes and entorhinal measures become abnormal first, followed by other structures within the temporal, parietal and frontal lobes.
+The problem of limited data in medical imaging has so far been addressed through transfer learning methods.
+These techniques successfully used transfer learning techniques to improve the accuracy of AD diagnosis \cite{hon2017towards,cheng2017multi} or prediction of MCI conversion \cite{cheng2015domain}, but have two key limitations.
+First, they use deep learning or other machine learning methods, which are not interpretable and don't allow us to understand underlying disease mechanisms that are either specific to rare diseases, or shared across related diseases.
+Secondly, these models cannot be used to forecast the future evolution of subjects at risk of disease, which is important for selecting the right subjects in clinical trials.
 
-Several data-driven methods have been proposed in order to reconstruct group-wise long term biomarker progressions from collections of short term individual biomarker measurements \cite{lorenzi2017disease,oxtoby2018}.
-These approaches mostly rely on the estimation of a latent time re-parameterization associated with each individual, often encoded by a translation (time shift) of the subject's measurements over the temporal axis.
-This time re-parameterization can be used as a proxy of disease staging associated with an individual biomarker profile relatively to the global disease progression model.
-
-% Flow: multimodal data needed for disease progression -> disease progression on rare dementias is challenging -> cannot apply model learned from other disease because of different spatial patterns -> but studies show at least some partial spatial overlap -> therefore transfer learning possible due to presence of "some overlap" -> benefits of transfer learning -> existing transfer learning literature
-The availability of multimodal collections of biomarkers' measurements across clinical groups is fundamental to the application of such disease progression models.
-For this reason, the application of disease progression modeling to rare dementia type is currently very challenging, due to the associated problem of missing biomarkers and low sample size.
-Moreover, an average model of disease progression estimated from sporadic AD cases may not generalize to specific disease variants.
-For example, in Posterior Cortical Atrophy (PCA), posterior regions such as the occipital lobe and superior parietal regions have been shown to be affected early.
-While the spatial patterns are distinct between various forms of dementia, recent studies \cite{Ossenkoppele2014atrophy} have showed at least partial spatial-overlap between different clinical phenotypes, while others have further suggested that PCA and typical AD subjects lie on a continuum of phenotypical variation \cite{crutch2012posterior}.
-The presence of overlap in pathology patterns across different dementias suggests that it should be theoretically possible to perform transfer learning across the diseases.
-There are two key benefits to performing transfer learning across dementia variants: 1.
-biomarker evolutions can be estimated for rare dementias for which there is not enough data (unimodal datasets, few subjects, cross-sectional data only) and 2.
-adding extra information from other datasets can help with staging and biomarker trajectory estimation.
-While recent studies \cite{hon2017towards} performed transfer learning from generic image datasets to Alzheimer's disease, we are not aware of any studies that have tried to transfer knowledge across different types of Alzheimer's variants, either for diagnosis classification or for estimating continuous biomarker trajectories.
-
-We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal progressions for multiple dementias simultaneously and which inherently performs transfer learning between the modelled dementia phenotypes.
-This is achieved by disentangling \emph{disease-specific} from \emph{disease-agnostic} biomarker relationships.
-We fit the DKT model to two datasets simultaneously: (1) the TADPOLE Challenge dataset containing subjects from the ADNI study with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy (PCA) from our local centre.
-We then used the fitted model to predict plausible non-MRI trajectories for PCA patients.
-We finally validated the DTI trajectories in PCA using a small test set of 20 DTI scans from the PCA patients and controls from our local centre.
-
-% * <alexandra.young@ucl.ac.uk> 2018-03-02T17:54:07.562Z:
-% 
-% > unsupervised
-% Is it unsupervised or does it know which disease/dataset is which, in which case it's semi-supervised?
-% 
-% ^ <mrazvan22@gmail.com> 2018-03-02T18:05:17.625Z:
-% 
-% Model is indeed supervised with respect to dataset/disease, but unsupervised with respect to diagnosis within same disease (e.g.
-Control, MCI and AD in ADNI).
-I removed the supervised-ness though in order not to create confusion.
-%
-% ^.
+We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal biomarker progressions for multiple neurodegenerative diseases simultaneously -- including rare neurodegenerative diseases -- and which inherently performs transfer learning between the modelled phenotypes.
+This is achieved by exploiting biomarker relationships that are shared across diseases, whilst accounting for differences in the spatial distribution of brain pathology.
+DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects at risk of diseases.
+We apply DKT on Alzheimer's variants and demonstrate its ability to predict non-MRI trajectories for patients with Posterior Cortical Atrophy, in lack of such data.
+This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from the Dementia Research Centre (DRC), UK.
+We first show that the estimated non-MRI trajectories for PCA subjects are plausible as they agree with previous literature findings.
+We further validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients, showing it yields favourable performance compared to standard approaches.
+Code for DKT is available online: \url{https://www.dropbox.com/s/yotur0ohvgwc8om/dkt.zip?dl=0} (anonymised link).
 
@@
  \includegraphics[width=0.8\textwidth,trim=0 0 0 0,clip]{figures/disease_knowledge_transfer.pdf}
- \caption{Outline of the proposed framework for joint modelling of multiple diseases.
+ \caption{Diagram of the proposed DKT framework.
 We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores (Y-axis, top row), each one related to different brain regions.
 Each region-specific dysfunctionality score then further models (X-axis, bottom row) the progression of several modality-specific biomarkers within that same region.
+For instance, the temporal dysfunction, modelled as a biomarker in the disease specific model (top row), is the X-axis in the disease agnostic model (temporal unit, bottom row), which aggregates together abnormality from amyloid, tau and MR imaging within the temporal lobe.
 The biomarker correlations within the bottom units are assumed to be disease agnostic and shared across all diseases modelled.
@@
 
+\newcommand{\lp}{\lambda_{d_i}^{\psi(k)}}
+\newcommand{\lpuu}{\lambda_{d_i}^{\psi(k),(u)}}
+\newcommand{\lpum}{\lambda_{d_i}^{\psi(k),(u-1)}}
+
 Fig.
 \ref{fig:diagram} shows the overall diagram of our proposed framework for joint modelling of diseases.
-We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores, each one related to different brain regions (top row).
-Each dysfunctionality score is then modelled as the progression of several biomarkers within that same region, but having different modalities (bottom row).
-Each group of biomarkers in the bottom row will be called a \emph{functional unit}, because the correlations between biomarkers are related though common "function" in a disease agnostic way, since they are related to the same underlying brain region.
-Biomarker groupings into functional units are defined a-priori.
-We model the correlations within each unit using a Gaussian-Process (GP) based disease progression model (DPM) \cite{lorenzi2017disease}, which allows us to reconstruct unit-specific \emph{dysfunction} progression manifolds (bottom row, X axis), which can be used for staging subjects.
-Finally, we use the same GP model to express the progression within each disease (Figure 1, top) in terms of the dysfunction scores estimated within each functional unit.
-More precisely, the X-axis dysfunction scores from the functional units become Y-axis measurements in the disease specific models.
+We assume that the progression of each disease (X-axis, top row) can be modelled as a unique evolution of abstract dysfunctionality scores, each one related to different brain regions (top row).
+Each dysfunctionality score is an abstract measure of pathology within that brain region, derived from several multimodal biomarkers (e.g.
+amyloid, tau, MRI).
 
-The model has a simple mathematical formulation.
+Each dysfunctionality score is modelled as the progression of several biomarkers within that same region, but acquired using different types of modalities (Fig.
+\ref{fig:diagram} bottom row).
+Each group of biomarkers in the bottom row will be called an \emph{agnostic unit}, because the biomarker dynamics are shared across all diseases modelled (i.e.
+disease agnostic).
+
+The assumption that the dynamics of some biomarkers are agnostic, or shared across diseases, is key to DKT.
+There are several reasons why we can make this assumption: first of all, neurodegenerative diseases are known
+
+We choose to model the correlations within each unit using the disease progression model (DPM) by Jedynak et al.
+2012, but any other DPM can also be used.
+The DPM allows us to reconstruct unit-specific dysfunction progression manifolds (bottom row, X axis), which can be used for staging subjects.
+Finally, we use the same model to express the progression within each disease (Figure 1, top) in terms of the dysfunction scores estimated within each agnostic unit.
+More precisely, the X-axis dysfunction scores from the agnostic units become Y-axis measurements in the disease specific models.
+
+\subsection{Mathematical formulation}
+
+The DKT framework has a generic mathematical formulation which uses several disease progression models as building blocks.
+We will first describe the generic framework, and in the following section we will present our chosen implementation of the building blocks.
 We assume a set of given biomarkers measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers at various visits.
 We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the time shift of subject $i$.
-We further denote by $\theta_k$ the parameters used to represent the trajectory for biomarker $k \in K$ within its functional unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ is a function that maps each biomarker $k$ to a unique functional unit $l \in \Lambda$, where $\Lambda$ is the set of functional units.
-Moreover, we denote by $\lambda_d^l$ the parameters for the trajectory of the dysfunction score corresponding to functional unit $l \in \Lambda$ in the space of disease $d$.
+We further denote by $\theta_k$ the parameters used to represent the trajectory for biomarker $k \in K$ within its agnostic unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ maps each biomarker $k$ to a unique agnostic unit $l \in \Lambda$, where $\Lambda$ is the set of agnostic units.
+Moreover, we denote by $\lambda_d^l$ the parameters for the trajectory of the dysfunction score corresponding to agnostic unit $l \in \Lambda$ in the space of disease $d$.
 These definitions allow us to formulate the likelihood for a single measurement $y_{ijk}$ as follows:
@@
 \begin{equation}
- p(y_{ijk}|\theta_k, \lambda_{d_i}^{\psi(k)}, \beta_i) = \sum_{\beta_i^{\psi(k)}} p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) p(\beta_i^{\psi(k)}| \lambda_{d_i}^{\psi(k)}, \beta_i)
+ p(y_{ijk}|\theta_k, \lp, \beta_i, \epsilon_k) = N(y_{ijk}| g(f(\beta_i + m_{ij}; \lp; \theta_k), \epsilon_k)
+\end{equation}
+where $g(\ .\ ; \theta_k)$ represents the trajectory of biomarker $k$ within agnostic unit $\psi(k)$ and $f(\ .\ ; \lambda_{d_i}^{\psi(k)})$ represents the trajectory of the agnostic unit $\psi(k)$ within the space of disease $d_i$.
+To be precise, $d_i \in \mathbb{D}$ represents the index of the disease space where subject $i$ belongs, where $\mathbb{D}$ is the set of all diseases modelled.
+For example, MCI and tAD subjects from ADNI as well as tAD subjects from the DRC cohort can all be assigned $d_i=1$, while PCA subjects from our own centre can be assigned $d_i=2$.
+Healthy controls can be assigned to either disease space, although a more precise assignment would take molecular biomarkers into account.
+Variable $\epsilon_k$ denotes the variance of measurements for biomarker $k$.
+
+We extend the above model to multiple subjects, visits and biomarkers to get the full model likelihood:
+\begin{equation}
+ p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \\ \prod_{(i,j,k) \in \Omega} p(y_{ijk}|\theta_k, \lp, \beta_i) 
 \end{equation}
 
-where $\beta_i^{\psi(k)}$ is a latent variable denoting the dysfunctionality score of subject $i$ in functional unit $\psi(k)$, where biomarker $k$ was assigned.
-Moreover, $d_i \in \mathbb{D}$ represents the index of the disease space where subject $i$ belongs, where $\mathbb{D}$ is the set of all diseases modelled.
-For example, MCI and tAD subjects from ADNI as well as tAD subjects from our local cohort can all be assigned $d_i=1$, while PCA subjects from our local dataset can be assigned $d_i=2$.
-Healthy controls can be assigned to either disease space.
-Extending the above model to multiple subjects, visits and biomarkers, we get the final model likelihood:
-\begin{equation}
- p(y_{.,.,.}|\theta, \lambda, \beta) = \\ \prod_{(i,j,k) \in \Omega} \sum_{\beta_i^{\psi(k)}} p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) p(\beta_i^{\psi(k)}| \lambda_{d_i}^{\psi(k)}, \beta_i)
-\end{equation}
-where $\theta = [\theta_1, ..., \theta_K]$, $\lambda = [\lambda_d^{l}|l \in \Lambda, d \in \mathbb{D}]$ and $\beta =[\beta_1, ..., \beta_N]$ represents the stacked parameters for the trajectories of biomarker, dysfunctionality scores in disease $d$ and subject-specific time shifts.
-Modelling of the likelihood of measurements $y_{ijk}$ within each functional unit is done using the non-parametric GP model by \cite{lorenzi2017disease}:
+where $\boldsymbol{y} = [y_{ijk} | \forall (i,j,k) \in \Omega ]$ is the vector of all biomarker measurements, while $\boldsymbol{\theta} = [\theta_1, ..., \theta_K]$ represents the stacked parameters for the trajectories of biomarkers in agnostic units, $\boldsymbol{\lambda} = [\lambda_d^{l}|l \in \Lambda, d \in \mathbb{D}]$ are the parameters of the dysfunctionality trajectories within the disease models, $\boldsymbol{\beta} =[\beta_1, ..., \beta_N]$ are the subject-specific time shifts and $\boldsymbol{\epsilon} = [\epsilon_k | k \in K]$  estimates biomarker measurement noise.
+Here we assumed independence across different subjects, but the biomarker measurements and visits are linked through the latent time-shift $\beta_i$ for each subject.
+The parameters of the model that need to be estimated are $[\boldsymbol{\theta}, \boldsymbol{\lambda}, \boldsymbol{\beta}, \boldsymbol{\epsilon}]$.
+
+\subsection{Modelling Biomarker Trajectories}
+\label{sec:dktBiomkTraj}
+
+So far we defined the DKT framework using generic models $g(\ .\ ; \theta_k)$ and $f(\ .\ ; \lp)$ for the biomarker trajectories within the agnostic units and the disease models.
+Now we choose to implement the $f$ and $g$ models as parametric sigmoidal curves, to enable a robust optimisation and because these models account for the floor and ceiling effects normally observed in AD biomarkers \cite{sabuncu2011dynamics}.
+The sigmoidal model for $f$ is defined as:
 
 \begin{equation}
-p(y_{ijk}| \beta_i^{\psi(k)}, \theta_k) = f(m_{ij} + \beta_i^{\psi(k)}) + \epsilon
+ f(s;\theta_k) = \frac{a_k}{1+exp(-b_k(s-c_k))} + d_k
 \end{equation}
-where $f \approx GP(0, \Sigma_G)$ models the population-level trajectory.
-The trajectory is also constrained to be monotonically increasing by enforcing a prohibit-likelihood on the derivative process $\frac{df(t)}{dt}$ \cite{lorenzi2017disease}.
-The same model is also applied to express the likelihood of $p(\beta_i^{\psi(k)}| \lambda_d^{\psi(k)}, \beta_i)$, which can give an estimate of the trajectories of dysfunctionality scores in the model specific to disease $d$.
-As already explained in \cite{lorenzi2017disease}, model fitting is done by Expectation-Propagation through conjugate gradient descent.
-In our hierarchical design, we first fit the models within each functional units independently, and then fit the disease specific models.
+
+where $s$ is the disease progression score of a subject and $\theta_k = [a_k, b_k, c_k, d_k]$ are parameters controlling the shape of the trajectory for biomarker $k$: $d_k$ and $d_k + a_k$ represent the lower and upper limits of the sigmoidal function, $c_k$ represents the inflection point and $a_k b_k/4$ represents the slope at the inflection point.
+A similar model is used also for $g$.
+
+\subsection{Parameter Estimation}
+
+\newcommand{\uu}{^{(u)}}
+\newcommand{\um}{^{(u-1)}}
+
+We estimate the model parameters using a two-stage approach.
+In the first stage, we perform belief propagation within each agnostic unit and then within each disease model.
+Each agnostic unit and disease model is assumed to be an independent disease progression model that we fit by alternatively optimising the fit of biomarker trajectories and subject-specific time-shifts, using the approach described in \cite{jedynak2012computational}.
+At this stage we assume the existence of a latent variable $\beta_i^{\psi(k)} = f(\beta_i + m_{ij}; \lp)$ representing the dysfunctionality score of subject $i$ within the agnostic unit $\psi(k)$, which represents a time-shift within that agnostic unit.
+
+In the second stage we jointly optimise across all agnostic units and disease models using loopy belief propagation.
+An overview of the algorithm is given in Figure \ref{fig:dktAlgo}.
+Given the initial parameters estimated from the first stage (line 1), the algorithm continuously updates the biomarker trajectories within the agnostic units (lines 4-5), dysfunctionality trajectories (line 9) and subject-specific time shifts (line 13) until convergence.
+The cost function for all parameters is nearly identical, the main difference being the measurements $(i,j,k)$ over subjects $i$, visits $j$ and biomarkers $k$ that are selected for computing the measurement error.
+For estimating the trajectory of biomarker $k$ within agnostic unit $\psi(k)$, measurements are taken from $\Omega_k$ representing all measurements of biomarker $k$ from all subjects and visits.
+For estimating the dysfunctionality trajectories,  $\Omega_{d,l}$ represents the measurement indices from all subjects with disease $d$ (i.e.
+$d_i = d$) and all biomarkers $k$ that belong to agnostic unit $l$ (i.e.
+$\psi(k) = l$).
+Finally, $\Omega_i$ (line 13) represents all measurements from subject $i$, for all biomarkers and visits.
+
+\begin{figure}
+\begin{algorithm}[H]
+\tiny
+ Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
+  \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
+   \tcp*[l]{Estimate biomarker trajectories (disease agnostic)}
+    \For{$k=1$ to $K$}{
+      ${\theta_k\uu = \argmin_{\theta_k} \sum_{(i,j) \in \Omega_k} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\right) \right]^2  - log\ p(\theta_k)}$\\
+      ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
+    }
+     \tcp*[l]{Estimate dysfunctionality trajectories (disease specific)} 
+    \For{$d=1 \in \mathbb{D}$}{
+      \For{$l=1 \in \Lambda$}{
+        ${\lambda_{d}^{l, (u)} = \argmin_{\lambda_{d}^{l}} \sum_{(i,j,k) \in \Omega_{d,l}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lambda_{d}^{l}) ; \theta_k\uu 
+        \right) \right]^2  - log\ p(\lambda_{d}^{l})}$\\
+      }
+    }
+    \tcp*[l]{Estimate subject-specific time shifts} 
+    \For{$i=1 \in [1, \dots, S]$}{
+      ${\beta_i\uu = \argmin_{\beta_i} \sum_{(j,k) \in \Omega_i} \left[y_{ijk} - g\left(f(\beta_i + m_{ij}; \lpuu) ; \theta_k\uu
+      \right) \right]^2  - log\ p(\beta_i)}$\\
+    }
+}
+\normalfont
+\end{algorithm}
+\caption[The algorithm for estimating the DKT parameters]{The algorithm for estimating the DKT parameters.
+The algorithm successively updates the biomarker trajectories within the agnostic units (disease agnostic models), dysfunctionality trajectories (disease specific) and subject-specific time shifts until convergence.}
+\label{fig:dktAlgo}
+\end{figure}
+
+\subsection{Generating Synthetic Data}
+\label{sec:dktMetSyn}
+
+We first test DKT on synthetic data, in order to assess the performance when ground truth is known. We generate synthetic data from two diseases as follows:
+\begin{itemize}
+ \item[] \textbf{Disease model}
+ \item We define two agnostic units $l_0$ and $l_1$ and 6 biomarkers $k_0-k_6$, which we allocate to agnostic units as follows: $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$.
+Within their units, we define the trajectory of each biomarker as a sigmoidal curves with the following $\theta_k$ parameters:
+ \begin{itemize}
+  \item agnostic unit $l_0$: $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$ and $\theta_4 = (1,5,0.9,0)$ 
+  \item agnostic unit $l_1$: $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$ and $\theta_5 = (1,10,0.9,0)$ 
+ \end{itemize}
+ \item We define two synthetic diseases, "synthetic AD" ($d=0$) and "synthetic PCA" ($d=1$).
+For each disease $d$, each agnostic unit $l$ has a distinct dysfunctionality trajectory defined as a sigmoidal curve with parameters $\lambda_d^l$ as follows:
+ \begin{itemize}
+  \item "synthetic AD" disease: $\lambda_0^0 = (1, 0.3, -4, 0)$  and $\lambda_0^1 = (1, 0.2, 6, 0)$.
+  \item "synthetic PCA" disease: $\lambda_1^0 = (1, 0.3, 6, 0)$ and $\lambda_1^1 = (1, 0.2, -4, 0)$.
+ \end{itemize}
+
+ \item[] \textbf{Subject model}
+ \item We generated time-shifts $\beta_i$ for 100 subjects (disease $d_0$) and 50 subjects (disease $d_1$) based on a uniform distribution with ranges $(-13, 10)$ years before/after disease onset. 
+ \item Within each disease, we generated the subjects' diagnosis (controls/patients) based on an exponential likelihood model with mean -4.5 (controls)/4.5 (patients) years before/after disease onset.
+ \item For each subject and each biomarker, we generated data for four consecutive visits, each visit one year apart, using a noise standard deviation of 0.05.
+\end{itemize}
+
+These trajectory and subject parameters were chosen to mimic the TADPOLE and DRC cohorts, described below.
+Before fitting DKT on the synthetic dataset, we discarded the data from biomarkers $k_0$, $k_1$, $k_4$ and $k_5$ for all subjects within the synthetic PCA cohort, to simulate the lack of multimodal data in these subjects.
+Remaining biomarkers $k_2$ and $k_3$, for which data was still available in the synthetic PCA cohort, are assumed to be of the same modality (e.g.
+MRI volume) but to represent measurements from different brain regions (e.g.
+temporal and occipital).
 
@@
 
-We used 87 controls and 76 PCA subjects from our local dataset, which only had MRI scans.
-From TADPOLE/ADNI, out of the subjects that contained at least one FDG PET, AV45, AV1451 or DTI scan, we randomly selected a subset of 230 subjects, in order not to bias the numbers towards the typical AD disease.
-Volumetric measures for each subject have been obtained using the Freesurfer software in both ADNI and our local dataset.
+We trained DKT on ADNI data from the TADPOLE challenge (Marinescu et al., arXiv, 2018), since it contained a large number of multimodal biomarkers already pre-processed and aggregated into one table.
+From the TADPOLE dataset we selected a subset of 230 subjects which had at least one FDG PET, AV45, AV1451 or DTI scan.
+Most subjects also had MRI scans and cognitive tests.
+In order to model another disease, we further included 76 PCA subjects from our own centre in the training set, along with 67 tAD and 87 age-matched controls, all of which only had MRI scans.
+
+For both datasets, volumetric measures for each subject have been obtained using the Freesurfer software.
 For FDG, AV45 and AV1451 PET, we used already extracted SUVR measures from ADNI.
 For DTI, we used fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe.
-For every lobe, we averaged the biomarker values for regions of interest within each lobe, regressed out age, gender, total intracranial volume (TIV) and dataset (ADNI vs our local dataset) and then normalised the final biomarker values to lie within the [0,1] range.
+For every lobe, we averaged the biomarker values for regions of interest within each lobe and regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs our own dataset).
+Finally, we normalised the biomarker values to lie within the [0,1] range.
+
+For validating DKT's performance at predicting missing biomarkers in PCA, we used a separate test set of DTI scans from our own controls and PCA subjects.
+As this validation set was acquired at a centre different from ADNI and on different scanners, we matched the FA mean and standard deviation of our own controls to be equal to the FA mean and standard deviation of the ADNI controls.
+No DTI data from PCA subjects was exposed to the algorithm at training time.
 
 \section{Results}
+\label{sec:dktRes}
+
+\subsection{Synthetic Results}
+\label{sec:dktResSyn}
+
+Fig.
+\ref{fig:dktSynthTrajCompTrue} shows the true and estimated subject shifts and trajectories for each agnostic unit $l$ and biomarker $k$.
+In the top-left figures we show scatter plots of the true shifts (y-axis) against estimated shifts (x-axis), for the 'synthetic AD' and 'synthetic PCA' diseases.
+On the top-right and middle-left figures, we show the trajectories of the agnostic units within disease $d=0$ (synthetic AD) and $d=1$ (synthetic PCA).
+In the middle-right and bottom-left figures, we show the biomarker trajectories within units $l_0$ and $l_1$.
+In Figure \ref{fig:dktSynthTrajDrcSpace}, we show the corresponding trajectories of PCA patients, which as opposed to Fig.
+\ref{fig:dktSynthTrajCompTrue}, are plotted directly against the time-shifts, as it is normally done in a classical disease progression model.
+We also show the true trajectories and the data of the synthetic PCA cohort.
+
+The results in Fig.
+\ref{fig:dktSynthTrajCompTrue} suggest that the DKT-estimated trajectories match closely (MAE $<$ 0.058) with the true trajectories, for both the unit-trajectories within the disease-specific models and the biomarker trajectories within the disease-agnostic models.
+Moreover, the subject time-shifts are very close ($R^2$ $>$ 0.98) to the true time-shifts.
+When plotted directly against the disease space, the estimated PCA trajectories also match the true trajectories, even when there is a complete lack of such data (Fig.
+\ref{fig:dktSynthTrajDrcSpace}, biomarkers 0,1,4 and 5).
+There are however small errors in  biomarkers 0 and 5 which are due to measurement noise (confirmed by experiments with smaller noise level -- not shown here).
+The equivalent trajectories estimated for the synthetic AD cohort also show very good agreement with the true trajectories (results not shown here).
+
+% \begin{figure}
+% \includegraphics[width=\textwidth]{images/dkt/plotHierData611_synth1Pen5_JMD.png}
+%  \caption{(top-left) (top-left) Scatter plots of the true shifts (y-axis) against estimated shifts (x-axis), for the 'synthetic AD' (left) and 'synthetic PCA' (right) diseases.
+(top-right and middle-left) Trajectories of unit }
+%  \label{fig:dktSynthTrajHierData}
+% \end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{\jmdFld/resfiles/synth/synth1_JMD/compTrueParams101_synth1_JMD.pdf}
+ \caption[DKT Simulation Results - Comparison between true and DKT-estimated biomarker trajectories and subject time-shifts.]{Comparison between true and DKT-estimated subject time-shifts and biomarker trajectories.
+(top-left) Scatter plots of the true shifts (y-axis) against estimated shifts (x-axis), for the 'synthetic AD' (left) and 'synthetic PCA' (right) diseases.
+We also show the DKT-estimated and true trajectories of the agnostic units within the 'synthetic AD' disease (top-right) and the 'synthetic PCA' disease (middle-left).
+For these figures, the x-axis measures the normalised disease progression score $s_i$ while the y-axis measures the dysfunctionality scores $f(s_i;\lambda_d^l)$.
+Finally, we also show the biomarker trajectories within unit 0 (middle-right) and unit 1 (bottom), where the x-axis represents the dysfunctionality scores $f(s_i;\lambda_d^l)$ and the y-axis represents the biomarker value.}
+ \label{fig:dktSynthTrajCompTrue}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{\jmdFld/resfiles/synth/synth1_JMD/trajDisSpaceDis1_101_synth1_JMD.pdf}
+ \caption[Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories]{Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories.
+Estimation of the trajectories in biomarkers 0,1,4 and 5 has been done without any data from the "synthetic PCA" disease, only based on the disease-agnostic correlations with biomarkers 2 and 3.}
+ \label{fig:dktSynthTrajDrcSpace}
+\end{figure}
+
+\subsection{Results on TADPOLE and DRC Datasets}
+\label{sec:dktResTadDrc}
 
 % Fig1: biomarker traj.
-over dysfunction scores in one functional unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
+over dysfunction scores in one agnostic unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
 Fig.
-\ref{fig:pcaTadDisSpace}A shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data.
+\ref{fig:pcaTadDisSpace} shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data.
 The X-axis shows the dysfunctionality scores within the occipital unit, which represent estimated time-shifts, in months, from an arbitrary reference X=0, while the Y-axis shows biomarker values normalised to [0,1] range.
-The model shows a good data fit, and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunctionality scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}.
-Fig.
-\ref{fig:pcaTadDisSpace} also shows the progression of dysfunctionality scores over the disease stage for (B) typical AD and (C) PCA.
-In typical AD, we see that temporal dysfunction becomes abnormal earliest, while PCA shows early parietal dysfunction, again in line with previous findings in the literature \cite{crutch2012posterior,Baron2001}.
+The model shows an unbiased data fit (Fig.
+\ref{fig:pcaTadDisSpace}A), and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunctionality scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}.
+We also show the progression of dysfunctionality scores over the disease stage for typical AD (Fig \ref{fig:pcaTadDisSpace}B) and PCA (Fig \ref{fig:pcaTadDisSpace}C).
+In typical AD, we see that hippocampal dysfunction becomes abnormal earliest, while PCA shows early hippocampal dysfunction, which is later exceeded by the dysfunction in the occipital, temporal and parietal regions, which are known to be affected in PCA \cite{crutch2012posterior,Baron2001}.
+
 In Fig.
@@
 We do this for five different modalities: MRI Volumes, DTI, FDG, AV45 and AV1451.
-The results again recapitulate known patterns in PCA, where trajectories from posterior regions become abnormal first in MRI, DTI and AV1451, with the exception of FDG.
-This might be due to the normalisation of the FDG trajectories on the Y-axis to the [0,1] range, which results in artificially more scaling to trajectories with less disease signal.
-We also notice two non-monotonic trajectories in the AV45 plot, which are due to bad starting points within parameter optimisation combined with low disease signal within these biomarkers.
+The results again recapitulate known patterns in PCA, where posterior regions are predominantly affected in all modalities.
+However, for MRI volumes and AV45, we also see early abnormalities, which we attribute to the models underestimating the biomarker measurement noise.
 
@@
 \end{subfigure}
-\caption{(A) Estimated biomarker trajectories in the occipital functional unit.
+\caption{(A) Estimated biomarker trajectories in the occipital agnostic unit.
 Subject data from ADNI and our local cohort are also shown.
@@
 
-\section{Validation}
+\section{Validation on DTI Data in PCA}
+\label{sec:dktResVal}
 
-We performed validation using a separate test set of 20 DTI scans from controls and PCA patients from our local cohort.
-Fig.
-\ref{fig:DTIvalid} shows the estimated DTI biomarker trajectories plotted directly against the disease progression time-shifts, along with the DTI data from the test set.
-The time shift of each subject was estimated only based on their MRI scans.
-The inferred trajectories in PCA show good agreement with the data in the cingulate, hippocampus and temporal lobes, but less agreement in the other regions.
-This might be because the tAD DTI dataset that was used to train the model had better disease signal in regions traditionally associated with typical AD.
+We validated our model using a separate test set of 20 DTI scans from controls and PCA patients from the DRC cohort.
+We used DKT to predict the DTI biomarker values for the subjects within the unseen test set, using only their MRI biomarkers.
+Table \ref{sec:dktPerfMetrics} shows the prediction mean squared error (MSE) and rank correlation between the DKT-predicted biomarker values and the measured values.
+We computed the rank correlation in order to remove the effect of any systemic biases due to the completely different progressive disease and dataset that we are predicting on.
+We also show similar performance metrics for two simpler models:
+\begin{itemize}
+ \item Latent stage model: a sigmoidal based disease progression model, as described in \cite{jedynak2012computational}. This model assumes all tAD and PCA subjects follow the same progression.
+ \item Linear model: a linear univariate regression model that predicts the DTI biomarker based on the corresponding MRI biomarker, independently for each region.
+\end{itemize}
 
-% DTI validation with local dataset
+The results from Table \ref{sec:dktPerfMetrics} indicate that the DKT predictions are relatively small for the frontal, occipital, temporal and parietal areas.
+Moreover, rank correlation of the DKT predicted values is also high for the cingulate lobe and the hippocampus.
+In terms of model comparison, DKT has better performance than the linear model (all results significant with $p < 0.002$, Bonferroni corrected), since it uses information across all brain regions instead of assuming independence across regions.
+While DKT has similar performance to the latent stage model, it does not assume that the diseases have the same progression and it allows us to understand mechanisms that are shared between related diseases.
+
+\newcommand{\cw}{c}
 
 \begin{table}
+\footnotesize
 \begin{tabular}{c | c c c c c c}
-\textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippocampus} & \textbf{Occipital} & \textbf{Parietal} & \textbf{Temporal}\\
+\textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippo.} & \textbf{Occip.} & \textbf{Parietal} & \textbf{Temporal}\\
 & \multicolumn{6}{c}{\textbf{Prediction Error (MSE)}}\\
@@
 \end{tabular}
+\normalfont
 \caption[Performance evaluation of DKT and two simpler models]{Performance evaluation of DKT and two simpler models: a latent stage disease progression model and a linear univariate model of DTI predicted from MRI.
@@
 \section{Discussion}
+\label{sec:dktDis}
 
-% summary: what we developed -> what the model can do
-We present here, for the first time, a framework for joint modelling of biomarker progression in multiple dementias simultaneously.
-The framework allows transferring biomarker trajectories to rare dementias for which there is not enough data to allow estimation of such trajectories, accounting for a different spatial distribution of pathology between dementia varieties.
+We presented DKT, a framework that enables, for the first time, joint modelling of biomarker progressions in multiple neurodegenerative diseases simultaneously.
+The framework allows the inference of biomarker trajectories in rare diseases, for which there is not enough data to allow estimation of such trajectories, and accounts for a different spatial distribution of pathology between distinct phenotypes.
+This further enables us to understand the complex mechanisms of rare diseases, as well as mechanisms shared between different types of related diseases.
 
-% limitations: each region needs to have disease signal -> validation did not work well in some regions -> not accounted for subject specific effects. 
-Our work has several limitations: (1) the model assumes that within each region there is enough disease signal to properly estimate the dysfunctionality scores (2) in the validation, the inferred trajectories in posterior regions traditionally associated with PCA do not match well with the data and (3) in terms of modelling, our framework did not account for subject-specific effects.
+We provided an example implementation of DKT using specific models of the biomarker trajectories, measurement noise and link function (the disease progression score).
+However, DKT should be considered as a general framework for joint modelling of biomarker trajectories within different diseases simultaneously.
+The actual implementation of DKT can thus be extended to use non-parametric trajectories, or more complex link functions that estimate not only subject time-shifts but also progression speed or higher order terms.
 
-% how to address aforementioned limitations + future work
-There are several potential avenues for further research.
-During training, we can account for missing disease signal in some brain regions by adding other types of dementias that affect other regions.
-The estimated PCA trajectories also need to be thoroughly validated using data from other modalities.
-The model formulation also extends naturally to include subject-specific effects.
+While in this work we have focused on Alzheimer's variants such as tAD and PCA, DKT can also be applied to other progressive neurodegenerative diseases of non-Alzheimer's type such as tauopathies (e.g.
+Frontotemporal dementia), synucleinopathies (e.g.
+Parkinson's disease), other neurodegenerative diseases such as Huntington's disease or Multiple Sclerosis, and even the normal ageing process.
+Cognitive tests can also be included in the disease-specific sub-models of DKT, or even allocated in the agnostic units of the regions that are responsible for those tasks, based on previous voxel-based morphometry studies.
+However, some care needs to be exercised when selecting the biomarkers and grouping them into agnostic units, as in some diseases the assumption of disease agnostic dynamics might not hold for some groups of molecular biomarkers.
+For example, some non-Alzheimer's tauopathies such as Frontotemporal dementia might show tau abnormalities but no corresponding amyloid abnormalities within the same region.
+In the case of Frontotemporal dementia, we recommend including higher-level biomarkers such as glucose metabolism from FDG, white matter degeneration from DTI or volume from structural MRI, but one should exclude amyloid markers.
+
+Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) DKT currently works only on extracted brain features, discarding important information present in the brain morphometry, 4) for validation, the synthetic experiment we ran was limited to only one setting of the parameters and 5) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
+
+There are several potential avenues for further research: 1) to account for heterogeneity, DKT can also be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) to account for brain morphometry and connectivity, DKT can be extended into a fully spatio-temporal model, by estimating continuous changes in volumetric brain images -- in this case, each voxel can have an associated dysfunctionality score that is derived from measurements of various modalities from that voxel; 4-5) DKT can be further validated on more complex synthetic experiments with variable parameter settings, and on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.
+On these subgroups, DKT can be used to transfer biomarker modalities that have been left out during training.
 
@@
 \begin{thebibliography}{5}
-
-\bibitem{jack2010hypothetical}
-Jack, C.R., Knopman, D.S., Jagust, W.J., Shaw, L.M., Aisen, P.S., Weiner, M.W., Petersen, R.C.
-and Trojanowski, J.Q., 2010.
-Hypothetical model of dynamic biomarkers of the Alzheimer's pathological cascade.
-The Lancet Neurology, 9(1), pp.119-128.
 
@@
 
+\bibitem{iturria2016early}
+Iturria-Medina, Y., Sotero, R.C., Toussaint, P.J., Mateos-Pérez, J.M., Evans, A.C., Weiner, M.W., Aisen, P., Petersen, R., Jack, C.R., Jagust, W.
+and Trojanowki, J.Q., 2016.
+Early role of vascular dysregulation on late-onset Alzheimer's disease based on multifactorial data-driven analysis.
+Nature communications, 7, p.11934.
+
+\bibitem{young2014data}
+Young, A.L., Oxtoby, N.P., Daga, P., Cash, D.M., Fox, N.C., Ourselin, S., Schott, J.M.
+and Alexander, D.C., 2014.
+A data-driven model of biomarker changes in sporadic Alzheimer's disease.
+Brain, 137(9), pp.2564-2577.
+
+\bibitem{young2018uncovering}
+Young, A.L., Marinescu, R.V., Oxtoby, N.P., Bocchetta, M., Yong, K., Firth, N.C., Cash, D.M., Thomas, D.L., Dick, K.M., Cardoso, J.
+and van Swieten, J., 2018.
+Uncovering the heterogeneity and temporal complexity of neurodegenerative diseases with Subtype and Stage Inference.
+Nature communications, 9(1), p.4273.
+
+\bibitem{scelsi2018genetic}
+Scelsi, M.A., Khan, R.R., Lorenzi, M., Christopher, L., Greicius, M.D., Schott, J.M., Ourselin, S.
+and Altmann, A., 2018.
+Genetic study of multimodal imaging Alzheimer's disease progression score implicates novel loci.
+Brain.
+
+\bibitem{hon2017towards}
+Hon, M. and Khan, N., 2017. Towards Alzheimer's Disease Classification through Transfer Learning. arXiv preprint arXiv:1711.11117.
+
+\bibitem{cheng2017multi}
+Cheng, B., Liu, M., Shen, D., Li, Z., Zhang, D.
+and Alzheimer's Disease Neuroimaging Initiative, 2017.
+Multi-domain transfer learning for early diagnosis of Alzheimer's disease.
+Neuroinformatics, 15(2), pp.115-132.
+
+\bibitem{cheng2015domain}
+Cheng, B., Liu, M., Zhang, D., Munsell, B.C. and Shen, D., 2015. Domain transfer learning for MCI conversion prediction. IEEE Transactions on Biomedical Engineering, 62(7), pp.1805-1817.
+
+\bibitem{marinescu2018tadpole}
+Marinescu, R.V., Oxtoby, N.P., Young, A.L., Bron, E.E., Toga, A.W., Weiner, M.W., Barkhof, F., Fox, N.C., Klein, S.
+and Alexander, D.C., 2018.
+TADPOLE Challenge: Prediction of Longitudinal Evolution in Alzheimer's Disease.
+arXiv preprint arXiv:1805.03909.
+
+\bibitem{sabuncu2011dynamics}
+Sabuncu, M.R., Desikan, R.S., Sepulcre, J., Yeo, B.T.T., Liu, H., Schmansky, N.J., Reuter, M., Weiner, M.W., Buckner, R.L., Sperling, R.A.
+and Fischl, B., 2011.
+The dynamics of cortical and hippocampal atrophy in Alzheimer disease.
+Archives of neurology, 68(8), pp.1040-1048.
+
+\bibitem{jedynak2012computational}
+Jedynak, B.M., Lang, A., Liu, B., Katz, E., Zhang, Y., Wyman, B.T., Raunig, D., Jedynak, C.P., Caffo, B., Prince, J.L.
+and Alzheimer's Disease Neuroimaging Initiative, 2012.
+A computational neurodegenerative disease progression score: method and results with the Alzheimer's disease Neuroimaging Initiative cohort.
+Neuroimage, 63(3), pp.1478-1486.
+
 \bibitem{Ossenkoppele2014atrophy}
@@
 
-\bibitem{hon2017towards}
-Hon, M. and Khan, N., 2017. Towards Alzheimer's Disease Classification through Transfer Learning. arXiv preprint arXiv:1711.11117.
-
 \bibitem{Baron2001}

### 2018-11-05 paper/paper.tex
regenerated diagram + fixed methods

@@
 The problem of limited data in medical imaging has so far been addressed through transfer learning methods.
-These techniques successfully used transfer learning techniques to improve the accuracy of AD diagnosis \cite{hon2017towards,cheng2017multi} or prediction of MCI conversion \cite{cheng2015domain}, but have two key limitations.
+These techniques successfully used transfer learning techniques to improve the accuracy of AD diagnosis \cite{hon2017towards} or prediction of MCI conversion \cite{cheng2015domain}, but have two key limitations.
 First, they use deep learning or other machine learning methods, which are not interpretable and don't allow us to understand underlying disease mechanisms that are either specific to rare diseases, or shared across related diseases.
@@
  \centering
- \includegraphics[width=0.8\textwidth,trim=0 0 0 0,clip]{figures/disease_knowledge_transfer.pdf}
+ \includegraphics[width=0.9\textwidth,trim=0 0 0 0,clip]{figures/disease_knowledge_transfer_symbols.pdf}
  \caption{Diagram of the proposed DKT framework.
-We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores (Y-axis, top row), each one related to different brain regions.
-Each region-specific dysfunctionality score then further models (X-axis, bottom row) the progression of several modality-specific biomarkers within that same region.
+We assume that each disease can be modelled as the evolution of abstract dysfunction scores (Y-axis, top row), each one related to different brain regions.
+Each region-specific dysfunction score then further models (X-axis, bottom row) the progression of several modality-specific biomarkers within that same region.
 For instance, the temporal dysfunction, modelled as a biomarker in the disease specific model (top row), is the X-axis in the disease agnostic model (temporal unit, bottom row), which aggregates together abnormality from amyloid, tau and MR imaging within the temporal lobe.
 The biomarker correlations within the bottom units are assumed to be disease agnostic and shared across all diseases modelled.
-Disease knowledge transfer can then be achieved via the disease-agnostic units.}
+Disease knowledge transfer can then be achieved via the disease-agnostic units.
+Mathematical notation from section \ref{sec:method} is shown in red to ease understanding.}
  \label{fig:diagram}
@@
 \section{Method}
+\label{sec:method}
+
+\subsection{Method Overview}
 
@@
 Fig.
-\ref{fig:diagram} shows the overall diagram of our proposed framework for joint modelling of diseases.
-We assume that the progression of each disease (X-axis, top row) can be modelled as a unique evolution of abstract dysfunctionality scores, each one related to different brain regions (top row).
-Each dysfunctionality score is an abstract measure of pathology within that brain region, derived from several multimodal biomarkers (e.g.
-amyloid, tau, MRI).
+\ref{fig:diagram} shows the overall diagram of the DKT framework.
+We assume that the progression of each disease (X-axis, top row) can be modelled as a unique evolution of dysfunction trajectories (top row) representing multimodal pathology within a specific brain region.
+Each dysfunction trajectory is modelled as the progression of several biomarkers within that same region, but acquired using different types of modalities (e.g.
+MRI, PET or DTI, see Fig.
+\ref{fig:diagram} bottom row).
+Each group of biomarkers in the bottom row will be called a \emph{disease-agnostic unit} or simply \emph{agnostic unit}, because the biomarker dynamics are assumed to be shared across all diseases modelled.
 
-Each dysfunctionality score is modelled as the progression of several biomarkers within that same region, but acquired using different types of modalities (Fig.
-\ref{fig:diagram} bottom row).
-Each group of biomarkers in the bottom row will be called an \emph{agnostic unit}, because the biomarker dynamics are shared across all diseases modelled (i.e.
-disease agnostic).
+The assumption that the dynamics of some biomarkers are disease-agnostic (i.e.
+shared across diseases), is key to DKT.
+There are several reasons why we can make this assumption.
+First of all, pathology in many related neurodegenerative diseases ( e.g.
+Alzheimer's variants) are hypothesized to share the same underlying mechanisms (e.g.
+amyloid and tau accumulation), and within one region, such accumulation leads to the same pathology dynamics across all the related disease variants: hypometabolism (FDG PET), neuronal damage (T1/DTI MRI) and region-specific cognitive decline -- the difference between these variants is that distinct regions are affected at different times and with different pathology rates and extent, which are accounted for by the disease specific models.
 
-The assumption that the dynamics of some biomarkers are agnostic, or shared across diseases, is key to DKT.
-There are several reasons why we can make this assumption: first of all, neurodegenerative diseases are known
+\subsection{Mathematical formulation}
 
-We choose to model the correlations within each unit using the disease progression model (DPM) by Jedynak et al.
-2012, but any other DPM can also be used.
+We choose to model the correlations within each agnostic unit using the disease progression model (DPM) by \cite{jedynak2012computational}, but any other DPM can also be used.
 The DPM allows us to reconstruct unit-specific dysfunction progression manifolds (bottom row, X axis), which can be used for staging subjects.
@@
 
-\subsection{Mathematical formulation}
-
 The DKT framework has a generic mathematical formulation which uses several disease progression models as building blocks.
 We will first describe the generic framework, and in the following section we will present our chosen implementation of the building blocks.
+
+\subsection{Disease-Specific Model}
+
+In this section we model the biomarker dynamics that are specific to each disease, by mapping the subjects' disease stages to dysfunction scores.
+We model the progression of related diseases as the unique dynamics of abstract dysfunction scores, which aggregate together data from multiple biomarkers.
+
+We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the time shift of subject $i$.
+We then assume that each subject $i$ has a dysfunction score $\gamma_i^l$ corresponding to multimodal pathology in brain region $l$, which is a function of its disease stage:
+
+\begin{equation}
+\label{eqDysfunctionScoreDef}
+ \gamma_i^l = f(\beta_i + m_{ij}; \lambda_{d_i}^l)
+\end{equation}
+where $f$ is a smooth monotonic function mapping each disease stage to a dysfunction score, having parameters $\lambda_{d_i}^l$ corresponding to agnostic unit $l \in \Lambda$ in the disease space $d_i$ of subject $i$.
+We later implement $f$ using a sigmoid function, as described in section \ref{sec:dktBiomkTraj}.
+Moreover, $d_i \in \mathbb{D}$ represents the index of the disease space where subject $i$ belongs, where $\mathbb{D}$ is the set of all diseases modelled.
+For example, MCI and tAD subjects from ADNI as well as tAD subjects from the DRC cohort can all be assigned $d_i=1$, while PCA subjects from our own centre can be assigned $d_i=2$.
+Healthy controls can be assigned to either disease space, although a more precise assignment would take molecular biomarkers into account.
+Variable $\epsilon_k$ denotes the variance of measurements for biomarker $k$.
+
+\subsection{Disease-Agnostic Model}
+
+In this section we model the biomarker dynamics that are disease-agnostic, by constructing the mapping from the dysfunction scores $\gamma_i^l$ to the biomarker measurements.
 We assume a set of given biomarkers measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers at various visits.
-We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the time shift of subject $i$.
 We further denote by $\theta_k$ the parameters used to represent the trajectory for biomarker $k \in K$ within its agnostic unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ maps each biomarker $k$ to a unique agnostic unit $l \in \Lambda$, where $\Lambda$ is the set of agnostic units.
-Moreover, we denote by $\lambda_d^l$ the parameters for the trajectory of the dysfunction score corresponding to agnostic unit $l \in \Lambda$ in the space of disease $d$.
 These definitions allow us to formulate the likelihood for a single measurement $y_{ijk}$ as follows:
@@
 \begin{equation}
- p(y_{ijk}|\theta_k, \lp, \beta_i, \epsilon_k) = N(y_{ijk}| g(f(\beta_i + m_{ij}; \lp; \theta_k), \epsilon_k)
+ p(y_{ijk}|\theta_k, \lp, \beta_i, \epsilon_k) = N(y_{ijk}| g( \gamma_i^{\psi(k)} ; \theta_k), \epsilon_k)
 \end{equation}
-where $g(\ .\ ; \theta_k)$ represents the trajectory of biomarker $k$ within agnostic unit $\psi(k)$ and $f(\ .\ ; \lambda_{d_i}^{\psi(k)})$ represents the trajectory of the agnostic unit $\psi(k)$ within the space of disease $d_i$.
-To be precise, $d_i \in \mathbb{D}$ represents the index of the disease space where subject $i$ belongs, where $\mathbb{D}$ is the set of all diseases modelled.
-For example, MCI and tAD subjects from ADNI as well as tAD subjects from the DRC cohort can all be assigned $d_i=1$, while PCA subjects from our own centre can be assigned $d_i=2$.
-Healthy controls can be assigned to either disease space, although a more precise assignment would take molecular biomarkers into account.
-Variable $\epsilon_k$ denotes the variance of measurements for biomarker $k$.
+where $g(\ .\ ; \theta_k)$ represents the trajectory of biomarker $k$ within agnostic unit $\psi(k)$.
+Parameters $\lp$ are used to define $\gamma_i^{\psi(k)}$ based on Eq.
+\ref{eqDysfunctionScoreDef}, where agnostic unit $l$ is now referred to as $\psi(k)$, to clarify that this is the unit where biomarker $k$ has been allocated.
+
+\subsection{Extending to Multiple Subjects and Biomarkers}
 
@@
 \end{equation}
-
-where $\boldsymbol{y} = [y_{ijk} | \forall (i,j,k) \in \Omega ]$ is the vector of all biomarker measurements, while $\boldsymbol{\theta} = [\theta_1, ..., \theta_K]$ represents the stacked parameters for the trajectories of biomarkers in agnostic units, $\boldsymbol{\lambda} = [\lambda_d^{l}|l \in \Lambda, d \in \mathbb{D}]$ are the parameters of the dysfunctionality trajectories within the disease models, $\boldsymbol{\beta} =[\beta_1, ..., \beta_N]$ are the subject-specific time shifts and $\boldsymbol{\epsilon} = [\epsilon_k | k \in K]$  estimates biomarker measurement noise.
+where $\boldsymbol{y} = [y_{ijk} | \forall (i,j,k) \in \Omega ]$ is the vector of all biomarker measurements, while $\boldsymbol{\theta} = [\theta_1, ..., \theta_K]$ represents the stacked parameters for the trajectories of biomarkers in agnostic units, $\boldsymbol{\lambda} = [\lambda_d^{l}|l \in \Lambda, d \in \mathbb{D}]$ are the parameters of the dysfunction trajectories within the disease models, $\boldsymbol{\beta} =[\beta_1, ..., \beta_N]$ are the subject-specific time shifts and $\boldsymbol{\epsilon} = [\epsilon_k | k \in K]$  estimates biomarker measurement noise.
 Here we assumed independence across different subjects, but the biomarker measurements and visits are linked through the latent time-shift $\beta_i$ for each subject.
 The parameters of the model that need to be estimated are $[\boldsymbol{\theta}, \boldsymbol{\lambda}, \boldsymbol{\beta}, \boldsymbol{\epsilon}]$.
+It should be noted that $\gamma_i^l$ are not model parameters, as they were defined for clarity of explanation.
 
@@
 So far we defined the DKT framework using generic models $g(\ .\ ; \theta_k)$ and $f(\ .\ ; \lp)$ for the biomarker trajectories within the agnostic units and the disease models.
-Now we choose to implement the $f$ and $g$ models as parametric sigmoidal curves, to enable a robust optimisation and because these models account for the floor and ceiling effects normally observed in AD biomarkers \cite{sabuncu2011dynamics}.
+Now we choose to implement the $f$ and $g$ models as parametric sigmoidal curves, to enable a robust optimisation and because these models account for the floor and ceiling effects present in AD biomarkers.
 The sigmoidal model for $f$ is defined as:
@@
 \end{equation}
-
 where $s$ is the disease progression score of a subject and $\theta_k = [a_k, b_k, c_k, d_k]$ are parameters controlling the shape of the trajectory for biomarker $k$: $d_k$ and $d_k + a_k$ represent the lower and upper limits of the sigmoidal function, $c_k$ represents the inflection point and $a_k b_k/4$ represents the slope at the inflection point.
@@
 In the first stage, we perform belief propagation within each agnostic unit and then within each disease model.
-Each agnostic unit and disease model is assumed to be an independent disease progression model that we fit by alternatively optimising the fit of biomarker trajectories and subject-specific time-shifts, using the approach described in \cite{jedynak2012computational}.
-At this stage we assume the existence of a latent variable $\beta_i^{\psi(k)} = f(\beta_i + m_{ij}; \lp)$ representing the dysfunctionality score of subject $i$ within the agnostic unit $\psi(k)$, which represents a time-shift within that agnostic unit.
+Each agnostic unit and disease model is assumed to be an independent disease progression model that we fit by alternatively optimising the fit of biomarker trajectories and subject-specific time-shifts.
+At this stage we assume the existence of a latent variable $\beta_i^{\psi(k)} = f(\beta_i + m_{ij}; \lp)$ representing the dysfunction score of subject $i$ within the agnostic unit $\psi(k)$, which represents a time-shift within that agnostic unit.
 
@@
 An overview of the algorithm is given in Figure \ref{fig:dktAlgo}.
-Given the initial parameters estimated from the first stage (line 1), the algorithm continuously updates the biomarker trajectories within the agnostic units (lines 4-5), dysfunctionality trajectories (line 9) and subject-specific time shifts (line 13) until convergence.
+Given the initial parameters estimated from the first stage (line 1), the algorithm continuously updates the biomarker trajectories within the agnostic units (lines 4-5), dysfunction trajectories (line 9) and subject-specific time shifts (line 13) until convergence.
 The cost function for all parameters is nearly identical, the main difference being the measurements $(i,j,k)$ over subjects $i$, visits $j$ and biomarkers $k$ that are selected for computing the measurement error.
 For estimating the trajectory of biomarker $k$ within agnostic unit $\psi(k)$, measurements are taken from $\Omega_k$ representing all measurements of biomarker $k$ from all subjects and visits.
-For estimating the dysfunctionality trajectories,  $\Omega_{d,l}$ represents the measurement indices from all subjects with disease $d$ (i.e.
+For estimating the dysfunction trajectories, $\Omega_{d,l}$ represents the measurement indices from all subjects with disease $d$ (i.e.
 $d_i = d$) and all biomarkers $k$ that belong to agnostic unit $l$ (i.e.
@@
     }
-     \tcp*[l]{Estimate dysfunctionality trajectories (disease specific)} 
+     \tcp*[l]{Estimate dysfunction trajectories (disease specific)} 
     \For{$d=1 \in \mathbb{D}$}{
@@
 \caption[The algorithm for estimating the DKT parameters]{The algorithm for estimating the DKT parameters.
-The algorithm successively updates the biomarker trajectories within the agnostic units (disease agnostic models), dysfunctionality trajectories (disease specific) and subject-specific time shifts until convergence.}
+The algorithm successively updates the biomarker trajectories within the agnostic units (disease agnostic models), dysfunction trajectories (disease specific) and subject-specific time shifts until convergence.}
 \label{fig:dktAlgo}
@@
 We first test DKT on synthetic data, in order to assess the performance when ground truth is known. We generate synthetic data from two diseases as follows:
+%\begin{tabular}{c c c}
+%
+%\end{tabular}
 \begin{itemize}
@@
  \item We define two synthetic diseases, "synthetic AD" ($d=0$) and "synthetic PCA" ($d=1$).
-For each disease $d$, each agnostic unit $l$ has a distinct dysfunctionality trajectory defined as a sigmoidal curve with parameters $\lambda_d^l$ as follows:
+For each disease $d$, each agnostic unit $l$ has a distinct dysfunction trajectory defined as a sigmoidal curve with parameters $\lambda_d^l$ as follows:
  \begin{itemize}
@@
 We also show the DKT-estimated and true trajectories of the agnostic units within the 'synthetic AD' disease (top-right) and the 'synthetic PCA' disease (middle-left).
-For these figures, the x-axis measures the normalised disease progression score $s_i$ while the y-axis measures the dysfunctionality scores $f(s_i;\lambda_d^l)$.
-Finally, we also show the biomarker trajectories within unit 0 (middle-right) and unit 1 (bottom), where the x-axis represents the dysfunctionality scores $f(s_i;\lambda_d^l)$ and the y-axis represents the biomarker value.}
+For these figures, the x-axis measures the normalised disease progression score $s_i$ while the y-axis measures the dysfunction scores $f(s_i;\lambda_d^l)$.
+Finally, we also show the biomarker trajectories within unit 0 (middle-right) and unit 1 (bottom), where the x-axis represents the dysfunction scores $f(s_i;\lambda_d^l)$ and the y-axis represents the biomarker value.}
  \label{fig:dktSynthTrajCompTrue}
@@
 \ref{fig:pcaTadDisSpace} shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data.
-The X-axis shows the dysfunctionality scores within the occipital unit, which represent estimated time-shifts, in months, from an arbitrary reference X=0, while the Y-axis shows biomarker values normalised to [0,1] range.
+The X-axis shows the dysfunction scores within the occipital unit, which represent estimated time-shifts, in months, from an arbitrary reference X=0, while the Y-axis shows biomarker values normalised to [0,1] range.
 The model shows an unbiased data fit (Fig.
-\ref{fig:pcaTadDisSpace}A), and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunctionality scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}.
-We also show the progression of dysfunctionality scores over the disease stage for typical AD (Fig \ref{fig:pcaTadDisSpace}B) and PCA (Fig \ref{fig:pcaTadDisSpace}C).
-In typical AD, we see that hippocampal dysfunction becomes abnormal earliest, while PCA shows early hippocampal dysfunction, which is later exceeded by the dysfunction in the occipital, temporal and parietal regions, which are known to be affected in PCA \cite{crutch2012posterior,Baron2001}.
+\ref{fig:pcaTadDisSpace}A), and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunction scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}.
+We also show the progression of dysfunction scores over the disease stage for typical AD (Fig \ref{fig:pcaTadDisSpace}B) and PCA (Fig \ref{fig:pcaTadDisSpace}C).
+In typical AD, we see that hippocampal dysfunction becomes abnormal earliest, while PCA shows early hippocampal dysfunction, which is later exceeded by the dysfunction in the occipital, temporal and parietal regions, which are known to be affected in PCA \cite{crutch2012posterior}.
 
@@
 Subject data from ADNI and our local cohort are also shown.
-The X-axis, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject.
+The X-axis, defined as the occipital dysfunction score, represents the time-shifts (in months) of each subject.
 Red lines represent samples from the trajectory posterior.
 The Y-axis measures biomarker values (normalised).
-(B-C) Progression of dysfunctionality scores for (B) typical AD and (C) PCA.}
+(B-C) Progression of dysfunction scores for (B) typical AD and (C) PCA.}
 \label{fig:pcaTadDisSpace}
@@
 
-There are several potential avenues for further research: 1) to account for heterogeneity, DKT can also be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) to account for brain morphometry and connectivity, DKT can be extended into a fully spatio-temporal model, by estimating continuous changes in volumetric brain images -- in this case, each voxel can have an associated dysfunctionality score that is derived from measurements of various modalities from that voxel; 4-5) DKT can be further validated on more complex synthetic experiments with variable parameter settings, and on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.
+There are several potential avenues for further research: 1) to account for heterogeneity, DKT can also be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) to account for brain morphometry and connectivity, DKT can be extended into a fully spatio-temporal model, by estimating continuous changes in volumetric brain images -- in this case, each voxel can have an associated dysfunction score that is derived from measurements of various modalities from that voxel; 4-5) DKT can be further validated on more complex synthetic experiments with variable parameter settings, and on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.
 On these subgroups, DKT can be used to transfer biomarker modalities that have been left out during training.
@@
 
-\bibitem{cheng2017multi}
-Cheng, B., Liu, M., Shen, D., Li, Z., Zhang, D.
-and Alzheimer's Disease Neuroimaging Initiative, 2017.
-Multi-domain transfer learning for early diagnosis of Alzheimer's disease.
-Neuroinformatics, 15(2), pp.115-132.
-
 \bibitem{cheng2015domain}
@@
 arXiv preprint arXiv:1805.03909.
-
-\bibitem{sabuncu2011dynamics}
-Sabuncu, M.R., Desikan, R.S., Sepulcre, J., Yeo, B.T.T., Liu, H., Schmansky, N.J., Reuter, M., Weiner, M.W., Buckner, R.L., Sperling, R.A.
-and Fischl, B., 2011.
-The dynamics of cortical and hippocampal atrophy in Alzheimer disease.
-Archives of neurology, 68(8), pp.1040-1048.
 
@@
 
-\bibitem{Baron2001}
-Baron, J.C., Ch\'{e}telat, G., Desgranges, B., Perchey, G., Landeau, B., {de la Sayette}, V. and Eustache, F., 2001. NeuroImage 14(2), pp.298-309.
-
 \end{thebibliography}

### 2018-11-08 paper/paper.tex
minor

@@
 Alzheimer's variants) are hypothesized to share the same underlying mechanisms (e.g.
-amyloid and tau accumulation), and within one region, such accumulation leads to the same pathology dynamics across all the related disease variants: hypometabolism (FDG PET), neuronal damage (T1/DTI MRI) and region-specific cognitive decline -- the difference between these variants is that distinct regions are affected at different times and with different pathology rates and extent, which are accounted for by the disease specific models.
+amyloid and tau accumulation), and within one region, such accumulation leads to the same pathology dynamics across all the related disease variants: hypometabolism (FDG PET), neuronal damage (T1/DTI MRI) and region-specific cognitive decline \cite{jack2010hypothetical} -- the difference between these variants is that distinct brain regions are affected at different times and with different pathology rates and extent, likely caused by selective vulnerability of networks within these regions \cite{seeley2009neurodegenerative}.
+Secondly, even if the diseases share different upstream mechanisms (e.g.
+amyloid vs tau accumulation), downstream biomarkers measuring hypometabolism, white matter degradation and structural markers follow the same pathological cascade and will have similar dynamics across diseases.
 
@@
 We first test DKT on synthetic data, in order to assess the performance when ground truth is known. We generate synthetic data from two diseases as follows:
-%\begin{tabular}{c c c}
-%
-%\end{tabular}
+
 \begin{itemize}
@@
  \item[] \textbf{Subject model}
- \item We generated time-shifts $\beta_i$ for 100 subjects (disease $d_0$) and 50 subjects (disease $d_1$) based on a uniform distribution with ranges $(-13, 10)$ years before/after disease onset. 
+ \item We generated time-shifts $\beta_i$ for 100 subjects (disease $d_0$) and 50 subjects (disease $d_1$), $\beta_i \sim U(-13,10)$ years before/after disease onset. 
  \item Within each disease, we generated the subjects' diagnosis (controls/patients) based on an exponential likelihood model with mean -4.5 (controls)/4.5 (patients) years before/after disease onset.
- \item For each subject and each biomarker, we generated data for four consecutive visits, each visit one year apart, using a noise standard deviation of 0.05.
+ \item For each subject and each biomarker, we generated data for four consecutive visits, each visit one year apart. The noise standard deviation $\epsilon_k$ = 0.05.
 \end{itemize}
@@
 
+\bibitem{jack2010hypothetical}
+Jack Jr, C.R., Knopman, D.S., Jagust, W.J., Shaw, L.M., Aisen, P.S., Weiner, M.W., Petersen, R.C.
+and Trojanowski, J.Q., 2010.
+Hypothetical model of dynamic biomarkers of the Alzheimer's pathological cascade.
+The Lancet Neurology, 9(1), pp.119-128.
+
+\bibitem{seeley2009neurodegenerative}
+Seeley, W.W., Crawford, R.K., Zhou, J., Miller, B.L. and Greicius, M.D., 2009. Neurodegenerative diseases target large-scale human brain networks. Neuron, 62(1), pp.42-52.
+
 \bibitem{jedynak2012computational}

### 2018-11-08 paper/paper.tex
more on DKT paper

@@
 
-We first test DKT on synthetic data, in order to assess the performance when ground truth is known. We generate synthetic data from two diseases as follows:
+We first test DKT on synthetic data, in order to assess its performance when ground truth is known.
+More precisely, we generate data that follows the DKT model exactly, and test the model's ability to recover the disease-specific (higher-level) and disease-agnostic (lower-level) trajectories, as well as the subject time-shifts.
 
-\begin{itemize}
- \item[] \textbf{Disease model}
- \item We define two agnostic units $l_0$ and $l_1$ and 6 biomarkers $k_0-k_6$, which we allocate to agnostic units as follows: $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$.
-Within their units, we define the trajectory of each biomarker as a sigmoidal curves with the following $\theta_k$ parameters:
- \begin{itemize}
-  \item agnostic unit $l_0$: $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$ and $\theta_4 = (1,5,0.9,0)$ 
-  \item agnostic unit $l_1$: $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$ and $\theta_5 = (1,10,0.9,0)$ 
- \end{itemize}
- \item We define two synthetic diseases, "synthetic AD" ($d=0$) and "synthetic PCA" ($d=1$).
-For each disease $d$, each agnostic unit $l$ has a distinct dysfunction trajectory defined as a sigmoidal curve with parameters $\lambda_d^l$ as follows:
- \begin{itemize}
-  \item "synthetic AD" disease: $\lambda_0^0 = (1, 0.3, -4, 0)$  and $\lambda_0^1 = (1, 0.2, 6, 0)$.
-  \item "synthetic PCA" disease: $\lambda_1^0 = (1, 0.3, 6, 0)$ and $\lambda_1^1 = (1, 0.2, -4, 0)$.
- \end{itemize}
+ \begin{table}
+ \begin{tabular}{c | c}
+& \textbf{Trajectory parameters} \\
+%\hline
+ Biomarker allocation &  $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$ \\
+Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$,  $\theta_4 = (1,5,0.9,0)$  \\
+Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$, $\theta_5 = (1,10,0.9,0)$ \\
 
- \item[] \textbf{Subject model}
- \item We generated time-shifts $\beta_i$ for 100 subjects (disease $d_0$) and 50 subjects (disease $d_1$), $\beta_i \sim U(-13,10)$ years before/after disease onset. 
- \item Within each disease, we generated the subjects' diagnosis (controls/patients) based on an exponential likelihood model with mean -4.5 (controls)/4.5 (patients) years before/after disease onset.
- \item For each subject and each biomarker, we generated data for four consecutive visits, each visit one year apart. The noise standard deviation $\epsilon_k$ = 0.05.
-\end{itemize}
+"Synthetic AD" & $\lambda_0^0 = (1, 0.3, -4, 0)$  and $\lambda_0^1 = (1, 0.2, 6, 0)$ \\
+ "Synthetic PCA" & $\lambda_1^0 = (1, 0.3, 6, 0)$ and $\lambda_1^1 = (1, 0.2, -4, 0)$ \\
+\hline
+& \textbf{Subject parameters} \\
+ Number of subjects & 100 (synthetic AD) and 50 (synthetic PCA) \\ 
+ Time-shifts $\beta_i$ & $\beta_i \sim U(-13,10)$ years \\
+ Diagnosis & $p(control) \propto Exp (-4.5)$,  $p(patient) \propto Exp (4.5)$\\
+ Data generation & 4 visits/subject, 1 year apart, $\epsilon_k = 0.05$\\ 
+\end{tabular}
+\caption{Parameters used for synthetic data generation, emulating the TADPOLE dataset and our own dataset.}
+\label{tab:synParams}
+\end{table}
 
-These trajectory and subject parameters were chosen to mimic the TADPOLE and DRC cohorts, described below.
+We generate synthetic data from two diseases (i.e.
+\emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE cohort and our own cohort.
+The six biomarkers ($k_1$-$k_6$) have been allocated to two agnostic units $l_0$ and $l_1$ a-priori.
 Before fitting DKT on the synthetic dataset, we discarded the data from biomarkers $k_0$, $k_1$, $k_4$ and $k_5$ for all subjects within the synthetic PCA cohort, to simulate the lack of multimodal data in these subjects.

### 2018-11-09 paper/paper.tex
DKT addressed the remaining feedback. Now need to to ct down to 12 pages

@@
 The estimation of accurate biomarker signatures in Alzheimer's disease (AD) and related neurodegenerative diseases is crucial for understanding underlying disease mechanisms, predicting subjects' progressions, and enrichment in clinical trials.
-Recently, several data-driven disease progression models were proposed to reconstruct long term biomarker signatures from collections of short term individual measurements \cite{lorenzi2017disease,oxtoby2018}.
+Recently, data-driven disease progression models were proposed to reconstruct long term biomarker signatures from collections of short term individual measurements \cite{lorenzi2017disease,oxtoby2018}.
 When applied to large datasets of typical AD, disease progression models have shown important benefits in understanding the earliest events in the AD cascade \cite{iturria2016early}, quantifying biomarkers' heterogeneity \cite{young2018uncovering}, helped discover novel genes involved in AD \cite{scelsi2018genetic} and they showed improved predictions over standard approaches \cite{oxtoby2018}.
@@
 We apply DKT on Alzheimer's variants and demonstrate its ability to predict non-MRI trajectories for patients with Posterior Cortical Atrophy, in lack of such data.
-This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from the Dementia Research Centre (DRC), UK.
+This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from our own centre.
 We first show that the estimated non-MRI trajectories for PCA subjects are plausible as they agree with previous literature findings.
@@
 Moreover, $d_i \in \mathbb{D}$ represents the index of the disease space where subject $i$ belongs, where $\mathbb{D}$ is the set of all diseases modelled.
-For example, MCI and tAD subjects from ADNI as well as tAD subjects from the DRC cohort can all be assigned $d_i=1$, while PCA subjects from our own centre can be assigned $d_i=2$.
+For example, MCI and tAD subjects from ADNI as well as tAD subjects from our own cohort can all be assigned $d_i=1$, while PCA subjects from our own centre can be assigned $d_i=2$.
 Healthy controls can be assigned to either disease space, although a more precise assignment would take molecular biomarkers into account.
@@
 So far we defined the DKT framework using generic models $g(\ .\ ; \theta_k)$ and $f(\ .\ ; \lp)$ for the biomarker trajectories within the agnostic units and the disease models.
-Now we choose to implement the $f$ and $g$ models as parametric sigmoidal curves, to enable a robust optimisation and because these models account for the floor and ceiling effects present in AD biomarkers.
-The sigmoidal model for $f$ is defined as:
+Now we choose to implement the $f$ and $g$ models as parametric sigmoidal curves similar to \cite{jedynak2012computational}, to enable a robust optimisation and because these models account for the floor and ceiling effects present in AD biomarkers.
 
-\begin{equation}
- f(s;\theta_k) = \frac{a_k}{1+exp(-b_k(s-c_k))} + d_k
-\end{equation}
-where $s$ is the disease progression score of a subject and $\theta_k = [a_k, b_k, c_k, d_k]$ are parameters controlling the shape of the trajectory for biomarker $k$: $d_k$ and $d_k + a_k$ represent the lower and upper limits of the sigmoidal function, $c_k$ represents the inflection point and $a_k b_k/4$ represents the slope at the inflection point.
+%The sigmoidal model for $f$ is defined as:
+%
+%\begin{equation}
+% f(s;\theta_k) = \frac{a_k}{1+exp(-b_k(s-c_k))} + d_k
+%\end{equation}
+%where $s$ is the disease progression score of a subject and $\theta_k = [a_k, b_k, c_k, d_k]$ are parameters controlling the shape of the trajectory for biomarker $k$: $d_k$ and $d_k + a_k$ represent the lower and upper limits of the sigmoidal function, $c_k$ represents the inflection point and $a_k b_k/4$ represents the slope at the inflection point.
 A similar model is used also for $g$.
@@
 In the first stage, we perform belief propagation within each agnostic unit and then within each disease model.
-Each agnostic unit and disease model is assumed to be an independent disease progression model that we fit by alternatively optimising the fit of biomarker trajectories and subject-specific time-shifts.
-At this stage we assume the existence of a latent variable $\beta_i^{\psi(k)} = f(\beta_i + m_{ij}; \lp)$ representing the dysfunction score of subject $i$ within the agnostic unit $\psi(k)$, which represents a time-shift within that agnostic unit.
-
 In the second stage we jointly optimise across all agnostic units and disease models using loopy belief propagation.
@@
 \emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE cohort and our own cohort.
-The six biomarkers ($k_1$-$k_6$) have been allocated to two agnostic units $l_0$ and $l_1$ a-priori.
-Before fitting DKT on the synthetic dataset, we discarded the data from biomarkers $k_0$, $k_1$, $k_4$ and $k_5$ for all subjects within the synthetic PCA cohort, to simulate the lack of multimodal data in these subjects.
+The six biomarkers ($k_1$-$k_6$) have been \emph{a-priori} allocated to two agnostic units $l_0$ and $l_1$.
+To simulate the lack of multimodal data in the synthetic PCA subjects, we discarded the data from biomarkers $k_0$, $k_1$, $k_4$ and $k_5$ for all these subjects.
 Remaining biomarkers $k_2$ and $k_3$, for which data was still available in the synthetic PCA cohort, are assumed to be of the same modality (e.g.
@@
 
-For validating DKT's performance at predicting missing biomarkers in PCA, we used a separate test set of DTI scans from our own controls and PCA subjects.
-As this validation set was acquired at a centre different from ADNI and on different scanners, we matched the FA mean and standard deviation of our own controls to be equal to the FA mean and standard deviation of the ADNI controls.
+For demonstrating DKT's performance at predicting missing biomarkers in PCA, we used a separate test set of 20 DTI scans from our own controls and PCA subjects.
+As this test set was acquired at a centre different from ADNI and on different scanners, we matched the FA mean and standard deviation of our own controls to be equal to the FA mean and standard deviation of the ADNI controls.
 No DTI data from PCA subjects was exposed to the algorithm at training time.
@@
 Fig.
-\ref{fig:dktSynthTrajCompTrue} shows the true and estimated subject shifts and trajectories for each agnostic unit $l$ and biomarker $k$.
-In the top-left figures we show scatter plots of the true shifts (y-axis) against estimated shifts (x-axis), for the 'synthetic AD' and 'synthetic PCA' diseases.
-On the top-right and middle-left figures, we show the trajectories of the agnostic units within disease $d=0$ (synthetic AD) and $d=1$ (synthetic PCA).
-In the middle-right and bottom-left figures, we show the biomarker trajectories within units $l_0$ and $l_1$.
-In Figure \ref{fig:dktSynthTrajDrcSpace}, we show the corresponding trajectories of PCA patients, which as opposed to Fig.
-\ref{fig:dktSynthTrajCompTrue}, are plotted directly against the time-shifts, as it is normally done in a classical disease progression model.
-We also show the true trajectories and the data of the synthetic PCA cohort.
+\ref{fig:dktSynthTrajCompTrue} shows the true and estimated subject shifts and trajectories for each agnostic unit $l$ and biomarker $k$ (see figure caption for detailed description).
+Results suggest that the DKT-estimated trajectories match closely (MAE $<$ 0.058) with the true trajectories, for both the unit-trajectories within the disease-specific models and the biomarker trajectories within the disease-agnostic models.
+Moreover, the subject time-shifts are very close ($R^2$ $>$ 0.98) to the true time-shifts.
 
-The results in Fig.
-\ref{fig:dktSynthTrajCompTrue} suggest that the DKT-estimated trajectories match closely (MAE $<$ 0.058) with the true trajectories, for both the unit-trajectories within the disease-specific models and the biomarker trajectories within the disease-agnostic models.
-Moreover, the subject time-shifts are very close ($R^2$ $>$ 0.98) to the true time-shifts.
-When plotted directly against the disease space, the estimated PCA trajectories also match the true trajectories, even when there is a complete lack of such data (Fig.
+%When plotted directly against the disease space, the estimated PCA trajectories also match the true trajectories, even when there is a complete lack of such data (Fig.
 \ref{fig:dktSynthTrajDrcSpace}, biomarkers 0,1,4 and 5).
@@
 The equivalent trajectories estimated for the synthetic AD cohort also show very good agreement with the true trajectories (results not shown here).
+
+%In Figure \ref{fig:dktSynthTrajDrcSpace}, we show the corresponding trajectories of PCA patients, which as opposed to Fig.
+\ref{fig:dktSynthTrajCompTrue}, are plotted directly against the time-shifts, as it is normally done in a classical disease progression model.
+We also show the true trajectories and the data of the synthetic PCA cohort.
 
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{\jmdFld/resfiles/synth/synth1_JMD/trajDisSpaceDis1_101_synth1_JMD.pdf}
- \caption[Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories]{Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories.
+%\begin{figure}
+%\includegraphics[width=\textwidth]{\jmdFld/resfiles/synth/synth1_JMD/trajDisSpaceDis1_101_synth1_JMD.pdf}
+% \caption[Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories]{Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories.
 Estimation of the trajectories in biomarkers 0,1,4 and 5 has been done without any data from the "synthetic PCA" disease, only based on the disease-agnostic correlations with biomarkers 2 and 3.}
- \label{fig:dktSynthTrajDrcSpace}
-\end{figure}
+% \label{fig:dktSynthTrajDrcSpace}
+%\end{figure}
 
-\subsection{Results on TADPOLE and DRC Datasets}
+\subsection{Results on TADPOLE and our own dataset}
 \label{sec:dktResTadDrc}
@@
 \ref{fig:pcaTadDisSpace} shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data.
-The X-axis shows the dysfunction scores within the occipital unit, which represent estimated time-shifts, in months, from an arbitrary reference X=0, while the Y-axis shows biomarker values normalised to [0,1] range.
 The model shows an unbiased data fit (Fig.
@@
 In Fig.
-\ref{fig:PCAtrajByModality}, we plot the inferred biomarker trajectories for PCA directly across the disease progression.
-We do this for five different modalities: MRI Volumes, DTI, FDG, AV45 and AV1451.
+\ref{fig:PCAtrajByModality}, we plot the inferred trajectories for PCA directly across the disease progression, for all five modalities analysed.
 The results again recapitulate known patterns in PCA, where posterior regions are predominantly affected in all modalities.
@@
 % PCA vs tAD disease space
-% \vspace{-2em}
-\begin{picture}(5,5)
-\put(0,-70){\textbf{\Large{A}}}
-\end{picture}
+
 \begin{figure}[H]
@@
 \centering
+\begin{picture}(5,5)
+\put(-10,170){\textbf{\Large{A}}}
+\end{picture}
 \includegraphics[width=0.8\textwidth, trim=90 0 110 0, clip]{\expFld/unit1_allTraj.png} 
@@
 The X-axis, defined as the occipital dysfunction score, represents the time-shifts (in months) of each subject.
-Red lines represent samples from the trajectory posterior.
 The Y-axis measures biomarker values (normalised).
@@
 
-We validated our model using a separate test set of 20 DTI scans from controls and PCA patients from the DRC cohort.
+We validated our model using a separate test set of 20 DTI scans from controls and PCA patients from our own cohort.
 We used DKT to predict the DTI biomarker values for the subjects within the unseen test set, using only their MRI biomarkers.
@@
 We computed the rank correlation in order to remove the effect of any systemic biases due to the completely different progressive disease and dataset that we are predicting on.
-We also show similar performance metrics for two simpler models:
-\begin{itemize}
- \item Latent stage model: a sigmoidal based disease progression model, as described in \cite{jedynak2012computational}. This model assumes all tAD and PCA subjects follow the same progression.
- \item Linear model: a linear univariate regression model that predicts the DTI biomarker based on the corresponding MRI biomarker, independently for each region.
+We also show the same performance metrics for two simpler models: (1) a \emph{latent stage model}, as described in \cite{jedynak2012computational}, which assumes all tAD and PCA subjects follow the same progression and (2) a \emph{linear univariate model} that predicts the DTI biomarker based on the corresponding MRI biomarker, independently for each region.
 \end{itemize}
 
-The results from Table \ref{sec:dktPerfMetrics} indicate that the DKT predictions are relatively small for the frontal, occipital, temporal and parietal areas.
-Moreover, rank correlation of the DKT predicted values is also high for the cingulate lobe and the hippocampus.
+The results from Table \ref{sec:dktPerfMetrics} indicate that the DKT prediction errors are relatively small for the frontal, occipital, temporal and parietal areas.
+Moreover, the rank correlations of the DKT predicted values are also high for the cingulate lobe and the hippocampus.
 In terms of model comparison, DKT has better performance than the linear model (all results significant with $p < 0.002$, Bonferroni corrected), since it uses information across all brain regions instead of assuming independence across regions.
@@
 DKT & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.18$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
-Latent stage model & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.17$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
-Linear Model & 0.05$\pm$0.02* & 0.15$\pm$0.04* & 0.09$\pm$0.03* & 0.07$\pm$0.03* & 0.07$\pm$0.02* & 0.07$\pm$0.02*\\
+Latent stage & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.17$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
+Linear & 0.05$\pm$0.02* & 0.15$\pm$0.04* & 0.09$\pm$0.03* & 0.07$\pm$0.03* & 0.07$\pm$0.02* & 0.07$\pm$0.02*\\
 & \multicolumn{6}{c}{\textbf{Rank Correlation (Spearman rho)}}\\
 DKT  & 0.76  & 0.48  & 0.76  & 0.55  & 0.55  & 0.33 \\
-Latent stage model  & 0.76  & 0.49  & 0.80*  & 0.56  & 0.51*  & 0.33 \\
-Linear Model  & 0.48*  & 0.31*  & 0.64*  & 0.61*  & 0.57*  & 0.27* \\
+Latent stage & 0.76  & 0.49  & 0.80*  & 0.56  & 0.51*  & 0.33 \\
+Linear & 0.48*  & 0.31*  & 0.64*  & 0.61*  & 0.57*  & 0.27* \\
 \end{tabular}
@@
 While in this work we have focused on Alzheimer's variants such as tAD and PCA, DKT can also be applied to other progressive neurodegenerative diseases of non-Alzheimer's type such as tauopathies (e.g.
-Frontotemporal dementia), synucleinopathies (e.g.
-Parkinson's disease), other neurodegenerative diseases such as Huntington's disease or Multiple Sclerosis, and even the normal ageing process.
+Frontotemporal dementia -- FTD), synucleinopathies (e.g.
+Parkinson's disease -- PD), other neurodegenerative diseases such as Huntington's disease (HD) or Multiple Sclerosis (MS), and even the normal ageing process.
 Cognitive tests can also be included in the disease-specific sub-models of DKT, or even allocated in the agnostic units of the regions that are responsible for those tasks, based on previous voxel-based morphometry studies.
@@
 
-Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) DKT currently works only on extracted brain features, discarding important information present in the brain morphometry, 4) for validation, the synthetic experiment we ran was limited to only one setting of the parameters and 5) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
+Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) DKT currently works only on extracted brain features, discarding important information present in the brain morphometry, 4) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and our own centre and 5) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
 
-There are several potential avenues for further research: 1) to account for heterogeneity, DKT can also be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) to account for brain morphometry and connectivity, DKT can be extended into a fully spatio-temporal model, by estimating continuous changes in volumetric brain images -- in this case, each voxel can have an associated dysfunction score that is derived from measurements of various modalities from that voxel; 4-5) DKT can be further validated on more complex synthetic experiments with variable parameter settings, and on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.
+There are several potential avenues for further research: 1) to account for heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) to account for brain morphometry and connectivity, DKT can be extended into a fully spatio-temporal model, by estimating continuous changes in volumetric brain images -- in this case, each voxel can have an associated dysfunction score that is derived from measurements of various modalities from that voxel; 4-5) DKT can be further validated on more complex synthetic experiments with a range of parameter settings, as well as on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.
 On these subgroups, DKT can be used to transfer biomarker modalities that have been left out during training.
+
+As a key direction of further research, DKT can be used to understand novel mechanisms within a disease (e.g.
+AD variants), as well as across diseases (e.g.
+AD vs FTD).
+For AD variants in particular, one can study whether early biomarkers such as amyloid and tau have different dynamics compared to typical AD, and what these differences are.
+This can be done by choosing different a-priori allocations of biomarkers to agnostic units, for example allocating amyloid and tau into one a shared agnostic unit, or to the disease specific models, thus making them dependent on the disease.
+The model with the best data fit or predictive power will indicate the most likely mechanism.
+For distinct neurodegenerative diseases, it is currently unclear where their corresponding event-cascades merge, and differences in these pathological events across brain regions.
+DKT can therefore be used to answer such questions, by exploring different allocations of upstream biomarkers to such agnostic units, while differences across brain regions can be studied using DKT's region-specific dysfunctionality scores.
 

### 2018-12-04 paper/paper.tex
dkt more cutting down on pages

@@
 \SetCommentSty{mycommfont}
+
+\setlength{\textfloatsep}{1pt}
+%\setlength{\floatsep}{1pt}
+%\setlength{\topmargin}{1pt}
+
+\renewcommand\floatpagefraction{.95}
+\renewcommand\topfraction{.95}
+\renewcommand\bottomfraction{.95}
+\renewcommand\textfraction{.05}   
+\setcounter{totalnumber}{50}
+\setcounter{topnumber}{50}
+\setcounter{bottomnumber}{50}
 
@@
 where $f$ is a smooth monotonic function mapping each disease stage to a dysfunction score, having parameters $\lambda_{d_i}^l$ corresponding to agnostic unit $l \in \Lambda$ in the disease space $d_i$ of subject $i$.
-We later implement $f$ using a sigmoid function, as described in section \ref{sec:dktBiomkTraj}.
+We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this models accounts for the floor and ceiling effects present in AD biomarkers.
 Moreover, $d_i \in \mathbb{D}$ represents the index of the disease space where subject $i$ belongs, where $\mathbb{D}$ is the set of all diseases modelled.
@@
 It should be noted that $\gamma_i^l$ are not model parameters, as they were defined for clarity of explanation.
-
-\subsection{Modelling Biomarker Trajectories}
-\label{sec:dktBiomkTraj}
-
-So far we defined the DKT framework using generic models $g(\ .\ ; \theta_k)$ and $f(\ .\ ; \lp)$ for the biomarker trajectories within the agnostic units and the disease models.
-Now we choose to implement the $f$ and $g$ models as parametric sigmoidal curves similar to \cite{jedynak2012computational}, to enable a robust optimisation and because these models account for the floor and ceiling effects present in AD biomarkers.
 
@@
 
-\subsection{Data acquisition and Preprocessing}
+\subsection{Data Acquisition and Preprocessing}
 
@@
 
-\section{Results}
+\section{Results on Synthetic and Patient Datasets}
 \label{sec:dktRes}
 
-\subsection{Synthetic Results}
-\label{sec:dktResSyn}
-
-Fig.
-\ref{fig:dktSynthTrajCompTrue} shows the true and estimated subject shifts and trajectories for each agnostic unit $l$ and biomarker $k$ (see figure caption for detailed description).
+Results on synthetic data are presented in Fig.
+\ref{fig:dktSynthTrajCompTrue}, showing the true and estimated subject shifts and trajectories for each agnostic unit $l$ and biomarker $k$ (see figure caption for detailed description).
 Results suggest that the DKT-estimated trajectories match closely (MAE $<$ 0.058) with the true trajectories, for both the unit-trajectories within the disease-specific models and the biomarker trajectories within the disease-agnostic models.
@@
 
-\begin{figure}
+\begin{figure}[htp]
 \includegraphics[width=\textwidth]{\jmdFld/resfiles/synth/synth1_JMD/compTrueParams101_synth1_JMD.pdf}
@@
 (top-left) Scatter plots of the true shifts (y-axis) against estimated shifts (x-axis), for the 'synthetic AD' (left) and 'synthetic PCA' (right) diseases.
-We also show the DKT-estimated and true trajectories of the agnostic units within the 'synthetic AD' disease (top-right) and the 'synthetic PCA' disease (middle-left).
+We also show the DKT-estimated and true trajectories of the agnostic units within the 'synthetic AD' disease (top-right, "Dis0") and the 'synthetic PCA' disease (middle-left, "Dis1").
 For these figures, the x-axis measures the normalised disease progression score $s_i$ while the y-axis measures the dysfunction scores $f(s_i;\lambda_d^l)$.
@@
 
-\subsection{Results on TADPOLE and our own dataset}
-\label{sec:dktResTadDrc}
+%\section{Results on TADPOLE and our own dataset}
+%\label{sec:dktResTadDrc}
 
@@
 over dysfunction scores in one agnostic unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
-Fig.
-\ref{fig:pcaTadDisSpace} shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data.
+Results on TADPOLE dataset and our own dataset are presented in Fig.
+\ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data.
 The model shows an unbiased data fit (Fig.
@@
 
-\begin{figure}[H]
+\begin{figure}[htp]
 \centering
@@
 % estimated (hypothetical) trajectories in PCA: DTI, FDG, AV45, AV1451.Volumetric trajectories were based on PCA MRI data.
-\begin{figure}
+\begin{figure}[htp]
  \includegraphics[width=\textwidth, trim=0 20 0 0, clip]{\expFld/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
@@
 We also show the same performance metrics for two simpler models: (1) a \emph{latent stage model}, as described in \cite{jedynak2012computational}, which assumes all tAD and PCA subjects follow the same progression and (2) a \emph{linear univariate model} that predicts the DTI biomarker based on the corresponding MRI biomarker, independently for each region.
-\end{itemize}
 
@@
 Cognitive tests can also be included in the disease-specific sub-models of DKT, or even allocated in the agnostic units of the regions that are responsible for those tasks, based on previous voxel-based morphometry studies.
-However, some care needs to be exercised when selecting the biomarkers and grouping them into agnostic units, as in some diseases the assumption of disease agnostic dynamics might not hold for some groups of molecular biomarkers.
+
+%However, some care needs to be exercised when selecting the biomarkers and grouping them into agnostic units, as in some diseases the assumption of disease agnostic dynamics might not hold for some groups of molecular biomarkers.
 For example, some non-Alzheimer's tauopathies such as Frontotemporal dementia might show tau abnormalities but no corresponding amyloid abnormalities within the same region.

### 2018-12-05 paper/paper.tex
dkt paper now 12 pages

@@
 
-\usepackage[linesnumbered]{algorithm2e}
+\usepackage[linesnumbered,noend]{algorithm2e}
 \newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{blue}{#1}}
@@
 
-\setlength{\textfloatsep}{1pt}
+\setlength{\textfloatsep}{20pt}
 %\setlength{\floatsep}{1pt}
@@
 %
-\titlerunning{Disease Knowledge Transfer across Alzheimer's Variants}  % abbreviated title (for running head)
+\titlerunning{Disease Knowledge Transfer across Neurodegenerative Diseases}  % abbreviated title (for running head)
 %                                     also used for the TOC unless
@@
 % 
-% Author list: Me, Marco (writing+model code), Pere (helped with validation), Alex (writing+tadpole), Neil (tadpole),  Arman, Keir (DTI data), Seb, Danny
+% Author list: Me, Marco (writing+model code), Stefano, Alex Young (writing+tadpole), Pere (helped with validation),Neil (tadpole),  Arman, Keir (DTI data), Seb, Danny
 % 
@@
 While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
-Our code is available online: \url{https://www.dropbox.com/s/yotur0ohvgwc8om/dkt.zip?dl=0} (anonymised link).
+Our code is available online: (URL not shown yet due to double-blind submission).
 
@@
 The estimation of accurate biomarker signatures in Alzheimer's disease (AD) and related neurodegenerative diseases is crucial for understanding underlying disease mechanisms, predicting subjects' progressions, and enrichment in clinical trials.
-Recently, data-driven disease progression models were proposed to reconstruct long term biomarker signatures from collections of short term individual measurements \cite{lorenzi2017disease,oxtoby2018}.
-When applied to large datasets of typical AD, disease progression models have shown important benefits in understanding the earliest events in the AD cascade \cite{iturria2016early}, quantifying biomarkers' heterogeneity \cite{young2018uncovering}, helped discover novel genes involved in AD \cite{scelsi2018genetic} and they showed improved predictions over standard approaches \cite{oxtoby2018}.
+Recently, data-driven disease progression models were proposed to reconstruct long term biomarker signatures from collections of short term individual measurements \cite{lorenzi2017disease,oxtoby2018,jedynak2012computational}.
+When applied to large datasets of typical AD, disease progression models have shown important benefits in understanding the earliest events in the AD cascade \cite{oxtoby2018}, quantifying biomarkers' heterogeneity \cite{young2018uncovering} and they showed improved predictions over standard approaches \cite{oxtoby2018}.
 However, by necessity these models require large datasets -- in addition they must be both multimodal and longitudinal.
@@
 Moreover, such a model estimated from common diseases such as typical AD may not generalise to specific variants.
-For example, in Posterior Cortical Atrophy (PCA) -- a neurodegenerative syndrome causing visual disruption -- posterior regions such as the occipital lobe and superior parietal regions are affected early, instead of the hippocampus and temporal regions that are affected early in typical AD.
+For example, in Posterior Cortical Atrophy (PCA) -- a neurodegenerative syndrome causing visual disruption -- posterior regions such as the occipital lobe are affected early, instead of the hippocampus and temporal regions that are affected early in typical AD.
 
@@
 This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from our own centre.
-We first show that the estimated non-MRI trajectories for PCA subjects are plausible as they agree with previous literature findings.
+We first show that the estimated trajectories for PCA and tAD subjects are plausible as they agree with previous literature findings.
 We further validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients, showing it yields favourable performance compared to standard approaches.
-Code for DKT is available online: \url{https://www.dropbox.com/s/yotur0ohvgwc8om/dkt.zip?dl=0} (anonymised link).
+Code for DKT is available online: (URL not shown yet).
 
@@
 
-\subsection{Mathematical formulation}
-
 We choose to model the correlations within each agnostic unit using the disease progression model (DPM) by \cite{jedynak2012computational}, but any other DPM can also be used.
@@
 More precisely, the X-axis dysfunction scores from the agnostic units become Y-axis measurements in the disease specific models.
-
-The DKT framework has a generic mathematical formulation which uses several disease progression models as building blocks.
-We will first describe the generic framework, and in the following section we will present our chosen implementation of the building blocks.
 
@@
 where $f$ is a smooth monotonic function mapping each disease stage to a dysfunction score, having parameters $\lambda_{d_i}^l$ corresponding to agnostic unit $l \in \Lambda$ in the disease space $d_i$ of subject $i$.
-We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this models accounts for the floor and ceiling effects present in AD biomarkers.
+We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this accounts for floor and ceiling effects present in AD biomarkers.
 Moreover, $d_i \in \mathbb{D}$ represents the index of the disease space where subject $i$ belongs, where $\mathbb{D}$ is the set of all diseases modelled.
 For example, MCI and tAD subjects from ADNI as well as tAD subjects from our own cohort can all be assigned $d_i=1$, while PCA subjects from our own centre can be assigned $d_i=2$.
-Healthy controls can be assigned to either disease space, although a more precise assignment would take molecular biomarkers into account.
 Variable $\epsilon_k$ denotes the variance of measurements for biomarker $k$.
@@
 
-In this section we model the biomarker dynamics that are disease-agnostic, by constructing the mapping from the dysfunction scores $\gamma_i^l$ to the biomarker measurements.
+We model the biomarker dynamics that are disease-agnostic, by constructing the mapping from the dysfunction scores $\gamma_i^l$ to the biomarker measurements.
 We assume a set of given biomarkers measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers at various visits.
-We further denote by $\theta_k$ the parameters used to represent the trajectory for biomarker $k \in K$ within its agnostic unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ maps each biomarker $k$ to a unique agnostic unit $l \in \Lambda$, where $\Lambda$ is the set of agnostic units.
+We further denote by $\theta_k$ the trajectory parameters for biomarker $k \in K$ within its agnostic unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ maps each biomarker $k$ to a unique agnostic unit $l \in \Lambda$, where $\Lambda$ is the set of agnostic units.
 These definitions allow us to formulate the likelihood for a single measurement $y_{ijk}$ as follows:
@@
 The parameters of the model that need to be estimated are $[\boldsymbol{\theta}, \boldsymbol{\lambda}, \boldsymbol{\beta}, \boldsymbol{\epsilon}]$.
-It should be noted that $\gamma_i^l$ are not model parameters, as they were defined for clarity of explanation.
 
@@
 An overview of the algorithm is given in Figure \ref{fig:dktAlgo}.
-Given the initial parameters estimated from the first stage (line 1), the algorithm continuously updates the biomarker trajectories within the agnostic units (lines 4-5), dysfunction trajectories (line 9) and subject-specific time shifts (line 13) until convergence.
+Given the initial parameters estimated from the first stage (line 1), the algorithm continuously updates the biomarker trajectories within the agnostic units (lines 4-5), dysfunction trajectories (line 8) and subject-specific time shifts (line 10) until convergence.
 The cost function for all parameters is nearly identical, the main difference being the measurements $(i,j,k)$ over subjects $i$, visits $j$ and biomarkers $k$ that are selected for computing the measurement error.
@@
 $\psi(k) = l$).
-Finally, $\Omega_i$ (line 13) represents all measurements from subject $i$, for all biomarkers and visits.
+Finally, $\Omega_i$ (line 10) represents all measurements from subject $i$, for all biomarkers and visits.
 
@@
 \end{algorithm}
-\caption[The algorithm for estimating the DKT parameters]{The algorithm for estimating the DKT parameters.
-The algorithm successively updates the biomarker trajectories within the agnostic units (disease agnostic models), dysfunction trajectories (disease specific) and subject-specific time shifts until convergence.}
+\caption[The algorithm for estimating the DKT parameters]{The algorithm used to estimate the DKT parameters, based on loopy belief-propagation.}
 \label{fig:dktAlgo}
@@
 
-We first test DKT on synthetic data, in order to assess its performance when ground truth is known.
-More precisely, we generate data that follows the DKT model exactly, and test the model's ability to recover the disease-specific (higher-level) and disease-agnostic (lower-level) trajectories, as well as the subject time-shifts.
+We first test DKT on synthetic data, to assess its performance when ground truth is known.
+More precisely, we generate data that follows the DKT model exactly, and test DKT's ability to recover biomarker trajectories and subject time-shifts.
+We generate synthetic data from two diseases (i.e.
+\emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE cohort and our own cohort.
+The six biomarkers ($k_1$-$k_6$) have been \emph{a-priori} allocated to two agnostic units $l_0$ and $l_1$.
+To simulate the lack of multimodal data in the synthetic PCA subjects, we discarded the data from biomarkers $k_0$, $k_1$, $k_4$ and $k_5$ for all these subjects.
+Remaining biomarkers $k_2$ and $k_3$, for which data was still available in the synthetic PCA cohort, are assumed to be of the same modality (e.g.
+MRI volume) but to represent measurements from different brain regions (e.g.
+temporal and occipital).
 
@@
 \end{table}
-
-We generate synthetic data from two diseases (i.e.
-\emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE cohort and our own cohort.
-The six biomarkers ($k_1$-$k_6$) have been \emph{a-priori} allocated to two agnostic units $l_0$ and $l_1$.
-To simulate the lack of multimodal data in the synthetic PCA subjects, we discarded the data from biomarkers $k_0$, $k_1$, $k_4$ and $k_5$ for all these subjects.
-Remaining biomarkers $k_2$ and $k_3$, for which data was still available in the synthetic PCA cohort, are assumed to be of the same modality (e.g.
-MRI volume) but to represent measurements from different brain regions (e.g.
-temporal and occipital).
 
@@
 Results on TADPOLE dataset and our own dataset are presented in Fig.
-\ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data.
+\ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with aligned subject data.
 The model shows an unbiased data fit (Fig.
@@
 We also show the progression of dysfunction scores over the disease stage for typical AD (Fig \ref{fig:pcaTadDisSpace}B) and PCA (Fig \ref{fig:pcaTadDisSpace}C).
-In typical AD, we see that hippocampal dysfunction becomes abnormal earliest, while PCA shows early hippocampal dysfunction, which is later exceeded by the dysfunction in the occipital, temporal and parietal regions, which are known to be affected in PCA \cite{crutch2012posterior}.
+While both typical AD and PCA show early hippocampal dysfunction, PCA shows higher dysfunction in the occipital, temporal and parietal regions in late stages, while tAD shows widespread dysfunction, in line with previous literature findings \cite{crutch2012posterior}.
 
@@
 \end{subfigure}
-\caption{(A) Estimated biomarker trajectories in the occipital agnostic unit.
+\caption{(A) Estimated biomarker trajectories in the occipital agnostic unit, emulating the bottom-right subfigure from Fig \ref{fig:diagram}.
 Subject data from ADNI and our local cohort are also shown.
@@
 The Y-axis measures biomarker values (normalised).
-(B-C) Progression of dysfunction scores for (B) typical AD and (C) PCA.}
+(B-C) Progression of dysfunction scores for (B) typical AD and (C) PCA, emulating top subfigures from Fig \ref{fig:diagram}.}
 \label{fig:pcaTadDisSpace}
@@
 The only data that were available were the MRI volumetric data.
-The dynamics of the other biomarkers has been inferred by the model using data from typical AD, and taking into account the different spatial distribution of pathology in PCA as compared to typical AD.}
+The dynamics of the other biomarkers has been inferred by the model using data from typical AD, and taking into account the different spatial distribution of pathology in PCA vs tAD.}
  \label{fig:PCAtrajByModality}
@@
 \normalfont
-\caption[Performance evaluation of DKT and two simpler models]{Performance evaluation of DKT and two simpler models: a latent stage disease progression model and a linear univariate model of DTI predicted from MRI.
-The top table shows the prediction error (MSE) while the bottom table shows the rank correlation measured by Spearman's rho.
+\caption[Performance evaluation of DKT and two simpler models]{Performance evaluation of DKT and two simpler models.
 (*) Statistically significant difference between the current model performance and DKT, based on a two-tailed t-test, Bonferroni corrected.}
@@
 Frontotemporal dementia -- FTD), synucleinopathies (e.g.
-Parkinson's disease -- PD), other neurodegenerative diseases such as Huntington's disease (HD) or Multiple Sclerosis (MS), and even the normal ageing process.
+Parkinson's disease -- PD) or other neurodegenerative diseases, and even the normal ageing process.
 Cognitive tests can also be included in the disease-specific sub-models of DKT, or even allocated in the agnostic units of the regions that are responsible for those tasks, based on previous voxel-based morphometry studies.
@@
 There are several potential avenues for further research: 1) to account for heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) to account for brain morphometry and connectivity, DKT can be extended into a fully spatio-temporal model, by estimating continuous changes in volumetric brain images -- in this case, each voxel can have an associated dysfunction score that is derived from measurements of various modalities from that voxel; 4-5) DKT can be further validated on more complex synthetic experiments with a range of parameter settings, as well as on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.
-On these subgroups, DKT can be used to transfer biomarker modalities that have been left out during training.
+On these subgroups, DKT can be used to transfer biomarker modalities left out during training.
 
@@
 The model with the best data fit or predictive power will indicate the most likely mechanism.
-For distinct neurodegenerative diseases, it is currently unclear where their corresponding event-cascades merge, and differences in these pathological events across brain regions.
-DKT can therefore be used to answer such questions, by exploring different allocations of upstream biomarkers to such agnostic units, while differences across brain regions can be studied using DKT's region-specific dysfunctionality scores.
+For distinct neurodegenerative diseases, it is currently unclear where their corresponding event-cascades merge, and the differences in pathological events across brain regions.
+DKT can be used to answer such questions, by exploring different allocations of upstream biomarkers to such agnostic units, while differences across brain regions can be studied using DKT's region-specific dysfunctionality scores.
 
@@
 
-\bibitem{iturria2016early}
-Iturria-Medina, Y., Sotero, R.C., Toussaint, P.J., Mateos-Pérez, J.M., Evans, A.C., Weiner, M.W., Aisen, P., Petersen, R., Jack, C.R., Jagust, W.
+\bibitem{jedynak2012computational}
+Jedynak, B.M., Lang, A., Liu, B., Katz, E., Zhang, Y., Wyman, B.T., Raunig, D., Jedynak, C.P., Caffo, B., Prince, J.L.
+and ADNI, 2012.
+A computational neurodegenerative disease progression score: method and results with the Alzheimer's disease Neuroimaging Initiative cohort.
+Neuroimage, 63(3), pp.1478-1486.
+
+%\bibitem{iturria2016early}
+%Iturria-Medina, Y., Sotero, R.C., Toussaint, P.J., Mateos-Pérez, J.M., Evans, A.C., Weiner, M.W., Aisen, P., Petersen, R., Jack, C.R., Jagust, W.
 and Trojanowki, J.Q., 2016.
@@
 
-\bibitem{young2014data}
-Young, A.L., Oxtoby, N.P., Daga, P., Cash, D.M., Fox, N.C., Ourselin, S., Schott, J.M.
+%\bibitem{young2014data}
+%Young, A.L., Oxtoby, N.P., Daga, P., Cash, D.M., Fox, N.C., Ourselin, S., Schott, J.M.
 and Alexander, D.C., 2014.
@@
 
-\bibitem{scelsi2018genetic}
-Scelsi, M.A., Khan, R.R., Lorenzi, M., Christopher, L., Greicius, M.D., Schott, J.M., Ourselin, S.
+%\bibitem{scelsi2018genetic}
+%Scelsi, M.A., Khan, R.R., Lorenzi, M., Christopher, L., Greicius, M.D., Schott, J.M., Ourselin, S.
 and Altmann, A., 2018.
@@
 
-\bibitem{jedynak2012computational}
-Jedynak, B.M., Lang, A., Liu, B., Katz, E., Zhang, Y., Wyman, B.T., Raunig, D., Jedynak, C.P., Caffo, B., Prince, J.L.
-and Alzheimer's Disease Neuroimaging Initiative, 2012.
-A computational neurodegenerative disease progression score: method and results with the Alzheimer's disease Neuroimaging Initiative cohort.
-Neuroimage, 63(3), pp.1478-1486.
-
-\bibitem{Ossenkoppele2014atrophy}
-Ossenkoppele, R., Cohn‐Sheehy, B.I., La Joie, R., Vogel, J.W., Möller, C., Lehmann, M., van Berckel, B.N., Seeley, W.W., Pijnenburg, Y.A., Gorno‐Tempini, M.L.
+%\bibitem{Ossenkoppele2014atrophy}
+%Ossenkoppele, R., Cohn‐Sheehy, B.I., La Joie, R., Vogel, J.W., Möller, C., Lehmann, M., van Berckel, B.N., Seeley, W.W., Pijnenburg, Y.A., Gorno‐Tempini, M.L.
 and Kramer, J.H., 2015.

### 2018-12-12 paper/paper.tex
final small edits

@@
 We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature.
-We further validate DKT in two situations: (1) on synthetic data, showing that it can accurately estimate the ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable performance compared to standard approaches.
+We further validate DKT in two situations: (1) on synthetic data, showing that it can accurately estimate the ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable predictive performance compared to standard approaches.
 While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
-Our code is available online: (URL not shown yet due to double-blind submission).
 
@@
 When applied to large datasets of typical AD, disease progression models have shown important benefits in understanding the earliest events in the AD cascade \cite{oxtoby2018}, quantifying biomarkers' heterogeneity \cite{young2018uncovering} and they showed improved predictions over standard approaches \cite{oxtoby2018}.
-However, by necessity these models require large datasets -- in addition they must be both multimodal and longitudinal.
-Such data is not available in rare neurodegenerative diseases.
+However, by necessity these models require large datasets -- in addition they should be both multimodal and longitudinal.
+Such data is not always available in rare neurodegenerative diseases.
 In particular, most datasets for rare neurodegenerative diseases come from local clinical centres, are unimodal (e.g.
@@
 Moreover, such a model estimated from common diseases such as typical AD may not generalise to specific variants.
-For example, in Posterior Cortical Atrophy (PCA) -- a neurodegenerative syndrome causing visual disruption -- posterior regions such as the occipital lobe are affected early, instead of the hippocampus and temporal regions that are affected early in typical AD.
+For example, in Posterior Cortical Atrophy (PCA) -- a neurodegenerative syndrome causing visual disruption -- posterior regions such as the occipital lobe are affected early, instead of the hippocampus and temporal regions in typical AD.
 
 The problem of limited data in medical imaging has so far been addressed through transfer learning methods.
-These techniques successfully used transfer learning techniques to improve the accuracy of AD diagnosis \cite{hon2017towards} or prediction of MCI conversion \cite{cheng2015domain}, but have two key limitations.
+These were successfully used to improve the accuracy of AD diagnosis \cite{hon2017towards} or prediction of MCI conversion \cite{cheng2015domain}, but have two key limitations.
 First, they use deep learning or other machine learning methods, which are not interpretable and don't allow us to understand underlying disease mechanisms that are either specific to rare diseases, or shared across related diseases.
@@
 
-We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal biomarker progressions for multiple neurodegenerative diseases simultaneously -- including rare neurodegenerative diseases -- and which inherently performs transfer learning between the modelled phenotypes.
+We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal biomarker progressions for multiple diseases simultaneously -- including rare neurodegenerative diseases -- and which inherently performs transfer learning between the modelled phenotypes.
 This is achieved by exploiting biomarker relationships that are shared across diseases, whilst accounting for differences in the spatial distribution of brain pathology.
@@
 We further validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients, showing it yields favourable performance compared to standard approaches.
-Code for DKT is available online: (URL not shown yet).
+Code for DKT is available online: (URL not shown yet for anonymity).
 
@@
 For instance, the temporal dysfunction, modelled as a biomarker in the disease specific model (top row), is the X-axis in the disease agnostic model (temporal unit, bottom row), which aggregates together abnormality from amyloid, tau and MR imaging within the temporal lobe.
-The biomarker correlations within the bottom units are assumed to be disease agnostic and shared across all diseases modelled.
+The biomarker relationships within the bottom units are assumed to be disease agnostic and shared across all diseases modelled.
 Disease knowledge transfer can then be achieved via the disease-agnostic units.
@@
 First of all, pathology in many related neurodegenerative diseases ( e.g.
-Alzheimer's variants) are hypothesized to share the same underlying mechanisms (e.g.
+Alzheimer's variants) are hypothesised to share the same underlying mechanisms (e.g.
 amyloid and tau accumulation), and within one region, such accumulation leads to the same pathology dynamics across all the related disease variants: hypometabolism (FDG PET), neuronal damage (T1/DTI MRI) and region-specific cognitive decline \cite{jack2010hypothetical} -- the difference between these variants is that distinct brain regions are affected at different times and with different pathology rates and extent, likely caused by selective vulnerability of networks within these regions \cite{seeley2009neurodegenerative}.
@@
 \end{equation}
-where $f$ is a smooth monotonic function mapping each disease stage to a dysfunction score, having parameters $\lambda_{d_i}^l$ corresponding to agnostic unit $l \in \Lambda$ in the disease space $d_i$ of subject $i$.
+where $f$ is a smooth monotonic function mapping each disease stage to a dysfunction score, having parameters $\lambda_{d_i}^l$ corresponding to agnostic unit $l \in \Lambda$, where $\Lambda$ is the set of all agnostic units.
+Moreover, $d_i \in \mathbb{D}$ represents the index of the disease corresponding to subject $i$, where $\mathbb{D}$ is the set of all diseases modelled.
+For example, MCI and tAD subjects from ADNI as well as tAD subjects from our own cohort can all be assigned $d_i=1$, while PCA subjects from our own centre can be assigned $d_i=2$.
 We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this accounts for floor and ceiling effects present in AD biomarkers.
-Moreover, $d_i \in \mathbb{D}$ represents the index of the disease space where subject $i$ belongs, where $\mathbb{D}$ is the set of all diseases modelled.
-For example, MCI and tAD subjects from ADNI as well as tAD subjects from our own cohort can all be assigned $d_i=1$, while PCA subjects from our own centre can be assigned $d_i=2$.
 Variable $\epsilon_k$ denotes the variance of measurements for biomarker $k$.
@@
 We model the biomarker dynamics that are disease-agnostic, by constructing the mapping from the dysfunction scores $\gamma_i^l$ to the biomarker measurements.
-We assume a set of given biomarkers measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers at various visits.
-We further denote by $\theta_k$ the trajectory parameters for biomarker $k \in K$ within its agnostic unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ maps each biomarker $k$ to a unique agnostic unit $l \in \Lambda$, where $\Lambda$ is the set of agnostic units.
+We assume a set of given biomarker measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers at various visits.
+We further denote by $\theta_k$ the trajectory parameters for biomarker $k \in K$ within its agnostic unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ maps each biomarker $k$ to a unique agnostic unit $l \in \Lambda$.
 These definitions allow us to formulate the likelihood for a single measurement $y_{ijk}$ as follows:
@@
 
-We trained DKT on ADNI data from the TADPOLE challenge (Marinescu et al., arXiv, 2018), since it contained a large number of multimodal biomarkers already pre-processed and aggregated into one table.
+We trained DKT on ADNI data from the TADPOLE challenge \cite{marinescu2018tadpole}, since it contained a large number of multimodal biomarkers already pre-processed and aggregated into one table.
 From the TADPOLE dataset we selected a subset of 230 subjects which had at least one FDG PET, AV45, AV1451 or DTI scan.
@@
 
-For both datasets, volumetric measures for each subject have been obtained using the Freesurfer software.
-For FDG, AV45 and AV1451 PET, we used already extracted SUVR measures from ADNI.
-For DTI, we used fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe.
-For every lobe, we averaged the biomarker values for regions of interest within each lobe and regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs our own dataset).
+For both datasets, we computed multimodal biomarker measurements corresponding to each brain lobe: MRI volumes using the Freesurfer software, FDG-, AV45- and AV1451-PET standardised uptake value ratios (SUVR) extracted with the standard ADNI pipeline, and DTI fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe.
+For every lobe, we regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs our own dataset).
 Finally, we normalised the biomarker values to lie within the [0,1] range.
+For each lobe, we allocated all multimodal biomarkers corresponding to that lobe to its own agnostic unit.
 
@@
 The framework allows the inference of biomarker trajectories in rare diseases, for which there is not enough data to allow estimation of such trajectories, and accounts for a different spatial distribution of pathology between distinct phenotypes.
-This further enables us to understand the complex mechanisms of rare diseases, as well as mechanisms shared between different types of related diseases.
+Our approach is also interpretable, enabling us to understand complex mechanisms underpinning these diseases.
 
@@
 
-While in this work we have focused on Alzheimer's variants such as tAD and PCA, DKT can also be applied to other progressive neurodegenerative diseases of non-Alzheimer's type such as tauopathies (e.g.
-Frontotemporal dementia -- FTD), synucleinopathies (e.g.
-Parkinson's disease -- PD) or other neurodegenerative diseases, and even the normal ageing process.
-Cognitive tests can also be included in the disease-specific sub-models of DKT, or even allocated in the agnostic units of the regions that are responsible for those tasks, based on previous voxel-based morphometry studies.
-
-%However, some care needs to be exercised when selecting the biomarkers and grouping them into agnostic units, as in some diseases the assumption of disease agnostic dynamics might not hold for some groups of molecular biomarkers.
-For example, some non-Alzheimer's tauopathies such as Frontotemporal dementia might show tau abnormalities but no corresponding amyloid abnormalities within the same region.
-In the case of Frontotemporal dementia, we recommend including higher-level biomarkers such as glucose metabolism from FDG, white matter degeneration from DTI or volume from structural MRI, but one should exclude amyloid markers.
+While here we have focused on Alzheimer's variants such as tAD and PCA, DKT can also be applied to other progressive neurodegenerative diseases of non-Alzheimer's type and even the normal ageing process.
+While we only used imaging data, cognitive tests can also be included in the disease-specific sub-models of DKT, or even allocated in the agnostic units of the regions that are responsible for those tasks.
 
@@
 For AD variants in particular, one can study whether early biomarkers such as amyloid and tau have different dynamics compared to typical AD, and what these differences are.
-This can be done by choosing different a-priori allocations of biomarkers to agnostic units, for example allocating amyloid and tau into one a shared agnostic unit, or to the disease specific models, thus making them dependent on the disease.
+This can be done by choosing different a-priori allocations of biomarkers to agnostic units -- e.g.
+allocating amyloid and tau into one a shared agnostic unit, or otherwise directly to the disease specific models, thus making them dependent on the disease.
 The model with the best data fit or predictive power will indicate the most likely mechanism.
-For distinct neurodegenerative diseases, it is currently unclear where their corresponding event-cascades merge, and the differences in pathological events across brain regions.
-DKT can be used to answer such questions, by exploring different allocations of upstream biomarkers to such agnostic units, while differences across brain regions can be studied using DKT's region-specific dysfunctionality scores.
+For distinct neurodegenerative diseases, it is also currently unclear where their corresponding event-cascades merge, and the differences in pathological events across brain regions.
+DKT can be used to answer such questions, by exploring different allocations of upstream biomarkers to such agnostic units.
 
@@
 \bibitem{lorenzi2017disease}
-Lorenzi, M., Filippone, M., Alexander, D.C.
-and Ourselin, S., 2017.
-Disease Progression Modeling and Prediction through Random Effect Gaussian Processes and Time Transformation.
-arXiv preprint arXiv:1701.01668.
+Lorenzi, M., Filippone, M., Frisoni, G.B., Alexander, D.C., Ourselin, S.
+and Alzheimer's Disease Neuroimaging Initiative, 2017.
+Probabilistic disease progression modeling to characterize diagnostic uncertainty: application to staging and prediction in Alzheimer's disease.
+NeuroImage.
 
@@
 and Alexander, D.C., 2018.
-Data-driven models of dominantly-inherited Alzheimer's disease progression.
-bioRxiv, p.250654.
+Data-driven models of dominantly-inherited Alzheimer’s disease progression.
+Brain, 141(5), pp.1529-1544.
 
@@
 TADPOLE Challenge: Prediction of Longitudinal Evolution in Alzheimer's Disease.
-arXiv preprint arXiv:1805.03909.
+arXiv:1805.03909.
 

### 2019-01-11 paper/paper.tex
submitted paper to arxiv

@@
 
-\setlength{\textfloatsep}{20pt}
+\setlength{\textfloatsep}{15pt}
 %\setlength{\floatsep}{1pt}
@@
 
-\author{*******************************************}
+\author{R\u{a}zvan V. Marinescu\inst{1} \and Marco Lorenzi\inst{1,4} \and Stefano Blumberg\inst{1} \and 
+ Alexandra L.
+Young\inst{1} \and Pere P.
+Morell\inst{1} \and Neil P.
+Oxtoby\inst{1} \and Arman Eshaghi\inst{1,2} \and Keir X.
+Yong\inst{3} \and Sebastian J.
+Crutch\inst{3} \and Daniel C.
+Alexander\inst{1}, for
+the  Alzheimer’s  Disease  Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu).
+As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report.
+A complete listing of ADNI investigators can be found at: http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
 
-\institute{***************************************}
+\authorrunning{R\u{a}zvan V. Marinescu et al.} % abbreviated author list (for running head)
+
+\institute{Centre for Medical Image Computing, University College London, UK
+\and 
+Queen Square MS Centre, UCL Institute of Neurology, UK
+\and 
+Dementia Research Centre, University College London, UK
+\and
+University of C\^{o}te d'Azur, Inria Sophia Antipolis, France
+% Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
+% F-91405 Orsay Cedex, France}
+}
 
@@
 
-\newcommand{\expFld}{../resfiles/tad-drcTiny_JMD}
-\newcommand{\jmdFld}{..}
+%\newcommand{\expFld}{../resfiles/tad-drcTiny_JMD}
+\newcommand{\expFld}{.}
 
@@
 Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects.
-For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD dataset (tAD) from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own centre, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
+For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from the Dementia Research Centre (DRC) UK, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
 We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature.
@@
 While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
+Source code for DKT is available online: \url{https://github.com/mrazvan22/dkt}.
 
@@
 We apply DKT on Alzheimer's variants and demonstrate its ability to predict non-MRI trajectories for patients with Posterior Cortical Atrophy, in lack of such data.
-This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from our own centre.
+This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from the Dementia Research Centre (DRC), UK.
 We first show that the estimated trajectories for PCA and tAD subjects are plausible as they agree with previous literature findings.
 We further validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients, showing it yields favourable performance compared to standard approaches.
-Code for DKT is available online: (URL not shown yet for anonymity).
+Code for DKT is available online: \url{https://github.com/mrazvan22/dkt}.
 
@@
  \centering
- \includegraphics[width=0.9\textwidth,trim=0 0 0 0,clip]{figures/disease_knowledge_transfer_symbols.pdf}
+ \includegraphics[width=0.9\textwidth,trim=0 0 0 0,clip]{disease_knowledge_transfer_symbols.pdf}
  \caption{Diagram of the proposed DKT framework.
@@
 Moreover, $d_i \in \mathbb{D}$ represents the index of the disease corresponding to subject $i$, where $\mathbb{D}$ is the set of all diseases modelled.
-For example, MCI and tAD subjects from ADNI as well as tAD subjects from our own cohort can all be assigned $d_i=1$, while PCA subjects from our own centre can be assigned $d_i=2$.
+For example, MCI and tAD subjects from ADNI as well as tAD subjects from the DRC cohort can all be assigned $d_i=1$, while PCA subjects can be assigned $d_i=2$.
 We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this accounts for floor and ceiling effects present in AD biomarkers.
@@
 We generate synthetic data from two diseases (i.e.
-\emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE cohort and our own cohort.
+\emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE and DRC cohorts.
 The six biomarkers ($k_1$-$k_6$) have been \emph{a-priori} allocated to two agnostic units $l_0$ and $l_1$.
@@
  \begin{table}
+ %\fontsize{7}{12}
+ \centering
  \begin{tabular}{c | c}
@@
 \end{tabular}
-\caption{Parameters used for synthetic data generation, emulating the TADPOLE dataset and our own dataset.}
+\caption{Parameters used for synthetic data generation, emulating the TADPOLE and DRC datasets.}
 \label{tab:synParams}
@@
 Most subjects also had MRI scans and cognitive tests.
-In order to model another disease, we further included 76 PCA subjects from our own centre in the training set, along with 67 tAD and 87 age-matched controls, all of which only had MRI scans.
+In order to model another disease, we further included 76 PCA subjects from the DRC centre in the training set, along with 67 tAD and 87 age-matched controls, all of which only had MRI scans.
 
 For both datasets, we computed multimodal biomarker measurements corresponding to each brain lobe: MRI volumes using the Freesurfer software, FDG-, AV45- and AV1451-PET standardised uptake value ratios (SUVR) extracted with the standard ADNI pipeline, and DTI fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe.
-For every lobe, we regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs our own dataset).
+For every lobe, we regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs DRC).
 Finally, we normalised the biomarker values to lie within the [0,1] range.
@@
 
-For demonstrating DKT's performance at predicting missing biomarkers in PCA, we used a separate test set of 20 DTI scans from our own controls and PCA subjects.
-As this test set was acquired at a centre different from ADNI and on different scanners, we matched the FA mean and standard deviation of our own controls to be equal to the FA mean and standard deviation of the ADNI controls.
+For demonstrating DKT's performance at predicting missing biomarkers in PCA, we used a separate test set of 20 DTI scans from controls and PCA subjects from the DRC dataset.
+As this test set was acquired at a centre different from ADNI and on different scanners, we matched the FA mean and standard deviation of the DRC controls to be equal to the FA mean and standard deviation of the ADNI controls.
 No DTI data from PCA subjects was exposed to the algorithm at training time.
@@
 \begin{figure}[htp]
-\includegraphics[width=\textwidth]{\jmdFld/resfiles/synth/synth1_JMD/compTrueParams101_synth1_JMD.pdf}
+%\includegraphics[width=\textwidth]{../resfiles/synth/synth1_JMD/compTrueParams101_synth1_JMD.pdf}
+\includegraphics[width=\textwidth]{compTrueParams101_synth1_JMD.pdf}
  \caption[DKT Simulation Results - Comparison between true and DKT-estimated biomarker trajectories and subject time-shifts.]{Comparison between true and DKT-estimated subject time-shifts and biomarker trajectories.
@@
 %\begin{figure}
-%\includegraphics[width=\textwidth]{\jmdFld/resfiles/synth/synth1_JMD/trajDisSpaceDis1_101_synth1_JMD.pdf}
+%\includegraphics[width=\textwidth]{../resfiles/synth/synth1_JMD/trajDisSpaceDis1_101_synth1_JMD.pdf}
 % \caption[Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories]{Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories.
@@
 over dysfunction scores in one agnostic unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
-Results on TADPOLE dataset and our own dataset are presented in Fig.
+Results on TADPOLE and DRC datasets are presented in Fig.
 \ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with aligned subject data.
@@
 \caption{(A) Estimated biomarker trajectories in the occipital agnostic unit, emulating the bottom-right subfigure from Fig \ref{fig:diagram}.
-Subject data from ADNI and our local cohort are also shown.
+Subject data from ADNI and DRC cohorts are also shown.
 The X-axis, defined as the occipital dysfunction score, represents the time-shifts (in months) of each subject.
@@
 
-We validated our model using a separate test set of 20 DTI scans from controls and PCA patients from our own cohort.
+We validated our model using a separate test set of 20 DTI scans from controls and PCA patients from the DRC.
 We used DKT to predict the DTI biomarker values for the subjects within the unseen test set, using only their MRI biomarkers.
@@
 
-We provided an example implementation of DKT using specific models of the biomarker trajectories, measurement noise and link function (the disease progression score).
-However, DKT should be considered as a general framework for joint modelling of biomarker trajectories within different diseases simultaneously.
+While we provided an example implementation of DKT using specific models of the biomarker trajectories, measurement noise and link function (the disease progression score), DKT should be considered as a general framework for joint modelling of biomarker trajectories within different diseases simultaneously.
 The actual implementation of DKT can thus be extended to use non-parametric trajectories, or more complex link functions that estimate not only subject time-shifts but also progression speed or higher order terms.
@@
 
-Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) DKT currently works only on extracted brain features, discarding important information present in the brain morphometry, 4) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and our own centre and 5) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
+Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) DKT currently works only on extracted brain features, discarding important information present in the brain morphometry, 4) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and the DRC centre and 5) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
 
 There are several potential avenues for further research: 1) to account for heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) to account for brain morphometry and connectivity, DKT can be extended into a fully spatio-temporal model, by estimating continuous changes in volumetric brain images -- in this case, each voxel can have an associated dysfunction score that is derived from measurements of various modalities from that voxel; 4-5) DKT can be further validated on more complex synthetic experiments with a range of parameter settings, as well as on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.
-On these subgroups, DKT can be used to transfer biomarker modalities left out during training.
 
@@
 The model with the best data fit or predictive power will indicate the most likely mechanism.
-For distinct neurodegenerative diseases, it is also currently unclear where their corresponding event-cascades merge, and the differences in pathological events across brain regions.
-DKT can be used to answer such questions, by exploring different allocations of upstream biomarkers to such agnostic units.
 

### 2019-03-07 paper/paper.tex
cutting down on pages

@@
 Alexander\inst{1}, for
-the  Alzheimer’s  Disease  Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu).
+the  Alzheimer’s  Disease  Neuroimaging Initiative
+% \thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu).
 As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report.
-A complete listing of ADNI investigators can be found at: http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
+A complete listing of ADNI investigators can be found at: http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}
+}
 
@@
 
+We estimate the model parameters using a two-stage approach.
+In the first stage, we perform belief propagation within each agnostic unit and then within each disease model.
+In the second stage we jointly optimise across all agnostic units and disease models using loopy belief propagation.
+
 %The sigmoidal model for $f$ is defined as:
@@
 
-\subsection{Parameter Estimation}
-
-\newcommand{\uu}{^{(u)}}
-\newcommand{\um}{^{(u-1)}}
-
-We estimate the model parameters using a two-stage approach.
+% \subsection{Parameter Estimation}
+% 
+% \newcommand{\uu}{^{(u)}}
+% \newcommand{\um}{^{(u-1)}}
+% 
+% We estimate the model parameters using a two-stage approach.
 In the first stage, we perform belief propagation within each agnostic unit and then within each disease model.
@@
 Finally, $\Omega_i$ (line 10) represents all measurements from subject $i$, for all biomarkers and visits.
-
-\begin{figure}
-\begin{algorithm}[H]
-\tiny
- Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
-  \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
-   \tcp*[l]{Estimate biomarker trajectories (disease agnostic)}
-    \For{$k=1$ to $K$}{
-      ${\theta_k\uu = \argmin_{\theta_k} \sum_{(i,j) \in \Omega_k} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\right) \right]^2  - log\ p(\theta_k)}$\\
-      ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
-    }
-     \tcp*[l]{Estimate dysfunction trajectories (disease specific)} 
-    \For{$d=1 \in \mathbb{D}$}{
-      \For{$l=1 \in \Lambda$}{
-        ${\lambda_{d}^{l, (u)} = \argmin_{\lambda_{d}^{l}} \sum_{(i,j,k) \in \Omega_{d,l}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lambda_{d}^{l}) ; \theta_k\uu 
-        \right) \right]^2  - log\ p(\lambda_{d}^{l})}$\\
-      }
-    }
-    \tcp*[l]{Estimate subject-specific time shifts} 
-    \For{$i=1 \in [1, \dots, S]$}{
-      ${\beta_i\uu = \argmin_{\beta_i} \sum_{(j,k) \in \Omega_i} \left[y_{ijk} - g\left(f(\beta_i + m_{ij}; \lpuu) ; \theta_k\uu
-      \right) \right]^2  - log\ p(\beta_i)}$\\
-    }
-}
-\normalfont
-\end{algorithm}
-\caption[The algorithm for estimating the DKT parameters]{The algorithm used to estimate the DKT parameters, based on loopy belief-propagation.}
-\label{fig:dktAlgo}
-\end{figure}
+% 
+% 
+% \begin{figure}
+% \begin{algorithm}[H]
+% \tiny
+%  Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
+%   \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
+%    \tcp*[l]{Estimate biomarker trajectories (disease agnostic)}
+%     \For{$k=1$ to $K$}{
+%       ${\theta_k\uu = \argmin_{\theta_k} \sum_{(i,j) \in \Omega_k} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\right) \right]^2  - log\ p(\theta_k)}$\\
+%       ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
+%     }
+%      \tcp*[l]{Estimate dysfunction trajectories (disease specific)} 
+%     \For{$d=1 \in \mathbb{D}$}{
+%       \For{$l=1 \in \Lambda$}{
+%         ${\lambda_{d}^{l, (u)} = \argmin_{\lambda_{d}^{l}} \sum_{(i,j,k) \in \Omega_{d,l}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lambda_{d}^{l}) ; \theta_k\uu 
+%         \right) \right]^2  - log\ p(\lambda_{d}^{l})}$\\
+%       }
+%     }
+%     \tcp*[l]{Estimate subject-specific time shifts} 
+%     \For{$i=1 \in [1, \dots, S]$}{
+%       ${\beta_i\uu = \argmin_{\beta_i} \sum_{(j,k) \in \Omega_i} \left[y_{ijk} - g\left(f(\beta_i + m_{ij}; \lpuu) ; \theta_k\uu
+%       \right) \right]^2  - log\ p(\beta_i)}$\\
+%     }
+% }
+% \normalfont
+% \end{algorithm}
+% \caption[The algorithm for estimating the DKT parameters]{The algorithm used to estimate the DKT parameters, based on loopy belief-propagation.}
+% \label{fig:dktAlgo}
+% \end{figure}
 
@@
  \begin{table}
- %\fontsize{7}{12}
+\fontsize{6}{8}\selectfont
  \centering
@@
 Most subjects also had MRI scans and cognitive tests.
-In order to model another disease, we further included 76 PCA subjects from the DRC centre in the training set, along with 67 tAD and 87 age-matched controls, all of which only had MRI scans.
+In order to model another disease, we further included 76 PCA subjects from the DRC centre in the training set, along with 67 tAD and 87 age- matched controls, all of which only had MRI scans.
 
@@
 \begin{table}
-\footnotesize
+\centering
+\fontsize{8}{10}\selectfont
 \begin{tabular}{c | c c c c c c}
@@
 \end{tabular}
-\normalfont
+% \normalfont
 \caption[Performance evaluation of DKT and two simpler models]{Performance evaluation of DKT and two simpler models.
 (*) Statistically significant difference between the current model performance and DKT, based on a two-tailed t-test, Bonferroni corrected.}
+% \end{footnotesize}
 \label{sec:dktPerfMetrics}
@@
 
-While we provided an example implementation of DKT using specific models of the biomarker trajectories, measurement noise and link function (the disease progression score), DKT should be considered as a general framework for joint modelling of biomarker trajectories within different diseases simultaneously.
+% While we provided an example implementation of DKT using specific models of the biomarker trajectories, measurement noise and link function (the disease progression score), DKT should be considered as a general framework for joint modelling of biomarker trajectories within different diseases simultaneously.
 The actual implementation of DKT can thus be extended to use non-parametric trajectories, or more complex link functions that estimate not only subject time-shifts but also progression speed or higher order terms.
-
-While here we have focused on Alzheimer's variants such as tAD and PCA, DKT can also be applied to other progressive neurodegenerative diseases of non-Alzheimer's type and even the normal ageing process.
+% 
+% While here we have focused on Alzheimer's variants such as tAD and PCA, DKT can also be applied to other progressive neurodegenerative diseases of non-Alzheimer's type and even the normal ageing process.
 While we only used imaging data, cognitive tests can also be included in the disease-specific sub-models of DKT, or even allocated in the agnostic units of the regions that are responsible for those tasks.
@@
 
-As a key direction of further research, DKT can be used to understand novel mechanisms within a disease (e.g.
+% As a key direction of further research, DKT can be used to understand novel mechanisms within a disease (e.g.
 AD variants), as well as across diseases (e.g.

### 2019-03-08 paper/paper.tex
minor

@@
 While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
-Source code for DKT is available online: \url{https://github.com/mrazvan22/dkt}.
 
@@
 We first show that the estimated trajectories for PCA and tAD subjects are plausible as they agree with previous literature findings.
-We further validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients, showing it yields favourable performance compared to standard approaches.
-Code for DKT is available online: \url{https://github.com/mrazvan22/dkt}.
+We finally validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients.
 
@@
 The biomarker relationships within the bottom units are assumed to be disease agnostic and shared across all diseases modelled.
-Disease knowledge transfer can then be achieved via the disease-agnostic units.
+Knowledge transfer between the two diseases can then be achieved via the disease-agnostic units.
 Mathematical notation from section \ref{sec:method} is shown in red to ease understanding.}
@@
 
-\subsection{Disease-Specific Model}
+% \subsection{Disease-Specific Model}
 
@@
 
-\subsection{Disease-Agnostic Model}
+% \subsection{Disease-Agnostic Model}
 
@@
 
-\subsection{Extending to Multiple Subjects and Biomarkers}
+% \subsection{Extending to Multiple Subjects and Biomarkers}
 
@@
 \ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with aligned subject data.
-The model shows an unbiased data fit (Fig.
-\ref{fig:pcaTadDisSpace}A), and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunction scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}.
+The population model shows a good data fit (Fig.
+\ref{fig:pcaTadDisSpace}A, MAE $\leq$ 0.25), and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunction scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}.
 We also show the progression of dysfunction scores over the disease stage for typical AD (Fig \ref{fig:pcaTadDisSpace}B) and PCA (Fig \ref{fig:pcaTadDisSpace}C).

### 2019-03-08 paper/paper.tex
m

@@
  \centering
- \includegraphics[width=0.9\textwidth,trim=0 0 0 0,clip]{disease_knowledge_transfer_symbols.pdf}
+ \includegraphics[width=0.8\textwidth,trim=0 0 0 0,clip]{disease_knowledge_transfer_symbols.pdf}
  \caption{Diagram of the proposed DKT framework.
@@
 We model the progression of related diseases as the unique dynamics of abstract dysfunction scores, which aggregate together data from multiple biomarkers.
-
 We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the time shift of subject $i$.
@@
 over dysfunction scores in one agnostic unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
-Results on TADPOLE and DRC datasets are presented in Fig.
+
+% Results on TADPOLE and DRC datasets are presented in Fig.
 \ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with aligned subject data.
@@
 
-\begin{figure}[htp]
-\centering
-\begin{subfigure}{\textwidth}
-\centering
-\begin{picture}(5,5)
-\put(-10,170){\textbf{\Large{A}}}
-\end{picture}
-\includegraphics[width=0.8\textwidth, trim=90 0 110 0, clip]{\expFld/unit1_allTraj.png} 
-% \caption{}
-% \label{fig:occipUnit}
-\end{subfigure}
-
-\begin{picture}(5,5)
-\put(0,20){\textbf{\Large{B}}}
-\end{picture}
-\begin{subfigure}{0.47\textwidth}
-\centering
-% typical AD\\
-\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis0_tAD_allTrajZeroOne.png} 
-% \caption{C}
-\end{subfigure}
-\begin{picture}(5,5)
-\put(0,20){\textbf{\Large{C}}}
-\end{picture}
-\begin{subfigure}{0.47\textwidth}
-\centering
-% PCA\\
-\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis1_PCA_allTrajZeroOne.png} 
-% \caption{PCA}
-\end{subfigure}
-\caption{(A) Estimated biomarker trajectories in the occipital agnostic unit, emulating the bottom-right subfigure from Fig \ref{fig:diagram}.
+% \begin{figure}[htp]
+% \centering
+% \begin{subfigure}{\textwidth}
+% \centering
+% \begin{picture}(5,5)
+% \put(-10,170){\textbf{\Large{A}}}
+% \end{picture}
+% \includegraphics[width=0.8\textwidth, trim=90 0 110 0, clip]{\expFld/unit1_allTraj.png} 
+% % \caption{}
+% % \label{fig:occipUnit}
+% \end{subfigure}
+% 
+% \begin{picture}(5,5)
+% \put(0,20){\textbf{\Large{B}}}
+% \end{picture}
+% \begin{subfigure}{0.47\textwidth}
+% \centering
+% % typical AD\\
+% \includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis0_tAD_allTrajZeroOne.png} 
+% % \caption{C}
+% \end{subfigure}
+% \begin{picture}(5,5)
+% \put(0,20){\textbf{\Large{C}}}
+% \end{picture}
+% \begin{subfigure}{0.47\textwidth}
+% \centering
+% % PCA\\
+% \includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis1_PCA_allTrajZeroOne.png}
+% % \caption{PCA}
+% \end{subfigure}
+% \caption{(A) Estimated biomarker trajectories in the occipital agnostic unit, emulating the bottom-right subfigure from Fig \ref{fig:diagram}.
 Subject data from ADNI and DRC cohorts are also shown.
@@
 (B-C) Progression of dysfunction scores for (B) typical AD and (C) PCA, emulating top subfigures from Fig \ref{fig:diagram}.}
-\label{fig:pcaTadDisSpace}
-\end{figure}
+% \label{fig:pcaTadDisSpace}
+% \end{figure}
 

### 2019-03-19 paper/paper.tex
DKT started ADNI subtype validation

@@
 
-\author{R\u{a}zvan V. Marinescu\inst{1} \and Marco Lorenzi\inst{1,4} \and Stefano Blumberg\inst{1} \and 
- Alexandra L.
-Young\inst{1} \and Pere P.
-Morell\inst{1} \and Neil P.
-Oxtoby\inst{1} \and Arman Eshaghi\inst{1,2} \and Keir X.
-Yong\inst{3} \and Sebastian J.
-Crutch\inst{3} \and Daniel C.
-Alexander\inst{1}, for
-the  Alzheimer’s  Disease  Neuroimaging Initiative
-% \thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu).
-As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report.
-A complete listing of ADNI investigators can be found at: http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}
-}
+% \author{******************}
 
-\authorrunning{R\u{a}zvan V. Marinescu et al.} % abbreviated author list (for running head)
+\author{}
 
-\institute{Centre for Medical Image Computing, University College London, UK
-\and 
-Queen Square MS Centre, UCL Institute of Neurology, UK
-\and 
-Dementia Research Centre, University College London, UK
-\and
-University of C\^{o}te d'Azur, Inria Sophia Antipolis, France
-% Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
-% F-91405 Orsay Cedex, France}
-}
+\authorrunning{} % abbreviated author list (for running head)
+
+\institute{}
+
+% \institute{Centre for Medical Image Computing, University College London, UK
+% \and 
+% Queen Square MS Centre, UCL Institute of Neurology, UK
+% \and 
+% Dementia Research Centre, University College London, UK
+% \and
+% University of C\^{o}te d'Azur, Inria Sophia Antipolis, France
+% % Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
+% % F-91405 Orsay Cedex, France}
+% }
 
@@
 DKT is a joint-disease generative model of biomarker progressions, which exploits biomarker relationships that are shared across diseases.
-As opposed to current deep learning approaches, DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects instead of solving simpler control vs diseased classification tasks.
 Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects.
-For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from the Dementia Research Centre (DRC) UK, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
+For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own clinical center (anonymized), for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
 We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature.
@@
 We apply DKT on Alzheimer's variants and demonstrate its ability to predict non-MRI trajectories for patients with Posterior Cortical Atrophy, in lack of such data.
-This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from the Dementia Research Centre (DRC), UK.
+This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from our own center.
 We first show that the estimated trajectories for PCA and tAD subjects are plausible as they agree with previous literature findings.
@@
 
-\subsection{Method Overview}
+% \subsection{Method Overview}
 
@@
 shared across diseases), is key to DKT.
-There are several reasons why we can make this assumption.
+We make this assumption for two reasons.
 First of all, pathology in many related neurodegenerative diseases ( e.g.
 Alzheimer's variants) are hypothesised to share the same underlying mechanisms (e.g.
-amyloid and tau accumulation), and within one region, such accumulation leads to the same pathology dynamics across all the related disease variants: hypometabolism (FDG PET), neuronal damage (T1/DTI MRI) and region-specific cognitive decline \cite{jack2010hypothetical} -- the difference between these variants is that distinct brain regions are affected at different times and with different pathology rates and extent, likely caused by selective vulnerability of networks within these regions \cite{seeley2009neurodegenerative}.
+amyloid and tau accumulation), and within one region, such mechanisms lead to the same pathology dynamics across all the disease variants modelled: hypometabolism, neuronal damage and region-specific cognitive decline \cite{jack2010hypothetical} -- the difference between these variants is that distinct brain regions are affected at different times and with different pathology rates and extent, likely caused by selective vulnerability of networks within these regions \cite{seeley2009neurodegenerative}.
 Secondly, even if the diseases share different upstream mechanisms (e.g.
@@
 
-We choose to model the correlations within each agnostic unit using the disease progression model (DPM) by \cite{jedynak2012computational}, but any other DPM can also be used.
+% We choose to model the correlations within each agnostic unit using the disease progression model (DPM) by \cite{jedynak2012computational}, but any other DPM can also be used.
 The DPM allows us to reconstruct unit-specific dysfunction progression manifolds (bottom row, X axis), which can be used for staging subjects.
@@
 In this section we model the biomarker dynamics that are specific to each disease, by mapping the subjects' disease stages to dysfunction scores.
-We model the progression of related diseases as the unique dynamics of abstract dysfunction scores, which aggregate together data from multiple biomarkers.
 We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the time shift of subject $i$.
@@
 Moreover, $d_i \in \mathbb{D}$ represents the index of the disease corresponding to subject $i$, where $\mathbb{D}$ is the set of all diseases modelled.
-For example, MCI and tAD subjects from ADNI as well as tAD subjects from the DRC cohort can all be assigned $d_i=1$, while PCA subjects can be assigned $d_i=2$.
+For example, MCI and tAD subjects from ADNI as well as tAD subjects from our own cohort can all be assigned $d_i=1$, while PCA subjects can be assigned $d_i=2$.
 We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this accounts for floor and ceiling effects present in AD biomarkers.
@@
 where $g(\ .\ ; \theta_k)$ represents the trajectory of biomarker $k$ within agnostic unit $\psi(k)$.
+We implement $g$ also using a parametric sigmoidal model, as such a model can account for floor and ceiling effects.
 Parameters $\lp$ are used to define $\gamma_i^{\psi(k)}$ based on Eq.
@@
 We generate synthetic data from two diseases (i.e.
-\emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE and DRC cohorts.
+\emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE cohort as well as our own cohort.
 The six biomarkers ($k_1$-$k_6$) have been \emph{a-priori} allocated to two agnostic units $l_0$ and $l_1$.
@@
 \end{tabular}
-\caption{Parameters used for synthetic data generation, emulating the TADPOLE and DRC datasets.}
+\caption{Parameters used for synthetic data generation, emulating the TADPOLE cohort and our own dataset.}
 \label{tab:synParams}
@@
 Most subjects also had MRI scans and cognitive tests.
-In order to model another disease, we further included 76 PCA subjects from the DRC centre in the training set, along with 67 tAD and 87 age- matched controls, all of which only had MRI scans.
+In order to model another disease, we further included MRI scans from 76 PCA subjects from our own cohort, along with scans from 67 tAD and 87 age-matched controls.
 
 For both datasets, we computed multimodal biomarker measurements corresponding to each brain lobe: MRI volumes using the Freesurfer software, FDG-, AV45- and AV1451-PET standardised uptake value ratios (SUVR) extracted with the standard ADNI pipeline, and DTI fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe.
-For every lobe, we regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs DRC).
+For every lobe, we regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs our own center).
 Finally, we normalised the biomarker values to lie within the [0,1] range.
 For each lobe, we allocated all multimodal biomarkers corresponding to that lobe to its own agnostic unit.
-
-For demonstrating DKT's performance at predicting missing biomarkers in PCA, we used a separate test set of 20 DTI scans from controls and PCA subjects from the DRC dataset.
-As this test set was acquired at a centre different from ADNI and on different scanners, we matched the FA mean and standard deviation of the DRC controls to be equal to the FA mean and standard deviation of the ADNI controls.
-No DTI data from PCA subjects was exposed to the algorithm at training time.
 
@@
 \ref{fig:dktSynthTrajCompTrue}, showing the true and estimated subject shifts and trajectories for each agnostic unit $l$ and biomarker $k$ (see figure caption for detailed description).
-Results suggest that the DKT-estimated trajectories match closely (MAE $<$ 0.058) with the true trajectories, for both the unit-trajectories within the disease-specific models and the biomarker trajectories within the disease-agnostic models.
+Results suggest that the DKT-estimated trajectories match closely (MAE $<$ 0.058) with the true trajectories.
 Moreover, the subject time-shifts are very close ($R^2$ $>$ 0.98) to the true time-shifts.
-
-%When plotted directly against the disease space, the estimated PCA trajectories also match the true trajectories, even when there is a complete lack of such data (Fig.
-\ref{fig:dktSynthTrajDrcSpace}, biomarkers 0,1,4 and 5).
-There are however small errors in  biomarkers 0 and 5 which are due to measurement noise (confirmed by experiments with smaller noise level -- not shown here).
-The equivalent trajectories estimated for the synthetic AD cohort also show very good agreement with the true trajectories (results not shown here).
-
-%In Figure \ref{fig:dktSynthTrajDrcSpace}, we show the corresponding trajectories of PCA patients, which as opposed to Fig.
-\ref{fig:dktSynthTrajCompTrue}, are plotted directly against the time-shifts, as it is normally done in a classical disease progression model.
-We also show the true trajectories and the data of the synthetic PCA cohort.
-
-% \begin{figure}
-% \includegraphics[width=\textwidth]{images/dkt/plotHierData611_synth1Pen5_JMD.png}
-%  \caption{(top-left) (top-left) Scatter plots of the true shifts (y-axis) against estimated shifts (x-axis), for the 'synthetic AD' (left) and 'synthetic PCA' (right) diseases.
-(top-right and middle-left) Trajectories of unit }
-%  \label{fig:dktSynthTrajHierData}
-% \end{figure}
 
@@
 We also show the DKT-estimated and true trajectories of the agnostic units within the 'synthetic AD' disease (top-right, "Dis0") and the 'synthetic PCA' disease (middle-left, "Dis1").
-For these figures, the x-axis measures the normalised disease progression score $s_i$ while the y-axis measures the dysfunction scores $f(s_i;\lambda_d^l)$.
-Finally, we also show the biomarker trajectories within unit 0 (middle-right) and unit 1 (bottom), where the x-axis represents the dysfunction scores $f(s_i;\lambda_d^l)$ and the y-axis represents the biomarker value.}
+Finally, we also show the biomarker trajectories within unit 0 (middle-right) and unit 1 (bottom).}
   \label{fig:dktSynthTrajCompTrue}
@@
 
-%\begin{figure}
-%\includegraphics[width=\textwidth]{../resfiles/synth/synth1_JMD/trajDisSpaceDis1_101_synth1_JMD.pdf}
-% \caption[Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories]{Estimated biomarker trajectories for the "synthetic PCA" disease, plotted alongside true trajectories.
-Estimation of the trajectories in biomarkers 0,1,4 and 5 has been done without any data from the "synthetic PCA" disease, only based on the disease-agnostic correlations with biomarkers 2 and 3.}
-% \label{fig:dktSynthTrajDrcSpace}
-%\end{figure}
-
-%\section{Results on TADPOLE and our own dataset}
-%\label{sec:dktResTadDrc}
-
-% Fig1: biomarker traj.
-over dysfunction scores in one agnostic unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
-
-% Results on TADPOLE and DRC datasets are presented in Fig.
-\ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with aligned subject data.
-The population model shows a good data fit (Fig.
-\ref{fig:pcaTadDisSpace}A, MAE $\leq$ 0.25), and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunction scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}.
-We also show the progression of dysfunction scores over the disease stage for typical AD (Fig \ref{fig:pcaTadDisSpace}B) and PCA (Fig \ref{fig:pcaTadDisSpace}C).
-While both typical AD and PCA show early hippocampal dysfunction, PCA shows higher dysfunction in the occipital, temporal and parietal regions in late stages, while tAD shows widespread dysfunction, in line with previous literature findings \cite{crutch2012posterior}.
-
 In Fig.
 \ref{fig:PCAtrajByModality}, we plot the inferred trajectories for PCA directly across the disease progression, for all five modalities analysed.
-The results again recapitulate known patterns in PCA, where posterior regions are predominantly affected in all modalities.
-However, for MRI volumes and AV45, we also see early abnormalities, which we attribute to the models underestimating the biomarker measurement noise.
-
-% PCA vs tAD disease space
-
-% \begin{figure}[htp]
-% \centering
-% \begin{subfigure}{\textwidth}
-% \centering
-% \begin{picture}(5,5)
-% \put(-10,170){\textbf{\Large{A}}}
-% \end{picture}
-% \includegraphics[width=0.8\textwidth, trim=90 0 110 0, clip]{\expFld/unit1_allTraj.png} 
-% % \caption{}
-% % \label{fig:occipUnit}
-% \end{subfigure}
-% 
-% \begin{picture}(5,5)
-% \put(0,20){\textbf{\Large{B}}}
-% \end{picture}
-% \begin{subfigure}{0.47\textwidth}
-% \centering
-% % typical AD\\
-% \includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis0_tAD_allTrajZeroOne.png} 
-% % \caption{C}
-% \end{subfigure}
-% \begin{picture}(5,5)
-% \put(0,20){\textbf{\Large{C}}}
-% \end{picture}
-% \begin{subfigure}{0.47\textwidth}
-% \centering
-% % PCA\\
-% \includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{\expFld/dis1_PCA_allTrajZeroOne.png}
-% % \caption{PCA}
-% \end{subfigure}
-% \caption{(A) Estimated biomarker trajectories in the occipital agnostic unit, emulating the bottom-right subfigure from Fig \ref{fig:diagram}.
-Subject data from ADNI and DRC cohorts are also shown.
-The X-axis, defined as the occipital dysfunction score, represents the time-shifts (in months) of each subject.
-The Y-axis measures biomarker values (normalised).
-(B-C) Progression of dysfunction scores for (B) typical AD and (C) PCA, emulating top subfigures from Fig \ref{fig:diagram}.}
-% \label{fig:pcaTadDisSpace}
-% \end{figure}
+The results recapitulate known patterns in PCA, where posterior regions are predominantly affected.
 
@@
 \begin{figure}[htp]
- \includegraphics[width=\textwidth, trim=0 20 0 0, clip]{\expFld/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
+\centering
+ \includegraphics[width=0.8\textwidth, trim=0 20 0 0, clip]{\expFld/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
  \caption{Estimated trajectories for the PCA cohort.
@@
 
-\section{Validation on DTI Data in PCA}
+\subsection{Validation on DTI Data in PCA}
 \label{sec:dktResVal}
 
-We validated our model using a separate test set of 20 DTI scans from controls and PCA patients from the DRC.
+We validated our model using a separate test set of 20 DTI scans from controls and PCA patients from our own cohort.
 We used DKT to predict the DTI biomarker values for the subjects within the unseen test set, using only their MRI biomarkers.
-Table \ref{sec:dktPerfMetrics} shows the prediction mean squared error (MSE) and rank correlation between the DKT-predicted biomarker values and the measured values.
-We computed the rank correlation in order to remove the effect of any systemic biases due to the completely different progressive disease and dataset that we are predicting on.
+To evaluate prediction accuracy, we computed the prediction mean squared error (MSE) and rank correlation between the DKT-predicted biomarker values and the measured values in PCA patients.
+
+% We computed the rank correlation in order to remove the effect of any systemic biases due to the completely different progressive disease and dataset that we are predicting on.
 We also show the same performance metrics for two simpler models: (1) a \emph{latent stage model}, as described in \cite{jedynak2012computational}, which assumes all tAD and PCA subjects follow the same progression and (2) a \emph{linear univariate model} that predicts the DTI biomarker based on the corresponding MRI biomarker, independently for each region.
@@
 In terms of model comparison, DKT has better performance than the linear model (all results significant with $p < 0.002$, Bonferroni corrected), since it uses information across all brain regions instead of assuming independence across regions.
-While DKT has similar performance to the latent stage model, it does not assume that the diseases have the same progression and it allows us to understand mechanisms that are shared between related diseases.
+While DKT has similar performance to the latent stage model, it allows us to understand mechanisms that are shared between related diseases.
 
@@
 \caption[Performance evaluation of DKT and two simpler models]{Performance evaluation of DKT and two simpler models.
-(*) Statistically significant difference between the current model performance and DKT, based on a two-tailed t-test, Bonferroni corrected.}
+(*) Statistically significant difference in the performance of DKT vs latent-stage/linear models, based on a two-tailed t-test, Bonferroni corrected.}
 % \end{footnotesize}
@@
 We presented DKT, a framework that enables, for the first time, joint modelling of biomarker progressions in multiple neurodegenerative diseases simultaneously.
-The framework allows the inference of biomarker trajectories in rare diseases, for which there is not enough data to allow estimation of such trajectories, and accounts for a different spatial distribution of pathology between distinct phenotypes.
+
+% The framework allows the inference of biomarker trajectories in rare diseases, for which there is not enough data to allow estimation of such trajectories, and accounts for a different spatial distribution of pathology between distinct phenotypes.
 Our approach is also interpretable, enabling us to understand complex mechanisms underpinning these diseases.
@@
 
-Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) DKT currently works only on extracted brain features, discarding important information present in the brain morphometry, 4) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and the DRC centre and 5) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
+Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and our local centre and 4) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
 
-There are several potential avenues for further research: 1) to account for heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) to account for brain morphometry and connectivity, DKT can be extended into a fully spatio-temporal model, by estimating continuous changes in volumetric brain images -- in this case, each voxel can have an associated dysfunction score that is derived from measurements of various modalities from that voxel; 4-5) DKT can be further validated on more complex synthetic experiments with a range of parameter settings, as well as on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.
+There are several potential avenues for further research: 1) to account for heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3-4) DKT can be further validated on more complex synthetic experiments with a range of parameter settings, as well as on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.
 

### 2019-03-19 tadpoleDrcPrepData.py

@@ -31,9 +31,6 @@ def addBiomks(biomkStruct, sourceDf, targetDf, collapseFunc):
     targetDf.loc[:, biomkLabels[b]] = pd.Series(np.sum(dataFrameCurrSubset,axis=1),
       index=targetDf.index)
     # targetDf = targetDf.assign(asd=np.sum(dataFrameCurrSubset,axis=1))
-    # print(targetDf.loc[:5, biomkLabels[b]])
-    # print(ads)
-
 
 
   return targetDf
@@ -200,6 +197,7 @@ def normaliseData(dataDfAll, validDf, allBiomkCols):
   #   # print(asda)
 
   return dataDfAll, validDf
+
 
 def prepareData(finalDataFile, tinyData, addExtraBiomk):
 
@@ -430,10 +428,7 @@ def prepareData(finalDataFile, tinyData, addExtraBiomk):
   # and only keep the DTI. Don't remove the MRI cols though, needs to be in
   # same format as dataDfAll
   # UPDATE May 2018: No, don't drop MRI values. I need them for prediction of DTI vals
-  # from simpler linear model
-  # print('validDf', validDf.loc[:, mriCols])
-  # validDf.loc[:,mriCols] = np.nan
-  # print('validDf', validDf.loc[:,mriCols])
+
 
   outFilePrefix = 'afterReg'
   visValidDf(validDf, outFilePrefix)

### 2019-03-19 tadpoleOnly.py

@@ -1001,8 +1001,6 @@ def main():
 
 
 
-
-
 def runAllExpTadpoleDrc(params, expName, dpmBuilder, compareTrueParamsFunc = None):
   """ runs all experiments"""
 

### 2019-03-19 tadpoleSubtypes.py

@@ -1,0 +1,721 @@
+import sys
+import numpy
+import numpy as np
+import colorsys
+from socket import gethostname
+import time
+import argparse
+import os
+import colorsys
+import copy
+import pandas as pd
+import pickle
+import random
+import auxFunc
+import scipy
+import scipy.io as sio
+import scipy.stats
+
+
+# sys.path.append(os.path.abspath("../diffEqModel/"))
+
+
+parser = argparse.ArgumentParser(description='Launches voxel-wise/point-wise DPM on ADNI'
+                                             'using cortical thickness maps derived from MRI')
+
+parser.add_argument('--agg', dest='agg', type=int, default=0,
+  help='agg=1 => plot figures without using Xwindows, for use on cluster where the plots cannot be displayed '
+       ' agg=0 => plot with Xwindows (for use on personal machine)')
+
+parser.add_argument('--runIndex', dest='runIndex', type=int,
+  default=1, help='index of run instance/process .. for cross-validation')
+
+parser.add_argument('--nrProc', dest='nrProc', type=int,
+  default=1, help='# of processes')
+
+parser.add_argument('--modelToRun', dest='modelToRun', type=int,
+  help='index of model to run')
+
+parser.add_argument('--cluster', action="store_true",
+  help='need to include this flag if runnin on cluster')
+
+parser.add_argument('--nrRows', dest='nrRows', type=int,
+  help='nr of subfigure rows to plot at every iteration')
+
+parser.add_argument('--nrCols', dest='nrCols', type=int,
+  help='nr of subfigure columns to plot at every iteration')
+
+parser.add_argument('--penalty', dest='penalty', type=float,
+  help='penalty value for non-monotonic trajectories. between 0 (no effect) and 10 (strong effect). ')
+
+parser.add_argument('--regData', action="store_true", default=False,
[…]

### 2019-03-26 paper/paper.tex
finished validation on ADNI subtypes

@@
 \textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippo.} & \textbf{Occip.} & \textbf{Parietal} & \textbf{Temporal}\\
-& \multicolumn{6}{c}{\textbf{Prediction Error (MSE)}}\\
-DKT & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.18$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
-Latent stage & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.17$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
-Linear & 0.05$\pm$0.02* & 0.15$\pm$0.04* & 0.09$\pm$0.03* & 0.07$\pm$0.03* & 0.07$\pm$0.02* & 0.07$\pm$0.02*\\
-& \multicolumn{6}{c}{\textbf{Rank Correlation (Spearman rho)}}\\
+& \multicolumn{6}{c}{\textbf{ADNI Subtypes: Hippocampal to Cortical}}\\
+DKT &      0.56 $\pm$ 0.23 &    0.35 $\pm$ 0.17 &        0.58 $\pm$ 0.14 &     -0.10 $\pm$ 0.29 &     0.71 $\pm$ 0.11 &     0.34 $\pm$ 0.26 \\
+Latent stage &      0.44 $\pm$ 0.25 &    0.34 $\pm$ 0.21 &       0.34 $\pm$ 0.24* &     -0.07 $\pm$ 0.22 &     0.64 $\pm$ 0.16 &    0.08 $\pm$ 0.24* \\
+Multivariate &      0.60 $\pm$ 0.18 &   0.11 $\pm$ 0.22* &       0.12 $\pm$ 0.29* &     -0.22 $\pm$ 0.22 &   -0.44 $\pm$ 0.14* &   -0.32 $\pm$ 0.29* \\
+Spline &    -0.24 $\pm$ 0.25* &  -0.06 $\pm$ 0.27* &        0.58 $\pm$ 0.17 &     -0.16 $\pm$ 0.27 &    0.23 $\pm$ 0.25* &    0.10 $\pm$ 0.25* \\
+Linear &    -0.24 $\pm$ 0.25* &   0.20 $\pm$ 0.25* &        0.58 $\pm$ 0.17 &     -0.16 $\pm$ 0.27 &    0.23 $\pm$ 0.25* &    0.13 $\pm$ 0.23* \\
+& \multicolumn{6}{c}{\textbf{tAD to Posterior Cortical Atrophy}}\\
 DKT  & 0.76  & 0.48  & 0.76  & 0.55  & 0.55  & 0.33 \\

### 2019-03-26 JointModel.py

@@ -527,8 +527,6 @@ class JointModel(DisProgBuilder.DPMInterface):
             print(unitModels[u].X[b][s].shape[0])
             raise ValueError('Shape of new array doesnt match')
 
-
-
       # dimension check
       assert newXarray[0].shape[0] == unitModels[u].X_array[0].shape[0]
 
@@ -550,7 +548,7 @@ class JointModel(DisProgBuilder.DPMInterface):
     XshiftedScaledDBS = [0 for _ in range(self.nrDis)]
     XdisDBSX = [0 for _ in range(self.nrDis)]
     for d in range(self.nrDis):
-      XshiftedScaledDBS[d], XdisDBSX[d], _, _ = disModels[d].getData(flagAllBiomkShouldBePresent=True)
+      XshiftedScaledDBS[d], XdisDBSX[d], _, _ = disModels[d].getData(flagAllBiomkShouldBePresent=False)
 
     XdisSX = [0 for _ in range(self.unitModels[0].nrSubj)]
     nrSubj = unitModels[0].nrSubj
@@ -602,9 +600,6 @@ class JointModel(DisProgBuilder.DPMInterface):
         # reset the traj parameters, so that they don't end up with decreasing trajectories
         self.unitModels[u].initialiseParams()
 
-      # self.unitModels[u].priors = self.params['priorsUnitModels'][u]
-      # self.unitModels[u].Set_penalty(2)
-      # self.unitModels[u].estimTrajParams(Niterat = 70)
       self.unitModels[u].estimTrajParams()
 
       # fig = self.unitModels[u].plotter.plotTraj(self.unitModels[u])

### 2019-03-26 Plotter.py

@@ -835,6 +835,9 @@ class PlotterGP(ABC):
       if YsubjData2BSX is not None:
         listsMin += [np.min(yS) for yS in YsubjData2BSX[b] if yS.shape[0] > 0]
         listsMax += [np.max(yS) for yS in YsubjData2BSX[b] if yS.shape[0] > 0]
+      if YsubjData3BSX is not None:
+        listsMin += [np.min(yS) for yS in YsubjData3BSX[b] if yS.shape[0] > 0]
+        listsMax += [np.max(yS) for yS in YsubjData3BSX[b] if yS.shape[0] > 0]
 
       if len(listsMin) == 0:
         min_yB[b] = 0

### 2019-03-26 SigmoidModel.py

@@ -326,7 +326,6 @@ class SigmoidModel(DPMModelGeneric.DPMModelGeneric):
 
   def predictBiomkWithParams(self, newX, params):
 
-
     deltaX = 5 * (self.maxScX - self.minScX)
     if not (self.minScX - deltaX <= np.min(newX) <= self.maxScX + deltaX):
       print('newX', newX)
@@ -337,8 +336,8 @@ class SigmoidModel(DPMModelGeneric.DPMModelGeneric):
     xsScaled = self.applyScalingXForward(newX.reshape(-1), biomk=0)  # arbitrary space ->[0,1]
 
     predictedBiomksXB = np.zeros((xsScaled.shape[0], self.nrBiomk))
-    print('self.nrBiomk',self.nrBiomk)
-    print(dasa)
+    # print('self.nrBiomk',self.nrBiomk)
+    # print(dasa)
     for b in range(self.nrBiomk):
       trajParams, variance = self.unpack_parameters(params[b])
       predictedBiomksXB[:, b] = self.sigFunc(xsScaled, trajParams)

### 2019-03-26 drcValidFuncs.py

@@ -366,20 +366,6 @@ def validateDRCBiomk(dpmObj, params):
   predTrajXB = dpmObj.predictBiomkSubjGivenXs(xsTrajX, disNr)
   trajSamplesBXS = dpmObj.sampleBiomkTrajGivenXs(xsTrajX, disNr, nrSamples=100)
 
-  print('XshiftedDisModelBS', XshiftedDisModelBS)
-  print('XvalidShifFilt', XvalidShifFilt)
-  print('predTrajXB', predTrajXB[:,0])
-  print('xsTrajX', xsTrajX)
-  print('ysPredBS', ysPredBS)
-  # print(ads)
-
-
-  # fig = dpmObj.plotterObj.plotTrajInDisSpace(xsTrajX, predTrajXB, trajSamplesBXS,
-  #   XshiftedDisModelBS, Yfilt, diagSubjCurrDis,
-  #   XvalidShifFilt, YvalidFilt, diagValidFilt,
-  #   XsubjData3BSX=None, YsubjData3BSX=None, diagData3S=None, labels=self.plotTrajParams['labels'], replaceFig=True)
-  # fig.savefig('%s/validPCA.png' % params['outFolder'])
-
 
   ### build a simpler linear predictor from MR to DTI for every ROI independently.
   # Train it on ADNI MR+DTI data and use it to predict DRC-DTI from DRC-MR.
@@ -389,8 +375,10 @@ def validateDRCBiomk(dpmObj, params):
   mriBiomksDf = [i for i in range(len(colsList)) if colsList[i].startswith('Volume')]
   dtiBiomksDf = [i for i in range(len(colsList)) if colsList[i].startswith('DTI')]
 
-  dataDfAllMat = dataDfAll.values
-  # print('dataDfAllMat', dataDfAllMat)
+  dataAllTrainDf = dataDfAll[dataDfAll.dataset == 1,:]
+
+  dataAllTrain = dataAllTrainDf.values
+  # print('dataAllTrain', dataAllTrain)
 
   nrDtiBiomk = len(mriBiomksDf)
 
@@ -425,14 +413,11 @@ def validateDRCBiomk(dpmObj, params):
 
   for f in range(nrDtiBiomk):
 
-    mriDataCurrCol = dataDfAllMat[:, mriBiomksDf[f]]
-    dtiDataCurrCol = dataDfAllMat[:, dtiBiomksDf[f]]
[…]

### 2019-03-26 tadpoleDrc.py

@@ -213,8 +213,6 @@ def main():
   labels = ds['list_biomarkers']
   diag = ds['diag']
 
-
-
   # visDataHist(dataDfAll)
   nrUnqDiags = np.unique(dataDfAll.diag)
   print(dataDfAll)
@@ -439,92 +437,94 @@ def main():
   modelNames, res = evaluationFramework.runModels(params, expName,
    args.modelToRun, runAllExpTadpoleDrc)
 
+  from tadpoleSubtypes import printRes
+
   if params['masterProcess']:
-    printRes(modelNames, res, plotTrajParams)
+    printRes(modelNames, res, plotTrajParams, params)
 
 
-def printRes(modelNames, res, plotTrajParams):
-  #nrModels = len(modelNames)
-
-  print(modelNames)
-
-  modelNames += ['Lin']
-
-  dktModelName = 'JMD'
-  sigModelName = 'Sig'
-  linModelName = 'Lin'
-
-  officialNames = {'JMD' :  'DKT', 'Sig' : 'Latent stage model', 'Lin' : 'Linear Model'}
-
-  dktIndex = 0
-  sigIndex = 1
-  linIndex = 2
-
-  # print('##### biomk prediction ######')
-  nrModels = len(officialNames)
-  mseMUB = list(range(nrModels))
-  mseMeanMU = list(range(nrModels))
-  mseStdMU = list(range(nrModels))
-
-  corrMUB = list(range(nrModels))
-  corrMeanMU = list(range(nrModels))
-  corrStdMU = list(range(nrModels))
-  pvalsMU = list(range(nrModels))
-
-  for m in range(nrModels-1):
-    mseMUB[m] = res[m]['metrics']['dpm']['mseUB']
-    mseMeanMU[m] = np.nanmean(mseMUB[m], axis=1)
-    mseStdMU[m] = np.nanstd(mseMUB[m], axis=1)
-
-    corrMUB[m] = res[m]['metrics']['dpm']['corrUB']
-    # pvalsMU[m] = res[m]['metrics']['dpm']['pValsUB']
-
-    corrMeanMU[m] = np.nanmean(corrMUB[m], axis=1)
-    corrStdMU[m] = np.nanstd(corrMUB[m], axis=1)
-
-  mseMUB[linIndex] = res[0]['metrics']['lin']['mseUB']
[…]

### 2019-03-26 tadpoleSubtypes.py

@@ -15,7 +15,7 @@ import auxFunc
 import scipy
 import scipy.io as sio
 import scipy.stats
-
+from scipy.interpolate import UnivariateSpline
 
 # sys.path.append(os.path.abspath("../diffEqModel/"))
 
@@ -356,66 +356,69 @@ def prepareData(finalDataFile, tinyData, addExtraBiomk):
 
   selectedBiomk = dataDfTadpole.loc[:, 'Volume Cingulate' : ].columns.tolist()
   if addExtraBiomk:
-    selectedBiomk += ['ADAS13', 'CDRSB', 'RAVLT_immediate']
+    selectedBiomk += ['ADAS13', 'CDRSB', 'RAVLT_immediate', 'MMSE', 'FAQ']
 
   for c in selectedBiomk:
     dataDfTadpole[c] = dataDfTadpole[c].astype(np.float128) # increase precision of floats to 128
 
 
   #### make a tiny dataset for testing. When model is ready, run on full data ###
+  print(dataDfTadpole.shape)
+  nrModPerSubj = (~np.isnan(dataDfTadpole['FDG Temporal'])).astype(int) + (~np.isnan(dataDfTadpole['DTI FA Temporal'])).astype(int) \
+                    + (~np.isnan(dataDfTadpole['AV45 Temporal'])).astype(int) + (~np.isnan(dataDfTadpole['AV1451 Temporal'])).astype(int)
+
+  hasMriPlusOneOtherImgInd = np.logical_and((~np.isnan(dataDfTadpole['Volume Temporal'])), nrModPerSubj >= 1)
+
+  print('nrModPerSubj', nrModPerSubj)
+  # print(das)
+
+  print('dataDfTadpole s2', dataDfTadpole.shape)
+
+  idxToDrop = np.logical_not(hasMriPlusOneOtherImgInd)
+  # print('idxToDrop', np.sum(idxToDrop))
+  dataDfTadpole.drop(dataDfTadpole.index[idxToDrop], inplace=True)
+  dataDfTadpole.reset_index(drop=True, inplace=True)
+
+  # print(dataDfTadpole.shape)
+  # print(adsa)
+
+  unqRID = np.unique(dataDfTadpole.RID)
+  g1RID = np.unique(dataDfTadpole.RID[dataDfTadpole.dataset == 1])
+  g2RID = np.unique(dataDfTadpole.RID[dataDfTadpole.dataset == 2])
+  g3RID = np.unique(dataDfTadpole.RID[dataDfTadpole.dataset == 3])
+
[…]

### 2019-03-26 LinearModel.py
added linear model

@@ -1,0 +1,327 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import sys
+import scipy.optimize
+import DPMModelGeneric
+import scipy.stats
+
+
+
+class LinearModel(DPMModelGeneric.DPMModelGeneric):
+  plt.interactive(False)
+
+  def __init__(self, X, Y, visitIndices, outFolder, plotter, names_biomarkers, params):
+    super().__init__(X, Y, visitIndices, outFolder, plotter, names_biomarkers, params)
+
+    minX = np.float(np.min([el for sublist in self.X_array for item in sublist for el in item]))
+    maxX = np.float(np.max([el for sublist in self.X_array for item in sublist for el in item]))
+    self.updateMinMax(minX, maxX)
+
+    self.parameters = [0 for b in range(self.nrBiomk)]
+    self.initialiseParams()
+
+    scaledYarrayB = [self.applyScalingY(self.Y_array[b], b) for b in range(self.nrBiomk)]
+    self.min_yB = np.array([np.min(scaledYarrayB[b].reshape(-1)) for b in range(self.nrBiomk)])
+    self.max_yB = np.array([np.max(scaledYarrayB[b].reshape(-1)) for b in range(self.nrBiomk)])
+    self.priors = None
+
+  def initialiseParams(self):
+
+    avgStdX = np.nanmean([np.std(self.X_array[b]) for b in range(self.nrBiomk) if np.std(self.X_array[b] != 0)])
+    for b in range(self.nrBiomk):
+      trajParams = [1, 0]
+      variance = np.var(self.Y_array[b])
+      self.parameters[b] = [trajParams, variance]
+
+
+
+  def applyScalingX(self, x_data, biomk=0):
+    return x_data
+
+  def applyScalingY(self, y_data, biomk):
+    return y_data
+
+  def applyScalingYInv(self, y_data, biomk):
+    return y_data
+
+  def applyScalingYAllBiomk(self, biomksXB):
+    return biomksXB
+
+  def applyScalingXForward(self, x_data, biomk=0):
+    return x_data
+
+  def applyScalingXInv(self, x_data, biomk=0):
+    return x_data
+
+
+  def modelFunc(self, xs, theta):
+    return theta[0] * xs + theta[1]
+
+  def computePriorTrajOneBiomk(self, params):
+    """ params = [theta_params, variance] """
+
[…]

### 2019-03-29 paper/paper.tex
almost finished paper draft

--- 2019-03-19 paper/paper.tex
@@
 
+% We introduce Disease Knowledge Transfer (DKT), a novel technique for transferring biomarker information between related neurodegenerative diseases.
+DKT infers robust multimodal biomarker trajectories in rare neurodegenerative diseases even when only limited, unimodal data is available, by transferring information from larger multimodal datasets from common neurodegenerative diseases.
+DKT is a joint-disease generative model of biomarker progressions, which exploits biomarker relationships that are shared across diseases.
+Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects.
+For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own clinical center, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
+We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature.
+We further validate DKT in two situations: (1) on synthetic data, showing it can accurately estimate ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable predictive performance compared to other models.
+While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
+
 \begin{abstract}
@@
 Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects.
-For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own clinical center (anonymized), for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
+For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own clinical center, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
 We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature.
-We further validate DKT in two situations: (1) on synthetic data, showing that it can accurately estimate the ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable predictive performance compared to standard approaches.
+We further validate DKT in two situations: (1) on synthetic data, showing it can accurately estimate ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable predictive performance compared to other models.
 While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
@@
 These were successfully used to improve the accuracy of AD diagnosis \cite{hon2017towards} or prediction of MCI conversion \cite{cheng2015domain}, but have two key limitations.
-First, they use deep learning or other machine learning methods, which are not interpretable and don't allow us to understand underlying disease mechanisms that are either specific to rare diseases, or shared across related diseases.
+First, they use deep learning or other machine learning methods, which are not easily interpretable and don't allow us to understand underlying disease mechanisms that are either specific to rare diseases, or shared across related diseases.
 Secondly, these models cannot be used to forecast the future evolution of subjects at risk of disease, which is important for selecting the right subjects in clinical trials.
 
-We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal biomarker progressions for multiple diseases simultaneously -- including rare neurodegenerative diseases -- and which inherently performs transfer learning between the modelled phenotypes.
+We propose Disease Knowledge Transfer (DKT), a generative model that estimates continuous multimodal biomarker progressions for multiple diseases simultaneously -- including rare neurodegenerative diseases -- and which inherently performs transfer learning between the modelled phenotypes.
 This is achieved by exploiting biomarker relationships that are shared across diseases, whilst accounting for differences in the spatial distribution of brain pathology.
@@
 We first show that the estimated trajectories for PCA and tAD subjects are plausible as they agree with previous literature findings.
-We finally validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients.
+We finally validate DKT on three datasets: 1) simulated data with known ground truth, 2) TADPOLE sub-populations with different progressions and 3) 20 DTI scans from controls and PCA patients from our clinical center.
 
@@
 shared across diseases), is key to DKT.
-We make this assumption for two reasons.
-First of all, pathology in many related neurodegenerative diseases ( e.g.
+We can make this assumption for two reasons.
+First, pathology in many related neurodegenerative diseases (e.g.
 Alzheimer's variants) are hypothesised to share the same underlying mechanisms (e.g.
-amyloid and tau accumulation), and within one region, such mechanisms lead to the same pathology dynamics across all the disease variants modelled: hypometabolism, neuronal damage and region-specific cognitive decline \cite{jack2010hypothetical} -- the difference between these variants is that distinct brain regions are affected at different times and with different pathology rates and extent, likely caused by selective vulnerability of networks within these regions \cite{seeley2009neurodegenerative}.
+amyloid and tau accumulation), and within one region, such mechanisms lead to similar pathology dynamics across all the disease variants modelled: hypometabolism, neuronal damage and region-specific cognitive decline \cite{jack2010hypothetical}.
+However, the main key difference between these variants is that distinct brain regions are affected at different times and with different pathology rates and extent, likely caused by selective vulnerability of networks within these regions \cite{seeley2009neurodegenerative}.
 Secondly, even if the diseases share different upstream mechanisms (e.g.
-amyloid vs tau accumulation), downstream biomarkers measuring hypometabolism, white matter degradation and structural markers follow the same pathological cascade and will have similar dynamics across diseases.
+amyloid vs tau accumulation), downstream biomarkers measuring hypometabolism, white matter degradation and structural markers are likely to follow the same pathological cascade and will have similar dynamics across diseases.
 
@@
 
-In this section we model the biomarker dynamics that are specific to each disease, by mapping the subjects' disease stages to dysfunction scores.
+We now model the biomarker dynamics that are specific to each disease, by mapping the subjects' disease stages to dysfunction scores.
 We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the time shift of subject $i$.
@@
 We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this accounts for floor and ceiling effects present in AD biomarkers.
-Variable $\epsilon_k$ denotes the variance of measurements for biomarker $k$.
 
@@
 We model the biomarker dynamics that are disease-agnostic, by constructing the mapping from the dysfunction scores $\gamma_i^l$ to the biomarker measurements.
-We assume a set of given biomarker measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers at various visits.
+We assume a set of given biomarker measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements.
 We further denote by $\theta_k$ the trajectory parameters for biomarker $k \in K$ within its agnostic unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ maps each biomarker $k$ to a unique agnostic unit $l \in \Lambda$.
@@
 \end{equation}
-where $g(\ .\ ; \theta_k)$ represents the trajectory of biomarker $k$ within agnostic unit $\psi(k)$.
+where $g(\ .\ ; \theta_k)$ represents the trajectory of biomarker $k$ within agnostic unit $\psi(k)$, with parameters $\theta_k$.
 We implement $g$ also using a parametric sigmoidal model, as such a model can account for floor and ceiling effects.
@@
 \ref{eqDysfunctionScoreDef}, where agnostic unit $l$ is now referred to as $\psi(k)$, to clarify that this is the unit where biomarker $k$ has been allocated.
+Variable $\epsilon_k$ denotes the variance of measurements for biomarker $k$.
 
@@
 \end{equation}
-where $\boldsymbol{y} = [y_{ijk} | \forall (i,j,k) \in \Omega ]$ is the vector of all biomarker measurements, while $\boldsymbol{\theta} = [\theta_1, ..., \theta_K]$ represents the stacked parameters for the trajectories of biomarkers in agnostic units, $\boldsymbol{\lambda} = [\lambda_d^{l}|l \in \Lambda, d \in \mathbb{D}]$ are the parameters of the dysfunction trajectories within the disease models, $\boldsymbol{\beta} =[\beta_1, ..., \beta_N]$ are the subject-specific time shifts and $\boldsymbol{\epsilon} = [\epsilon_k | k \in K]$  estimates biomarker measurement noise.
-Here we assumed independence across different subjects, but the biomarker measurements and visits are linked through the latent time-shift $\beta_i$ for each subject.
-The parameters of the model that need to be estimated are $[\boldsymbol{\theta}, \boldsymbol{\lambda}, \boldsymbol{\beta}, \boldsymbol{\epsilon}]$.
+where $\boldsymbol{y} = [y_{ijk} | \forall (i,j,k) \in \Omega ]$ is the vector of all biomarker measurements, while $\boldsymbol{\theta} = [\theta_1, ..., \theta_K]$ represents the stacked parameters for the trajectories of biomarkers in agnostic units, $\boldsymbol{\lambda} = [\lambda_d^{l}|l \in \Lambda, d \in \mathbb{D}]$ are the parameters of the dysfunction trajectories within the disease models, $\boldsymbol{\beta} =[\beta_1, ..., \beta_N]$ are the subject-specific time shifts and $\boldsymbol{\epsilon} = [\epsilon_k | k \in K]$ estimates measurement noise.
 
-We estimate the model parameters using a two-stage approach.
+We estimate the model parameters $[\boldsymbol{\theta}, \boldsymbol{\lambda}, \boldsymbol{\beta}, \boldsymbol{\epsilon}]$ using a two-stage approach.
 In the first stage, we perform belief propagation within each agnostic unit and then within each disease model.
@@
 
-We first test DKT on synthetic data, to assess its performance when ground truth is known.
+We first test DKT on synthetic data, to assess its performance against known ground truth.
 More precisely, we generate data that follows the DKT model exactly, and test DKT's ability to recover biomarker trajectories and subject time-shifts.
-We generate synthetic data from two diseases (i.e.
-\emph{synthetic PCA} and \emph{synthetic AD}) using parameters from Table \ref{tab:synParams}, emulating the TADPOLE cohort as well as our own cohort.
+We generate synthetic data from two diseases (50 subjects with "\emph{synthetic PCA}" and 100 subjects with "\emph{synthetic AD}") using the parameters from the bottom-left table in Fig.
+\ref{fig:dktSynthTrajCompTrue}, emulating the TADPOLE cohort as well as our own cohort.
 The six biomarkers ($k_1$-$k_6$) have been \emph{a-priori} allocated to two agnostic units $l_0$ and $l_1$.
 To simulate the lack of multimodal data in the synthetic PCA subjects, we discarded the data from biomarkers $k_0$, $k_1$, $k_4$ and $k_5$ for all these subjects.
-Remaining biomarkers $k_2$ and $k_3$, for which data was still available in the synthetic PCA cohort, are assumed to be of the same modality (e.g.
+
+% Remaining biomarkers $k_2$ and $k_3$, for which data was still available in the synthetic PCA cohort, are assumed to be of the same modality (e.g.
 MRI volume) but to represent measurements from different brain regions (e.g.
@@
 
- \begin{table}
-\fontsize{6}{8}\selectfont
- \centering
- \begin{tabular}{c | c}
-& \textbf{Trajectory parameters} \\
-%\hline
- Biomarker allocation &  $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$ \\
-Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$,  $\theta_4 = (1,5,0.9,0)$  \\
-Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$, $\theta_5 = (1,10,0.9,0)$ \\
-
-"Synthetic AD" & $\lambda_0^0 = (1, 0.3, -4, 0)$  and $\lambda_0^1 = (1, 0.2, 6, 0)$ \\
- "Synthetic PCA" & $\lambda_1^0 = (1, 0.3, 6, 0)$ and $\lambda_1^1 = (1, 0.2, -4, 0)$ \\
-\hline
-& \textbf{Subject parameters} \\
- Number of subjects & 100 (synthetic AD) and 50 (synthetic PCA) \\ 
- Time-shifts $\beta_i$ & $\beta_i \sim U(-13,10)$ years \\
- Diagnosis & $p(control) \propto Exp (-4.5)$,  $p(patient) \propto Exp (4.5)$\\
- Data generation & 4 visits/subject, 1 year apart, $\epsilon_k = 0.05$\\ 
-\end{tabular}
-\caption{Parameters used for synthetic data generation, emulating the TADPOLE cohort and our own dataset.}
-\label{tab:synParams}
-\end{table}
+%  \begin{table}
+% \fontsize{6}{8}\selectfont
+%  \centering
+%  \begin{tabular}{c | c}
+% & \textbf{Trajectory parameters} \\
+% %\hline
+%  Biomarker allocation &  $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$ \\
+% Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$,  $\theta_4 = (1,5,0.9,0)$  \\
+% Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$, $\theta_5 = (1,10,0.9,0)$ \\
+%   
+% "Synthetic AD" & $\lambda_0^0 = (1, 0.3, -4, 0)$  and $\lambda_0^1 = (1, 0.2, 6, 0)$ \\
+%  "Synthetic PCA" & $\lambda_1^0 = (1, 0.3, 6, 0)$ and $\lambda_1^1 = (1, 0.2, -4, 0)$ \\
+% \hline
+% & \textbf{Subject parameters} \\
+%  Number of subjects & 100 (synthetic AD) and 50 (synthetic PCA) \\ 
+%  Time-shifts $\beta_i$ & $\beta_i \sim U(-13,10)$ years \\
+%  Diagnosis & $p(control) \propto Exp (-4.5)$,  $p(patient) \propto Exp (4.5)$\\
+%  Data generation & 4 visits/subject, 1 year apart, $\epsilon_k = 0.05$\\ 
+% \end{tabular}
+% \caption{Parameters used for synthetic data generation, emulating the TADPOLE cohort and our own dataset.}
+% \label{tab:synParams}
+% \end{table}
 
@@
 Finally, we normalised the biomarker values to lie within the [0,1] range.
-For each lobe, we allocated all multimodal biomarkers corresponding to that lobe to its own agnostic unit.
+
+% For each lobe, we allocated all multimodal biomarkers corresponding to that lobe to its own agnostic unit.
 
@@
 \includegraphics[width=\textwidth]{compTrueParams101_synth1_JMD.pdf}
+\begin{flushright}
+\vspace{-8em}
+\fontsize{5}{8.5}\selectfont
+ \begin{tabular}{c | c}
+% & \textbf{Trajectory parameters} \\
+%\hline
+ Biomarker alloc. &  $l_0:\{k_0, k_2, k_4\}$ $l_1: \{k_1, k_3, k_5\}$ \\
+Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$ $\theta_2 = (1,5,0.55,0)$\\
+& $\theta_4 = (1,5,0.9,0)$  \\
+Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$ $\theta_3 = (1,10,0.55,0)$\\
+& $\theta_5 = (1,10,0.9,0)$ \\
+
+Synthetic AD & $\lambda_0^0 = (1, 0.3, -4, 0)$  $\lambda_0^1 = (1, 0.2, 6, 0)$ \\
+ Synthetic PCA & $\lambda_1^0 = (1, 0.3, 6, 0)$ $\lambda_1^1 = (1, 0.2, -4, 0)$ \\
+% \hline
+% & \textbf{Subject parameters} \\
+%  Number of subjects & 100 (synthetic AD) and 50 (synthetic PCA) \\ 
+%  Time-shifts $\beta_i$ & $\beta_i \sim U(-13,10)$ years \\
+%  Diagnosis & $p(control) \propto Exp (-4.5)$,  $p(patient) \propto Exp (4.5)$\\
+%  Data generation & 4 visits/subject, 1 year apart, $\epsilon_k = 0.05$\\ 
+\end{tabular}
+\end{flushright}
 \caption[DKT Simulation Results - Comparison between true and DKT-estimated biomarker trajectories and subject time-shifts.]{Comparison between true and DKT-estimated subject time-shifts and biomarker trajectories.
@@
 We also show the DKT-estimated and true trajectories of the agnostic units within the 'synthetic AD' disease (top-right, "Dis0") and the 'synthetic PCA' disease (middle-left, "Dis1").
-Finally, we also show the biomarker trajectories within unit 0 (middle-right) and unit 1 (bottom).}
+Finally, we also show the biomarker trajectories within unit 0 (middle-right) and unit 1 (bottom-left).
+Parameters used for data generation are shown in the bottom-right table.}
   \label{fig:dktSynthTrajCompTrue}
@@
 In Fig.
-\ref{fig:PCAtrajByModality}, we plot the inferred trajectories for PCA directly across the disease progression, for all five modalities analysed.
-The results recapitulate known patterns in PCA, where posterior regions are predominantly affected.
+\ref{fig:PCAtrajByModality}, we plot the inferred trajectories for PCA directly across the disease progression.
+The results recapitulate known patterns in PCA, where posterior regions are predominantly affected, and shows that DKT can predict plausible non-MRI biomarkers for PCA in lack of such data for these subjects.
 
@@
 
-\subsection{Validation on DTI Data in PCA}
+\subsection{Validation on DTI Data in tAD and PCA}
 \label{sec:dktResVal}
 
-We validated our model using a separate test set of 20 DTI scans from controls and PCA patients from our own cohort.
-We used DKT to predict the DTI biomarker values for the subjects within the unseen test set, using only their MRI biomarkers.
-To evaluate prediction accuracy, we computed the prediction mean squared error (MSE) and rank correlation between the DKT-predicted biomarker values and the measured values in PCA patients.
+We validated DKT by predicting unseen DTI data from two datasets: 1) TADPOLE subjects with a different progression from the training subjects, and 2) a separate test set of 20 DTI scans from controls and PCA patients from our own cohort.
+To split TADPOLE into subgroups with different progression, we used the SuStaIn model by \cite{young2018uncovering}, which resulted into three subgroups: hippocampal, cortical and subcortical, with prominent early atrophy in those respective regions.
+This allows us to validate DKT by transferring multimodal data between subgroups with different progressions.
+To evaluate prediction accuracy, we computed the rank correlation between the DKT-predicted biomarker values and the measured values in the test data.
 
@@
 
-The results from Table \ref{sec:dktPerfMetrics} indicate that the DKT prediction errors are relatively small for the frontal, occipital, temporal and parietal areas.
-Moreover, the rank correlations of the DKT predicted values are also high for the cingulate lobe and the hippocampus.
-In terms of model comparison, DKT has better performance than the linear model (all results significant with $p < 0.002$, Bonferroni corrected), since it uses information across all brain regions instead of assuming independence across regions.
-While DKT has similar performance to the latent stage model, it allows us to understand mechanisms that are shared between related diseases.
+Validation results are shown in Table \ref{sec:dktPerfMetrics}, for TADPOLE subgroups (hippocampal to cortical, other combinations of subgroups not shown due to lack of space) as well as PCA subjects.
+When predicting missing DTI markers of the TADPOLE cortical subgroup as well as PCA subjects from our local cohort (Table \ref{sec:dktPerfMetrics}), the DKT correlations are generally high for the cingulate, hippocampus and parietal, and lower for the frontal lobe.
+DKT also shows favourable performance to four other models: the latent-stage model from \cite{jedynak2012computational}, a multivariate non-linear model that predicts a DTI ROI marker from multiple MRI markers from different ROIs, as well as spline and linear models that predict DTI biomarkers directly from the MRI marker corresponding to the same ROI.
+In particular for predicting DTI FA in the parietal and temporal lobes, DKT scores has significantly better predictions that almost all methods tested.
 
@@
 \centering
-\fontsize{8}{10}\selectfont
+\fontsize{7}{10}\selectfont
 \begin{tabular}{c | c c c c c c}
-\textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippo.} & \textbf{Occip.} & \textbf{Parietal} & \textbf{Temporal}\\
-& \multicolumn{6}{c}{\textbf{Prediction Error (MSE)}}\\
-DKT & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.18$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
-Latent stage & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.17$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
-Linear & 0.05$\pm$0.02* & 0.15$\pm$0.04* & 0.09$\pm$0.03* & 0.07$\pm$0.03* & 0.07$\pm$0.02* & 0.07$\pm$0.02*\\
-& \multicolumn{6}{c}{\textbf{Rank Correlation (Spearman rho)}}\\
-DKT  & 0.76  & 0.48  & 0.76  & 0.55  & 0.55  & 0.33 \\
-Latent stage & 0.76  & 0.49  & 0.80*  & 0.56  & 0.51*  & 0.33 \\
-Linear & 0.48*  & 0.31*  & 0.64*  & 0.61*  & 0.57*  & 0.27* \\
+\textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippocam.} & \textbf{Occipital} & \textbf{Parietal} & \textbf{Temporal}\\
+& \multicolumn{6}{c}{\textbf{TADPOLE subgroups: Hippocampal subgroup to Cortical subgroup}}\\
+DKT (ours) &      0.56 $\pm$ 0.23 &    \textbf{0.35 $\pm$ 0.17} &        \textbf{0.58 $\pm$ 0.14} &     -0.10 $\pm$ 0.29 &     \textbf{0.71 $\pm$ 0.11} &     \textbf{0.34 $\pm$ 0.26} \\
+Latent stage &      0.44 $\pm$ 0.25 &    0.34 $\pm$ 0.21 &       0.34 $\pm$ 0.24* &     \textbf{-0.07 $\pm$ 0.22} &     0.64 $\pm$ 0.16 &    0.08 $\pm$ 0.24* \\
+Multivariate &      \textbf{0.60 $\pm$ 0.18} &   0.11 $\pm$ 0.22* &       0.12 $\pm$ 0.29* &     -0.22 $\pm$ 0.22 &   -0.44 $\pm$ 0.14* &   -0.32 $\pm$ 0.29* \\
+Spline &    -0.24 $\pm$ 0.25* &  -0.06 $\pm$ 0.27* &        0.58 $\pm$ 0.17 &     -0.16 $\pm$ 0.27 &    0.23 $\pm$ 0.25* &    0.10 $\pm$ 0.25* \\
+Linear &    -0.24 $\pm$ 0.25* &   0.20 $\pm$ 0.25* &        0.58 $\pm$ 0.17 &     -0.16 $\pm$ 0.27 &    0.23 $\pm$ 0.25* &    0.13 $\pm$ 0.23* \\
+& \multicolumn{6}{c}{\textbf{typical Alzheimer's to Posterior Cortical Atrophy}}\\
+DKT (ours) &    0.77 $\pm$ 0.11 &    0.39 $\pm$ 0.26 &      0.75 $\pm$ 0.09 &    0.60 $\pm$ 0.14 &    \textbf{0.55 $\pm$ 0.24} &    \textbf{0.35 $\pm$ 0.22} \\
+Latent stage &    \textbf{0.80 $\pm$ 0.09} &    \textbf{0.53 $\pm$ 0.17} &      \textbf{0.80 $\pm$ 0.12} &    0.56 $\pm$ 0.18 &    0.50 $\pm$ 0.21 &    0.32 $\pm$ 0.24 \\
+Multivariate &   0.44 $\pm$ 0.19* &  -0.21 $\pm$ 0.25* &    -0.53 $\pm$ 0.26* &  -0.39 $\pm$ 0.20* &  -0.27 $\pm$ 0.28* &  -0.13 $\pm$ 0.26* \\
+Spline &   0.52 $\pm$ 0.20* &  -0.03 $\pm$ 0.35* &     0.66 $\pm$ 0.11* &   0.09 $\pm$ 0.25* &    0.53 $\pm$ 0.20 &   0.30 $\pm$ 0.21* \\
+Linear &   0.52 $\pm$ 0.20* &    0.34 $\pm$ 0.27 &     0.66 $\pm$ 0.11* &    \textbf{0.64 $\pm$ 0.17} &    0.54 $\pm$ 0.22 &   0.30 $\pm$ 0.21* \\
 \end{tabular}
 % \normalfont
-\caption[Performance evaluation of DKT and two simpler models]{Performance evaluation of DKT and two simpler models.
-(*) Statistically significant difference in the performance of DKT vs latent-stage/linear models, based on a two-tailed t-test, Bonferroni corrected.}
+\caption[Performance evaluation of DKT and other models]{Performance evaluation of DKT and four other statistical models of decreasing complexity.
+We show the rank correlation between predicted biomarkers and measured biomarkers in (top) TADPOLE subgroups -- transfer from hippocampal subgroup to cortical subgroup -- and (bottom) PCA.
+(*) Statistically significant difference in the performance of DKT vs the other models, based on a two-tailed t-test, Bonferroni corrected.}
 % \end{footnotesize}
@@
 
-We presented DKT, a framework that enables, for the first time, joint modelling of biomarker progressions in multiple neurodegenerative diseases simultaneously.
+We presented DKT, a framework that enables joint modelling of biomarker progressions in multiple neurodegenerative diseases simultaneously.
+Our work has however several limitations, which can be addressed in future research: 1) to account for population heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) DKT can be further validated on more complex synthetic experiments with a range of datasets generated with different parameters.
 
@@
 
-Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and our local centre and 4) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
-
-There are several potential avenues for further research: 1) to account for heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3-4) DKT can be further validated on more complex synthetic experiments with a range of parameter settings, as well as on patient data from ADNI, where the population could be \emph{a-priori} split into sub-groups with different progressions.
+% Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and our local centre and 4) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
 

### 2019-03-29 JointModel.py

@@ -119,24 +119,24 @@ class JointModel(DisProgBuilder.DPMInterface):
         self.loadCheckpoint(i-1, 5)
 
       # estimate subject latent variables
-      # self.estimSubjShifts(self.unitModels, self.disModels)
-      # self.makePlots(plotFigs, i, 0)
-      # self.saveCheckpoint(i, 0)
+      self.estimSubjShifts(self.unitModels, self.disModels)
+      self.makePlots(plotFigs, i, 0)
+      self.saveCheckpoint(i, 0)
       # self.loadCheckpoint(i, 0)
 
       while i < nrIt:
 
         # estimate biomk trajectories - disease agnostic
-        # self.estimBiomkTraj(self.unitModels, self.disModels, i)
-        # self.makePlots(plotFigs, i, 1)
-        # self.saveCheckpoint(i, 1)
-        # # self.loadCheckpoint(i, 1)
-        #
-        # # estimate subject latent variables
-        # self.estimSubjShifts(self.unitModels, self.disModels)
-        # self.makePlots(plotFigs, i, 2)
-        # self.saveCheckpoint(i, 2)
-        self.loadCheckpoint(i, 2)
+        self.estimBiomkTraj(self.unitModels, self.disModels, i)
+        self.makePlots(plotFigs, i, 1)
+        self.saveCheckpoint(i, 1)
+        # self.loadCheckpoint(i, 1)
+
+        # estimate subject latent variables
+        self.estimSubjShifts(self.unitModels, self.disModels)
+        self.makePlots(plotFigs, i, 2)
+        self.saveCheckpoint(i, 2)
+        # self.loadCheckpoint(i, 2)
         #
         # # estimate unit trajectories - disease specific
         self.estimTrajWithinDisModel(self.unitModels, self.disModels)
@@ -172,7 +172,7 @@ class JointModel(DisProgBuilder.DPMInterface):
 
 
       # for SuStaIn validation
-      self.loadCheckpoint(3, 1)  # for the real data
+      # self.loadCheckpoint(3, 1)  # for the real data
       # self.makePlots(plotFigs, i, 2)
       # print(adsa)
 
@@ -194,7 +194,8 @@ class JointModel(DisProgBuilder.DPMInterface):
     for d in range(self.nrDis):
       self.disModels[d].priors = self.priorsDisModels[d]
       self.disModels[d].plotter.plotTrajParams['allTrajOverlap'] = self.params['plotTrajParams']['allTrajOverlap']
-
[…]

### 2019-03-29 JointModelOnePass.py

@@ -109,12 +109,12 @@ class JDMOnePass(DisProgBuilder.DPMInterface):
     disModelsFile = '%s/disModels.npz' % self.outFolder
     nrSubj = self.unitModels[0].nrSubj
 
-    for s in range(len(self.params['X'][0])):
-      entriesCurrSubj = [self.params['X'][b][s].shape[0] > 0 for b in range(30)]
-      nrEntriesPerSubj = np.sum(entriesCurrSubj)
-      if nrEntriesPerSubj == 0:
-        print(s, entriesCurrSubj)
-        print(dadsa)
+    # for s in range(len(self.params['X'][0])):
+    #   entriesCurrSubj = [self.params['X'][b][s].shape[0] > 0 for b in range(30)]
+    #   nrEntriesPerSubj = np.sum(entriesCurrSubj)
+    #   if nrEntriesPerSubj == 0:
+    #     print(s, entriesCurrSubj)
+    #     print(dadsa)
 
     # print(labels)
     # print(dasda)
@@ -223,6 +223,9 @@ class JDMOnePass(DisProgBuilder.DPMInterface):
         self.disModels[disNr] = self.disModelObj(xDysfunSubjCurrDisUSX, dysfuncScoresCurrDisUSX, visitIndicesCurrDisUSX,
           outFolderCurDis, plotterCurrDis, plotTrajParamsDis['labels'], self.params)
         self.disModels[disNr].priors = self.priorsDisModels[disNr]
+        # manually set limits on Y axis for the disease models
+        self.disModels[disNr].min_yB = [0 for b in range(self.disModels[disNr].nrBiomk)]
+        self.disModels[disNr].min_yB = [1 for b in range(self.disModels[disNr].nrBiomk)]
         self.disModels[disNr].Optimize(nrGlobIterDis, Plot=True)
 
         XshiftedScaledDBS, XdisDBSX, _, _ = self.disModels[disNr].getData(flagAllBiomkShouldBePresent=False)

### 2019-03-29 Plotter.py

@@ -178,6 +178,7 @@ class PlotterJDM:
       trajStruct = plotterDis.getTrajStruct(disModels[d])
       nrPlotsSoFar = plotterDis.subplotAllTrajWithModelData(disModels[d], trajStruct, nrPlotsSoFar, nrRows, nrCols)
 
+
     ######### compare biomk traj within functional units ##########
 
     for f in range(nrFuncUnits):
@@ -185,8 +186,8 @@ class PlotterJDM:
       trajStruct = plotterFunc.getTrajStruct(unitModels[f])
       nrPlotsSoFar = plotterFunc.subplotAllTrajWithModelData(unitModels[f], trajStruct, nrPlotsSoFar, nrRows, nrCols)
 
-    pl.tight_layout(pad=self.plotTrajParams['padTightLayout'])
-    pl.subplots_adjust(wspace=0, hspace=0)
+    pl.tight_layout(pad=self.plotTrajParams['padTightLayout'], w_pad=0.5, h_pad=1.0)
+    pl.subplots_adjust(wspace=0.3, hspace=0.6)
 
     if replaceFig:
       fig.show()

### 2019-03-29 auxFunc.py

@@ -208,6 +208,14 @@ def getGammShapeRateFromTranTime(transitionTimePriorMean, transitionTimePriorMin
 
   return bPriorShape, bPriorRate
 
+def getMeanStdBFromTranTime(transitionTimePriorMean, transitionTimePriorMin, transitionTimePriorMax):
+
+  bPriorMean = 16 / (1 * transitionTimePriorMean)
+  bPriorStd = np.abs(16 / (1 * transitionTimePriorMax) - 16 / (1 * transitionTimePriorMin))
+
+
+  return bPriorMean, bPriorStd
+
 def applyScalingToBiomk(dataCrossSB, scalingBiomk2B):
   scaledData = dataCrossSB * scalingBiomk2B[1,:][None, :] + scalingBiomk2B[0,:][None, :]
   return scaledData

### 2019-03-29 drcValidFuncs.py

@@ -313,16 +313,16 @@ def validateDRCBiomk(dpmObj, params):
 
   labels = params['labels']
   print('labels', labels)
-  dtiBiomksList = [i for i in range(len(labels)) if labels[i].startswith('DTI')]
+  nonMriBiomksList = [i for i in range(len(labels)) if labels[i].startswith('DTI')]
   mriBiomksList = [i for i in range(len(labels)) if labels[i].startswith('Volume')]
 
   assert len(ysPredBS) == len(Yvalid)
 
-  nrDtiCols = len(dtiBiomksList)
-  mse = [0 for b in dtiBiomksList]
+  nrDtiCols = len(nonMriBiomksList)
+  mse = [0 for b in nonMriBiomksList]
 
   # subjects who have DTI validation
-  subjWithValidIndx = np.where([ys.shape[0] > 0 for ys in Yvalid[dtiBiomksList[0]]])[0]
+  subjWithValidIndx = np.where([ys.shape[0] > 0 for ys in Yvalid[nonMriBiomksList[0]]])[0]
   nrSubjWithValid = subjWithValidIndx.shape[0]
   XvalidFilt, YvalidFilt = DPMModelGeneric.DPMModelGeneric.filterXYsubjInd(Xvalid, Yvalid, subjWithValidIndx)
   diagValidFilt = diagValid[subjWithValidIndx]
@@ -339,19 +339,19 @@ def validateDRCBiomk(dpmObj, params):
       idxCurrDis = np.where(RIDvalidFilt[s] == ridCurrDis)[0][0]
       xsOrigFromModel = xsOrigPred1S[idxCurrDis]
 
-      assert np.where(xsOrigFromModel == XvalidFilt[dtiBiomksList[b2]][s][0])[0].shape[0] == 1
-      idxXsWithValid = np.where(xsOrigFromModel == XvalidFilt[dtiBiomksList[b2]][s][0])[0][0]
-      ysPredCurrSubj = ysPredBS[dtiBiomksList[b2]][idxCurrDis][idxXsWithValid]
+      assert np.where(xsOrigFromModel == XvalidFilt[nonMriBiomksList[b2]][s][0])[0].shape[0] == 1
+      idxXsWithValid = np.where(xsOrigFromModel == XvalidFilt[nonMriBiomksList[b2]][s][0])[0][0]
[…]

### 2019-03-29 hierSynth.py

@@ -219,8 +219,8 @@ def main():
         XemptyListsAllBiomk[b][s] = paramsDisTwo['Xtrue'][b][s]
         YemptyListsAllBiomk[b][s] = paramsDisTwo['Ytrue'][b][s]
       else:
-        XemptyListsAllBiomk[b][s] = []
-        YemptyListsAllBiomk[b][s] = []
+        XemptyListsAllBiomk[b][s] = np.array([])
+        YemptyListsAllBiomk[b][s] = np.array([])
 
   paramsDisTwo['XemptyListsAllBiomk'] = XemptyListsAllBiomk
   paramsDisTwo['YemptyListsAllBiomk'] = YemptyListsAllBiomk

### 2019-03-29 tadpoleDrc.py

@@ -111,7 +111,7 @@ plotTrajParams['diagLabels'] = {CTL:'CTL ADNI', MCI:'MCI ADNI', AD:'tAD ADNI',
 # plotTrajParams['freesurfPath'] = freesurfPath
 # plotTrajParams['blenderPath'] = blenderPath
 plotTrajParams['isSynth'] = False
-plotTrajParams['padTightLayout'] = 0.0
+plotTrajParams['padTightLayout'] = 0.4
 
 if args.agg:
   plotTrajParams['agg'] = True
@@ -125,7 +125,9 @@ else: #if hostName == 'razvan-Precision-T1700':
   height = 450
 
 
-
+pd.set_option('display.max_rows', 100)
+pd.set_option('display.max_columns', 50)
+pd.set_option('display.width', 5000)
 
 
 def visDataHist(dataDfAll):
@@ -329,19 +331,34 @@ def main():
   bPriorShape, bPriorRate = getGammShapeRateFromTranTime(
     transitionTimePriorMean, transitionTimePriorMin, transitionTimePriorMax)
 
-  transitionTimePriorMeanDis = 300 # in DPS 0-1 space, prior mean
-  transitionTimePriorMinDis = 50
-  transitionTimePriorMaxDis = 500
+  transitionTimePriorMeanAD = 0.1 # using months instead of years
+  transitionTimePriorMinAD = 0.09
+  transitionTimePriorMaxAD = 0.11
 
-  bPriorShapeDis, bPriorRateDis = getGammShapeRateFromTranTime(
-    transitionTimePriorMean, transitionTimePriorMin, transitionTimePriorMax)
+  bPriorShapeDisAD, bPriorRateDisAD = getGammShapeRateFromTranTime(
+    transitionTimePriorMeanAD, transitionTimePriorMinAD, transitionTimePriorMaxAD)
 
-  # print('bPriorShape', bPriorShape)
-  # print('bPriorRate', bPriorRate)
-  # print(adsas)
+  _, bPriorStdAD = getMeanStdBFromTranTime(
+    transitionTimePriorMeanAD, transitionTimePriorMinAD, transitionTimePriorMaxAD)
 
-  params['priorsDisModels'] = [dict(meanA=1, stdA=1e-20, meanD=0, stdD=1e-20,
[…]

### 2019-03-29 tadpoleSubtypes.py

@@ -767,8 +767,8 @@ def printRes(modelNames, res, plotTrajParams, params):
 
     disNrsValid = [d2 for d2 in range(nrDis) if d2 not in [d]]
     biomkNames = res[0]['metrics'][d][disNrsValid[0]]['labelsNonMri']
-
-    resDf = pd.DataFrame(index=range(24), columns=['Model'] + biomkNames)
+    nrBiomk = len(biomkNames)
+    resDf = pd.DataFrame(index=range(12 * len(disNrsValid)), columns=['Model'] + biomkNames)
 
     c = 0
 
@@ -833,7 +833,7 @@ def printRes(modelNames, res, plotTrajParams, params):
       corrStdMU[multivarIndex] = np.nanstd(corrMUB[multivarIndex], axis=1)
 
       # Perform Bonferroni correction
-      sigLevel = 0.05/(6*2*2)
+      sigLevel = 0.05/(6*2*nrModels)
 
       # print('mseMUB[dktIndex][u, :]', np.nanmean(mseMUB[dktIndex][1, :]), mseMUB[dktIndex][1, :])
       # print('mseMUB[linIndex][u, :]', np.nanmean(mseMUB[linIndex][1, :]), mseMUB[linIndex][1, :])

### 2019-03-29 paper/paper.tex
almost finished paper draft

@@
 DKT is a joint-disease generative model of biomarker progressions, which exploits biomarker relationships that are shared across diseases.
-Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects.
+Our proposed method allows, for the first time, the estimation of plausible \emph{multimodal} biomarker trajectories in Posterior Cortical Atrophy (PCA), a rare neurodegenerative disease where only unimodal MRI data is available.
 For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own clinical center, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
-We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature.
-We further validate DKT in two situations: (1) on synthetic data, showing it can accurately estimate ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable predictive performance compared to other models.
+While validation is challenging due to lack of data in PCA, we validate DKT on synthetic data and two patient datasets (TADPOLE and PCA cohorts), showing it can estimate the ground truth parameters in the simulation and predict unseen biomarkers on the two patient datasets.
 While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
 
-\keywords{Disease Progression Model, Transfer Learning, Manifold Learning, Alzheimer's Disease, Posterior Cortical Atrophy}
+% \keywords{Disease Progression Model, Transfer Learning, Manifold Learning, Alzheimer's Disease, Posterior Cortical Atrophy}
 \end{abstract}
@@
 The estimation of accurate biomarker signatures in Alzheimer's disease (AD) and related neurodegenerative diseases is crucial for understanding underlying disease mechanisms, predicting subjects' progressions, and enrichment in clinical trials.
-Recently, data-driven disease progression models were proposed to reconstruct long term biomarker signatures from collections of short term individual measurements \cite{lorenzi2017disease,oxtoby2018,jedynak2012computational}.
+Recently, data-driven disease progression models were proposed to reconstruct long term biomarker signatures from collections of short term individual measurements \cite{oxtoby2018,jedynak2012computational}.
 When applied to large datasets of typical AD, disease progression models have shown important benefits in understanding the earliest events in the AD cascade \cite{oxtoby2018}, quantifying biomarkers' heterogeneity \cite{young2018uncovering} and they showed improved predictions over standard approaches \cite{oxtoby2018}.
@@
 This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from our own center.
-We first show that the estimated trajectories for PCA and tAD subjects are plausible as they agree with previous literature findings.
 We finally validate DKT on three datasets: 1) simulated data with known ground truth, 2) TADPOLE sub-populations with different progressions and 3) 20 DTI scans from controls and PCA patients from our clinical center.
@@
 We assume that each disease can be modelled as the evolution of abstract dysfunction scores (Y-axis, top row), each one related to different brain regions.
-Each region-specific dysfunction score then further models (X-axis, bottom row) the progression of several modality-specific biomarkers within that same region.
+Each region-specific dysfunction score then further models (X-axis, bottom row) the progression of several multimodal biomarkers within that same region.
 For instance, the temporal dysfunction, modelled as a biomarker in the disease specific model (top row), is the X-axis in the disease agnostic model (temporal unit, bottom row), which aggregates together abnormality from amyloid, tau and MR imaging within the temporal lobe.
@@
 Fig.
-\ref{fig:diagram} shows the overall diagram of the DKT framework.
-We assume that the progression of each disease (X-axis, top row) can be modelled as a unique evolution of dysfunction trajectories (top row) representing multimodal pathology within a specific brain region.
-Each dysfunction trajectory is modelled as the progression of several biomarkers within that same region, but acquired using different types of modalities (e.g.
-MRI, PET or DTI, see Fig.
-\ref{fig:diagram} bottom row).
-Each group of biomarkers in the bottom row will be called a \emph{disease-agnostic unit} or simply \emph{agnostic unit}, because the biomarker dynamics are assumed to be shared across all diseases modelled.
+\ref{fig:diagram} shows the diagram of the DKT framework.
+We assume that the progression of each disease can be modelled as a unique evolution of dysfunction trajectories representing region-specific multimodal pathology, further modelled as the progression of several biomarkers within that same region, but acquired using different modalities (Fig.
+\ref{fig:diagram} bottom).
+Each group of biomarkers in the bottom row will be called a \emph{disease-agnostic unit} or simply \emph{agnostic unit}, because biomarker dynamics here are assumed to be shared across all diseases modelled.
 
@@
 Alzheimer's variants) are hypothesised to share the same underlying mechanisms (e.g.
-amyloid and tau accumulation), and within one region, such mechanisms lead to similar pathology dynamics across all the disease variants modelled: hypometabolism, neuronal damage and region-specific cognitive decline \cite{jack2010hypothetical}.
-However, the main key difference between these variants is that distinct brain regions are affected at different times and with different pathology rates and extent, likely caused by selective vulnerability of networks within these regions \cite{seeley2009neurodegenerative}.
+amyloid and tau accumulation), and within one region, such mechanisms lead to similar pathology dynamics across all the disease variants modelled \cite{jack2010hypothetical}, with the key difference that distinct brain regions are affected at different times and with different pathology rates and extent, likely caused by selective vulnerability of networks within these regions \cite{seeley2009neurodegenerative}.
 Secondly, even if the diseases share different upstream mechanisms (e.g.
@@
 
-We model the biomarker dynamics that are disease-agnostic, by constructing the mapping from the dysfunction scores $\gamma_i^l$ to the biomarker measurements.
-We assume a set of given biomarker measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements.
+We further model the biomarker dynamics that are disease-agnostic, by constructing the mapping from the dysfunction scores $\gamma_i^l$ to the biomarker measurements.
+We assume a set of given biomarker measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is the set of available biomarker measurements.
 We further denote by $\theta_k$ the trajectory parameters for biomarker $k \in K$ within its agnostic unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ maps each biomarker $k$ to a unique agnostic unit $l \in \Lambda$.
@@
 \end{equation}
-where $g(\ .\ ; \theta_k)$ represents the trajectory of biomarker $k$ within agnostic unit $\psi(k)$, with parameters $\theta_k$.
-We implement $g$ also using a parametric sigmoidal model, as such a model can account for floor and ceiling effects.
+where $g(\ .\ ; \theta_k)$ represents the trajectory of biomarker $k$ within agnostic unit $\psi(k)$, with parameters $\theta_k$, and is again implemented using a sigmoidal function.
 Parameters $\lp$ are used to define $\gamma_i^{\psi(k)}$ based on Eq.
-\ref{eqDysfunctionScoreDef}, where agnostic unit $l$ is now referred to as $\psi(k)$, to clarify that this is the unit where biomarker $k$ has been allocated.
+\ref{eqDysfunctionScoreDef}, where agnostic unit $l$ is now referred to as $\psi(k)$, to clarify this is the unit where biomarker $k$ has been allocated.
 Variable $\epsilon_k$ denotes the variance of measurements for biomarker $k$.
@@
 We trained DKT on ADNI data from the TADPOLE challenge \cite{marinescu2018tadpole}, since it contained a large number of multimodal biomarkers already pre-processed and aggregated into one table.
-From the TADPOLE dataset we selected a subset of 230 subjects which had at least one FDG PET, AV45, AV1451 or DTI scan.
-Most subjects also had MRI scans and cognitive tests.
+From the TADPOLE dataset we selected a subset of 230 subjects which had an MRI scan and at least one FDG PET, AV45, AV1451 or DTI scan.
 In order to model another disease, we further included MRI scans from 76 PCA subjects from our own cohort, along with scans from 67 tAD and 87 age-matched controls.
 
-For both datasets, we computed multimodal biomarker measurements corresponding to each brain lobe: MRI volumes using the Freesurfer software, FDG-, AV45- and AV1451-PET standardised uptake value ratios (SUVR) extracted with the standard ADNI pipeline, and DTI fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe.
+For both datasets, we computed multimodal biomarker measurements corresponding to each brain lobe: MRI volumes using the Freesurfer software, FDG-, AV45- and AV1451-PET standardised uptake value ratios (SUVR) extracted with the standard ADNI pipeline, and DTI fractional anisotropy (FA) measures from adjacent white-matter regions.
 For every lobe, we regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs our own center).
-Finally, we normalised the biomarker values to lie within the [0,1] range.
+Finally, biomarkers were normalized to the [0,1] range.
 
@@
 Moreover, the subject time-shifts are very close ($R^2$ $>$ 0.98) to the true time-shifts.
+While some errors in trajectory estimation can be noticed, these are due to the informed priors on the model parameters in order to ensure identifiability and convergence of parameters.
 
@@
 %\includegraphics[width=\textwidth]{../resfiles/synth/synth1_JMD/compTrueParams101_synth1_JMD.pdf}
-\includegraphics[width=\textwidth]{compTrueParams101_synth1_JMD.pdf}
+\includegraphics[width=0.8\textwidth]{../resfiles/synth/synth1_JMD/compTrueParams105_synth1_JMD.pdf}
 \begin{flushright}
-\vspace{-8em}
-\fontsize{5}{8.5}\selectfont
- \begin{tabular}{c | c}
+\vspace{-16.0em}
+\fontsize{5}{8.0}\selectfont
+ \begin{tabular}{c}
 % & \textbf{Trajectory parameters} \\
 %\hline
- Biomarker alloc. &  $l_0:\{k_0, k_2, k_4\}$ $l_1: \{k_1, k_3, k_5\}$ \\
-Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$ $\theta_2 = (1,5,0.55,0)$\\
-& $\theta_4 = (1,5,0.9,0)$  \\
-Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$ $\theta_3 = (1,10,0.55,0)$\\
-& $\theta_5 = (1,10,0.9,0)$ \\
-
-Synthetic AD & $\lambda_0^0 = (1, 0.3, -4, 0)$  $\lambda_0^1 = (1, 0.2, 6, 0)$ \\
- Synthetic PCA & $\lambda_1^0 = (1, 0.3, 6, 0)$ $\lambda_1^1 = (1, 0.2, -4, 0)$ \\
+ Biomarker allocation:\\
+ $l_0:\{k_0, k_2, k_4\}$ \\
+ $l_1: \{k_1, k_3, k_5\}$ \\
+\hline
+Agnostic unit $l_0$:\\
+$\theta_0 = (1,5,0.20,0)$\\
+$\theta_2 = (1,5,0.55,0)$\\
+$\theta_4 = (1,5,0.90,0)$ \\
+\hline
+Agnostic unit $l_1$:\\
+$\theta_1 = (1,10,0.20,0)$\\
+$\theta_3 = (1,10,0.55,0)$\\
+$\theta_5 = (1,10,0.90,0)$ \\
+\hline  
+Synthetic AD:\\
+$\lambda_0^0 = (1, 0.3, -4, 0)$\\
+$\lambda_0^1 = (1, 0.2,\ \ 6, 0)$ \\
+\hline
+Synthetic PCA:\\ 
+$\lambda_1^0 = (1, 0.3,\ \ 6, 0)$\\
+$\lambda_1^1 = (1, 0.2, -4, 0)$\\
 % \hline
@@
 \end{flushright}
+\vspace{-1em}
 \caption[DKT Simulation Results - Comparison between true and DKT-estimated biomarker trajectories and subject time-shifts.]{Comparison between true and DKT-estimated subject time-shifts and biomarker trajectories.
-(top-left) Scatter plots of the true shifts (y-axis) against estimated shifts (x-axis), for the 'synthetic AD' (left) and 'synthetic PCA' (right) diseases.
-We also show the DKT-estimated and true trajectories of the agnostic units within the 'synthetic AD' disease (top-right, "Dis0") and the 'synthetic PCA' disease (middle-left, "Dis1").
-Finally, we also show the biomarker trajectories within unit 0 (middle-right) and unit 1 (bottom-left).
-Parameters used for data generation are shown in the bottom-right table.}
+(top-left/top-middle) Scatter plots of the true shifts (y-axis) against estimated shifts (x-axis), for the 'synthetic AD' and 'synthetic PCA' diseases.
+We then show the DKT-estimated and true trajectories of the agnostic units within the 'synthetic AD' disease (top-right, "Dis0") and the 'synthetic PCA' disease (bottom-left, "Dis1").
+Finally, we also show the biomarker trajectories within unit 0 (bottom-center) and unit 1 (bottom-right).
+Parameters used for generating the trajectory shapes are shown in the table on the right.}
   \label{fig:dktSynthTrajCompTrue}
@@
 
-In Fig.
-\ref{fig:PCAtrajByModality}, we plot the inferred trajectories for PCA directly across the disease progression.
-The results recapitulate known patterns in PCA, where posterior regions are predominantly affected, and shows that DKT can predict plausible non-MRI biomarkers for PCA in lack of such data for these subjects.
+We then apply DKT to real patient data, with the aim of transferring multimodal biomarker trajectories from tAD (TADPOLE dataset) to PCA.
+The inferred PCA trajectories, shown in Fig.
+\ref{fig:PCAtrajByModality}, recapitulate known patterns in PCA, where posterior regions are predominantly affected.
+We leave further interpretation of the results here for future work -- our aim here was only to demonstrate the possibility of inferring plausible non-MRI biomarkers in a rare neurodegenerative disease, in lack of such data for these subjects.
+As far as we are aware, this is the first time a continuous signature of non-MRI biomarkers is estimated for PCA, due to the rarity of the disease and the lack of data.
 
@@
 The only data that were available were the MRI volumetric data.
-The dynamics of the other biomarkers has been inferred by the model using data from typical AD, and taking into account the different spatial distribution of pathology in PCA vs tAD.}
+The dynamics of the other biomarkers has been inferred by the model using data from typical AD, and taking into account the different spatial distribution of pathology in PCA vs tAD.
+}
  \label{fig:PCAtrajByModality}
@@
 
-We validated DKT by predicting unseen DTI data from two datasets: 1) TADPOLE subjects with a different progression from the training subjects, and 2) a separate test set of 20 DTI scans from controls and PCA patients from our own cohort.
+We further validated DKT by predicting unseen DTI data from two patient datasets: 1) TADPOLE subjects with a different progression from the training subjects, and 2) a separate test set of 20 DTI scans from controls and PCA patients from our own cohort.
 To split TADPOLE into subgroups with different progression, we used the SuStaIn model by \cite{young2018uncovering}, which resulted into three subgroups: hippocampal, cortical and subcortical, with prominent early atrophy in those respective regions.
-This allows us to validate DKT by transferring multimodal data between subgroups with different progressions.
 To evaluate prediction accuracy, we computed the rank correlation between the DKT-predicted biomarker values and the measured values in the test data.
@@
 When predicting missing DTI markers of the TADPOLE cortical subgroup as well as PCA subjects from our local cohort (Table \ref{sec:dktPerfMetrics}), the DKT correlations are generally high for the cingulate, hippocampus and parietal, and lower for the frontal lobe.
-DKT also shows favourable performance to four other models: the latent-stage model from \cite{jedynak2012computational}, a multivariate non-linear model that predicts a DTI ROI marker from multiple MRI markers from different ROIs, as well as spline and linear models that predict DTI biomarkers directly from the MRI marker corresponding to the same ROI.
+DKT also shows favourable performance to four other models: the latent-stage model from \cite{jedynak2012computational}, a multivariate non-linear model that predicts a DTI ROI marker from multiple MRI markers, as well as spline and linear models that predict a regional DTI biomarker directly from its corresponding MRI marker.
 In particular for predicting DTI FA in the parietal and temporal lobes, DKT scores has significantly better predictions that almost all methods tested.
@@
 \end{tabular}
-% \normalfont
+\vspace{0.5em}
 \caption[Performance evaluation of DKT and other models]{Performance evaluation of DKT and four other statistical models of decreasing complexity.
-We show the rank correlation between predicted biomarkers and measured biomarkers in (top) TADPOLE subgroups -- transfer from hippocampal subgroup to cortical subgroup -- and (bottom) PCA.
+We show the rank correlation between predicted biomarkers and measured biomarkers in (top) TADPOLE subgroups -- infromation transfer from hippocampal subgroup to cortical subgroup -- and (bottom) PCA.
 (*) Statistically significant difference in the performance of DKT vs the other models, based on a two-tailed t-test, Bonferroni corrected.}
@@
 
-We presented DKT, a framework that enables joint modelling of biomarker progressions in multiple neurodegenerative diseases simultaneously.
+In this work we made initial steps at the challenging problem of transfer learning between different neurodegenerative diseases.
+Our proposed DKT method enabled the estimation of quantitative non-MRI trajectories in a rare disease (PCA) where very limited data was avialable.
+To our knowledge, this is the first time such a multimodal continuous signature is derived for PCA, as the only other longitudinal study of PCA only computed atrophy measures from MRI scans \cite{lehmann2011cortical}.
 Our work has however several limitations, which can be addressed in future research: 1) to account for population heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) DKT can be further validated on more complex synthetic experiments with a range of datasets generated with different parameters.
@@
 
-\bibitem{lorenzi2017disease}
-Lorenzi, M., Filippone, M., Frisoni, G.B., Alexander, D.C., Ourselin, S.
+% \bibitem{lorenzi2017disease}
+% Lorenzi, M., Filippone, M., Frisoni, G.B., Alexander, D.C., Ourselin, S.
 and Alzheimer's Disease Neuroimaging Initiative, 2017.
@@
 
+\bibitem{lehmann2011cortical}
+Lehmann, M., Crutch, S.J., Ridgway, G.R., Ridha, B.H., Barnes, J., Warrington, E.K., Rossor, M.N.
+and Fox, N.C., 2011.
+Cortical thickness and voxel-based morphometry in posterior cortical atrophy and typical Alzheimer's disease.
+Neurobiology of aging, 32(8), pp.1466-1476.
+
 \end{thebibliography}

### 2019-03-29 JointModel.py

@@ -160,21 +160,23 @@ class JointModel(DisProgBuilder.DPMInterface):
         i += 1
     else:
 
-      # i = 70
-      # self.loadCheckpoint(i, 2)
-      # self.makePlots(plotFigs, i, 2)
+      if self.params['plotTrajParams']['isSynth']:
+        i = 10
+        self.loadCheckpoint(i, 5)
+        self.makePlots(plotFigs, i, 5)
+      else:
 
-      i = 0
+        i = 0
 
-      self.loadCheckpoint(i, 2) # for the real data
-      # self.makePlots(plotFigs, i, 2)
-      # print(adsa)
+        self.loadCheckpoint(i, 2) # for the real data
+        # self.makePlots(plotFigs, i, 2)
+        # print(adsa)
 
 
-      # for SuStaIn validation
-      # self.loadCheckpoint(3, 1)  # for the real data
-      # self.makePlots(plotFigs, i, 2)
-      # print(adsa)
+        # for SuStaIn validation
+        # self.loadCheckpoint(3, 1)  # for the real data
+        # self.makePlots(plotFigs, i, 2)
+        # print(adsa)
 
 
     res = None
@@ -207,7 +209,8 @@ class JointModel(DisProgBuilder.DPMInterface):
   def makePlots(self, plotFigs, iterNr, picNr):
     if plotFigs:
       if self.params['plotTrajParams']['isSynth']:
-        fig = self.plotter.plotCompWithTrueParams(self.unitModels, self.disModels, replaceFig=True)
+        fig = self.plotter.plotCompWithTrueParams(self.unitModels, self.disModels,
+            replaceFig=True)
         fig.savefig('%s/compTrueParams%d%d_%s.pdf' % (self.outFolder, iterNr, picNr, self.expName))
         fig.savefig('%s/compTrueParams%d%d_%s.png' % (self.outFolder, iterNr, picNr, self.expName))
         pl.clf()

### 2019-03-29 Plotter.py

@@ -276,6 +276,10 @@ class PlotterJDM:
     else:
       nrRows, nrCols = auxFunc.findOptimalRowsCols(nrDis * 2 + nrFuncUnits * 2 + 2)
 
+
+    # nrRows = 2
+    # nrCols = 4
+
     # if (nrDis*2 + nrFuncUnits*2) > nrRows * nrCols:
     #   print('nrRows', nrRows)
     #   print('nrCols', nrCols)
@@ -1236,6 +1240,7 @@ class PlotterGP(ABC):
     trueTrajScaledXB = trajStruct['trueTrajScaledXB']
     yMinAll = trajStruct['yMinAll']
     yMaxAll = trajStruct['yMaxAll']
+    deltaY = (yMaxAll - yMinAll)/4
     min_yB = trajStruct['min_yB']
     max_yB = trajStruct['max_yB']
 
@@ -1252,16 +1257,16 @@ class PlotterGP(ABC):
     if self.plotTrajParams['allTrajOverlap']:
       ax2 = pl.subplot(nrRows, nrCols, nrPlotsSoFar+1)
       pl.title('%s all trajectories' % self.plotTrajParams['title'])
-      ax2.set_ylim([yMinAll, yMaxAll])
+      ax2.set_ylim([yMinAll - deltaY, yMaxAll])
       for b in range(gpModel.nrBiomk):
         ax2.plot(newXTrajScaledZeroOne, predTrajScaledXB[:, b], '-', lw=2
-                 , c=self.plotTrajParams['colorsTraj'][b], label=self.plotTrajParams['labels'][b])
+                 , c=self.plotTrajParams['colorsTraj'][b], label=self.plotTrajParams['labels'][b] + ' estim.')
         print('trueXsScaledZeroOne trueTrajScaledXB', trueXsScaledZeroOne.shape, trueTrajScaledXB.shape)
         ax2.plot(trueXsScaledZeroOne, trueTrajScaledXB[:, b], '--', lw=2
-                 , c=self.plotTrajParams['colorsTraj'][b], label=self.plotTrajParams['labels'][b])
+                 , c=self.plotTrajParams['colorsTraj'][b], label=self.plotTrajParams['labels'][b] + ' true')
 
       ax2.text(0.0, 0.95, 'MAE = %.3f' % np.mean(maeTraj))
-      ax2.legend(loc='lower right',ncol=1)
[…]

### 2019-03-29 jointSynth.py

@@ -69,23 +69,9 @@ from auxFunc import *
 import evaluationFramework
 from matplotlib import pyplot as pl
 
-hostName = gethostname()
-if hostName == 'razvan-Inspiron-5547':
-  homeDir = '/home/razvan'
-  blenderPath = 'blender'
-elif hostName == 'razvan-Precision-T1700':
-  homeDir = '/home/razvan'
-  blenderPath = 'blender'
-elif args.cluster:
-  homeDir = '/home/rmarines'
-  blenderPath = '/share/apps/blender-2.75/blender'
-else:
-  raise ValueError('Wrong hostname. If running on new machine, add '
-                   'application paths in python code above')
-
 
 plotTrajParams = {}
-plotTrajParams['SubfigTrajWinSize'] = (1600,900)
+plotTrajParams['SubfigTrajWinSize'] = (1200,600)
 plotTrajParams['nrRows'] = args.nrRows
 plotTrajParams['nrCols'] = args.nrCols
 plotTrajParams['diagColors'] = {CTL:'g', MCI:'y', AD:'r',
@@ -95,7 +81,7 @@ plotTrajParams['diagScatterMarkers'] = {CTL:'o', MCI:'o', AD:'o',
 plotTrajParams['legendCols'] = 4
 plotTrajParams['diagLabels'] = {CTL:'CTL', AD:'AD', PCA:'PCA', CTL2:'CTL2'}
 # plotTrajParams['ylimitsRandPoints'] = (-3,2)
-plotTrajParams['blenderPath'] = blenderPath
+# plotTrajParams['blenderPath'] = blenderPath
 plotTrajParams['isSynth'] = True
 plotTrajParams['padTightLayout'] = 1
 
@@ -164,7 +150,7 @@ def main():
   plotTrajParams['yNormMode'] = 'unscaled'
 
   # if False, plot estimated traj. in separate plot from true traj.
-  plotTrajParams['allTrajOverlap'] = False
+  plotTrajParams['allTrajOverlap'] = True
 
   params['unitNames'] = ['Unit%d' % f for f in range(nrFuncUnits)]
 
@@ -176,8 +162,11 @@ def main():
   params['penaltyUnits'] = 20
   params['penaltyDis'] = 1
   params['nrFuncUnits'] = nrFuncUnits
+  params['nrFuncUnitsImgOnly'] = nrFuncUnits
   params['biomkInFuncUnit'] = biomkInFuncUnit
[…]

### 2019-04-01 paper/paper.tex
minor

@@
 \newcommand{\expFld}{.}
-
-% We introduce Disease Knowledge Transfer (DKT), a novel technique for transferring biomarker information between related neurodegenerative diseases.
-DKT infers robust multimodal biomarker trajectories in rare neurodegenerative diseases even when only limited, unimodal data is available, by transferring information from larger multimodal datasets from common neurodegenerative diseases.
-DKT is a joint-disease generative model of biomarker progressions, which exploits biomarker relationships that are shared across diseases.
-Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects.
-For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own clinical center, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
-We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature.
-We further validate DKT in two situations: (1) on synthetic data, showing it can accurately estimate ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable predictive performance compared to other models.
-While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
 
@@
 
-Results on synthetic data are presented in Fig.
-\ref{fig:dktSynthTrajCompTrue}, showing the true and estimated subject shifts and trajectories for each agnostic unit $l$ and biomarker $k$ (see figure caption for detailed description).
-Results suggest that the DKT-estimated trajectories match closely (MAE $<$ 0.058) with the true trajectories.
-Moreover, the subject time-shifts are very close ($R^2$ $>$ 0.98) to the true time-shifts.
+Results on synthetic data in the presence of ground truth (Fig.
+\ref{fig:dktSynthTrajCompTrue}) suggest that DKT can robustly estimate the trajectory parameters (MAE $<$ 0.058) as well as the subject-specific time-shifts ($R^2$ $>$ 0.98).
 While some errors in trajectory estimation can be noticed, these are due to the informed priors on the model parameters in order to ensure identifiability and convergence of parameters.
@@
 The inferred PCA trajectories, shown in Fig.
-\ref{fig:PCAtrajByModality}, recapitulate known patterns in PCA, where posterior regions are predominantly affected.
+\ref{fig:PCAtrajByModality}, recapitulate known patterns in PCA \cite{crutch2012posterior}, where posterior regions such as occipital and parietal lobes are predominantly affected in later stages.
 We leave further interpretation of the results here for future work -- our aim here was only to demonstrate the possibility of inferring plausible non-MRI biomarkers in a rare neurodegenerative disease, in lack of such data for these subjects.
@@
 We further validated DKT by predicting unseen DTI data from two patient datasets: 1) TADPOLE subjects with a different progression from the training subjects, and 2) a separate test set of 20 DTI scans from controls and PCA patients from our own cohort.
-To split TADPOLE into subgroups with different progression, we used the SuStaIn model by \cite{young2018uncovering}, which resulted into three subgroups: hippocampal, cortical and subcortical, with prominent early atrophy in those respective regions.
+To split TADPOLE into subgroups with different progression, we used the SuStaIn model by \cite{young2018uncovering}, which resulted into three subgroups: hippocampal, cortical and subcortical, with prominent early atrophy in the hippocampus, cortical and subcortical regions respectively.
 To evaluate prediction accuracy, we computed the rank correlation between the DKT-predicted biomarker values and the measured values in the test data.
@@
 
-Validation results are shown in Table \ref{sec:dktPerfMetrics}, for TADPOLE subgroups (hippocampal to cortical, other combinations of subgroups not shown due to lack of space) as well as PCA subjects.
+Validation results are shown in Table \ref{sec:dktPerfMetrics}, for hippocampal to cortical TADPOLE subgroups (other pairs of subgroups not shown due to lack of space) as well as PCA subjects.
 When predicting missing DTI markers of the TADPOLE cortical subgroup as well as PCA subjects from our local cohort (Table \ref{sec:dktPerfMetrics}), the DKT correlations are generally high for the cingulate, hippocampus and parietal, and lower for the frontal lobe.
 DKT also shows favourable performance to four other models: the latent-stage model from \cite{jedynak2012computational}, a multivariate non-linear model that predicts a DTI ROI marker from multiple MRI markers, as well as spline and linear models that predict a regional DTI biomarker directly from its corresponding MRI marker.
-In particular for predicting DTI FA in the parietal and temporal lobes, DKT scores has significantly better predictions that almost all methods tested.
+In particular for predicting DTI FA in the parietal and temporal lobes, DKT has significantly better predictions that almost all methods tested.
 
@@
 \caption[Performance evaluation of DKT and other models]{Performance evaluation of DKT and four other statistical models of decreasing complexity.
-We show the rank correlation between predicted biomarkers and measured biomarkers in (top) TADPOLE subgroups -- infromation transfer from hippocampal subgroup to cortical subgroup -- and (bottom) PCA.
+We show the rank correlation between predicted biomarkers and measured biomarkers in (top) TADPOLE subgroups -- information transfer from hippocampal subgroup to cortical subgroup -- and (bottom) PCA.
 (*) Statistically significant difference in the performance of DKT vs the other models, based on a two-tailed t-test, Bonferroni corrected.}
@@
 In this work we made initial steps at the challenging problem of transfer learning between different neurodegenerative diseases.
-Our proposed DKT method enabled the estimation of quantitative non-MRI trajectories in a rare disease (PCA) where very limited data was avialable.
-To our knowledge, this is the first time such a multimodal continuous signature is derived for PCA, as the only other longitudinal study of PCA only computed atrophy measures from MRI scans \cite{lehmann2011cortical}.
+Our proposed DKT method enabled the estimation of quantitative non-MRI trajectories in a rare disease (PCA) where very limited data was available.
+To our knowledge, this is the first time a multimodal continuous signature is derived for PCA, as the only other longitudinal study of PCA only computed atrophy measures from MRI scans \cite{lehmann2011cortical}.
 Our work has however several limitations, which can be addressed in future research: 1) to account for population heterogeneity, DKT can be easily extended to include subject-specific effects; 2) improved schemes for biomarker allocation to agnostic units can take connectivity into account, or derive it from the data automatically; 3) DKT can be further validated on more complex synthetic experiments with a range of datasets generated with different parameters.

### 2019-04-03 paper/paper.tex
minor

@@
 For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own clinical center, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
-While validation is challenging due to lack of data in PCA, we validate DKT on synthetic data and two patient datasets (TADPOLE and PCA cohorts), showing it can estimate the ground truth parameters in the simulation and predict unseen biomarkers on the two patient datasets.
+Although validation is challenging due to lack of data in PCA, we validate DKT on synthetic data and two patient datasets (TADPOLE and PCA cohorts), showing it can estimate the ground truth parameters in the simulation and predict unseen biomarkers on the two patient datasets.
 While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
@@
 To evaluate prediction accuracy, we computed the rank correlation between the DKT-predicted biomarker values and the measured values in the test data.
+We compute the rank correlation instead of mean squared error as it is not susceptible to systemic biases of the models when predicting "unseen data" in a certain disease.
 
@@
 When predicting missing DTI markers of the TADPOLE cortical subgroup as well as PCA subjects from our local cohort (Table \ref{sec:dktPerfMetrics}), the DKT correlations are generally high for the cingulate, hippocampus and parietal, and lower for the frontal lobe.
-DKT also shows favourable performance to four other models: the latent-stage model from \cite{jedynak2012computational}, a multivariate non-linear model that predicts a DTI ROI marker from multiple MRI markers, as well as spline and linear models that predict a regional DTI biomarker directly from its corresponding MRI marker.
+DKT also shows favourable performance to four other models: the latent-stage model from \cite{jedynak2012computational}, a multivariate Gaussian Process model with RBF kernel that predicts a DTI ROI marker from multiple MRI markers, as well as cubic spline and linear models that predict a regional DTI biomarker directly from its corresponding MRI marker.
 In particular for predicting DTI FA in the parietal and temporal lobes, DKT has significantly better predictions that almost all methods tested.
@@
 Latent stage &    \textbf{0.80 $\pm$ 0.09} &    \textbf{0.53 $\pm$ 0.17} &      \textbf{0.80 $\pm$ 0.12} &    0.56 $\pm$ 0.18 &    0.50 $\pm$ 0.21 &    0.32 $\pm$ 0.24 \\
-Multivariate &   0.44 $\pm$ 0.19* &  -0.21 $\pm$ 0.25* &    -0.53 $\pm$ 0.26* &  -0.39 $\pm$ 0.20* &  -0.27 $\pm$ 0.28* &  -0.13 $\pm$ 0.26* \\
+Multivariate &   0.73 $\pm$ 0.09 &   0.45 $\pm$ 0.22  &    0.71 $\pm$ 0.08 & -0.28 $\pm$ 0.21* &  0.53 $\pm$ 0.22  &  0.25 $\pm$ 0.23* \\
 Spline &   0.52 $\pm$ 0.20* &  -0.03 $\pm$ 0.35* &     0.66 $\pm$ 0.11* &   0.09 $\pm$ 0.25* &    0.53 $\pm$ 0.20 &   0.30 $\pm$ 0.21* \\

### 2019-04-03 drcValidFuncs.py

@@ -447,9 +447,22 @@ def validateDRCBiomk(dpmObj, params):
     # print(nonMriDataCurrCol[nnMaskMulti].shape[0])
     assert nnDataMri[:,0].shape[0] == nnDataMri[:, 1].shape[0]
     assert nnDataMri[:,0].shape[0] == nonMriDataCurrCol[nnMaskMulti].shape[0]
-    multivarModel = Rbf(nnDataMri[:,0], nnDataMri[:,1], nnDataMri[:,2], nnDataMri[:,3], nnDataMri[:,4],
-               nnDataMri[:,5], nonMriDataCurrCol[nnMaskMulti], function='cubic')
+    # multivarModel = Rbf(nnDataMri[:,0], nnDataMri[:,1], nnDataMri[:,2], nnDataMri[:,3], nnDataMri[:,4],
+    #            nnDataMri[:,5], nonMriDataCurrCol[nnMaskMulti], function='cubic')
     # print(ads)
+
+    from sklearn.gaussian_process import GaussianProcessRegressor
+    from sklearn.gaussian_process.kernels import RBF, WhiteKernel
+
+    multivarModelRbf = Rbf(nnDataMri[:,0], nnDataMri[:,1], nnDataMri[:,2], nnDataMri[:,3], nnDataMri[:,4],
+               nnDataMri[:,5], nonMriDataCurrCol[nnMaskMulti], function='cubic')
+
+    kernel = 1.0 * RBF(length_scale=1, length_scale_bounds=(0.1, 10)) \
+             + WhiteKernel(noise_level=0.3, noise_level_bounds=(0.1, 0.1))
+    # print('nnDataMri[:,0:5].shape', nnDataMri[:,0:6].shape)
+    # print(nonMriDataCurrCol[nnMaskMulti].shape)
+    multivarModel = GaussianProcessRegressor(kernel=kernel,
+                                  alpha=0.0).fit(nnDataMri[:,0:6], nonMriDataCurrCol[nnMaskMulti])
 
 
     YvalidDktNonMri[b] = [[] for s in range(nrSubjWithValid)]  # Non-MRI predictions of DKT model for subj in validation set
@@ -475,9 +488,11 @@ def validateDRCBiomk(dpmObj, params):
                                  for u in range(nrMriBiomks)]
       # print('mrValsToPredictCurrSubj', mrValsToPredictCurrSubj)
[…]

### 2019-04-03 tadpoleSubtypes.py

@@ -1169,10 +1169,19 @@ def validateSubtypes(dpmObj, params, disNrTrain, disNrValid):
     # print(nonMriDataCurrCol[nnMaskMulti].shape[0])
     assert nnDataMri[:,0].shape[0] == nnDataMri[:, 1].shape[0]
     assert nnDataMri[:,0].shape[0] == nonMriDataCurrCol[nnMaskMulti].shape[0]
-    multivarModel = Rbf(nnDataMri[:,0], nnDataMri[:,1], nnDataMri[:,2], nnDataMri[:,3], nnDataMri[:,4],
-               nnDataMri[:,5], nonMriDataCurrCol[nnMaskMulti], function='cubic')
-    # print(ads)
 
+    from sklearn.gaussian_process import GaussianProcessRegressor
+    from sklearn.gaussian_process.kernels import RBF, WhiteKernel
+
+    multivarModelRbf = Rbf(nnDataMri[:,0], nnDataMri[:,1], nnDataMri[:,2], nnDataMri[:,3], nnDataMri[:,4],
+               nnDataMri[:,5], nonMriDataCurrCol[nnMaskMulti], function='cubic')
+
+    kernel = 1.0 * RBF(length_scale=1, length_scale_bounds=(0.1, 10)) \
+             + WhiteKernel(noise_level=0.3, noise_level_bounds=(0.1, 0.1))
+    # print('nnDataMri[:,0:5].shape', nnDataMri[:,0:6].shape)
+    # print(nonMriDataCurrCol[nnMaskMulti].shape)
+    multivarModel = GaussianProcessRegressor(kernel=kernel,
+                                  alpha=0.0).fit(nnDataMri[:,0:6], nonMriDataCurrCol[nnMaskMulti])
 
     YvalidDktNonMri[b] = [[] for s in range(nrSubjWithValid)]  # Non-MRI predictions of DKT model for subj in validation set
     YvalidLinModelNonMri[b] = [[] for s in range(nrSubjWithValid) ] # Non-MRI predictions of linear model for subj in validation set
@@ -1198,9 +1207,16 @@ def validateSubtypes(dpmObj, params, disNrTrain, disNrValid):
                                      for u in range(nrMriBiomks)]
           # print('mrValsToPredictCurrSubj', mrValsToPredictCurrSubj)
           assert len(mrValsToPredictCurrSubj) == nrMriBiomks
[…]

### 2019-06-09 paper/paper.tex
added feedback miccai 2019

@@
 When predicting missing DTI markers of the TADPOLE cortical subgroup as well as PCA subjects from our local cohort (Table \ref{sec:dktPerfMetrics}), the DKT correlations are generally high for the cingulate, hippocampus and parietal, and lower for the frontal lobe.
-DKT also shows favourable performance to four other models: the latent-stage model from \cite{jedynak2012computational}, a multivariate Gaussian Process model with RBF kernel that predicts a DTI ROI marker from multiple MRI markers, as well as cubic spline and linear models that predict a regional DTI biomarker directly from its corresponding MRI marker.
+DKT also shows favourable performance compared to four other models: the latent-stage model from \cite{jedynak2012computational}, a multivariate Gaussian Process model with RBF kernel that predicts a DTI ROI marker from multiple MRI markers, as well as cubic spline and linear models that predict a regional DTI biomarker directly from its corresponding MRI marker.
 In particular for predicting DTI FA in the parietal and temporal lobes, DKT has significantly better predictions that almost all methods tested.

### 2019-07-12 paper/paper.tex
added copyright form for miccai, re-orgnanised some files

@@
 %\newcommand{\expFld}{../resfiles/tad-drcTiny_JMD}
-\newcommand{\expFld}{.}
+\newcommand{\expFld}{figures}
 
@@
  \centering
- \includegraphics[width=0.8\textwidth,trim=0 0 0 0,clip]{disease_knowledge_transfer_symbols.pdf}
+ \includegraphics[width=0.8\textwidth,trim=0 0 0 0,clip]{figures/disease_knowledge_transfer_symbols.pdf}
  \caption{Diagram of the proposed DKT framework.
@@
 
-%  \begin{table}
-% \fontsize{6}{8}\selectfont
-%  \centering
-%  \begin{tabular}{c | c}
-% & \textbf{Trajectory parameters} \\
-% %\hline
-%  Biomarker allocation &  $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$ \\
-% Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$,  $\theta_4 = (1,5,0.9,0)$  \\
-% Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$, $\theta_5 = (1,10,0.9,0)$ \\
-%   
-% "Synthetic AD" & $\lambda_0^0 = (1, 0.3, -4, 0)$  and $\lambda_0^1 = (1, 0.2, 6, 0)$ \\
-%  "Synthetic PCA" & $\lambda_1^0 = (1, 0.3, 6, 0)$ and $\lambda_1^1 = (1, 0.2, -4, 0)$ \\
-% \hline
-% & \textbf{Subject parameters} \\
-%  Number of subjects & 100 (synthetic AD) and 50 (synthetic PCA) \\ 
-%  Time-shifts $\beta_i$ & $\beta_i \sim U(-13,10)$ years \\
-%  Diagnosis & $p(control) \propto Exp (-4.5)$,  $p(patient) \propto Exp (4.5)$\\
-%  Data generation & 4 visits/subject, 1 year apart, $\epsilon_k = 0.05$\\ 
-% \end{tabular}
-% \caption{Parameters used for synthetic data generation, emulating the TADPOLE cohort and our own dataset.}
-% \label{tab:synParams}
-% \end{table}
-
 \subsection{Data Acquisition and Preprocessing}
@@
 %\includegraphics[width=\textwidth]{../resfiles/synth/synth1_JMD/compTrueParams101_synth1_JMD.pdf}
-\includegraphics[width=0.8\textwidth]{../resfiles/synth/synth1_JMD/compTrueParams105_synth1_JMD.pdf}
+\includegraphics[width=0.8\textwidth]{figures/compTrueParams105_synth1_JMD.pdf}
 \begin{flushright}

### 2019-07-12 paper/supplementary.tex
started supplementary and modifications txt file

\documentclass{llncs}
\usepackage{amsmath,amssymb,calc,ifthen}
\usepackage{float}
%\usepackage{cancel}
\usepackage[table,usenames,dvipsnames]{xcolor} % for coloured cells in tables
\usepackage{tikz}
% Allows us to click on links and references!
\usepackage{hyperref}
\usepackage{url}
\hypersetup{
colorlinks,
citecolor=black,
filecolor=black,
linkcolor=black,
urlcolor=black
}
% Nice package for plotting graphs
% See excellent guide:
% http://www.tug.org/TUGboat/tb31-1/tb97wright-pgfplots.pdf
\usetikzlibrary{plotmarks,shapes}
\usepackage{amsmath,graphicx}
\usepackage{epstopdf}
%\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
% highlight - useful for TODOs and similar
\usepackage{color}
\newcommand{\hilight}[1]{\colorbox{yellow}{#1}}
\newcommand\ci{\perp\!\!\!\perp} % perpendicular sign
\newcommand*\rfrac[2]{{}^{#1}\!/_{#2}} % diagonal fraction
\newcommand\SLASH{\char`\\}
\usepackage{listings}
% margin size
\usepackage{pdfpages}
\usepackage{enumitem} % for nested enumerate numbers 1 1.1 1.1.1
% \usepackage{breqn}
% \usepackage[linesnumbered]{algorithm2e}
% \usepackage{algorithmicx,algpseudocode}
% \usepackage{wrapfig} % for allowing text wrapped around the algorithm
% \newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{blue}{#1}}
% \SetCommentSty{mycommfont}

% \usepackage{titlesec}
% \titlespacing*{\section}
% {0pt}{5.5ex plus 1ex minus .2ex}{4.3ex plus .2ex}

\usepackage[linesnumbered,noend]{algorithm2e}
\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{blue}{#1}}
\SetCommentSty{mycommfont}

\setlength{\textfloatsep}{15pt}
%\setlength{\floatsep}{1pt}
%\setlength{\topmargin}{1pt}

\renewcommand\floatpagefraction{.95}
\renewcommand\topfraction{.95}
\renewcommand\bottomfraction{.95}
\renewcommand\textfraction{.05}   
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\begin{document}

\definecolor{blue3}{HTML}{86B7FC} % med blue
\definecolor{blue1}{HTML}{B5F1FF} % light blue
\definecolor{blue2}{HTML}{E0F9FF} % very light blue

\title{Supplementary material: Disease Knowledge Transfer across Neurodegenerative Diseases}
%
\titlerunning{Supplementary -- DKT}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%

% * <mrazvan22@gmail.com> 2018-03-02T16:41:47.321Z:
% 
% Author list: Me, Marco (writing+model code), Stefano, Alex Young (writing+tadpole), Pere (helped with validation),Neil (tadpole),  Arman, Keir (DTI data), Seb, Danny
% 
% ^ <mrazvan22@gmail.com> 2018-03-02T16:55:06.003Z.

% \author{******************}

\author{}

\authorrunning{} % abbreviated author list (for running head)

\institute{}

% \institute{Centre for Medical Image Computing, University College London, UK
% \and 
% Queen Square MS Centre, UCL Institute of Neurology, UK
% \and 
% Dementia Research Centre, University College London, UK
% \and
% University of C\^{o}te d'Azur, Inria Sophia Antipolis, France
% % Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
% % F-91405 Orsay Cedex, France}
% }

\maketitle              % typeset the title of the contribution

%\newcommand{\expFld}{../resfiles/tad-drcTiny_JMD}
\newcommand{\expFld}{figures}

\newcommand{\lp}{\lambda_{d_i}^{\psi(k)}}
\newcommand{\lpuu}{\lambda_{d_i}^{\psi(k),(u)}}
\newcommand{\lpum}{\lambda_{d_i}^{\psi(k),(u-1)}}

\section{Parameter Estimation}

\newcommand{\uu}{^{(u)}}
\newcommand{\um}{^{(u-1)}}

We estimate the model parameters using a two-stage approach. In the first stage, we perform belief propagation within each agnostic unit and then within each disease model. In the second stage we jointly optimise across all agnostic units and disease models using loopy belief propagation. An overview of the algorithm is given in Figure \ref{fig:dktAlgo}. Given the initial parameters estimated from the first stage (line 1), the algorithm continuously updates the biomarker trajectories within the agnostic units (lines 4-5), dysfunction trajectories (line 8) and subject-specific time shifts (line 10) until convergence. The cost function for all parameters is nearly identical, the main difference being the measurements $(i,j,k)$ over subjects $i$, visits $j$ and biomarkers $k$ that are selected for computing the measurement error. For estimating the trajectory of biomarker $k$ within agnostic unit $\psi(k)$, measurements are taken from $\Omega_k$ representing all measurements of biomarker $k$ from all subjects and visits. For estimating the dysfunction trajectories, $\Omega_{d,l}$ represents the measurement indices from all subjects with disease $d$ (i.e. $d_i = d$) and all biomarkers $k$ that belong to agnostic unit $l$ (i.e. $\psi(k) = l$). Finally, $\Omega_i$ (line 10) represents all measurements from subject $i$, for all biomarkers and visits.

\begin{figure}
\begin{algorithm}[H]
\scriptsize
 Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
  \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
   \tcp*[l]{Estimate biomarker trajectories (disease agnostic)}
    \For{$k=1$ to $K$}{
      ${\theta_k\uu = \argmin_{\theta_k} \sum_{(i,j) \in \Omega_k} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\right) \right]^2  - log\ p(\theta_k)}$\\
      ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
    }
     \tcp*[l]{Estimate dysfunction trajectories (disease specific)} 
    \For{$d=1 \in \mathbb{D}$}{
      \For{$l=1 \in \Lambda$}{
        ${\lambda_{d}^{l, (u)} = \argmin_{\lambda_{d}^{l}} \sum_{(i,j,k) \in \Omega_{d,l}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lambda_{d}^{l}) ; \theta_k\uu 
        \right) \right]^2  - log\ p(\lambda_{d}^{l})}$\\
      }
    }
    \tcp*[l]{Estimate subject-specific time shifts} 
    \For{$i=1 \in [1, \dots, S]$}{
      ${\beta_i\uu = \argmin_{\beta_i} \sum_{(j,k) \in \Omega_i} \left[y_{ijk} - g\left(f(\beta_i + m_{ij}; \lpuu) ; \theta_k\uu
      \right) \right]^2  - log\ p(\beta_i)}$\\
    }
}
\normalfont
\end{algorithm}
\caption[The algorithm for estimating the DKT parameters]{The algorithm used to estimate the DKT parameters, based on loopy belief-propagation.}
\label{fig:dktAlgo}
\end{figure}

\section{Generation of synthetic dataset}

We tested DKT on synthetic data, to assess its performance against known
ground truth. More precisely, we generated data that follows the DKT model
exactly, and tested DKT's ability to recover biomarker trajectories and subject time-shifts.

We generated the synthetic data as follows, using parameters from Table \ref{tab:synParams}:
\begin{itemize}
 \item We simulate two synthetic diseases, "synthetic PCA" and "synthetic AD"
 \item We define 6 biomarkers that we allocate to agnostic units $l_0$ and $l_1$ (Table \ref{tab:synParams} top)
 \item Within each agnostic unit, we define the parameters $theta_0$ -- $theta_5$ corresponding to biomarker trajectories within the agnostic unit.
 \item For each disease, we define the parameters $\lambda$ corresponding to trajectories of dysfunction scores.
 \item We then sample data from 100 synthetic AD subjects and 50 PCA subjects with $\beta_i$ as given in Table \ref{tab:synParams} bottom using the model likelihood (Eq. 2 from main paper). For each subject, we generate data for 4 visits, each 1 year apart.
\end{itemize}

 \begin{table}
 %\fontsize{7}{12}
 \centering
 \begin{tabular}{c | c}
& \textbf{Trajectory parameters} \\
%\hline
 Biomarker allocation &  $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$ \\
Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$,  $\theta_4 = (1,5,0.9,0)$  \\
Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$, $\theta_5 = (1,10,0.9,0)$ \\

"Synthetic AD" & $\lambda_0^0 = (1, 0.3, -4, 0)$  and $\lambda_0^1 = (1, 0.2, 6, 0)$ \\
 "Synthetic PCA" & $\lambda_1^0 = (1, 0.3, 6, 0)$ and $\lambda_1^1 = (1, 0.2, -4, 0)$ \\
\hline
& \textbf{Subject parameters} \\
 Number of subjects & 100 (synthetic AD) and 50 (synthetic PCA) \\ 
 Time-shifts $\beta_i$ & $\beta_i \sim U(-13,10)$ years \\
 Diagnosis & $p(control) \propto Exp (-4.5)$,  $p(patient) \propto Exp (4.5)$\\
 Data generation & 4 visits/subject, 1 year apart, $\epsilon_k = 0.05$\\ 
\end{tabular}
\caption{Parameters used for synthetic data generation, emulating the TADPOLE and DRC datasets.}
\label{tab:synParams}
\end{table}

\section{Demographics of test sets}

The cohort from the Dementia Research Center UK used for validation had the following demongraphics: 10 subjects diagnosed with Posterior Cortical Atrophy, with a mean age of 59.4, 40\% females, as well as 10 age-matched controls with a mean age of 59.3, 50\% females.

For the validation on TADPOLE subtypes, we used applied the SuStaIn model on TADPOLE to split the population into three subtypes with different progression: a hippocampal subtype \TODODODO
When splitting the TADPOLE subjects into three populations with different progressions, the resulting groups had the following demographics:

\begin{table}
 \begin{tabular}{c | c c c c}
\textbf{Cohort} & \textbf{Nr. subjects} & \textbf{Nr. visits} & \textbf{Age (baseline)} & \textbf{Gender (\%F)}\\
\hline
Controls (Hippocampal) & 31 & 2.3 $\pm$ 1.8 & 74.4 $\pm$ 6.9 & 38\%\\
AD (Hippocampal) & 21 & 1.5 $\pm$ 0.8 & 74.5 $\pm$ 5.5 & 42\%\\
\hline
Controls (cortical) & 21 & 2.3 $\pm$ 1.3 & 70.9 $\pm$ 5.4 & 42\%\\
AD (cortical) & 35 & 1.7 $\pm$ 0.9 & 72.8 $\pm$ 7.4 & 28\%\\
\hline
Controls (subcortical) & 28 & 3.0 $\pm$ 1.5 & 73.7 $\pm$ 6.5 & 42\%\\
AD (subcortical) & 27 & 1.6 $\pm$ 0.9 & 73.7 $\pm$ 7.5 & 33\%\\
 \end{tabular}
 \caption{Demo}
 \label{tab:demogTadSubtypes}
\end{table}

% \begin{table}
%  \begin{tabular}{c | c c c}
%   Cohort & Number of scans & Age & Gender (\%F)\\
%   Controls & 10 & 59.4 & 50\%\\
%   PCA & 10 & 59.3 & 40\%\\
%  \end{tabular}
% \end{table}

\bibliographystyle{unsrtnat}
\begin{thebibliography}{5}

\end{thebibliography}

\clearpage

\end{document}

### 2019-07-12 tadpoleSubtypes.py

@@ -271,8 +271,8 @@ def prepareData(finalDataFile, tinyData, addExtraBiomk):
 
   # map controls to the three datasets based on sustain probability.
   # there will be 3 datasets:
-  # 1. CTL-cortical + patients-cortical
-  # 2. CTL-hippocampal + patients-hippocampal
+  # 1. CTL-hippocampal + patients-hippocampal
+  # 2. CTL-cortical + patients-cortical
   # 3. CTL-subcortical + patients-subcortical
 
   # dataDfTadpole.dataset[dataDfTadpole.diag == CTL] = dataDfTadpole.subtype[dataDfTadpole.diag == CTL]
@@ -398,6 +398,7 @@ def prepareData(finalDataFile, tinyData, addExtraBiomk):
   else:
     nrSubjToKeep = minNr
 
+
   assert nrSubjToKeep <= minNr
 
   np.random.seed(1)
@@ -420,6 +421,32 @@ def prepareData(finalDataFile, tinyData, addExtraBiomk):
   allBiomkCols = [x for x in allBiomkCols if not x.startswith('AV1451')]
   dataDfTadpole = dataDfTadpole.loc[:, 'RID': 'AV45 Temporal']
 
+  print(dataDfTadpole)
+
+  # print(np.logical_and(dataDfTadpole.Month_bl == 0, dataDfTadpole.dataset=1))
+
+  dfBl = dataDfTadpole.loc[dataDfTadpole.Month_bl == 0, :]
+  print('Group & Number & Visits & Age & Gender (\\%%F)\\\\')
+  grNames = ['Controls (Hippocampal)', 'Controls (cortical)', 'Controls (subcortical)',
+             'AD (Hippocampal)', 'AD (cortical)', 'AD (subcortical)']
+  for i in [1,4,2,5,3,6]:
+    # print('dfBl\n', dfBl)
+    dfCurr = dfBl.loc[dfBl.diag == i, :]
+    # print('dfCurr\n',dfCurr)
+    # print(np.nanmean(dfCurr.age))
+    # print(np.nanmean(dfCurr.loc[:,'gender-0f1m']))
+    dfGrByRID = dataDfTadpole.loc[dataDfTadpole.diag == i,:].groupby(['RID'], as_index=False)
+    avgNrVisits = np.mean(dfGrByRID.count().age)
+    stdNrVisits = np.std(dfGrByRID.count().age)
+
+
[…]

### 2019-07-16 paper/paper.tex
submitted camera-ready versiong

@@
 
-% * <mrazvan22@gmail.com> 2018-03-02T16:41:47.321Z:
-% 
-% Author list: Me, Marco (writing+model code), Stefano, Alex Young (writing+tadpole), Pere (helped with validation),Neil (tadpole),  Arman, Keir (DTI data), Seb, Danny
-% 
-% ^ <mrazvan22@gmail.com> 2018-03-02T16:55:06.003Z.
+\author{R\u{a}zvan V.
+Marinescu\inst{1,2} \and Marco Lorenzi\inst{5} \and Stefano Blumberg\inst{1} \and Alexandra L.
+Young\inst{1} \and Pere Planell-Morell\inst{1} \and Neil P.
+Oxtoby\inst{1} \and Arman Eshaghi\inst{1,3} \and Keir X.
+Young\inst{4} \and Sebastian J.
+Crutch\inst{4} \and Polina Golland\inst{2} \and Daniel C.
+Alexander\inst{1}, for the Alzheimer's Disease Neuroimaging Initiative}
 
-% \author{******************}
-
-\author{}
+% index{Marinescu, Razvan}
+% index{Lorenzi, Marco}
+% index{Blumberg, Stefano}
+% index{Young, Alexandra}
+% index{Planell-Morell, Pere}
+% index{Oxtoby, Neil}
+% index{Eshaghi, Arman}
+% index{Yong, Keir}
+% index{Crutch, Sebastian}
+% index{Goland, Polina}
+% index{Alexander, Daniel}
 
@@
 
-\institute{}
+% \institute{}
 
-% \institute{Centre for Medical Image Computing, University College London, UK
-% \and 
-% Queen Square MS Centre, UCL Institute of Neurology, UK
-% \and 
-% Dementia Research Centre, University College London, UK
-% \and
-% University of C\^{o}te d'Azur, Inria Sophia Antipolis, France
-% % Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
-% % F-91405 Orsay Cedex, France}
-% }
+\institute{Centre for Medical Image Computing, University College London, UK
+\and 
+Computer Science and Artificial Intelligence Laboratory, MIT, USA
+% \email{razvan@csail.mit.edu}
+\and
+Queen Square MS Centre, UCL Institute of Neurology, UK
+\and 
+Dementia Research Centre, University College London, UK
+\and
+University of C\^{o}te d'Azur, Inria Sophia Antipolis, France
+% Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
+% F-91405 Orsay Cedex, France}
+}
 
@@
 Our proposed method allows, for the first time, the estimation of plausible \emph{multimodal} biomarker trajectories in Posterior Cortical Atrophy (PCA), a rare neurodegenerative disease where only unimodal MRI data is available.
-For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from our own clinical center, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
+For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from the Dementia Research Centre (DRC), for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available.
 Although validation is challenging due to lack of data in PCA, we validate DKT on synthetic data and two patient datasets (TADPOLE and PCA cohorts), showing it can estimate the ground truth parameters in the simulation and predict unseen biomarkers on the two patient datasets.
 While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.
+Source code for DKT is available online: \url{https://github.com/mrazvan22/dkt}.
 
-% \keywords{Disease Progression Model, Transfer Learning, Manifold Learning, Alzheimer's Disease, Posterior Cortical Atrophy}
+\keywords{Disease Progression Modelling, Transfer Learning, Manifold Learning, Alzheimer's Disease, Posterior Cortical Atrophy}
 \end{abstract}
@@
 We apply DKT on Alzheimer's variants and demonstrate its ability to predict non-MRI trajectories for patients with Posterior Cortical Atrophy, in lack of such data.
-This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from our own center.
+This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from the Dementia Research Centre (DRC), UK.
 We finally validate DKT on three datasets: 1) simulated data with known ground truth, 2) TADPOLE sub-populations with different progressions and 3) 20 DTI scans from controls and PCA patients from our clinical center.
@@
  \centering
- \includegraphics[width=0.8\textwidth,trim=0 0 0 0,clip]{figures/disease_knowledge_transfer_symbols.pdf}
+ \includegraphics[width=0.9\textwidth,trim=0 0 0 0,clip]{figures/disease_knowledge_transfer_symbols.pdf}
  \caption{Diagram of the proposed DKT framework.
@@
 First, pathology in many related neurodegenerative diseases (e.g.
-Alzheimer's variants) are hypothesised to share the same underlying mechanisms (e.g.
+Alzheimer's variants) is hypothesised to share the same underlying mechanisms (e.g.
 amyloid and tau accumulation), and within one region, such mechanisms lead to similar pathology dynamics across all the disease variants modelled \cite{jack2010hypothetical}, with the key difference that distinct brain regions are affected at different times and with different pathology rates and extent, likely caused by selective vulnerability of networks within these regions \cite{seeley2009neurodegenerative}.
 Secondly, even if the diseases share different upstream mechanisms (e.g.
-amyloid vs tau accumulation), downstream biomarkers measuring hypometabolism, white matter degradation and structural markers are likely to follow the same pathological cascade and will have similar dynamics across diseases.
+amyloid vs tau accumulation), downstream biomarkers measuring hypometabolism, white matter degradation and atrophy are likely to follow the same pathological cascade and will have similar dynamics.
 
@@
 We now model the biomarker dynamics that are specific to each disease, by mapping the subjects' disease stages to dysfunction scores.
-We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the time shift of subject $i$.
-We then assume that each subject $i$ has a dysfunction score $\gamma_i^l$ corresponding to multimodal pathology in brain region $l$, which is a function of its disease stage:
+We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_{i} + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_{i}$ represents the time shift of subject $i$.
+We then assume that each subject $i$ at visit $j$ has a dysfunction score $\gamma_{ij}^l$ corresponding to multimodal pathology in brain region $l$, which is a function of its disease stage:
 
@@
 \label{eqDysfunctionScoreDef}
- \gamma_i^l = f(\beta_i + m_{ij}; \lambda_{d_i}^l)
+ \gamma_{ij}^l = f(\beta_{i} + m_{ij}; \lambda_{d_i}^l)
 \end{equation}
@@
 Moreover, $d_i \in \mathbb{D}$ represents the index of the disease corresponding to subject $i$, where $\mathbb{D}$ is the set of all diseases modelled.
-For example, MCI and tAD subjects from ADNI as well as tAD subjects from our own cohort can all be assigned $d_i=1$, while PCA subjects can be assigned $d_i=2$.
-We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this accounts for floor and ceiling effects present in AD biomarkers.
+For example, MCI and tAD subjects from ADNI as well as tAD subjects from the DRC cohort can all be assigned $d_i=1$, while PCA subjects can be assigned $d_i=2$.
+We implement $f$ as a parametric sigmoidal curve similar to \cite{jedynak2012computational}, to enable a robust optimisation and because this accounts for floor and ceiling effects present in AD biomarkers -- the monotonicity of this sigmoidal family is also very appropriate for many neurodegenerative diseases due to irreversability.
 
@@
 
-We further model the biomarker dynamics that are disease-agnostic, by constructing the mapping from the dysfunction scores $\gamma_i^l$ to the biomarker measurements.
+We further model the biomarker dynamics that are disease-agnostic, by constructing the mapping from the dysfunction scores $\gamma_{ij}^l$ to the biomarker measurements.
 We assume a set of given biomarker measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is the set of available biomarker measurements.
@@
 \begin{equation}
- p(y_{ijk}|\theta_k, \lp, \beta_i, \epsilon_k) = N(y_{ijk}| g( \gamma_i^{\psi(k)} ; \theta_k), \epsilon_k)
+ p(y_{ijk}|\theta_k, \lp, \beta_{i}, \epsilon_k) = N(y_{ijk}| g( \gamma_{ij}^{\psi(k)} ; \theta_k), \epsilon_k)
 \end{equation}
-where $g(\ .\ ; \theta_k)$ represents the trajectory of biomarker $k$ within agnostic unit $\psi(k)$, with parameters $\theta_k$, and is again implemented using a sigmoidal function.
-Parameters $\lp$ are used to define $\gamma_i^{\psi(k)}$ based on Eq.
+where $g(\ .\ ; \theta_k)$ represents the trajectory of biomarker $k$ within agnostic unit $\psi(k)$, with parameters $\theta_k$, and is again implemented using a sigmoidal function for reasons outlined above.
+Parameters $\lp$ are used to define $\gamma_{ij}^{\psi(k)}$ based on Eq.
 \ref{eqDysfunctionScoreDef}, where agnostic unit $l$ is now referred to as $\psi(k)$, to clarify this is the unit where biomarker $k$ has been allocated.
@@
 \begin{equation}
- p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \\ \prod_{(i,j,k) \in \Omega} p(y_{ijk}|\theta_k, \lp, \beta_i) 
+\label{eq:dktFinal}
+ p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \\ \prod_{(i,j,k) \in \Omega} p(y_{ijk}|\theta_k, \lp, \beta_{i}) 
 \end{equation}
@@
 
-We estimate the model parameters $[\boldsymbol{\theta}, \boldsymbol{\lambda}, \boldsymbol{\beta}, \boldsymbol{\epsilon}]$ using a two-stage approach.
-In the first stage, we perform belief propagation within each agnostic unit and then within each disease model.
-In the second stage we jointly optimise across all agnostic units and disease models using loopy belief propagation.
+We estimate the model parameters $[\boldsymbol{\theta}, \boldsymbol{\lambda}, \boldsymbol{\beta}, \boldsymbol{\epsilon}]$ using loopy belief propagation -- see algorithm in supplementary material.
+One key advantage of DKT is that the subject's time shift $\beta_{i}$ can be estimated using only a subset (e.g.
+MRI) of the subject's data -- the model can then infer the missing modalities (e.g.
+non-MRI) using Eq.
+\ref{eq:dktFinal}.
 
@@
 We generate synthetic data from two diseases (50 subjects with "\emph{synthetic PCA}" and 100 subjects with "\emph{synthetic AD}") using the parameters from the bottom-left table in Fig.
-\ref{fig:dktSynthTrajCompTrue}, emulating the TADPOLE cohort as well as our own cohort.
+\ref{fig:dktSynthTrajCompTrue}, emulating the TADPOLE and DRC cohorts -- see supplementary material for full details.
 The six biomarkers ($k_1$-$k_6$) have been \emph{a-priori} allocated to two agnostic units $l_0$ and $l_1$.
@@
 From the TADPOLE dataset we selected a subset of 230 subjects which had an MRI scan and at least one FDG PET, AV45, AV1451 or DTI scan.
-In order to model another disease, we further included MRI scans from 76 PCA subjects from our own cohort, along with scans from 67 tAD and 87 age-matched controls.
+In order to model another disease, we further included MRI scans from 76 PCA subjects from the DRC cohort, along with scans from 67 tAD and 87 age-matched controls.
 
 For both datasets, we computed multimodal biomarker measurements corresponding to each brain lobe: MRI volumes using the Freesurfer software, FDG-, AV45- and AV1451-PET standardised uptake value ratios (SUVR) extracted with the standard ADNI pipeline, and DTI fractional anisotropy (FA) measures from adjacent white-matter regions.
-For every lobe, we regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs our own center).
+For every lobe, we regressed out the following covariates: age, gender, total intracranial volume (TIV) and dataset (ADNI vs DRC).
 Finally, biomarkers were normalized to the [0,1] range.
@@
 
-We then apply DKT to real patient data, with the aim of transferring multimodal biomarker trajectories from tAD (TADPOLE dataset) to PCA.
+We then apply DKT to real patient data, with the aim of transferring multimodal biomarker trajectories from tAD to PCA.
 The inferred PCA trajectories, shown in Fig.
 \ref{fig:PCAtrajByModality}, recapitulate known patterns in PCA \cite{crutch2012posterior}, where posterior regions such as occipital and parietal lobes are predominantly affected in later stages.
-We leave further interpretation of the results here for future work -- our aim here was only to demonstrate the possibility of inferring plausible non-MRI biomarkers in a rare neurodegenerative disease, in lack of such data for these subjects.
-As far as we are aware, this is the first time a continuous signature of non-MRI biomarkers is estimated for PCA, due to the rarity of the disease and the lack of data.
+As opposed to typical AD, we find that the hippocampus is affected later on, further suggesting the model did not transfer too much tAD specific information.
+Here, we demonstrate the possibility of inferring plausible non-MRI biomarkers in a rare neurodegenerative disease, in lack of such data for these subjects.
+As far as we are aware, this is the first time a continuous signature of non-MRI biomarkers is estimated for PCA, due to its rarity and lack of data.
 
@@
 
-We further validated DKT by predicting unseen DTI data from two patient datasets: 1) TADPOLE subjects with a different progression from the training subjects, and 2) a separate test set of 20 DTI scans from controls and PCA patients from our own cohort.
+We further validated DKT by predicting unseen DTI data from two patient datasets: 1) TADPOLE subjects with a different progression from the training subjects, and 2) a separate test set of 20 DTI scans from controls and PCA patients from the DRC -- full demographics are given in the supplementary material.
 To split TADPOLE into subgroups with different progression, we used the SuStaIn model by \cite{young2018uncovering}, which resulted into three subgroups: hippocampal, cortical and subcortical, with prominent early atrophy in the hippocampus, cortical and subcortical regions respectively.
@@
 Validation results are shown in Table \ref{sec:dktPerfMetrics}, for hippocampal to cortical TADPOLE subgroups (other pairs of subgroups not shown due to lack of space) as well as PCA subjects.
-When predicting missing DTI markers of the TADPOLE cortical subgroup as well as PCA subjects from our local cohort (Table \ref{sec:dktPerfMetrics}), the DKT correlations are generally high for the cingulate, hippocampus and parietal, and lower for the frontal lobe.
+When predicting missing DTI markers of the TADPOLE cortical subgroup as well as PCA subjects from the DRC cohort (Table \ref{sec:dktPerfMetrics}), the DKT correlations are generally high for the cingulate, hippocampus and parietal, and lower for the frontal lobe.
 DKT also shows favourable performance compared to four other models: the latent-stage model from \cite{jedynak2012computational}, a multivariate Gaussian Process model with RBF kernel that predicts a DTI ROI marker from multiple MRI markers, as well as cubic spline and linear models that predict a regional DTI biomarker directly from its corresponding MRI marker.
@@
 \textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippocam.} & \textbf{Occipital} & \textbf{Parietal} & \textbf{Temporal}\\
-& \multicolumn{6}{c}{\textbf{TADPOLE subgroups: Hippocampal subgroup to Cortical subgroup}}\\
+& \multicolumn{6}{c}{\textbf{TADPOLE: Hippocampal subgroup to Cortical subgroup}}\\
 DKT (ours) &      0.56 $\pm$ 0.23 &    \textbf{0.35 $\pm$ 0.17} &        \textbf{0.58 $\pm$ 0.14} &     -0.10 $\pm$ 0.29 &     \textbf{0.71 $\pm$ 0.11} &     \textbf{0.34 $\pm$ 0.26} \\
@@
 \caption[Performance evaluation of DKT and other models]{Performance evaluation of DKT and four other statistical models of decreasing complexity.
-We show the rank correlation between predicted biomarkers and measured biomarkers in (top) TADPOLE subgroups -- information transfer from hippocampal subgroup to cortical subgroup -- and (bottom) PCA.
+We show the rank correlation between predicted biomarkers and measured biomarkers in (top) TADPOLE subgroups and (bottom) PCA.
 (*) Statistically significant difference in the performance of DKT vs the other models, based on a two-tailed t-test, Bonferroni corrected.}
@@
 
+\section{Acknowledgements}
+
+This work was supported by the EPSRC Centre For Doctoral Training in Medical Imaging with grant EP/L016478/1 and in part by the Neuroimaging Analysis Center through NIH grant NIH NIBIB NAC P41EB015902.
+Data collection and sharing for this project was funded by the Alzheimer’s Disease Neuroimaging Initiative (ADNI) (National Institutes of Health Grant U01 AG024904) and DOD ADNI (Department of Defense award number W81XWH-12-2-0012).
+The Dementia Research Centre is an ARUK coordination center.
+
 % The framework allows the inference of biomarker trajectories in rare diseases, for which there is not enough data to allow estimation of such trajectories, and accounts for a different spatial distribution of pathology between distinct phenotypes.
@@
 
-% Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and our local centre and 4) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
+% Our work has several limitations: 1) DKT assumes all subjects within a disease follow the same trajectory, without considering heterogeneity within the disease population, 2) the allocation of biomarkers into agnostic units has to be done using \emph{a-priori} human knowledge, 3) for validation, the synthetic experiment we ran was limited to only one setting of parameters mimicking the datasets from TADPOLE and the DRC centre and 4) the validation on patient data was also done only on a small set of 20 DTI scans, due to lack of multimodal data in PCA.
 

### 2019-07-16 paper/supplementary.tex
submitted camera-ready versiong

@@
 
-% * <mrazvan22@gmail.com> 2018-03-02T16:41:47.321Z:
-% 
-% Author list: Me, Marco (writing+model code), Stefano, Alex Young (writing+tadpole), Pere (helped with validation),Neil (tadpole),  Arman, Keir (DTI data), Seb, Danny
-% 
-% ^ <mrazvan22@gmail.com> 2018-03-02T16:55:06.003Z.
-
-% \author{******************}
-
-\author{}
+\author{R\u{a}zvan V.
+Marinescu\inst{1,2} \and Marco Lorenzi\inst{5} \and Stefano Blumberg\inst{1} \and Alexandra L.
+Young\inst{1} \and Pere Planell-Morell\inst{1} \and Neil P.
+Oxtoby\inst{1} \and Arman Eshaghi\inst{1,3} \and Keir X.
+Young\inst{4} \and Sebastian J.
+Crutch\inst{4} \and Polina Golland\inst{2} \and Daniel C.
+Alexander\inst{1}, for the Alzheimer's Disease Neuroimaging Initiative}
 
@@
 
-\institute{}
+% \institute{}
 
-% \institute{Centre for Medical Image Computing, University College London, UK
-% \and 
-% Queen Square MS Centre, UCL Institute of Neurology, UK
-% \and 
-% Dementia Research Centre, University College London, UK
-% \and
-% University of C\^{o}te d'Azur, Inria Sophia Antipolis, France
-% % Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
-% % F-91405 Orsay Cedex, France}
-% }
+\institute{Centre for Medical Image Computing, University College London, UK
+\and 
+Computer Science and Artificial Intelligence Laboratory, MIT, USA
+\email{razvan@csail.mit.edu}
+\and
+Queen Square MS Centre, UCL Institute of Neurology, UK
+\and 
+Dementia Research Centre, University College London, UK
+\and
+University of C\^{o}te d'Azur, Inria Sophia Antipolis, France
+% Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
+% F-91405 Orsay Cedex, France}
+}
 
@@
  \item We define 6 biomarkers that we allocate to agnostic units $l_0$ and $l_1$ (Table \ref{tab:synParams} top)
- \item Within each agnostic unit, we define the parameters $theta_0$ -- $theta_5$ corresponding to biomarker trajectories within the agnostic unit.
+ \item Within each agnostic unit, we define the parameters $\{\theta_0$, ..., $\theta_5\}$ corresponding to biomarker trajectories within the agnostic unit.
  \item For each disease, we define the parameters $\lambda$ corresponding to trajectories of dysfunction scores.
@@
 
-The cohort from the Dementia Research Center UK used for validation had the following demongraphics: 10 subjects diagnosed with Posterior Cortical Atrophy, with a mean age of 59.4, 40\% females, as well as 10 age-matched controls with a mean age of 59.3, 50\% females.
+The cohort from the Dementia Research Centre UK used for validation consisted of 10 subjects diagnosed with Posterior Cortical Atrophy, with a mean age of 59.4, 40\% females, as well as 10 age-matched controls with a mean age of 59.3, 50\% females.
 
-For the validation on TADPOLE subtypes, we used applied the SuStaIn model on TADPOLE to split the population into three subtypes with different progression: a hippocampal subtype \TODODODO
-When splitting the TADPOLE subjects into three populations with different progressions, the resulting groups had the following demographics:
+For the validation on TADPOLE subgroups, we used applied the SuStaIn model on TADPOLE to split the population into three subgroups with different progression: hippocampal, cortical and subcortical subypes with prominent atrophy in the hippocampus, cortical and subcortical areas respectively.
+The resulting subgroups had the following demographics:
 
@@
  \end{tabular}
- \caption{Demo}
+ \caption{Demographics of the subjects in the three TADPOLE subgroups.}
  \label{tab:demogTadSubtypes}
@@
 
-% \begin{table}
-%  \begin{tabular}{c | c c c}
-%   Cohort & Number of scans & Age & Gender (\%F)\\
-%   Controls & 10 & 59.4 & 50\%\\
-%   PCA & 10 & 59.3 & 40\%\\
-%  \end{tabular}
-% \end{table}
-
-\bibliographystyle{unsrtnat}
-\begin{thebibliography}{5}
-
-\end{thebibliography}
-
-\clearpage
+% \bibliographystyle{unsrtnat}
+% \begin{thebibliography}{5}
+% 
+% \end{thebibliography}
+% 
+% \clearpage
 

### 2019-10-02 paper/pres/poster-dkt.tex
started DKT miccai poster

\documentclass[portrait,a0,final,19pt]{a0poster}
%%%Load packages
\usepackage{multicol} 			%2-column layout
% \setlength\columnsep{8pt} % This is the default columnsep for all pages
\usepackage[left=2cm,right=2cm,bottom=2cm,top=0cm]{geometry}			%Reset margins
\usepackage{helvet}				%Load Helvetica font & CM math
\usepackage{color}				%Needed for colour boxes & coloured text
% \usepackage{graphics}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[font=Large,labelfont=bf]{caption}
\usepackage{float}
\usepackage{tikz}
\usetikzlibrary{arrows,positioning, shapes.symbols,shapes.callouts,patterns,shapes,chains,calc,backgrounds,fadings}

\usepackage{titlesec}
\usepackage{tabularx, booktabs} % make width of table columns evenly distributed (see http://tex.stackexchange.com/questions/60601/evenly-distributing-column-widths)

\usepackage{enumitem}
\usepackage{animate}[2017/05/18]

\usepackage{hyperref}
\usepackage{xcolor}

% \usepackage{titlepic}
% \usepackage{columns}

\titlespacing\section{0pt}{10pt plus 4pt minus 2pt}{5pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}

\titleformat*{\section}{\Huge\bfseries}
\titleformat*{\subsection}{\Large\bfseries}
\titleformat*{\subsubsection}{\large\bfseries}
\titleformat*{\paragraph}{\large\bfseries}
\titleformat*{\subparagraph}{\large\bfseries}

%%%Define colours and lengths
\definecolor{headingcol}{rgb}{0,0,0}			%Colour of main title
\definecolor{boxcol}{rgb}{0.7,0.2,0.2}		%Edge-colour of box and top banner
\fboxsep=1cm							%Padding between box and text
\setlength{\columnsep}{2cm}				%Set spacing between columns
\renewcommand{\familydefault}{\sfdefault}	%Set main text to sans-serif

%%%Format title
\makeatletter							%Needed to include code in main file
\renewcommand\@maketitle{%
\null									%Sets position marker
% \vspace{5em}
{
\color{headingcol}\sffamily\VeryHuge		%Set title font and colour
\@title \par}%
\vskip 0.6em%
{
\color{black}\sffamily\LARGE				%Set author font and colour
\lineskip .5em%
\begin{tabular}[t]{l}%
\@author
\end{tabular}\par}%
\vskip 1cm
\par
}
\makeatother

\title{Disease Knowledge Transfer across Neurodegenerative Diseases}

\newcommand{\inst}[1]{$^{#1}$}

\author{\Large{R\u{a}zvan V. Marinescu\inst{1}, Neil P. Oxtoby\inst{1}, Alexandra L. Young\inst{1}, Arman Eshaghi\inst{2}, Peter A. Wijeratne\inst{1}, Daniel C. Alexander\inst{1}}\\
\begin{tabular}{l p{1cm} l p{1cm} l}
\large{$^1$Centre for Medical Image Computing, University College London}  & & \large{$^2$Queen Square MS Centre, UCL Institute of Neurology} \\
\end{tabular}
}

% \author{R\u{a}zvan V. Marinescu\inst{1,2} \and Marco Lorenzi\inst{5} \and Stefano B. Blumberg\inst{1} \and Alexandra L. Young\inst{1} \and Pere Planell-Morell\inst{1} \and Neil P. Oxtoby\inst{1} \and Arman Eshaghi\inst{1,3} \and Keir X. Young\inst{4} \and Sebastian J. Crutch\inst{4} \and Polina Golland\inst{2} \and Daniel C. Alexander\inst{1}\\
% \begin{tabular}{l p{1cm} l p{1cm} l}
% \large{$^1$Centre for Medical Image Computing, University College London}  & & \large{$^2$Queen Square MS Centre, UCL Institute of Neurology} \\
% \end{tabular}
% }

\begin{document}
% \hspace{-3cm}								%Align with edge of page, not margin
% \vspace{2cm}
% \includegraphics[scale=1,trim=0 0 0 200, clip]{Black_Landscape.pdf}

%\colorbox{boxcol}{							%Coloured banner across top
\begin{minipage}{50cm}					%Minipage for title contents
% \vspace{-18cm}
\maketitle
\end{minipage}
%}
% \vspace{-4cm}

\begin{multicols}{2}							
\raggedcolumns							%Don't stretch contents vertically

\pagenumbering{gobble}

%%%Column1
\vspace{-3em}
\section*{Introduction}

\begin{itemize}
 \item Recently disease progression models work well on large, multimodal datasets
 \item Datasets on rare neurodegenerative diseases (e.g., Posterior Cortical Atrophy) are limited and cross-sectional.
\end{itemize}

% \vspace{-0.5em}

\section*{Aim}

Develop a transfer learning model to transfer knowledge between diseases

% \begin{figure}[H]
% \includegraphics[width=0.31\textwidth,trim=0 50 0 0, clip]{adniDiseaseProgression}
% \caption{Hypothetical progression of typical Alzheimer's disease. The earliest biomarkers that become abnormal are the accumulation of abnormal amyloid-beta and tau proteins, which have a misfolded (i.e. abnormal) 3D structure. This is followed by structural integrity from MRI and finally cognitive decline as measured by cognitive tests.}
% \end{figure}

\section*{Method}

\begin{itemize}
 \item We propose Disease Knowledge Transfer:
 \begin{itemize}
   \item 
 \end{itemize}

\end{itemize}

% \subsection{Method Overview}

\newcommand{\lp}{\lambda_{d_i}^{\psi(k)}}
\newcommand{\lpuu}{\lambda_{d_i}^{\psi(k),(u)}}
\newcommand{\lpum}{\lambda_{d_i}^{\psi(k),(u-1)}}

\begin{equation}
\label{eqDysfunctionScoreDef}
 \gamma_{ij}^l = f(\beta_{i} + m_{ij}; \lambda_{d_i}^l)
\end{equation}

\begin{equation}
 p(y_{ijk}|\theta_k, \lp, \beta_{i}, \epsilon_k) = N(y_{ijk}| g( \gamma_{ij}^{\psi(k)} ; \theta_k), \epsilon_k)
\end{equation}

\begin{equation}
\label{eq:dktFinal}
 p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \\ \prod_{(i,j,k) \in \Omega} p(y_{ijk}|\theta_k, \lp, \beta_{i}) 
\end{equation}

\begin{figure}[H]
 \centering
 \includegraphics[width=0.4\textwidth,]{../figures/disease_knowledge_transfer_symbols.pdf}
 \caption{}
 \label{fig:diagram}
\end{figure}

\section*{Results}

\begin{figure}[H]
%\includegraphics[width=\textwidth]{../resfiles/synth/synth1_JMD/compTrueParams101_synth1_JMD.pdf}
\includegraphics[width=0.4\textwidth]{../figures/compTrueParams105_synth1_JMD.pdf}
\caption[DKT Simulation Results - Comparison between true and DKT-estimated biomarker trajectories and subject time-shifts.]{Comparison between true and DKT-estimated subject time-shifts and biomarker trajectories.}
  \label{fig:dktSynthTrajCompTrue}
\end{figure}

\begin{figure}[H]
\centering
 \includegraphics[width=0.4\textwidth, trim=0 20 0 0, clip]{../figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
 \caption{Estimated trajectories for the PCA cohort. The only data that were available were the MRI volumetric data. The dynamics of the other biomarkers has been inferred by the model using data from typical AD, and taking into account the different spatial distribution of pathology in PCA vs tAD. }
 \label{fig:PCAtrajByModality}
\end{figure}

\section*{Conclusion}

\end{multicols}
\hrule

\begin{multicols}{2}
 							%Use 3-column layout
\raggedcolumns

\subsection*{References}

\large{
\begin{tabular}{l p{0.5cm} l}
 1. Fontejin et al., Neuroimg., 2012 & & 2. Young et al., Nat. Comms., 2018\\
 3. Villemagne et al., Lancet Neurol., 2013 & & 4. Marinescu et al., IPMI, 2017\\
\end{tabular}

\columnbreak

\subsection*{Weblinks}
\begin{itemize}
\item UCL Progression of Neurodegenerative Disease (POND): cmic.cs.ucl.ac.uk/pond/
\item UCL Centre for Medical Image Computing: www.ucl.ac.uk/cmic/
\end{itemize}
}

\vspace{1em}
% \hspace{2em}
\includegraphics[height=4.0cm]{epsrc_logo.jpg}
\hspace{0.5em}
\includegraphics[height=4.0cm]{pondLogo.png} 
\hspace{0.5em}
\includegraphics[height=4cm]{cdt_logo.png} 
\hspace{0.5em}
\includegraphics[height=4cm]{aruk_logo}

\end{multicols}

\end{document}

### 2019-10-03 paper/pres/poster-dkt.tex
miccai dtk poster

@@
-\documentclass[portrait,a0,final,19pt]{a0poster}
+\documentclass[portrait,a0,final,20pt]{a0poster}
 %%%Load packages
@@
 \usepackage{xcolor}
+\usepackage{newtxtext,newtxmath}
 
@@
 
-\author{\Large{R\u{a}zvan V. Marinescu\inst{1}, Neil P. Oxtoby\inst{1}, Alexandra L. Young\inst{1}, Arman Eshaghi\inst{2}, Peter A. Wijeratne\inst{1}, Daniel C. Alexander\inst{1}}\\
+% \author{\Large{R\u{a}zvan V. Marinescu\inst{1}, Neil P. Oxtoby\inst{1}, Alexandra L. Young\inst{1}, Arman Eshaghi\inst{2}, Peter A. Wijeratne\inst{1}, Daniel C. Alexander\inst{1}}\\
+% \begin{tabular}{l p{1cm} l p{1cm} l}
+% \large{$^1$Centre for Medical Image Computing, University College London}  & & \large{$^2$Queen Square MS Centre, UCL Institute of Neurology} \\
+% \end{tabular}
+% }
+
+\author{R\u{a}zvan V.
+Marinescu\inst{1,2}, Marco Lorenzi\inst{5}, Stefano B.
+Blumberg\inst{1}, Alexandra L.
+Young\inst{1}, Pere Planell-Morell\inst{1}, Neil P.
+Oxtoby\inst{1}, Arman Eshaghi\inst{1,3}, Keir X.
+Young\inst{4}, Sebastian J.
+Crutch\inst{4}, Polina Golland\inst{2}, Daniel C.
+Alexander\inst{1}\\
 \begin{tabular}{l p{1cm} l p{1cm} l}
@@
 
-% \author{R\u{a}zvan V.
-Marinescu\inst{1,2} \and Marco Lorenzi\inst{5} \and Stefano B.
-Blumberg\inst{1} \and Alexandra L.
-Young\inst{1} \and Pere Planell-Morell\inst{1} \and Neil P.
-Oxtoby\inst{1} \and Arman Eshaghi\inst{1,3} \and Keir X.
-Young\inst{4} \and Sebastian J.
-Crutch\inst{4} \and Polina Golland\inst{2} \and Daniel C.
-Alexander\inst{1}\\
-% \begin{tabular}{l p{1cm} l p{1cm} l}
-% \large{$^1$Centre for Medical Image Computing, University College London}  & & \large{$^2$Queen Square MS Centre, UCL Institute of Neurology} \\
-% \end{tabular}
-% }
+\newcommand{\fnt}[1]{\LARGE{#1}}
 
@@
 
+\fnt{
+
 \begin{multicols}{2}							
@@
 \vspace{-3em}
-\section*{Introduction}
-
-\begin{itemize}
- \item Recently disease progression models work well on large, multimodal datasets
- \item Datasets on rare neurodegenerative diseases (e.g., Posterior Cortical Atrophy) are limited and cross-sectional.
-\end{itemize}
-
-% \vspace{-0.5em}
 
@@
 
-Develop a transfer learning model to transfer knowledge between diseases
+Propose mechanism to infer progression of non-MRI biomarkers in rare neurodegenerative diseases by leveraging larger datasets of common neurodegenerative diseases.
 
-% \begin{figure}[H]
-% \includegraphics[width=0.31\textwidth,trim=0 50 0 0, clip]{adniDiseaseProgression}
-% \caption{Hypothetical progression of typical Alzheimer's disease.
-The earliest biomarkers that become abnormal are the accumulation of abnormal amyloid-beta and tau proteins, which have a misfolded (i.e.
-abnormal) 3D structure.
-This is followed by structural integrity from MRI and finally cognitive decline as measured by cognitive tests.}
-% \end{figure}
-
-\section*{Method}
-
-\begin{itemize}
- \item We propose Disease Knowledge Transfer:
- \begin{itemize}
-   \item 
- \end{itemize}
-
-\end{itemize}
-
-% \subsection{Method Overview}
+\section*{Why}
 
@@
 
-\begin{equation}
-\label{eqDysfunctionScoreDef}
- \gamma_{ij}^l = f(\beta_{i} + m_{ij}; \lambda_{d_i}^l)
-\end{equation}
+\begin{itemize}
+ \item Datasets on rare neurodegenerative diseases (e.g., Posterior Cortical Atrophy) are unimodal (MRI only), cross-sectional and small.
+ \item The continuous progression of non-MRI markers in rare neurodegenerative diseases is not well understood
+\end{itemize}
 
-\begin{equation}
- p(y_{ijk}|\theta_k, \lp, \beta_{i}, \epsilon_k) = N(y_{ijk}| g( \gamma_{ij}^{\psi(k)} ; \theta_k), \epsilon_k)
-\end{equation}
+% \vspace{-0.5em}
 
-\begin{equation}
-\label{eq:dktFinal}
- p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \\ \prod_{(i,j,k) \in \Omega} p(y_{ijk}|\theta_k, \lp, \beta_{i}) 
-\end{equation}
+% \begin{itemize}
+%  \item by transferring knowledge from larger multimodal datasets of typical Alzheimer's disease
+%  \item jointly model multiple diseases and multiple modalities
+% \end{itemize}
+
+\section*{Method}
+
+% \begin{itemize}
+%  \item We propose a joint-disease model called Disease Knowledge Transfer:
+ \vspace{1em}
+ \begin{figure}[H]
+ \centering
+  \begin{subfigure}{0.49\columnwidth}
+   \centering
+   \fnt{1. Each disease characterised by region-specific dysfunction profile}\\
+   $ \gamma_{ij}^l = f(\beta_{i} + m_{ij}; \lambda_{d_i}^l) $\\
+%    \vspace{0.9em}
+   \includegraphics[width=\textwidth]{diseaseEvol}
+
+  \end{subfigure}
+  \begin{subfigure}{0.49\columnwidth}
+   \centering
+%    \vspace{-1.2em}
+   \fnt{2. Dysfunction score modelled using region-specific biomarkers}\\
+   $ p(y_{ijk}|\theta_k, \lp, \beta_{i}, \epsilon_k) = N(y_{ijk}| g( \gamma_{ij}^{\psi(k)} ; \theta_k), \epsilon_k) $
+%    \vspace{0.6em}
+   \includegraphics[width=\textwidth]{dysfuncEvol}
+  \end{subfigure}
+
+ \end{figure}
+
+%    \item each disease has a unique evolution of region-specific dysfunction scores
+%    \item each dysfunction score modelled as the unique evolution of biomarkers within that region
+
+% \end{itemize}
+
+% \subsection{Method Overview}
+
+% $ p(y_{ijk}|\theta_k, \lp, \beta_{i}, \epsilon_k) = N(y_{ijk}| g( \gamma_{ij}^{\psi(k)} ; \theta_k), \epsilon_k) $
 
@@
  \centering
- \includegraphics[width=0.4\textwidth,]{../figures/disease_knowledge_transfer_symbols.pdf}
- \caption{}
- \label{fig:diagram}
+  \fnt{3. Extend to multiple subjects, biomarkers and diseases}\\
+ $ p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \prod_{(i,j,k) \in \Omega} p(y_{ijk}|\theta_k, \lp, \beta_{i}) $
+
+ \includegraphics[width=\columnwidth]{../figures/disease_knowledge_transfer_symbols.pdf}
+%  \caption{}
+%  \label{fig:diagram}
 \end{figure}
@@
 \begin{figure}[H]
-%\includegraphics[width=\textwidth]{../resfiles/synth/synth1_JMD/compTrueParams101_synth1_JMD.pdf}
+\fnt{Synthetic experiment shows that the model can recover the underlying parameters}\\
 \includegraphics[width=0.4\textwidth]{../figures/compTrueParams105_synth1_JMD.pdf}
-\caption[DKT Simulation Results - Comparison between true and DKT-estimated biomarker trajectories and subject time-shifts.]{Comparison between true and DKT-estimated subject time-shifts and biomarker trajectories.}
   \label{fig:dktSynthTrajCompTrue}
@@
 \centering
+\fnt{Inferred trajectories for Posterior Cortical Atrophy}\\
  \includegraphics[width=0.4\textwidth, trim=0 20 0 0, clip]{../figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
- \caption{Estimated trajectories for the PCA cohort.
+%  \caption{Estimated trajectories for the PCA cohort.
 The only data that were available were the MRI volumetric data.
@@
 }
- \label{fig:PCAtrajByModality}
+%  \label{fig:PCAtrajByModality}
 \end{figure}
+
+\begin{table}[H]
+\centering
+\fnt{Our model has favourable performance compared to other models}\\
+\vspace{1em}
+\fontsize{27}{30}\selectfont
+\begin{tabular}{c | c c c c c c}
+\textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippocam.} & \textbf{Occipital} & \textbf{Parietal} & \textbf{Temporal}\\
+& \multicolumn{6}{c}{\textbf{TADPOLE: Hippocampal subgroup to Cortical subgroup}}\\
+DKT (ours) &      0.56 $\pm$ 0.23 &    \textbf{0.35 $\pm$ 0.17} &        \textbf{0.58 $\pm$ 0.14} &     -0.10 $\pm$ 0.29 &     \textbf{0.71 $\pm$ 0.11} &     \textbf{0.34 $\pm$ 0.26} \\
+Latent stage &      0.44 $\pm$ 0.25 &    0.34 $\pm$ 0.21 &       0.34 $\pm$ 0.24* &     \textbf{-0.07 $\pm$ 0.22} &     0.64 $\pm$ 0.16 &    0.08 $\pm$ 0.24* \\
+Multivariate &      \textbf{0.60 $\pm$ 0.18} &   0.11 $\pm$ 0.22* &       0.12 $\pm$ 0.29* &     -0.22 $\pm$ 0.22 &   -0.44 $\pm$ 0.14* &   -0.32 $\pm$ 0.29* \\
+Spline &    -0.24 $\pm$ 0.25* &  -0.06 $\pm$ 0.27* &        0.58 $\pm$ 0.17 &     -0.16 $\pm$ 0.27 &    0.23 $\pm$ 0.25* &    0.10 $\pm$ 0.25* \\
+Linear &    -0.24 $\pm$ 0.25* &   0.20 $\pm$ 0.25* &        0.58 $\pm$ 0.17 &     -0.16 $\pm$ 0.27 &    0.23 $\pm$ 0.25* &    0.13 $\pm$ 0.23* \\
+& \multicolumn{6}{c}{\textbf{typical Alzheimer's to Posterior Cortical Atrophy}}\\
+DKT (ours) &    0.77 $\pm$ 0.11 &    0.39 $\pm$ 0.26 &      0.75 $\pm$ 0.09 &    0.60 $\pm$ 0.14 &    \textbf{0.55 $\pm$ 0.24} &    \textbf{0.35 $\pm$ 0.22} \\
+Latent stage &    \textbf{0.80 $\pm$ 0.09} &    \textbf{0.53 $\pm$ 0.17} &      \textbf{0.80 $\pm$ 0.12} &    0.56 $\pm$ 0.18 &    0.50 $\pm$ 0.21 &    0.32 $\pm$ 0.24 \\
+Multivariate &   0.73 $\pm$ 0.09 &   0.45 $\pm$ 0.22  &    0.71 $\pm$ 0.08 & -0.28 $\pm$ 0.21* &  0.53 $\pm$ 0.22  &  0.25 $\pm$ 0.23* \\
+Spline &   0.52 $\pm$ 0.20* &  -0.03 $\pm$ 0.35* &     0.66 $\pm$ 0.11* &   0.09 $\pm$ 0.25* &    0.53 $\pm$ 0.20 &   0.30 $\pm$ 0.21* \\
+Linear &   0.52 $\pm$ 0.20* &    0.34 $\pm$ 0.27 &     0.66 $\pm$ 0.11* &    \textbf{0.64 $\pm$ 0.17} &    0.54 $\pm$ 0.22 &   0.30 $\pm$ 0.21* \\
+\end{tabular}
+\vspace{0.5em}
+% \caption[Performance evaluation of DKT and other models]{Performance evaluation of DKT and four other statistical models of decreasing complexity. }
+% \end{footnotesize}
+\label{sec:dktPerfMetrics}
+\end{table}
 
@@
 \hrule
+
+}
 

### 2019-10-04 paper/pres/poster-dkt.tex
more on dkt poster

@@
 
+\usepackage[absolute,overlay]{textpos} % for logo next ti title
+
 \usepackage{titlesec}
@@
 
-\usepackage{hyperref}
+\usepackage[hidelinks, colorlinks = true,
+            linkcolor = blue,
+            urlcolor  = blue,
+            citecolor = blue,
+            anchorcolor = blue]{hyperref}
 \usepackage{xcolor}
 \usepackage{newtxtext,newtxmath}
+
+\usepackage[noend]{algorithm2e}
+\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{blue}{#1}}
+\SetCommentSty{mycommfont}
+
+\usepackage{amssymb} 
+\usepackage{pifont}% for red cross and green tick
 
@@
 {
-\color{headingcol}\sffamily\VeryHuge		%Set title font and colour
+\color{headingcol}\sffamily\VeryHuge \textbf		%Set title font and colour
 \@title \par}%
-\vskip 0.6em%
+% \vskip 0.6em%
 {
@@
 \author{R\u{a}zvan V.
-Marinescu\inst{1,2}, Marco Lorenzi\inst{5}, Stefano B.
+Marinescu\inst{1,2}, Marco Lorenzi\inst{3}, Stefano B.
 Blumberg\inst{1}, Alexandra L.
 Young\inst{1}, Pere Planell-Morell\inst{1}, Neil P.
-Oxtoby\inst{1}, Arman Eshaghi\inst{1,3}, Keir X.
-Young\inst{4}, Sebastian J.
-Crutch\inst{4}, Polina Golland\inst{2}, Daniel C.
+Oxtoby\inst{1},\\ Arman Eshaghi\inst{1,4}, Keir X.
+Young\inst{5}, Sebastian J.
+Crutch\inst{5}, Polina Golland\inst{2}, Daniel C.
 Alexander\inst{1}\\
 \begin{tabular}{l p{1cm} l p{1cm} l}
-\large{$^1$Centre for Medical Image Computing, University College London}  & & \large{$^2$Queen Square MS Centre, UCL Institute of Neurology} \\
+\large{$^1$Centre for Medical Image Computing, UCL, UK}  & & \large{$^3$University of C\^{o}te d'Azur, Inria Sophia Antipolis, France} & & \large{$^5$Dementia Research Centre, UCL, UK}\\
+\large{$^2$Computer Science and Artificial Intelligence Laboratory, MIT, USA} & & \large{$^4$Queen Square MS Centre, UCL Institute of Neurology, UK} & & 
+ \\
 \end{tabular}
@@
 
+% \institute{Centre for Medical Image Computing, University College London, UK
+% \and 
+% Computer Science and Artificial Intelligence Laboratory, MIT, USA
+% \email{razvan@csail.mit.edu}
+% \and
+% Queen Square MS Centre, UCL Institute of Neurology, UK
+% \and 
+% Dementia Research Centre, University College London, UK
+% \and
+% University of C\^{o}te d'Azur, Inria Sophia Antipolis, France
+% % Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
+% % F-91405 Orsay Cedex, France}
+% }
+
 \newcommand{\fnt}[1]{\LARGE{#1}}
+
+\DeclareMathOperator*{\argmin}{arg\,min}
+\DeclareMathOperator*{\argmax}{arg\,max}
 
@@
 
-%\colorbox{boxcol}{							%Coloured banner across top
 \begin{minipage}{50cm}					%Minipage for title contents
@@
 \end{minipage}
-%}
-% \vspace{-4cm}
+\begin{textblock}{0}(10,0.2)
+\begin{tabular}{l}
+\includegraphics[height=5cm]{ucl_logo}\\
+\includegraphics[height=5cm]{mit_logo}\\
+\end{tabular}
+\includegraphics[height=6cm]{qr_dkt.png}
+
+\end{textblock}
 
 \fnt{
+
+\newcommand{\lp}{\lambda_{d_i}^{\psi(k)}}
+\newcommand{\lpuu}{\lambda_{d_i}^{\psi(k),(u)}}
+\newcommand{\lpum}{\lambda_{d_i}^{\psi(k),(u-1)}}
+
+\newcommand{\cmark}{\ding{51}}%
+\newcommand{\xmark}{\ding{55}}%
+
+\newcommand{\yes}{{\LARGE \textcolor{green!50!black}{\cmark} \par}}
+\newcommand{\no}{{\LARGE \textcolor{red}{\xmark} \par}}
+
+\hrule
+
+\newcommand{\heading}[1]{\textbf{\Huge{#1}}}
 
@@
 
-\section*{Aim}
-
-Propose mechanism to infer progression of non-MRI biomarkers in rare neurodegenerative diseases by leveraging larger datasets of common neurodegenerative diseases.
+\heading{Aim}: Infer progression of non-MRI biomarkers in rare neurodegenerative diseases by leveraging larger datasets of common neurodegenerative diseases.
 
@@
 
-\newcommand{\lp}{\lambda_{d_i}^{\psi(k)}}
-\newcommand{\lpuu}{\lambda_{d_i}^{\psi(k),(u)}}
-\newcommand{\lpum}{\lambda_{d_i}^{\psi(k),(u-1)}}
+\begin{itemize}
+ \item Rare neurodegenerative diseases not well understood
+ \item Identify outcome measures and subjects for clinical trials
+\end{itemize}
+
+\columnbreak
+
+\section*{Challenges}
+
+% \begin{itemize}
+%  \item Datasets on rare neurodegenerative diseases (e.g., Posterior Cortical Atrophy) are unimodal (MRI only), cross-sectional and small.
+%  \item The continuous progression of non-MRI markers in rare neurodegenerative diseases is not well understood
+% \end{itemize}
+
+\begin{figure}[H]
+\fnt{
+\begin{subfigure}{0.48\columnwidth}
+  Typical Neurodeg. Diseases
 
   \begin{itemize}
- \item Datasets on rare neurodegenerative diseases (e.g., Posterior Cortical Atrophy) are unimodal (MRI only), cross-sectional and small.
- \item The continuous progression of non-MRI markers in rare neurodegenerative diseases is not well understood
+   \item Large datasets \yes
+   \item Multimodal imaging \yes
+   \item Longitudinal \yes
+%    \item Well-understood \yes
+
+  \end{itemize}
+\end{subfigure}
+\begin{subfigure}{0.48\columnwidth}
+
+   Rare Neurodeg. Diseases
+
+  \begin{itemize}
+   \item Small datasets \no
+   \item MRI only \no
+   \item Cross-sectional \no
+%    \item Poorly understood \no
   \end{itemize}
 
-% \vspace{-0.5em}
+\end{subfigure}
+ }
+\end{figure}
 
-% \begin{itemize}
-%  \item by transferring knowledge from larger multimodal datasets of typical Alzheimer's disease
-%  \item jointly model multiple diseases and multiple modalities
-% \end{itemize}
+\end{multicols}
+\hrule
+
+\begin{multicols}{2}							
+\raggedcolumns
 
@@
 
-% \begin{itemize}
-%  \item We propose a joint-disease model called Disease Knowledge Transfer:
 \vspace{1em}
@@
 %    \vspace{0.9em}
-   \includegraphics[width=\textwidth]{diseaseEvol}
+%    \includegraphics[width=\textwidth]{diseaseEvol}
 
@@
    \fnt{2. Dysfunction score modelled using region-specific biomarkers}\\
-   $ p(y_{ijk}|\theta_k, \lp, \beta_{i}, \epsilon_k) = N(y_{ijk}| g( \gamma_{ij}^{\psi(k)} ; \theta_k), \epsilon_k) $
+   $ y_{ijk} = g( \gamma_{ij}^{\psi(k)} ; \theta_k) + N(0,\epsilon_k) $
 %    \vspace{0.6em}
-   \includegraphics[width=\textwidth]{dysfuncEvol}
+%    \includegraphics[width=\textwidth]{dysfuncEvol}
   \end{subfigure}
 
+ \includegraphics[width=\columnwidth]{disease_knowledge_transfer_poster.pdf}
+  \fnt{3. Extend to multiple subjects, biomarkers and diseases}\\
+ $ p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \prod_{(i,j,k) \in \Omega} p(y_{ijk}|\theta_k, \lp, \beta_{i}) $
  \end{figure}
 
-%    \item each disease has a unique evolution of region-specific dysfunction scores
-%    \item each dysfunction score modelled as the unique evolution of biomarkers within that region
+\heading{Inference}: Perform loopy belief propagation
 
-% \end{itemize}
+% TODO bold out the diferences
 
-% \subsection{Method Overview}
+\newcommand{\uu}{^{(u)}}
+\newcommand{\um}{^{(u-1)}}
 
-% $ p(y_{ijk}|\theta_k, \lp, \beta_{i}, \epsilon_k) = N(y_{ijk}| g( \gamma_{ij}^{\psi(k)} ; \theta_k), \epsilon_k) $
+\newcommand{\algoFnt}[1]{\Large{#1}}
 
 \begin{figure}[H]
- \centering
-  \fnt{3. Extend to multiple subjects, biomarkers and diseases}\\
- $ p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \prod_{(i,j,k) \in \Omega} p(y_{ijk}|\theta_k, \lp, \beta_{i}) $
+\begin{algorithm}[H]
+\algoFnt
+ Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
+  \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
+   \tcp*[l]{\algoFnt{Estimate biomarker trajectories (disease agnostic)}}
+%     \For{$k=1$ to $K$}{
+      ${\theta_k\uu = \argmin_{\theta_k} \textcolor{red}{\sum_{(i,j) \in \Omega_k}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\right) \right]^2 }$\\
+%       ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
+%     }
+     \tcp*[l]{\algoFnt{Estimate dysfunction trajectories (disease specific)}} 
+%     \For{$d=1 \in \mathbb{D}$}{
+%       \For{$l=1 \in \Lambda$}{
+        ${\lambda_{d}^{l, (u)} = \argmin_{\lambda_{d}^{l}} \textcolor{red}{\sum_{(i,j,k) \in \Omega_{d,l}}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lambda_{d}^{l}) ; \theta_k\uu 
+        \right) \right]^2 }$\\
+%       }
+%     }
+    \tcp*[l]{\algoFnt{Estimate subject-specific time shifts}} 
+%     \For{$i=1 \in [1, \dots, S]$}{
+      ${\beta_i\uu = \argmin_{\beta_i} \textcolor{red}{\sum_{(j,k) \in \Omega_i}} \left[y_{ijk} - g\left(f(\beta_i + m_{ij}; \lpuu) ; \theta_k\uu
+      \right) \right]^2 }$\\
+%     }
+}
+% \normalfont
+\end{algorithm}
+% \caption[The algorithm for estimating the DKT parameters]{The algorithm used to estimate the DKT parameters, based on loopy belief-propagation.}
+\label{fig:dktAlgo}
+\end{figure}
 
- \includegraphics[width=\columnwidth]{../figures/disease_knowledge_transfer_symbols.pdf}
-%  \caption{}
-%  \label{fig:diagram}
-\end{figure}
+\heading{Demographics}
+
+\begin{itemize}
+ \item Dementia Research Center cohort: 76 PCA, 67 tAD, 87 age-matched controls.
+ \item TADPOLE Challenge dataset split into three cohorts with different progressions: hippocampal, cortical and subcortical.
+ \item Synthetic dataset mimicking the DRC cohort.
+\end{itemize}
+
+% The cohort from the Dementia Research Centre UK used for validation consisted of 10 subjects diagnosed with Posterior Cortical Atrophy, with a mean age of 59.4, 40\% females, as well as 10 age-matched controls with a mean age of 59.3, 50\% females.
+% 
+% 
+% For the validation on TADPOLE subgroups, we used applied the SuStaIn model on TADPOLE to split the population into three subgroups with different progression: hippocampal, cortical and subcortical subypes with prominent atrophy in the hippocampus, cortical and subcortical areas respectively.
+% The resulting subgroups had the following demographics:
+% 
+% \begin{table}
+%  \begin{tabular}{c | c c c c}
+% \textbf{Cohort} & \textbf{Nr. subjects} & \textbf{Nr. visits} & \textbf{Age (baseline)} & \textbf{Gender (\%F)}\\
+% \hline
+% Controls (Hippocampal) & 31 & 2.3 $\pm$ 1.8 & 74.4 $\pm$ 6.9 & 38\%\\
+% AD (Hippocampal) & 21 & 1.5 $\pm$ 0.8 & 74.5 $\pm$ 5.5 & 42\%\\
+% \hline
+% Controls (cortical) & 21 & 2.3 $\pm$ 1.3 & 70.9 $\pm$ 5.4 & 42\%\\
+% AD (cortical) & 35 & 1.7 $\pm$ 0.9 & 72.8 $\pm$ 7.4 & 28\%\\
+% \hline
+% Controls (subcortical) & 28 & 3.0 $\pm$ 1.5 & 73.7 $\pm$ 6.5 & 42\%\\
+% AD (subcortical) & 27 & 1.6 $\pm$ 0.9 & 73.7 $\pm$ 7.5 & 33\%\\
+%  \end{tabular}
+%  \caption{Demographics of the subjects in the three TADPOLE subgroups.}
+%  \label{tab:demogTadSubtypes}
+% \end{table}
 
@@
 \begin{figure}[H]
-\fnt{Synthetic experiment shows that the model can recover the underlying parameters}\\
-\includegraphics[width=0.4\textwidth]{../figures/compTrueParams105_synth1_JMD.pdf}
+\fnt{In synthetic experiment, the estimated parameters are close to the true parameters}\\
+\includegraphics[width=\columnwidth]{../figures/compTrueParams105_synth1_JMD.pdf}
   \label{fig:dktSynthTrajCompTrue}
@@
 \begin{figure}[H]
-\centering
-\fnt{Inferred trajectories for Posterior Cortical Atrophy}\\
- \includegraphics[width=0.4\textwidth, trim=0 20 0 0, clip]{../figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
+% \centering
+\fnt{Inferred multimodal trajectories for Posterior Cortical Atrophy are plausible, suggesting late-stage parietal and occipital damage. }\\
+ \includegraphics[width=\columnwidth, trim=0 20 0 0, clip]{../figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
 %  \caption{Estimated trajectories for the PCA cohort.
@@
 
-\begin{table}[H]
-\centering
-\fnt{Our model has favourable performance compared to other models}\\
-\vspace{1em}
+\fnt{Our model has favourable performance compared to other models, on two different datasets}\\
+\begin{figure}[H]
 \fontsize{27}{30}\selectfont
@@
 \label{sec:dktPerfMetrics}
-\end{table}
+\end{figure}
 
 \section*{Conclusion}
+
+\begin{itemize}
+ \item Developed a novel methodology and model for transfer learning across different diseases
+%  \item Inferred, likely for the first time, continuous multimodal trajectories for Posterior Cortical Atrophy
+%  \item 
+\end{itemize}
 
@@
 \begin{multicols}{2}
- 							%Use 3-column layout
+
+\Large{
+
 \raggedcolumns
 
-\subsection*{References}
+\subsection*{Weblinks}
+\begin{itemize}
+\item Source code: \url{https://github.com/mrazvan22/dkt}
+\item Personal website: \url{https://people.csail.mit.edu/razvan/}
+\end{itemize}
+}
 
-\large{
-\begin{tabular}{l p{0.5cm} l}
- 1. Fontejin et al., Neuroimg., 2012 & & 2. Young et al., Nat. Comms., 2018\\
- 3. Villemagne et al., Lancet Neurol., 2013 & & 4. Marinescu et al., IPMI, 2017\\
-\end{tabular}
+% \vspace{}
 
@@
 
-\subsection*{Weblinks}
-\begin{itemize}
-\item UCL Progression of Neurodegenerative Disease (POND): cmic.cs.ucl.ac.uk/pond/
-\item UCL Centre for Medical Image Computing: www.ucl.ac.uk/cmic/
-\end{itemize}
-}
+\subsection*{Funders and Grants}
 
+% \begin{figure}[H]
+% \begin{subfigure}{0.48\columnwidth}
 \vspace{1em}
 % \hspace{2em}
-\includegraphics[height=4.0cm]{epsrc_logo.jpg}
+\includegraphics[height=4.0cm]{epsrc_logo}
 \hspace{0.5em}
-\includegraphics[height=4.0cm]{pondLogo.png} 
+\includegraphics[height=4cm]{nac_logo} 
 \hspace{0.5em}
-\includegraphics[height=4cm]{cdt_logo.png} 
+\includegraphics[height=4cm]{cdt_logo} 
+\hspace{0.5em}
+\includegraphics[height=4cm]{nih_logo} 
 \hspace{0.5em}
 \includegraphics[height=4cm]{aruk_logo} 
+\hspace{0.5em}
+% \end{subfigure}
+% \begin{subfigure}{0.48\columnwidth}
+%  \begin{itemize}
+%  \item  EPSRC grant EP/L016478/1
+%  \item  NIH grant NIBIB NAC P41EB015902
+% \end{itemize}
+% \end{subfigure}
+
+% \end{figure}
 

### 2019-10-04 paper/pres/poster-dkt.tex
more edits on dkt poster

@@
 
-\author{R\u{a}zvan V.
-Marinescu\inst{1,2}, Marco Lorenzi\inst{3}, Stefano B.
-Blumberg\inst{1}, Alexandra L.
-Young\inst{1}, Pere Planell-Morell\inst{1}, Neil P.
-Oxtoby\inst{1},\\ Arman Eshaghi\inst{1,4}, Keir X.
-Young\inst{5}, Sebastian J.
-Crutch\inst{5}, Polina Golland\inst{2}, Daniel C.
-Alexander\inst{1}\\
+\author{\LARGE{R\u{a}zvan Marinescu\inst{1,2}, Marco Lorenzi\inst{3}, Stefano Blumberg\inst{1}, Alexandra Young\inst{1}, Pere Morell\inst{1}},\\ \LARGE{Neil Oxtoby\inst{1}, Arman Eshaghi\inst{1,4}, Keir Young\inst{5}, Sebastian Crutch\inst{5}, Polina Golland\inst{2}, Daniel Alexander\inst{1}}\\
 \begin{tabular}{l p{1cm} l p{1cm} l}
@@
 \end{minipage}
-\begin{textblock}{0}(10,0.2)
+\begin{textblock}{0}(12.2,0.2)
 \begin{tabular}{l}
+\includegraphics[height=5.5cm]{mit_logo}\includegraphics[height=6cm]{qr_dkt.png}\\
 \includegraphics[height=5cm]{ucl_logo}\\
-\includegraphics[height=5cm]{mit_logo}\\
 \end{tabular}
-\includegraphics[height=6cm]{qr_dkt.png}
 
@@
 
-\newcommand{\heading}[1]{\textbf{\Huge{#1}}}
+\newcommand{\heading}[1]{\noindent\textbf{\Huge{#1}}}
 
@@
 
-\heading{Aim}: Infer progression of non-MRI biomarkers in rare neurodegenerative diseases by leveraging larger datasets of common neurodegenerative diseases.
+\heading{Aim} Infer progression of non-MRI biomarkers in rare neurodegenerative diseases (NDs) by leveraging larger datasets of common NDs.
 
-\section*{Why}
-
-\begin{itemize}
- \item Rare neurodegenerative diseases not well understood
- \item Identify outcome measures and subjects for clinical trials
-\end{itemize}
+\heading{Why} Posterior Cortical Atrophy (PCA): progression of non-MRI biomarkers not known $\rightarrow$ Identify outcome measures and suitable subjects for PCA clinical trials
 
@@
 
-\section*{Challenges}
+\heading{Challenges}
 
@@
 
-\section*{Method}
+\heading{Method}
 
-\vspace{1em}
+\vspace{0.6em}
 \begin{figure}[H]
@@
 
-\heading{Inference}: Perform loopy belief propagation
+\heading{Inference:} Loopy belief propagation
 
@@
 \begin{itemize}
- \item Dementia Research Center cohort: 76 PCA, 67 tAD, 87 age-matched controls.
- \item TADPOLE Challenge dataset split into three cohorts with different progressions: hippocampal, cortical and subcortical.
- \item Synthetic dataset mimicking the DRC cohort.
+ \item Dementia Research Center cohort: 76 PCA, 67 tAD, 87 age-matched controls for training, 10 PCA with DTI for validation
+ \item TADPOLE Challenge dataset split into three subgroups with different progressions: 21 hippocampal, 35 cortical, 27 subcortical.
+ \item Synthetic dataset mimicking the DRC cohort: 50 subjects with "synthetic PCA", 100 subjects with "synthetic AD".
 \end{itemize}
@@
 
-\section*{Results}
+\columnbreak
 
+\heading{Results}
+
+\vspace{0.6em}
 \begin{figure}[H]
@@
 % \centering
-\fnt{Inferred multimodal trajectories for Posterior Cortical Atrophy are plausible, suggesting late-stage parietal and occipital damage. }\\
- \includegraphics[width=\columnwidth, trim=0 20 0 0, clip]{../figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
+\fnt{Inferred multimodal trajectories for Posterior Cortical Atrophy are plausible, suggesting late-stage parietal and occipital damage. }
+ \includegraphics[width=\columnwidth, trim=0 20 0 35, clip]{../figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
 %  \caption{Estimated trajectories for the PCA cohort.
@@
 
-\fnt{Our model has favourable performance compared to other models, on two different datasets}\\
+\noindent \fnt{Our model has favourable performance compared to other models, on two different datasets}
+\vspace{0.3em}
 \begin{figure}[H]
@@
 
-\section*{Conclusion}
+\heading{Conclusion}
 
@@
  \item Developed a novel methodology and model for transfer learning across different diseases
-%  \item Inferred, likely for the first time, continuous multimodal trajectories for Posterior Cortical Atrophy
+ \item Inferred multimodal trajectories for Posterior Cortical Atrophy
 %  \item 
@@
 
+\vspace{-1em}
 \subsection*{Weblinks}
@@
 \item Source code: \url{https://github.com/mrazvan22/dkt}
-\item Personal website: \url{https://people.csail.mit.edu/razvan/}
+\item Website: \url{https://people.csail.mit.edu/razvan/}
 \end{itemize}
@@
 
-% \vspace{}
-
 \columnbreak
 
-\subsection*{Funders and Grants}
+\subsection*{Funders}
 
@@
 % \begin{subfigure}{0.48\columnwidth}
-\vspace{1em}
+\vspace{0em}
 % \hspace{2em}
-\includegraphics[height=4.0cm]{epsrc_logo}
+\newcommand{\heightLogos}{3.5cm}
+\includegraphics[height=\heightLogos]{epsrc_logo}
 \hspace{0.5em}
-\includegraphics[height=4cm]{nac_logo} 
+\includegraphics[height=\heightLogos]{nac_logo} 
 \hspace{0.5em}
-\includegraphics[height=4cm]{cdt_logo} 
+\includegraphics[height=\heightLogos]{cdt_logo} 
 \hspace{0.5em}
-\includegraphics[height=4cm]{nih_logo} 
+\includegraphics[height=\heightLogos]{nih_logo} 
 \hspace{0.5em}
-\includegraphics[height=4cm]{aruk_logo} 
+\includegraphics[height=\heightLogos]{aruk_logo} 
 \hspace{0.5em}

### 2019-10-04 paper/pres/poster-dkt.tex
updates

@@
 
-\author{\LARGE{R\u{a}zvan Marinescu\inst{1,2}, Marco Lorenzi\inst{3}, Stefano Blumberg\inst{1}, Alexandra Young\inst{1}, Pere Morell\inst{1}},\\ \LARGE{Neil Oxtoby\inst{1}, Arman Eshaghi\inst{1,4}, Keir Young\inst{5}, Sebastian Crutch\inst{5}, Polina Golland\inst{2}, Daniel Alexander\inst{1}}\\
+\author{\LARGE{R\u{a}zvan Marinescu\inst{1,2}, Marco Lorenzi\inst{3}, Stefano Blumberg\inst{1}, Alexandra Young\inst{1}, Pere Morell\inst{1}},\\ \LARGE{Neil Oxtoby\inst{1}, Arman Eshaghi\inst{1,4}, Keir Yong\inst{5}, Sebastian Crutch\inst{5}, Polina Golland\inst{2}, Daniel Alexander\inst{1}}\\
 \begin{tabular}{l p{1cm} l p{1cm} l}
@@
 \begin{document}
-% \hspace{-3cm}								%Align with edge of page, not margin
+							%Align with edge of page, not margin
 % \vspace{2cm}
 % \includegraphics[scale=1,trim=0 0 0 200, clip]{Black_Landscape.pdf}
-
+\hspace{-1cm}	
 \begin{minipage}{50cm}					%Minipage for title contents
@@
    \centering
-   \fnt{1. Each disease characterised by region-specific dysfunction profile}\\
+   \fnt{1. Each disease characterised by region-specific dysfunction trajectories}\\
    $ \gamma_{ij}^l = f(\beta_{i} + m_{ij}; \lambda_{d_i}^l) $\\
@@
 %    \vspace{-1.2em}
-   \fnt{2. Dysfunction score modelled using region-specific biomarkers}\\
+   \fnt{2. Dysfunction trajectory modelled using region-specific biomarkers}\\
    $ y_{ijk} = g( \gamma_{ij}^{\psi(k)} ; \theta_k) + N(0,\epsilon_k) $
@@
 
-\heading{Demographics}
+\heading{Datasets}
 
 \begin{itemize}
- \item Dementia Research Center cohort: 76 PCA, 67 tAD, 87 age-matched controls for training, 10 PCA with DTI for validation
- \item TADPOLE Challenge dataset split into three subgroups with different progressions: 21 hippocampal, 35 cortical, 27 subcortical.
- \item Synthetic dataset mimicking the DRC cohort: 50 subjects with "synthetic PCA", 100 subjects with "synthetic AD".
+ \item Dementia Research Center cohort: MRI scans from 76 PCA, 67 tAD, 87 controls for training, 10 PCA with DTI for validation.
+ \item TADPOLE dataset (ADNI) split into three subgroups with different progressions: 21 hippocampal, 35 cortical, 27 subcortical.
+ \item Synthetic dataset mimicking the cohorts above: 50 subjects with "synthetic PCA", 100 subjects with "synthetic AD".
 \end{itemize}
@@
 \vspace{0.6em}
+\begin{itemize}
+ \item \fnt{In synthetic experiment, the estimated parameters are close to the true parameters.}
 \begin{figure}[H]
-\fnt{In synthetic experiment, the estimated parameters are close to the true parameters}\\
 \includegraphics[width=\columnwidth]{../figures/compTrueParams105_synth1_JMD.pdf}
@@
 
+ \item \fnt{Inferred multimodal trajectories for PCA in lack of such data.}
+ \item \fnt{Results are plausible, suggesting late-stage posterior damage. }
 \begin{figure}[H]
 % \centering
-\fnt{Inferred multimodal trajectories for Posterior Cortical Atrophy are plausible, suggesting late-stage parietal and occipital damage. }
+
  \includegraphics[width=\columnwidth, trim=0 20 0 35, clip]{../figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
@@
 
-\noindent \fnt{Our model has favourable performance compared to other models, on two different datasets}
+\item \fnt{Our model has favourable performance compared to other models, on two different datasets.}
 \vspace{0.3em}
@@
 \end{figure}
+
+\end{itemize}
 

### 2019-10-08 paper/pres/poster-dkt.tex
2nd rehearsal DKT

@@
 
-\author{\LARGE{R\u{a}zvan Marinescu\inst{1,2}, Marco Lorenzi\inst{3}, Stefano Blumberg\inst{1}, Alexandra Young\inst{1}, Pere Morell\inst{1}},\\ \LARGE{Neil Oxtoby\inst{1}, Arman Eshaghi\inst{1,4}, Keir Yong\inst{5}, Sebastian Crutch\inst{5}, Polina Golland\inst{2}, Daniel Alexander\inst{1}}\\
-\begin{tabular}{l p{1cm} l p{1cm} l}
-\large{$^1$Centre for Medical Image Computing, UCL, UK}  & & \large{$^3$University of C\^{o}te d'Azur, Inria Sophia Antipolis, France} & & \large{$^5$Dementia Research Centre, UCL, UK}\\
-\large{$^2$Computer Science and Artificial Intelligence Laboratory, MIT, USA} & & \large{$^4$Queen Square MS Centre, UCL Institute of Neurology, UK} & & 
- \\
-\end{tabular}
-}
+% \author{\LARGE{R\u{a}zvan Marinescu\inst{1,2}, Marco Lorenzi\inst{3}, Stefano Blumberg\inst{1}, Alexandra Young\inst{1}, Pere Morell\inst{1}},\\ \LARGE{Neil Oxtoby\inst{1}, Arman Eshaghi\inst{1,4}, Keir Yong\inst{5}, Sebastian Crutch\inst{5}, Polina Golland\inst{2}, Daniel Alexander\inst{1}}\\}
+
+\author{\LARGE{R\u{a}zvan V.
+Marinescu, Marco Lorenzi, Stefano B.
+Blumberg, Alexandra L.
+Young, Pere Morell},\\ \LARGE{Neil Oxtoby, Arman Eshaghi, Keir Yong, Sebastian Crutch, Polina Golland, Daniel C.
+Alexander}\\}
+
+% \begin{tabular}{l p{1cm} l p{1cm} l}
+% \large{$^1$Centre for Medical Image Computing, UCL, UK}  & & \large{$^3$University of C\^{o}te d'Azur, Inria Sophia Antipolis, France} & & \large{$^5$Dementia Research Centre, UCL, UK}\\
+% \large{$^2$Computer Science and Artificial Intelligence Laboratory, MIT, USA} & & \large{$^4$Queen Square MS Centre, UCL Institute of Neurology, UK} & & 
+%  \\
+% \end{tabular}
 
@@
 \end{minipage}
-\begin{textblock}{0}(12.2,0.2)
+\begin{textblock}{0}(9.95,0.2)
 \begin{tabular}{l}
-\includegraphics[height=5.5cm]{mit_logo}\includegraphics[height=6cm]{qr_dkt.png}\\
-\includegraphics[height=5cm]{ucl_logo}\\
+\includegraphics[height=5.5cm]{mit_logo}\includegraphics[height=5.5cm]{inria_logo}\includegraphics[height=6cm]{qr_dkt.png}\\
+\hspace{11.45cm}\includegraphics[height=5cm]{ucl_logo}\\
 \end{tabular}
@@
 
-\newcommand{\lp}{\lambda_{d_i}^{\psi(k)}}
-\newcommand{\lpuu}{\lambda_{d_i}^{\psi(k),(u)}}
-\newcommand{\lpum}{\lambda_{d_i}^{\psi(k),(u-1)}}
+\newcommand{\lp}{\lambda_{d_i}^{k}}
+\newcommand{\lpuu}{\lambda_{d_i}^{k,(u)}}
+\newcommand{\lpum}{\lambda_{d_i}^{k,(u-1)}}
 
@@
 
+\vspace{0.2em}
 \hrule
+\vspace{0.2em}
 
@@
 
-\heading{Aim} Infer progression of non-MRI biomarkers in rare neurodegenerative diseases (NDs) by leveraging larger datasets of common NDs.
+\heading{Aim}\;\;\;\;Infer progression of multimodal biomarkers in rare neurodegenerative diseases (NDs) by leveraging larger datasets of common NDs.
 
-\heading{Why} Posterior Cortical Atrophy (PCA): progression of non-MRI biomarkers not known $\rightarrow$ Identify outcome measures and suitable subjects for PCA clinical trials
+\heading{Why}\;\;\;\;Posterior Cortical Atrophy (PCA): progression of \underline{multimodal} biomarkers not known $\rightarrow$ Identify outcome measures and suitable subjects for PCA clinical trials
 
@@
 
-\heading{Challenges}
-
+% \heading{Challenges}
+% 
+% \begin{figure}[H]
+% \fnt{
+% 
+% \begin{subfigure}{0.48\columnwidth}
+%  
+%    Rare Neurodeg. Diseases
+%   
 %   \begin{itemize}
-%  \item Datasets on rare neurodegenerative diseases (e.g., Posterior Cortical Atrophy) are unimodal (MRI only), cross-sectional and small.
-%  \item The continuous progression of non-MRI markers in rare neurodegenerative diseases is not well understood
+%    \item Small datasets \no
+%    \item MRI only \no
+%    \item Cross-sectional \no
+% %    \item Poorly understood \no
 %   \end{itemize}
+%  
+% \end{subfigure}
+%  }
+% \end{figure}
 
 \begin{figure}[H]
-\fnt{
-\begin{subfigure}{0.48\columnwidth}
-  Typical Neurodeg. Diseases
-
-  \begin{itemize}
-   \item Large datasets \yes
-   \item Multimodal imaging \yes
-   \item Longitudinal \yes
-%    \item Well-understood \yes
-
-  \end{itemize}
-\end{subfigure}
-\begin{subfigure}{0.48\columnwidth}
-
-   Rare Neurodeg. Diseases
-
-  \begin{itemize}
-   \item Small datasets \no
-   \item MRI only \no
-   \item Cross-sectional \no
-%    \item Poorly understood \no
-  \end{itemize}
-
-\end{subfigure}
- }
+ \centering
+ \includegraphics[width=\columnwidth]{DKT_data_availability}
 \end{figure}
@@
 \hrule
+\vspace{0.2em}
 
@@
 
-\heading{Method}
+\heading{1. Intuition}
+
+\begin{itemize}
+ \item Diseases affect different brain regions un-equally, but underlyining mechanisms are the same (amyloid cascade)
+ \item \textbf{Idea:} each brain region follows ``its own disease course'', common across diseases $\rightarrow$ dysfunction score
+\end{itemize}
+
+\begin{figure}[H]
+
+\includegraphics[width=\columnwidth]{DKT_intuition}
+
+\end{figure}
+
+\heading{2. Method}
 
@@
    \fnt{2. Dysfunction trajectory modelled using region-specific biomarkers}\\
-   $ y_{ijk} = g( \gamma_{ij}^{\psi(k)} ; \theta_k) + N(0,\epsilon_k) $
+   $ y_{ijk} = g( \gamma_{ij}^{k} ; \theta_k) + N(0,\epsilon_k) $
 %    \vspace{0.6em}
@@
 
-\heading{Inference:} Loopy belief propagation
+\columnbreak
+
+\heading{3. Inference with belief propagation}
 
@@
 
+% \begin{itemize}
+%  \item Uses loopy belief propagation
+% \end{itemize}
+
 \begin{figure}[H]
 \begin{algorithm}[H]
-\algoFnt
- Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
+\algoFnt{
+%  Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
   \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
@@
 }
+}
 % \normalfont
@@
 
-\heading{Datasets}
+\heading{4. Datasets}
 
@@
  \item TADPOLE dataset (ADNI) split into three subgroups with different progressions: 21 hippocampal, 35 cortical, 27 subcortical.
- \item Synthetic dataset mimicking the cohorts above: 50 subjects with "synthetic PCA", 100 subjects with "synthetic AD".
+%  \item Synthetic dataset mimicking the cohorts above: 50 subjects with "synthetic PCA", 100 subjects with "synthetic AD".
 \end{itemize}
 
-% The cohort from the Dementia Research Centre UK used for validation consisted of 10 subjects diagnosed with Posterior Cortical Atrophy, with a mean age of 59.4, 40\% females, as well as 10 age-matched controls with a mean age of 59.3, 50\% females.
-% 
-% 
-% For the validation on TADPOLE subgroups, we used applied the SuStaIn model on TADPOLE to split the population into three subgroups with different progression: hippocampal, cortical and subcortical subypes with prominent atrophy in the hippocampus, cortical and subcortical areas respectively.
-% The resulting subgroups had the following demographics:
-% 
-% \begin{table}
-%  \begin{tabular}{c | c c c c}
-% \textbf{Cohort} & \textbf{Nr. subjects} & \textbf{Nr. visits} & \textbf{Age (baseline)} & \textbf{Gender (\%F)}\\
-% \hline
-% Controls (Hippocampal) & 31 & 2.3 $\pm$ 1.8 & 74.4 $\pm$ 6.9 & 38\%\\
-% AD (Hippocampal) & 21 & 1.5 $\pm$ 0.8 & 74.5 $\pm$ 5.5 & 42\%\\
-% \hline
-% Controls (cortical) & 21 & 2.3 $\pm$ 1.3 & 70.9 $\pm$ 5.4 & 42\%\\
-% AD (cortical) & 35 & 1.7 $\pm$ 0.9 & 72.8 $\pm$ 7.4 & 28\%\\
-% \hline
-% Controls (subcortical) & 28 & 3.0 $\pm$ 1.5 & 73.7 $\pm$ 6.5 & 42\%\\
-% AD (subcortical) & 27 & 1.6 $\pm$ 0.9 & 73.7 $\pm$ 7.5 & 33\%\\
-%  \end{tabular}
-%  \caption{Demographics of the subjects in the three TADPOLE subgroups.}
-%  \label{tab:demogTadSubtypes}
-% \end{table}
+\vspace{0.6em}
+\heading{5. Results}
 
-\columnbreak
-
-\heading{Results}
-
-\vspace{0.6em}
+% \vspace{0.6em}
 \begin{itemize}
- \item \fnt{In synthetic experiment, the estimated parameters are close to the true parameters.}
-\begin{figure}[H]
-\includegraphics[width=\columnwidth]{../figures/compTrueParams105_synth1_JMD.pdf}
-  \label{fig:dktSynthTrajCompTrue}
-\end{figure}
+%  \item \fnt{In synthetic experiment, the estimated parameters are close to the true parameters.}
+% \begin{figure}[H]
+% \includegraphics[width=\columnwidth]{../figures/compTrueParams105_synth1_JMD.pdf}
+%   \label{fig:dktSynthTrajCompTrue}
+% \end{figure}
 
@@
 DKT (ours) &      0.56 $\pm$ 0.23 &    \textbf{0.35 $\pm$ 0.17} &        \textbf{0.58 $\pm$ 0.14} &     -0.10 $\pm$ 0.29 &     \textbf{0.71 $\pm$ 0.11} &     \textbf{0.34 $\pm$ 0.26} \\
-Latent stage &      0.44 $\pm$ 0.25 &    0.34 $\pm$ 0.21 &       0.34 $\pm$ 0.24* &     \textbf{-0.07 $\pm$ 0.22} &     0.64 $\pm$ 0.16 &    0.08 $\pm$ 0.24* \\
+AD model &      0.44 $\pm$ 0.25 &    0.34 $\pm$ 0.21 &       0.34 $\pm$ 0.24* &     \textbf{-0.07 $\pm$ 0.22} &     0.64 $\pm$ 0.16 &    0.08 $\pm$ 0.24* \\
 Multivariate &      \textbf{0.60 $\pm$ 0.18} &   0.11 $\pm$ 0.22* &       0.12 $\pm$ 0.29* &     -0.22 $\pm$ 0.22 &   -0.44 $\pm$ 0.14* &   -0.32 $\pm$ 0.29* \\
@@
 DKT (ours) &    0.77 $\pm$ 0.11 &    0.39 $\pm$ 0.26 &      0.75 $\pm$ 0.09 &    0.60 $\pm$ 0.14 &    \textbf{0.55 $\pm$ 0.24} &    \textbf{0.35 $\pm$ 0.22} \\
-Latent stage &    \textbf{0.80 $\pm$ 0.09} &    \textbf{0.53 $\pm$ 0.17} &      \textbf{0.80 $\pm$ 0.12} &    0.56 $\pm$ 0.18 &    0.50 $\pm$ 0.21 &    0.32 $\pm$ 0.24 \\
+AD model &    \textbf{0.80 $\pm$ 0.09} &    \textbf{0.53 $\pm$ 0.17} &      \textbf{0.80 $\pm$ 0.12} &    0.56 $\pm$ 0.18 &    0.50 $\pm$ 0.21 &    0.32 $\pm$ 0.24 \\
 Multivariate &   0.73 $\pm$ 0.09 &   0.45 $\pm$ 0.22  &    0.71 $\pm$ 0.08 & -0.28 $\pm$ 0.21* &  0.53 $\pm$ 0.22  &  0.25 $\pm$ 0.23* \\
@@
 
-\heading{Conclusion}
+\heading{6. Conclusion}
 

### 2019-10-08 paper/pres/poster-dkt.tex
minor

@@
 
-\heading{Aim}\;\;\;\;Infer progression of multimodal biomarkers in rare neurodegenerative diseases (NDs) by leveraging larger datasets of common NDs.
+\heading{Aim:}\;\;\;\;Infer progression of multimodal biomarkers in rare neurodegenerative diseases (NDs) by leveraging larger datasets of common NDs.
 
-\heading{Why}\;\;\;\;Posterior Cortical Atrophy (PCA): progression of \underline{multimodal} biomarkers not known $\rightarrow$ Identify outcome measures and suitable subjects for PCA clinical trials
+\heading{Why:}\;\;\;\;Posterior Cortical Atrophy (PCA): progression of \underline{multimodal} biomarkers not known $\rightarrow$ Identify outcome measures and suitable subjects for PCA clinical trials.
 
@@
 \begin{itemize}
- \item Diseases affect different brain regions un-equally, but underlyining mechanisms are the same (amyloid cascade)
- \item \textbf{Idea:} each brain region follows ``its own disease course'', common across diseases $\rightarrow$ dysfunction score
+ \item Diseases affect different brain regions un-equally, but underlyining mechanisms are the same (amyloid cascade).
+ \item \textbf{Idea:} each brain region follows ``its own disease course'', common across diseases.
 \end{itemize}

### 2019-10-09 paper/pres/poster-dkt.tex
last changes for poster

--- 2019-10-08 paper/pres/poster-dkt.tex
@@
 
-\heading{Aim}\;\;\;\;Infer progression of multimodal biomarkers in rare neurodegenerative diseases (NDs) by leveraging larger datasets of common NDs.
+\heading{Aim}\;\;\;\;Infer progression of multimodal biomarkers in rare neurodegenerative diseases (NDs) by leveraging larger datasets of common, related variants.
 
@@
 \begin{figure}[H]
-
-\includegraphics[width=\columnwidth]{DKT_intuition}
+\centering
+\includegraphics[width=1\columnwidth]{DKT_intuition}
 
@@
    \fnt{2. Dysfunction trajectory modelled using region-specific biomarkers}\\
-   $ y_{ijk} = g( \gamma_{ij}^{k} ; \theta_k) + N(0,\epsilon_k) $
+   $ y_{ij}^k = g( \gamma_{ij}^{k} ; \theta_k) + N(0,\epsilon_k) $\\
+%    \fnt{$y_{ijk}$ = measure in subject $i$, visit $j$, region $k$.}
+
 %    \vspace{0.6em}
@@
   \end{subfigure}
+% \vspace{-2em}
 
- \includegraphics[width=\columnwidth]{disease_knowledge_transfer_poster.pdf}
+ \includegraphics[width=\columnwidth,trim=0 0 0 00,clip]{disease_knowledge_transfer_poster.pdf}
+
   \fnt{3. Extend to multiple subjects, biomarkers and diseases}\\
- $ p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \prod_{(i,j,k) \in \Omega} p(y_{ijk}|\theta_k, \lp, \beta_{i}) $
+ $ p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \prod_{(i,j,k) \in \Omega} p(y_{ij}^k|\theta_k, \lp, \beta_{i}) $
  \end{figure}
@@
 \begin{itemize}
+ \item TADPOLE dataset (ADNI) split into three subgroups with different progressions: 21 hippocampal, 35 cortical, 27 subcortical.
 \item Dementia Research Center cohort: MRI scans from 76 PCA, 67 tAD, 87 controls for training, 10 PCA with DTI for validation.
- \item TADPOLE dataset (ADNI) split into three subgroups with different progressions: 21 hippocampal, 35 cortical, 27 subcortical.
 %  \item Synthetic dataset mimicking the cohorts above: 50 subjects with "synthetic PCA", 100 subjects with "synthetic AD".
@@
 % \end{figure}
-
- \item \fnt{Inferred multimodal trajectories for PCA in lack of such data.}
- \item \fnt{Results are plausible, suggesting late-stage posterior damage. }
-\begin{figure}[H]
-% \centering
-
- \includegraphics[width=\columnwidth, trim=0 20 0 35, clip]{../figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
-%  \caption{Estimated trajectories for the PCA cohort.
-The only data that were available were the MRI volumetric data.
-The dynamics of the other biomarkers has been inferred by the model using data from typical AD, and taking into account the different spatial distribution of pathology in PCA vs tAD.
-}
-%  \label{fig:PCAtrajByModality}
-\end{figure}
 
@@
 \label{sec:dktPerfMetrics}
+\end{figure}
+
+ \item \fnt{Inferred multimodal trajectories for PCA in lack of such data.}
+ \item \fnt{Results are plausible, suggesting late-stage posterior damage. }
+\begin{figure}[H]
+% \centering
+
+ \includegraphics[width=\columnwidth, trim=0 20 0 35, clip]{../figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
+%  \caption{Estimated trajectories for the PCA cohort.
+The only data that were available were the MRI volumetric data.
+The dynamics of the other biomarkers has been inferred by the model using data from typical AD, and taking into account the different spatial distribution of pathology in PCA vs tAD.
+}
+%  \label{fig:PCAtrajByModality}
 \end{figure}

### 2019-10-10 Plotter.py
more on dkt poster

@@ -456,7 +456,7 @@ class PlotterJDM:
 
 
 
-    font = {'family': 'normal', 'size': 13}
+    font = {'family': 'normal', 'size': 22}
 
     import matplotlib
     matplotlib.rc('font', **font)
@@ -484,10 +484,14 @@ class PlotterJDM:
     colorsTraj = [colorsys.hsv_to_rgb(hue, 1, 1) for hue in
       np.linspace(0, 1, num=nrBiomkPerGroup, endpoint=False)]
 
+    groupReplaceMap = {'AV45': 'Amyloid (AV45)', 'AV1451': 'Tau (AV1451)', 'Volume': 'MRI Atrophy'}
+
     for g in range(nrBiomkGroups):
       biomkIndInGr[g] = np.array(range(g * nrBiomkPerGroup, (g+1)*nrBiomkPerGroup))
 
       groupName = params['labels'][biomkIndInGr[g][0]].split(' ')[0]
+      if groupName in groupReplaceMap.keys():
+        groupName = groupReplaceMap[groupName]
 
       # print('groupName', groupName)
       # print('biomkIndInGr[g]', biomkIndInGr[g])
@@ -553,8 +557,11 @@ class PlotterJDM:
       #   for x in group:
       #     x.set_visible(False)
 
+
+    pl.subplots_adjust(wspace=0.3, hspace=0.3)
+
     fig.text(0.55, 0.04, 'Disease Progression (months)', ha='center')
-    fig.text(0.08, 0.5, 'Biomarker Value (normalised)', va='center', rotation='vertical')
+    fig.text(0.04, 0.5, 'Biomarker Value (normalised)', va='center', rotation='vertical')
 
     if replaceFig:
       fig.show()

### 2019-10-10 drcValidFuncs.py

@@ -690,13 +690,16 @@ def validateDRCBiomk(dpmObj, params):
   diagValidFiltDktModel[diagValidFiltDktModel == CTL2] = CTL_DKT
   diagValidFiltDktModel[diagValidFiltDktModel == PCA] = PCA_DKT
 
-  # plot just the trajectories by modality groups
-  # for d in range(dpmObj.nrDis):
-  #   fig = dpmObj.plotter.plotTrajInDisSpaceOverlap(dpmObj, d, params, replaceFig=True)
-  #   fig.savefig('%s/trajDisSpaceOverlap_%s_%s.png' % (params['outFolder'],
-  #     params['disLabels'][d], params['expName']))
+  #plot just the trajectories by modality groups
+  for d in range(dpmObj.nrDis):
+    fig = dpmObj.plotter.plotTrajInDisSpaceOverlap(dpmObj, d, params, replaceFig=True)
+    figName = '%s/trajDisSpaceOverlap_%s_%s' % (params['outFolder'],
+      params['disLabels'][d], params['expName'])
+    fig.savefig('%s.png' % figName)
+    fig.savefig('%s.pdf' % figName)
+    print('Fig saved:%s.pdf ' % figName)
 
-  plotFigs = True
+  plotFigs = False
   if plotFigs:
 
     # for u in range(dpmObj.nrFuncUnits):

### 2021-03-12 paper/pres/poster-dkt.tex
more on dkt poster
un-commited edits to poster

@@
 \usepackage[noend]{algorithm2e}
-\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{blue}{#1}}
+\definecolor{comCol}{rgb}{0.0,0.0,0.0}
+\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{comCol}{#1}}
 \SetCommentSty{mycommfont}
@@
 
-\newcommand{\lp}{\lambda_{d_i}^{k}}
-\newcommand{\lpuu}{\lambda_{d_i}^{k,(u)}}
-\newcommand{\lpum}{\lambda_{d_i}^{k,(u-1)}}
+\definecolor{thetaCol}{rgb}{0.8,0.2,0.2}
+\definecolor{lambdaCol}{rgb}{0.0,0.5,0.0}
+\definecolor{betaCol}{rgb}{0,0.2,0.8}
+
+\newcommand{\lp}{\textcolor{lambdaCol}{\lambda_{d_i}^{\psi(k)}}}
+\newcommand{\lpuu}{\textcolor{lambdaCol}{\lambda_{d_i}^{\psi(k),(u)}}}
+\newcommand{\lpum}{\textcolor{lambdaCol}{\lambda_{d_i}^{\psi(k),(u-1)}}}
 
@@
 
-\heading{Why}\;\;\;\;Posterior Cortical Atrophy (PCA): progression of \underline{multimodal} biomarkers not known $\rightarrow$ Identify outcome measures and suitable subjects for PCA clinical trials
+\heading{Why}\;\;\;\;Posterior Cortical Atrophy (PCA): progression of \underline{multimodal} biomarkers not known $\rightarrow$ Identify outcome measures and suitable subjects for PCA clinical trials.
 
@@
 \begin{itemize}
- \item Diseases affect different brain regions un-equally, but underlyining mechanisms are the same (amyloid cascade)
- \item \textbf{Idea:} each brain region follows ``its own disease course'', common across diseases $\rightarrow$ dysfunction score
+ \item Diseases affect different brain regions un-equally, but underlyining mechanisms are the same (amyloid cascade).
+ \item \textbf{Idea:} a brain region follows ``the same pathology course'' for both diseases, but to different extents and with different speed.
 \end{itemize}
@@
    \centering
-   \fnt{1. Each disease characterised by region-specific dysfunction trajectories}\\
-   $ \gamma_{ij}^l = f(\beta_{i} + m_{ij}; \lambda_{d_i}^l) $\\
+   \fnt{1. Each disease characterised by region-specific \textcolor{lambdaCol}{dysfunction trajectories}}\\
+   $ \gamma_{ij}^l = f(\textcolor{betaCol}{\beta_{i}} + m_{ij}; \textcolor{lambdaCol}{\lambda_{d_i}^l}) $\\
 %    \vspace{0.9em}
@@
 %    \vspace{-1.2em}
-   \fnt{2. Dysfunction trajectory modelled using region-specific biomarkers}\\
-   $ y_{ij}^k = g( \gamma_{ij}^{k} ; \theta_k) + N(0,\epsilon_k) $\\
-%    \fnt{$y_{ijk}$ = measure in subject $i$, visit $j$, region $k$.}
+   \fnt{2. Dysfunction trajectory modelled using region-specific \textcolor{thetaCol}{biomarker trajectories}}\\
+   $ y_{ij}^k = g( \gamma_{ij}^{\psi(k)} ; \textcolor{thetaCol}{\theta_k}) + N(0,\epsilon_k) $\\
+%    \fnt{$y_{ij}^k$ = measure in subject $i$, visit $j$, region $k$.}
 
@@
   \fnt{3. Extend to multiple subjects, biomarkers and diseases}\\
- $ p(\boldsymbol{y}|\theta, \lambda, \beta , \epsilon) = \prod_{(i,j,k) \in \Omega} p(y_{ij}^k|\theta_k, \lp, \beta_{i}) $
+ $ p(\boldsymbol{y}|\textcolor{thetaCol}{\theta}, \textcolor{lambdaCol}{\lambda}, \textcolor{betaCol}{\beta}, \epsilon) = \prod_{(i,j,k) \in \Omega} p(y_{ij}^k|\textcolor{thetaCol}{\theta_k}, \lp, \textcolor{betaCol}{\beta_{i}}) $
  \end{figure}
@@
 %  Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
-  \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
-   \tcp*[l]{\algoFnt{Estimate biomarker trajectories (disease agnostic)}}
+  \While{$\boldsymbol{\textcolor{thetaCol}{\theta}}$, $\boldsymbol{\textcolor{lambdaCol}{\lambda}}$, $\boldsymbol{\textcolor{betaCol}{\beta}}$ not converged}{
+   \tcp*[l]{\algoFnt{Estimate \textcolor{thetaCol}{biomarker trajectories} (disease agnostic)}}
 %     \For{$k=1$ to $K$}{
-      ${\theta_k\uu = \argmin_{\theta_k} \textcolor{red}{\sum_{(i,j) \in \Omega_k}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\right) \right]^2 }$\\
-%       ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
+      ${\textcolor{thetaCol}{\theta_k\uu} = \argmin_{\textcolor{thetaCol}{\theta_k}} \textcolor{black}{\sum_{(i,j) \in \Omega_k}} \left[y_{ij}^k - g\left(f(\textcolor{betaCol}{\beta_i\um} + m_{ij}; \lpum) ; \textcolor{thetaCol}{\theta_k}\right) \right]^2 }$\\
+%       ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ij}^k - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
 %     }
-     \tcp*[l]{\algoFnt{Estimate dysfunction trajectories (disease specific)}} 
+     \tcp*[l]{\algoFnt{Estimate \textcolor{lambdaCol}{dysfunction trajectories} (disease specific)}} 
 %     \For{$d=1 \in \mathbb{D}$}{
 %       \For{$l=1 \in \Lambda$}{
-        ${\lambda_{d}^{l, (u)} = \argmin_{\lambda_{d}^{l}} \textcolor{red}{\sum_{(i,j,k) \in \Omega_{d,l}}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lambda_{d}^{l}) ; \theta_k\uu 
+        ${\textcolor{lambdaCol}{\lambda_{d}^{l, (u)}} = \argmin_{\textcolor{lambdaCol}{\lambda_{d}^{l}}} \textcolor{black}{\sum_{(i,j,k) \in \Omega_{d,l}}} \left[y_{ij}^k - g\left(f(\textcolor{betaCol}{\beta_i\um} + m_{ij}; \textcolor{lambdaCol}{\lambda_{d}^{l}}) ; \textcolor{thetaCol}{\theta_k\uu}
         \right) \right]^2 }$\\
@@
 %     }
-    \tcp*[l]{\algoFnt{Estimate subject-specific time shifts}} 
+    \tcp*[l]{\algoFnt{Estimate \textcolor{betaCol}{subject-specific time shifts}}} 
 %     \For{$i=1 \in [1, \dots, S]$}{
-      ${\beta_i\uu = \argmin_{\beta_i} \textcolor{red}{\sum_{(j,k) \in \Omega_i}} \left[y_{ijk} - g\left(f(\beta_i + m_{ij}; \lpuu) ; \theta_k\uu
+      ${\textcolor{betaCol}{\beta_i\uu} = \argmin_{\textcolor{betaCol}{\beta_i}} \textcolor{black}{\sum_{(j,k) \in \Omega_i}} \left[y_{ij}^k - g\left(f(\textcolor{betaCol}{\beta_i} + m_{ij}; \lpuu) ; \textcolor{thetaCol}{\theta_k\uu}
       \right) \right]^2 }$\\
@@
 
+%  \includegraphics[width=\columnwidth, trim=0 20 0 35, clip]{trajDisSpaceOverlap_PCA_tadDrcTiny_JMD}
   \includegraphics[width=\columnwidth, trim=0 20 0 35, clip]{../figures/trajDisSpaceOverlap_PCA_tad-drcTinyPen5_JMD.png}
+
 %  \caption{Estimated trajectories for the PCA cohort.
